---
title: "Deeper dive into vH4 model"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vH4"
  target_classes: 
    - GRADEGPA
output:
  html_document:
    keep_md: true
---

# MODIFY THIS TEXT TO FIT vH4

Work on course predictions in the "Course Guide" and then come back to this.

Spiffy up the boxplots DONE

Now I need to use this model to make predictions in the "Course Guide" --

  - It would be cool if I could reasonably demonstrate that the people predicted as "low" could have improved their grade by switching to a closer course.  That would be a little bit of an exercise, but probably worth it.
  
  - I need to find the report where I did all the work describing how I cleaned the data.  
  
  - I'd like to show the essentially precision/recall for all of the grades in my two categories  
  


**PURPOSE:** This report describes a model trained to predict initial math grades for first time freshmen in 2024 based on prior twenty years' of data on variables like ACT math test scores, high school GPA, and high school AP credits.  It includes the distance of the current  It performs with an r-squared value of 0.26.  Various aspects of the model prediction and actual performance are visualized and compared.  

EXECUTIVE SUMMARY

METHODOLOGY

To be copied from elsewhere.  
Importantly, 'W' or "withdraw" is removed from this analysis. 

DISCUSSION  

*ACCURACY*  

Predicting GPA directly performs with an R-squared value of 0.26.  Lumping the grades together into four bins, and predicting these bins, results in an R-squared value of 0.28.  Due to the similarity in performance, and a preference for the increased detail, this report focuses on the model that predicted grades directly.  

As far as distinguishing a binary grouping between "high" and "low", the model had an optimum Kappa of 0.39 at a cut-off value of predicting a GPA of 2.4 or less.  This draws a dividing line between a C+ and a B-, and identified "low" performers with 0.47 precision and 0.67 recall.  These "low" performers [[find out how they actually did, and shouldn't your histogram show that?]]

*PREDICTABLY FAILED*  

The model can identify a few students who were not expected to pass their first math class, and indeed did fail it, with good precision.  Due to their predictable failure, these students might be considered admissions mistakes, or are candidates for having their admissions determination reviewed.  

However, very few students are in this category.  The model predicted only 17 students would score 1.1 or less, and 11 of the students (65% precision) did indeed score 1.1 or less.  (Of the remaining six, one student scored 1.7, one student achieved a 4.0, and the remaining scored between a 2.0 and a 3.0.)  This very low number of students represented a recall metric of only 0.054.

A few conclusions can be drawn from this good-precision/low-recall result:  

  * That math grading practices are not mis-aligned with admissions standards.  (If they were mis-aligned, there would be more students expected to fail and the recall metric would be higher.)  
  * That practically every student admitted in 2024 was reasonably expected to pass a math class.   
  * That variables outside of information available upon admission is predicting or determining or causing math class failures.  This could include things like the number of credits taken, or a "poisonous" combination of difficult classes, or a mid-semester move.  It would be interesting to open up the class information to determine how quickly eventual failures can be identified. For example, to see if the first week's homework results or the first test's grade is highly predictive. 



  

NEXT STEPS

RESULTS 

What this report shows:

- There are a handful of students who predictably failed (with good precision.)  This could suggest: 
  - That these students shouldn't have been admitted and constitute an admissions mistake.  (If we knew they had a high chance of failure, should we have admitted them?)    
  - That the very small number of students in this category suggests that we don't have a large admissions failure.   
  
- An R-squared model of 0.28 is a decently predictive model on very few variables.  It can predict the student's grade with a mean absolute error of 0.84.  So incoming math ability and study ability can explain about 28% of grade variation.    

- However, it still leaves a lot un-explained.  It is not mechanically deterministic and doesn't suggest strong cut-offs.  The histograms compare the ranges of actual vs predicted values.  

- Failure (<1.3) seems to be about a lot more than math ability and study ability (test scores and high school GPA) as students fail despite predicted grades as high as an A.  (What's my metric that says this?)

- The optimum split is 2.7 (B-).  This is where the model performs at its best (least-random) ability to sort students into high and low grades. 

- Unexpectedly, "course" disappears as an explanatory variable. 

  - Maybe this means that the courses are roughly equivalent and course selection doesn't matter. 
  - Maybe this means that some unknown sorting mechanism adequately places students into math courses appropriate to their ability, so that only their studiousness or "study ability" (as indicated by their high school GPA) separates academic performance.  

- The top two predictors of grade are (overwhelmingly) the high school GPA and the ACT math test scores.  

  - A heat map of expected values based on twenty years' of data is very reasonable, with lower grades expected upon moving diagonally to the lower left. 
  
  - Grade inflation?  A lot redder than expected; Way more 'A's than expected.  Looks a lot easier to top off on this scale.  Looking a lot more like a pass-fail.  (Like---what's the number here?  How many did I expect and how many did I get?  It's a simple question of precision I guess.)  
  
    Looking at this: createCM(aligned[["GRADEGPA"]], cuts = c(0,3.3,3.7,4)) |> drawCM()
    
 Look how it "leans" right towards the higher end.  Way more people get >3.7 who are predicted to get <3.7 .
 
 Math looks very, very, very forgiving.  It is graded very generously.
 
Look at how low the precision is for  
createCM(aligned[["GRADEGPA"]], cuts = c(0,3.3,4)) |> drawCM().

They are really pushing grades up here. 

...or is that a correct interpretation?  Does it just mean my model is bad?  Can I manually adjust the prediction and improve the prediction?  If I can, why doesn't the model do it automatically?




  
  
  





And what makes these reports hard to complete 

This report needs some TLC, changes to plot sizes, some words and write-up, and then it needs a comparison with the grade-only prediction (maybe a separate report?).

Possibly develop a <= accuracy metric, a modification of recall, as I don't care how many people got better than a C+ but I care how many people got a C+ or less.

I'll polish up this report a little more before adding that.  


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(ggplot2)
library(patchwork)
library(kableExtra)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great!


source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", params$version, " ", params$target_classes, " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  params$target_classes
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", params$version, " ", params$target_classes, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  params$target_classes
)

```


```{r eval = FALSE}


# This will change based on the model prep.
# I might just want to delete this for a template "results"
# report.

## RAW DATA ##
ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))


##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

#
# identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
                          )
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
#  SATMATH, # leaving out SAT because it's a smaller proportion
#  SATVERBAL,
#  SATWRTG        
)

popCourses <- merge(popCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

popCourses <- merge(popCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
popCourses$cohort_year <- year(popCourses$COHORT_DT)

# Class year
popCourses$class_year <- year(popCourses$EOTDATE)

# Year to class
popCourses$yr_diff <- time_length(interval(popCourses$COHORT_DT, popCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
popCourses$season <- season_map[substr(popCourses$TERM, 4, 4)]


# print("Prep complete")

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

popCourses$cleanGrade <- clean_grade(popCourses$GRADE) 

popCourses$cleanGrade <- factor(popCourses$cleanGrade,
                                levels = c(
                                  "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                  "C", "C_minus", "D",   
                                  "E", "EU", "W", "I", "CR",  "NC", "V", 
                                  "missing", NA
                                )
)

# RARE AND UNUSUAL VALUES

rareGrades <- popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
# popCourses$wdraw_binary <- NA
# popCourses$wdraw_binary <- ifelse(popCourses$GRADE == "W","withdraw","not_withdraw")


# Combine grades
# BINARY
popCourses$grade_binary <- NA
hiFilter <- popCourses$GRADE %in% c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
loFilter <- popCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
popCourses$grade_binary[hiFilter] <- 1
popCourses$grade_binary[loFilter] <- 0

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
popCourses$grade_trinary <- NA
hiFilter <- popCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- popCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- popCourses$GRADE %in% c( "D+", "D", "D-", "E" ) 
popCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
popCourses$grade_trinary[medFilter] <- 1 # "med_grade"
popCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD (EXPANDED FAILURE)
popCourses$grade_quad <- NA
hiFilter <- popCourses$GRADE %in% c("A")
medPlusFilter <- popCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- popCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- popCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU", "W" ) 
popCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
popCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
popCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
popCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

# print("Transformations complete")


```

```{r}

## PREDICTIONS ##

thePredictions <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  # print(x)
  # print(table(missingFilter))
}
  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions) <- params$target_classes

```

```{r}


theResults <- lapply(params$target_classes, function(x) {
  
  # print(x)
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
#  print(x)
#  print(table(missingFilter))
}
  
  postResample(pred = thePredictions[[x]],
                  obs = data_list[[x]][["testing"]][[x]][missingFilter]
  
               )
})
names(theResults) <- params$target_classes



```

### Models 

"Quad" model was not evaluated and is not covered in this report. 

```{r}

data.frame(Letter = c("E, EU", "D-","D","D+", "C-","C","C+", "B-","B","B+","A-","A"),
           GPA = c(0,0.7,1,1.3,1.7,2.0,2.3,2.7,3.0,3.3,3.7,4.0),
           Quad =c(0,0,0,0,1,1,1,2,2,2,2,3)
           ) |>
  kbl(align = c("l", "c", "c"),
      col.names = c("Letter Grade", "GPA", "Quad"),
      caption = "Grade Mapping") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "left") |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, background = "#f0f0f0")

```



# Results

```{r}


# I'd like to normalize the errors with the range and standard deviation

# but it doesn't change that these aren't very good models

# normalized errors

normalized_metrics <- function(pred, obs) {
  rmse <- sqrt(mean((pred - obs)^2))
  mae  <- mean(abs(pred - obs))
  r2   <- cor(pred, obs)^2
  
  range_y <- diff(range(obs, na.rm = TRUE))
  sd_y    <- sd(obs, na.rm = TRUE)
  
  data.frame(
    RMSE = rmse,
    MAE  = mae,
    R2   = r2,
    NRMSE_range = rmse / range_y,
    NMAE_range  = mae  / range_y,
    NRMSE_sd    = rmse / sd_y,
    NMAE_sd     = mae  / sd_y
  )
}  
  
normResults <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
#  print(x)
#  print(table(missingFilter))
}
  
  normalized_metrics(
    pred = thePredictions[[x]],
    obs = data_list[[x]][["testing"]][[x]][missingFilter]
  )
  
})  
names(normResults) <- params$target_classes

library(kableExtra)

normResults |>
  (\(x){do.call(rbind,x)})() |>
  (\(x){round(x,2)})()|>
    kbl(
    caption = "Normalized Regression Results Across Grading Buckets",
    align = "c",
    col.names = c("",
      "RMSE", "MAE", "R-sq", 
      "NRMSE (Range)", "NMAE (Range)", 
      "NRMSE (SD)", "NMAE (SD)"
    )
  ) |>
  kable_classic(full_width = FALSE, html_font = "Calibri") |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>  # header
  row_spec(1:length(params$target_classes), background = c("#ffffff", "#f9f9f9")) |> # alternating rows
  add_header_above(c(" " = 1, "Raw Metrics" = 3, "Normalized Metrics" = 4))

```


```{r}

regression_diagnostics <- function(pred, reference, plot_title = "Predicted vs Actual") {
  if (length(pred) != length(reference)) {
    stop("`pred` and `reference` must be the same length.")
  }
  
  incoming.par <- par(mfrow= c(2,1), mar = c(3,2.5,0,1), oma = c(1.5,1.5,2.25,0))
  
  on.exit(par(incoming.par))
  
  # Compute metrics
  rmse <- sqrt(mean((pred - reference)^2))
  mae  <- mean(abs(pred - reference))
  r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2) 
  
  # Print metrics
#  cat("Regression diagnostics:\n")
#  cat(sprintf("RMSE: %.4f\n", rmse))
#  cat(sprintf("MAE:  %.4f\n", mae))
#  cat(sprintf("R-squared (variance decomposition): %.4f\n", r2))

  
  # Scatter plot
  plot(reference, pred,
      # xlab = "Actual",
      # ylab = "Predicted",
       # main = plot_title,
       pch = 19, col = "steelblue",
       las = 1)
  abline(a = 0, b = 1, col = "red", lwd = 2)  # 45-degree reference line
  grid()
  
  # boxplot
  bp <- boxplot(pred ~ reference,
        # xlab = "Actual",
        # ylab = "Predicted",
         # main = plot_title,
         col = "lightblue",
         border = "steelblue",
         las=1)
  
    # Add a red reference line for each box
        for (i in seq_along(bp$names)) {
          actual_val <- as.numeric(bp$names[i])
          segments(x0 = i - 0.4, y0 = actual_val,
                   x1 = i + 0.4, y1 = actual_val,
                   col = "red", lwd = 2)
        }
  
mtext(side = 3, line =0.75, text = plot_title, cex = 1.5, font = 2, outer = TRUE)  
  
mtext(side = 2, line =0.25, text = "Predicted", cex = 1.25, font = 1, outer = TRUE)  

mtext(side = 1, line =0.25, text = "Actual", cex = 1.25, font = 1, outer = TRUE)  

  # Return metrics invisibly as a list
  invisible(list(RMSE = rmse, MAE = mae, R2 = r2))
}


invisible(lapply(params$target_classes, function(x){
  
      # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  
regression_diagnostics(pred = thePredictions[[x]],
                       reference = data_list[[x]][["testing"]][[x]][missingFilter],
                       plot_title = x)
})
)

```



```{r fig.height=4.5, fig.width=9}

aligned <- alignPrediction(prediction_list = thePredictions, reference_list = data_list)

# ACTUAL vs PREDICTED HEATMAP
predictionMap(data = aligned[["GRADEGPA"]], 
              plot_title = "Predicted vs Actual for 2024, all courses")



```


```{r eval=FALSE}

# I'm not seeing a real need to display this for each grade 

suppressWarnings(
predictionHistogram(data = aligned[["GRADEGPA"]],
                    title = NA,
                    show = "actual",
                    cap = TRUE, 
                    hist_params = list(freq = TRUE))
)

suppressWarnings(
predictionHistogram(data = aligned[["GRADEGPA"]],
                    title = NA,
                    show = "predict",
                    cap = TRUE, 
                    hist_params = list(freq = TRUE))
)

```


```{r}

suppressWarnings(
predictionHistogram(data = aligned[["GRADEGPA"]],
                    title = NA,
                    show = "actual",
                    cap = TRUE,
                    cuts = c(0,2.7,4),
                    hist_params = list(freq = TRUE))
)

suppressWarnings(
predictionHistogram(data = aligned[["GRADEGPA"]],
                    title = NA,
                    show = "predict",
                    cap = TRUE, 
                    cuts = c(0,2.7,4),
                    hist_params = list(freq = TRUE))
)

```  

```{r eval=FALSE}

# density plot sandbox

# eh let's do the density plot  
density_grade <- density(aligned[["GRADEGPA"]][["GRADEGPA"]])
plot(density_grade)

density_pred <- aligned[["GRADEGPA"]][["pred"]] |>
  (\(x){x[!is.na(x)]})() |>
  density()
lines(density_pred, col = "blue", lwd =2)

aligned[["GRADEGPA"]][["pred_cut_2.7"]] <- cut(aligned[["GRADEGPA"]][["pred"]], breaks = c(0,2.7,4), include.lowest = TRUE)


lowCutFilter <-  aligned[["GRADEGPA"]][["cut"]] == "(0,2.7]" & !is.na(aligned[["GRADEGPA"]][["cut"]])
density_grade_low <- aligned[["GRADEGPA"]][["GRADEGPA"]] |>
  (\(x){x[lowCutFilter] })() |>
  density()
lines(density_grade_low, col = "red", lty = 3, lwd = 2)

density_pred_low <- aligned[["GRADEGPA"]][["pred"]] |>
  (\(x){x[lowCutFilter]})() |>
  density()
lines(density_pred_low, col = "blue", lty=3, lwd =2)


hiCutFilter <-  aligned[["GRADEGPA"]][["cut"]] == "(2.7,4]" & !is.na(aligned[["GRADEGPA"]][["cut"]])
density_grade_hi <- aligned[["GRADEGPA"]][["GRADEGPA"]] |>
  (\(x){x[hiCutFilter] })() |>
  density()
lines(density_grade_hi, col = "red", lty = 3, lwd = 2)

density_pred_hi <- aligned[["GRADEGPA"]][["pred"]] |>
  (\(x){x[hiCutFilter]})() |>
  density()
lines(density_pred_hi, col = "blue", lty=3, lwd =2)

# losing track on what I'm trying to show here

# let's try and get me some confusion matrices
# Not sure this is worth it, actually

cm_1.7 <- confusionMatrix(data = aligned[["GRADEGPA"]][["GRADEGPA"]] |> 
                          (\(x){x[lowCutFilter]})() |>
                          (\(x){cut(x, c(0,1.7,4))})(),

reference = aligned[["GRADEGPA"]][["pred"]] |> 
                          (\(x){x[lowCutFilter]})() |>
                          (\(x){cut(x, c(0,1.7,4))})()
)

drawCM(cm_1.7)

# not sure that's what I want at all
# I need to include the <1.7 at the high prediction as well

aligned[["GRADEGPA"]][["cut_1.7"]] <- cut(aligned[["GRADEGPA"]][["GRADEGPA"]], c(0,1.7,4), include.lowest = TRUE, right = FALSE )

table(aligned[["GRADEGPA"]][,c("pred_cut_2.7", "cut_1.7")])

#            cut_1.7
# pred_cut_2.7 [0,1.7) [1.7,4]
#     [0,2.7]     190     636
#     (2.7,4]      34    1054

#             cut_1.7
# pred_cut_2.7 [0,1.7] (1.7,4]
#     [0,2.7]     230     596
#     (2.7,4]      38    1050

# This is actually a different optimization problem than
# my optimum Kappa, which currently requires the same cut on
# both grade and pred.

# Seems like I want to play around as a pred cut somewhere between 2.3 and 2.7 (inclusive) will probably be my optimum precision/recall for failures (<1.7, not inclusive)

# Not sure it's worth spending an hour to figure this out.
# In fact, eh I'll publish this graphic but not spend more time on this optimum value.


table(aligned[["GRADEGPA"]][,c("pred_cut_2.7", "cut_1.7")]) |>
  kbl()

```

```{r}


aligned[["GRADEGPA"]][["pred_cut_2.4"]] <- cut(aligned[["GRADEGPA"]][["pred"]], breaks = c(0,2.4,4), include.lowest = TRUE)

aligned[["GRADEGPA"]][["cut_1.7"]] <- cut(aligned[["GRADEGPA"]][["GRADEGPA"]], c(0,1.7,4), include.lowest = TRUE, right = FALSE )


# Create confusion matrix with proportions
conf_table <- table(aligned[["GRADEGPA"]][,c("pred_cut_2.4", "cut_1.7")])

# Convert to proportions and percentages
prop_table <- prop.table(conf_table) * 100

# Combine counts and percentages
combined <- matrix(
  paste0(conf_table, "\n(", round(prop_table, 1), "%)"),
  nrow = nrow(conf_table)
)

dimnames(combined) <- dimnames(conf_table)

combined |>
  kbl(caption = "Predicted vs Actual Grade Categories",
      col.names = c("Actual: Fail (<1.7)", "Actual: Pass (≥1.7)"),
      align = "c",
      escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                font_size = 18) |>
  add_header_above(c("Predicted Grade" = 1, "Actual Grade" = 2),
                   bold = TRUE, font_size = 20) |>
  row_spec(0, bold = TRUE, font_size = 20, background = "#e6f2ff")

recall <- (conf_table |> as.matrix() |> (\(x){x[1,1]})()  / conf_table |> as.matrix() |> (\(x){sum(x[,1])})()) |> round(2) |> (\(x){x*100})()

precision <- (conf_table |> as.matrix() |> (\(x){x[1,1]})()  / conf_table |> as.matrix() |> (\(x){sum(x[1,])})()) |> round(2) |> (\(x){x*100})()

```   

Students predicted to have a grade less than 2.4 are a high-risk group of students. `r precision`% of these `r conf_table |> as.matrix() |> (\(x){sum(x[1,])})()` students actually did fail, or `r recall`% of the total of students who failed.  (`r recall`% recall and `r precision`% precision).





```{r fig.height=5, fig.width=11}

# Cumulative accuracy
# Where I calculate a binary confusion matrix, and gradually increase the upper bound,
# so that everything equal-or-less-than is considered accurate

cumAcc <- lapply(round(seq(from = 0.1, to = 3.9, by = 0.1),1), function(x){
  
  createCM(aligned[["GRADEGPA"]], cuts = c(0,x,4))
  
})

par(mar=c(5.1,4.1,4.1,7.6), bg="ivory")

plot(x = 0.5,
  ylim = c(0,1),
     xlim = c(0,length(cumAcc)),
  xaxt = "n",
  las = 2,
     ylab = "",
     xlab = "",
  type = "n")

rect(xleft = par("usr")[1] ,
     ybottom = par("usr")[3] ,
     xright = par("usr")[2],
     ytop = par("usr")[4],
     col = "grey95")

grid(col="gray90", lwd = 2, lty = 1)

sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> lines(col = "firebrick", lwd = 2)
sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> lines(col = "dodgerblue", lwd = 2)
sapply(cumAcc, function(x) {x[["byClass"]][["Recall"]]}) |> lines(col = "gold", lwd = 2)

axis(side = 1, at = 1:(length(cumAcc)+1), labels = round(seq(from = 0.1, to = 4.0, by = 0.1),1))

legend("topright",
       col = c("firebrick","dodgerblue","gold"),
       legend = c("Kappa","Precision","Recall"),
       lwd =4,
       inset = c(-0.175,0),
       xpd = TRUE,
       bg = "snow"
       )

mtext(side=3, font = 2, cex = 1.619, line = 1.619, text = "Confusion matrix metrics from binary discretization")
mtext(side=1, line = 3, "Upper GPA Bound")

# Admission mistake point
points(x = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){which(x == max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE))})(),
         y = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE)})(),
         col = "dodgerblue",
       pch = 16,
       cex = 1.1
)

text(x = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){which(x == max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE))})(),
         y = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE)})(),
  labels = "Admission 'mistake' candidates\n     (predictably failed)",
  pos = 4)


# Optimum (Kappa)
points(x = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> (\(x){which(x == max(x, na.rm = TRUE))})(),
       y = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> max(na.rm=TRUE),
       col = "firebrick",
       pch = "|",
       cex = 1.319
)

text(x = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> (\(x){which(x == max(x, na.rm = TRUE))})(),
     y = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> max(na.rm=TRUE),
     labels = "Model\noptimum",
     pos = 1)


```

Examining the precision for predictions of grades <0.9 looks like these few people could be an admissions error, or a very low probability of success for this handful of people. 


```{r}

createCM(aligned[["GRADEGPA"]], cuts = c(0,2.4,4)) |> drawCM(title = "Binary optimum")
createCM(aligned[["GRADEGPA"]], cuts = c(0,0.9,4)) |> drawCM(title = "Predictably failed with good precision")

```


# Variable importance

### Predictors

SPEND SOME MORE TIME TALKING ABOUT THIS.
REMOVE "load" EXPLANATION

`r colnames(data_list[[1]][["training"]])[-1]`. 

"load" is the credit load per student per term. 
"yr_diff" is the time (expressed as fractions of a year) between the cohort date and the end of term date for the course.  

The other variables are either self-explanatory or taken directly from OBIA tables.  







```{r fig.height=5}

invisible(
  lapply(params$target_classes, function(x){

  print(plot(
    varImp(model_list[[x]]),
    15,
    main = x
       )  
  )
})
)

```


# Relationship between distance to median successful students (as a z-score) and grade


```{r}

actDensity <- density(data_list[["GRADEGPA"]][["training"]][["ACTMATH.z"]])
hsgpaDensity <- density(data_list[["GRADEGPA"]][["training"]][["HSGPA.z"]])
distDensity <- density(data_list[["GRADEGPA"]][["training"]][["dist"]])

hist(data_list[["GRADEGPA"]][["training"]][["ACTMATH.z"]],
     col = "pink",
     main = "Histogram of ACT math z-scores",
     xlab = "Z-score (ACT Math)\n[bins include left endpoint]",
     right = FALSE,
     bg = "ivory")

plot(hsgpaDensity,
     bg = "ivory",
     main = "Density distribution of high school grade z-scores"
     )

polygon(hsgpaDensity, 
        col = "aliceblue", 
        border = NA)  # No border around the fill

lines(hsgpaDensity, lwd = 2)  # Redraw the line on top

plot(distDensity,
     bg = "ivory",
     main = "Density of combined z-score distances"
     )

polygon(distDensity, 
        col = "beige", 
        border = NA)  # No border around the fill

lines(distDensity, lwd = 2)  # Redraw the line on top

```

```{r}

# Box plots of z-scores vs grades

zBox <- function(data, column, 
                 box_params = list(),
                 axis_params = list(),
                 title_params = list(),
                 title = NA
                 ){
  
  thePlot <- boxplot(get(column) ~ GRADEGPA,
                     data = data,
                     ylab = "",
                     xlab = "",
                     outline = FALSE
                     )
  
  rect(xleft = par("usr")[1],
       ybottom = par("usr")[3],
       xright = par("usr")[2],
       ytop = par("usr")[4],
       col = "gray95")

  grid(col = "gray90", lty =1, lwd = 2)
  
  default_box_params <- list( 
    formula = as.formula(paste(column, " ~ GRADEGPA")),
    data = data,
    ylab = "",
    xlab = "",
    outline = FALSE,
    col = "lavender",
    add = TRUE
    )
  
  box_params <- modifyList(default_box_params, box_params)
  do.call(boxplot, box_params)
  
  default_axis_params <- list(
    side = c(1,2),
    font = c(1,1),
    cex = c(1.2,1.2),
    line = c(2.32,2.31),
    text = c("High school grade", "Z-score")
  )

  axis_params <- modifyList(default_axis_params, axis_params)
  do.call(mtext, axis_params)
    
  default_title_params <- list(
        side = c(3,3),
    font = c(2,3),
    cex = c(2.05, 1.25),
    line = c(1.69, 0.5),
    text = c(title, "Z-score distance to median successful student")
  )
  
  title_params <- modifyList(default_title_params, title_params)
  do.call(mtext, title_params)

  
}


zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "ACTMATH.z",
     title = "ACT Math",
     box_params = list(col="pink"))


zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "HSGPA.z",
     title = "High school GPA",
     box_params = list(col="aliceblue"))

zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "dist",
     title = "Combined distance",
     box_params = list(col="beige"))

# boxplot(ACTMATH.z ~ GRADEGPA, 
#        data = data_list[["GRADEGPA"]][["training"]], 
#        outline=FALSE)

# boxplot(HSGPA.z ~ GRADEGPA, data = data_list[["GRADEGPA"]][["training"]], outline=FALSE)

# boxplot(dist ~ GRADEGPA, data = data_list[["GRADEGPA"]][["training"]], outline=FALSE)

```

not a bad series of boxplots.  Def something happening here

Man I've done so many boxplots -- don't I have a custom function to manage this?  

I should do the incoming plot hex.

Also, the hist is great -- but maybe the density plot? Nah, what I got is fine. 




```{r eval=FALSE}

for(x in params$target_classes) {
  
    print(plot(
    varImp(model_list[[x]]),
    15,
    main = x
       )  
    )
}

```

# Confusion matrix

```{r eval=FALSE}

GPA_grade_levels <- unique(data_list[["GRADEGPA"]][["testing"]][["GRADEGPA"]]) |> 
  (\(x){ x[order(x)]})()
intermediate_steps <- GPA_grade_levels |> (\(x){ x[order(x)]})() |> diff() |> (\(x){x*0.5})()
GRADEGPA_thresholds <- GPA_grade_levels |> 
  (\(x){ x[order(x)]})() |>
  (\(x){x+intermediate_steps})()
GRADEGPA_thresholds[GRADEGPA_thresholds == max(GRADEGPA_thresholds)] <- 10 # upper limit is high so any high predictions are leveled back as "A"

GRADEGPA_thresholds <- c(-10, GRADEGPA_thresholds) # lower limit is low so any low prediction is leveled back to "F"

quad_grade_levels <- unique(data_list[["grade_quad"]][["testing"]][["grade_quad"]]) |> 
  (\(x){ x[order(x)]})()
intermediate_steps <- quad_grade_levels |> (\(x){ x[order(x)]})() |> diff() |> (\(x){x*0.5})()
grade_quad_thresholds <- quad_grade_levels |> 
  (\(x){ x[order(x)]})() |>
  (\(x){x+intermediate_steps})()
grade_quad_thresholds[grade_quad_thresholds == max(grade_quad_thresholds)] <- 10 # upper limit is high so any high predictions are leveled back as "A"

grade_quad_thresholds <- c(-10, grade_quad_thresholds)

missingInTraining <- function(target) {
  
  # Removing courses that are in testing but not in training 
  
    # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[target]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[target]][["testing"]][[courseColumn]],data_list[[target]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[target]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[target]][["testing"]][[courseColumn]] %in% missing
}
  return(missingFilter)
  
}

  
createConfusionMatrix <- function(pred, reference, thresholds = NA, ...){
  
  # set thresholds if not provided
  if(any(is.na(thresholds))) {
  grade_levels <- unique(reference)
  pred_levels <- sapply(pred, function(x){
    grade_levels[which.min(abs(x-grade_levels))]
  })
  
  }
  
  # convert to discrete values
  discrete_predictions <- cut(pred, thresholds, include.lowest = TRUE, right = FALSE)
  discrete_references <- cut(reference, thresholds, include.lowest=TRUE, right = FALSE)
  
  # create uniform factor levels

all_levels <- sort(unique(c(as.character(discrete_predictions), 
                            as.character(discrete_references))))
  
discrete_predictions <- factor(discrete_predictions, levels = all_levels)
discrete_references <- factor(discrete_references, levels = all_levels)

  # create confusion matrix
  
  cm <- confusionMatrix(discrete_predictions, discrete_references, mode = "everything", ... )
  
  return(cm)
  
}

gradeMissing <- missingInTraining("GRADEGPA")  
GRADEGPA_cm <- createConfusionMatrix(pred = thePredictions[["GRADEGPA"]],
                                      reference = data_list[["GRADEGPA"]][["testing"]][["GRADEGPA"]][gradeMissing],
                                      thresholds = GRADEGPA_thresholds
                                      )

drawCM(GRADEGPA_cm) # fixed!

quadMissing <- missingInTraining("grade_quad")  
quad_cm <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = grade_quad_thresholds
                                      )
drawCM(quad_cm, nameMatrix = cbind(original = c("[-10,0.5]", "(0.5,1.5]", "(1.5,2.5]", "(2.5,10]"),
                                display =  c("[-10,0.5]", "(0.5,1.5]", "(1.5,2.5]", "(2.5,10]"))
)


# let's turn that into a binary"
# to make this work, you'll need to modify the "cut" to "right = TRUE"

quad_cm_bin <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = c(-10,2,10),
                                     positive = "[-10,2]"
                                      )
drawCM(quad_cm_bin, nameMatrix = cbind(original = c("[-10,2]", "(2,10]"),
                                display =  c("[-10,2]", "(2,10]"))
)

# This one had an EXCELLENT kappa value-- like 0.6

# starting to look a lot better
# Maybe I use different thresholds for the prediction as the reference?
# Why not?

# and let's use better thresholds --- interested in flunking, here--

quad_cm_bin <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = c(-10,1,10),
                                     positive = "[-10,1)"
                                      )
drawCM(quad_cm_bin, nameMatrix = cbind(original = c("[-10,1)", "[1,10]"),
                                display =  c("[-10,1)", "[1,10]"))
)


quad_cm_bin <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = c(-10,2,10),
                                     positive = "[-10,2)"
                                      )
drawCM(quad_cm_bin, nameMatrix = cbind(original = c("[-10,2)", "[2,10]"),
                                display =  c("[-10,2)", "[2,10)"))
)

# ...and I need to update drawCM again, from vC6 to include "grepl(any....)"


# I should cough up my best one, but unfortunately it's not the most useful.



``` 


```{r}

drawGuide()

```


# 

# Training data

Presentation order: 

```{r eval=FALSE}



lapply(params$target_classes, function(x) {
  
  print(x)

  skim(data_list[[x]][["training"]])
  
})


```

