---
title: "Deeper Dive into vH4 (quarto)"
format: 
  html: default
  gfm: default
  pdf: default
  docx: default 
editor: source
execute: 
  echo: false
  warning: false  
  message: false
  error: false
knitr: 
  opts_chunk: 
    fig.height: 5
    fig.width: 7
---

```{r}
#| include: false


############
## SET UP ##
############

# knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
#                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(ggplot2)
library(patchwork)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

is_html <- knitr::is_html_output() # Renders plotly in HTML renderings, otherwise renders base R

model_version <- "vH4"

```


```{r}

##########
## LOAD ##
##########

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", model_version, " ", "GRADEGPA", " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  "GRADEGPA"
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", model_version, " ", "GRADEGPA", " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  "GRADEGPA"
)

```


```{r}

## PREDICTIONS ##

thePredictions <- lapply("GRADEGPA", function(x) {
  
  # identify course values in testing that aren't in training
  
  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  # print(x)
  # print(table(missingFilter))
}
  
  # drop 2025
  remove2025 <- data_list[[x]][["testing"]][["class_year"]] != 2025
  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]][missingFilter & remove2025,])
  
} )
names(thePredictions) <- "GRADEGPA"

```

```{r}


theResults <- lapply("GRADEGPA", function(x) {
  
  # print(x)
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
#  print(x)
#  print(table(missingFilter))
}
  
    # remove 2025
  remove2025 <- data_list[[x]][["testing"]][["class_year"]] != 2025
  
  
  postResample(pred = thePredictions[[x]],
                  obs = data_list[[x]][["testing"]][[x]][missingFilter & remove2025]
  
               )
})
names(theResults) <- "GRADEGPA"

```




```{r}

# normalized errors

normalized_metrics <- function(pred, obs) {
  rmse <- sqrt(mean((pred - obs)^2))
  mae  <- mean(abs(pred - obs))
  r2   <- cor(pred, obs)^2
  
  range_y <- diff(range(obs, na.rm = TRUE))
  sd_y    <- sd(obs, na.rm = TRUE)
  
  data.frame(
    RMSE = rmse,
    MAE  = mae,
    R2   = r2,
    NRMSE_range = rmse / range_y,
    NMAE_range  = mae  / range_y,
    NRMSE_sd    = rmse / sd_y,
    NMAE_sd     = mae  / sd_y
  )
}  
  
normResults <- lapply("GRADEGPA", function(x) {
  
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
#  print(x)
#  print(table(missingFilter))
}
  
      # remove 2025
  remove2025 <- data_list[[x]][["testing"]][["class_year"]] != 2025
  
  normalized_metrics(
    pred = thePredictions[[x]],
    obs = data_list[[x]][["testing"]][[x]][missingFilter & remove2025]
  )
  
})  
names(normResults) <- "GRADEGPA"

```

# SUMMARY 

# METHODLOGY 

# DISCUSSION 

# CONCLUSIONS 

# POSSIBLE NEXT STEPS 

# RESULTS 

```{r}
#| # eval: !expr is_html
#| label: tbl-results
#| tbl-cap: "Normalized Regression Results"


library(kableExtra)

normResults |>
  (\(x){do.call(rbind,x)})() |>
  (\(x){round(x,2)})()|>
    kbl(
    caption = "",
    align = "c",
    col.names = c("",
      "RMSE", "MAE", "R-sq", 
      "NRMSE (Range)", "NMAE (Range)", 
      "NRMSE (SD)", "NMAE (SD)"
    )
  ) |>
  kable_classic(full_width = FALSE, html_font = "Calibri") |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>  # header
  row_spec(1:1, background = c("#ffffff", "#f9f9f9")) |> # alternating rows
  add_header_above(c(" " = 1, "Raw Metrics" = 3, "Normalized Metrics" = 4))

```


```{r}

regression_diagnostics <- function(pred, reference, plot_title = "Predicted vs Actual") {
  if (length(pred) != length(reference)) {
    stop("`pred` and `reference` must be the same length.")
  }
  
  incoming.par <- par(mfrow= c(2,1), mar = c(3,2.5,0,1), oma = c(1.5,1.5,2.25,0))
  
  on.exit(par(incoming.par))
  
  # Compute metrics
  rmse <- sqrt(mean((pred - reference)^2))
  mae  <- mean(abs(pred - reference))
  r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2) 
  
  # Print metrics
#  cat("Regression diagnostics:\n")
#  cat(sprintf("RMSE: %.4f\n", rmse))
#  cat(sprintf("MAE:  %.4f\n", mae))
#  cat(sprintf("R-squared (variance decomposition): %.4f\n", r2))

  
  # Scatter plot
  plot(reference, pred,
      # xlab = "Actual",
      # ylab = "Predicted",
       # main = plot_title,
       pch = 19, col = "steelblue",
       las = 1)
  abline(a = 0, b = 1, col = "red", lwd = 2)  # 45-degree reference line
  grid()
  
  # boxplot
  bp <- boxplot(pred ~ reference,
        # xlab = "Actual",
        # ylab = "Predicted",
         # main = plot_title,
         col = "lightblue",
         border = "steelblue",
         las=1)
  
    # Add a red reference line for each box
        for (i in seq_along(bp$names)) {
          actual_val <- as.numeric(bp$names[i])
          segments(x0 = i - 0.4, y0 = actual_val,
                   x1 = i + 0.4, y1 = actual_val,
                   col = "red", lwd = 2)
        }
  
mtext(side = 3, line =0.75, text = plot_title, cex = 1.5, font = 2, outer = TRUE)  
  
mtext(side = 2, line =0.25, text = "Predicted", cex = 1.25, font = 1, outer = TRUE)  

mtext(side = 1, line =0.25, text = "Actual", cex = 1.25, font = 1, outer = TRUE)  

  # Return metrics invisibly as a list
  invisible(list(RMSE = rmse, MAE = mae, R2 = r2))
}

```


```{r}
#| label: fig-results_scatter_box
#| fig-cap: "These plots show what the correct prediction value would be with a red line."
#| fig-cap-location: top

invisible(lapply("GRADEGPA", function(x){
  
      # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
        # remove 2025
  remove2025 <- data_list[[x]][["testing"]][["class_year"]] != 2025

regression_diagnostics(pred = thePredictions[[x]],
                       reference = data_list[[x]][["testing"]][[x]][missingFilter & remove2025],
                       plot_title = "Predicted vs Actual")
})
)

```



```{r}

# manually align test data (reference) with prediction, removing 2025

# Identify mis-match in courses between training and testing sets

missing <- setdiff(data_list[["GRADEGPA"]][["testing"]][["course"]], data_list[["GRADEGPA"]][["training"]][["course"]])

missingFilter <- !data_list[["GRADEGPA"]][["testing"]][["course"]] %in% missing

# remove 2025
  remove2025 <- data_list[["GRADEGPA"]][["testing"]][["class_year"]] != 2025

aligned <- data_list[["GRADEGPA"]][["testing"]][missingFilter & remove2025,]

aligned$pred <- thePredictions[["GRADEGPA"]]

```


```{r}
#| label: fig-density
#| fig-cap: "The model tends to predict towards the center, while most students earn an 'A'."
#| fig-cap-location: top
    
# Density plot
xLim <- range(aligned[,c("GRADEGPA", "pred")])

ref <- aligned[,"GRADEGPA"]
pred <- aligned[,"pred"]

plot(density(ref), 
     col = "dodgerblue3", 
     lwd = 2, 
     xlim = xLim,
     main = "", 
     xlab = "GPA")
lines(density(pred), 
      col = "darkorange2", 
      lwd = 2)
legend("topleft", legend = c("Actual", "Predicted"), 
       col = c("dodgerblue3", "darkorange2"), lwd = 2)

mtext(text = "Predicted vs Actual",
      side = 3,
      line = 1.7,
      font = 2,
      cex = 2,
      )

mtext(text = "as a density plot",
      side = 3,
      line = 0.3,
      font = 1,
      cex = 1.3,
      )

```


```{r}
#| label: fig-heatmap
#| fig-height: 3.5
#| fig-width: 7
#| fig-cap: "The heatmap shows grade expectations increase with ACT Math and high school GPA."
#| fig-cap-location: top


# ACTUAL vs PREDICTED HEATMAP
predictionMap(data = aligned, 
              plot_title = "Predicted vs Actual for 2024, all courses")


```

## Selection of model optimum binary cutoff or threshold

The model has the highest accuracy in separating into high/low (or succeed/at-risk) populations at a predicted grade of 2.4, between an actual grade of C-plus (2.3) and B-minus (2.7).  


```{r}
#| label: fig-threshold
#| fig-height: 5
#| fig-width: 11
#| fig-cap: "Selection of model optimum cutoff"
#| fig-cap-location: top


# Cumulative accuracy
# Where I calculate a binary confusion matrix, and gradually increase the upper bound,
# so that everything equal-or-less-than is considered accurate

cumAcc <- lapply(round(seq(from = 0.1, to = 3.9, by = 0.1),1), function(x){
  
  createCM(aligned, cuts = c(0,x,4))
  
})

par(mar=c(5.1,4.1,4.1,7.6), bg="ivory")

plot(x = 0.5,
  ylim = c(0,1),
     xlim = c(0,length(cumAcc)),
  xaxt = "n",
  las = 2,
     ylab = "",
     xlab = "",
  type = "n")

rect(xleft = par("usr")[1] ,
     ybottom = par("usr")[3] ,
     xright = par("usr")[2],
     ytop = par("usr")[4],
     col = "grey95")

grid(col="gray90", lwd = 2, lty = 1)

sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> lines(col = "firebrick", lwd = 2)
sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> lines(col = "dodgerblue", lwd = 2)
sapply(cumAcc, function(x) {x[["byClass"]][["Recall"]]}) |> lines(col = "gold", lwd = 2)

axis(side = 1, at = 1:(length(cumAcc)+1), labels = round(seq(from = 0.1, to = 4.0, by = 0.1),1))

legend("topright",
       col = c("firebrick","dodgerblue","gold"),
       legend = c("Kappa","Precision","Recall"),
       lwd =4,
       inset = c(-0.175,0),
       xpd = TRUE,
       bg = "snow"
       )

mtext(side=3, font = 2, cex = 1.619, line = 1.619, text = "Confusion matrix metrics from binary discretization")
mtext(side=1, line = 3, "Upper GPA Bound")

# Admission mistake point
points(x = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){which(x == max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE))})(),
         y = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE)})(),
         col = "dodgerblue",
       pch = 16,
       cex = 1.1
)

text(x = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){which(x == max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE))})(),
         y = sapply(cumAcc, function(x) {x[["byClass"]][["Precision"]]}) |> (\(x){max(x[x<max(x,na.rm=TRUE)], na.rm = TRUE)})(),
  labels = "Admission 'mistake' candidates\n     (predictably failed)",
  pos = 4)


# Optimum (Kappa)
points(x = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> (\(x){which(x == max(x, na.rm = TRUE))})(),
       y = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> max(na.rm=TRUE),
       col = "firebrick",
       pch = "|",
       cex = 1.319
)

text(x = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> (\(x){which(x == max(x, na.rm = TRUE))})(),
     y = sapply(cumAcc, function(x) {x[["overall"]][["Kappa"]]}) |> max(na.rm=TRUE),
     labels = "Model\noptimum",
     pos = 1)

```


```{r}
#| eval: !expr is_html
#| label: tbl-risk
#| tbl-cap: "Actual failure rates of 'Success' and 'At Risk' populations at model binary optimum"

aligned[["pred_cut_2.4"]] <- cut(aligned[["pred"]], breaks = c(0,2.4,4), include.lowest = TRUE, right = TRUE)

aligned[["cut_1.7"]] <- cut(aligned[["GRADEGPA"]], c(0,1.7,4), include.lowest = TRUE, right = FALSE )


# Create confusion matrix with proportions
conf_table <- table(aligned[,c("pred_cut_2.4", "cut_1.7")])

# Convert to proportions and percentages
prop_table <- prop.table(conf_table) * 100

# Combine counts and percentages
combined <- matrix(
  paste0(conf_table, "\n(", round(prop_table, 1), "%)"),
  nrow = nrow(conf_table)
)

dimnames(combined) <- dimnames(conf_table)

combined |>
  kbl(caption = "Predicted vs Actual Grade Categories",
      col.names = c("Actual: Fail (<1.7)", "Actual: Pass (≥1.7)"),
      align = "c",
      escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                font_size = 18) |>
  add_header_above(c("Predicted Grade" = 1, "Actual Grade" = 2),
                   bold = TRUE, font_size = 20) |>
  row_spec(0, bold = TRUE, font_size = 20, background = "#e6f2ff")

recall <- (conf_table |> as.matrix() |> (\(x){x[1,1]})()  / conf_table |> as.matrix() |> (\(x){sum(x[,1])})()) |> round(2) |> (\(x){x*100})()

precision <- (conf_table |> as.matrix() |> (\(x){x[1,1]})()  / conf_table |> as.matrix() |> (\(x){sum(x[1,])})()) |> round(2) |> (\(x){x*100})()



```

Students predicted to have a grade less than 2.4 are a high-risk group of students. `r precision`% of these `r conf_table |> as.matrix() |> (\(x){sum(x[1,])})()` students actually did fail, or `r recall`% of the total of students who failed.  (`r recall`% recall and `r precision`% precision).

This compares to the risk of students predicted to have a grade more than 2.4. Only `r conf_table[2,1]` of these `r conf_table |> as.matrix() |> (\(x){sum(x[1,])})()` students failed, or `r conf_table[2,1] |> (\(x){x/sum(conf_table[2,]) })() |> round(2) |> (\(x){100*x})()  `%.

```{r}
#| label: fig-binary_cm
#| fig-cap: "Confusion matrix at model binary optimum" 
#| fig-cap-location: top

createCM(aligned, cuts = c(0,2.4,4)) |> drawCM(title = "Binary optimum")

```

Some words here describing the prediction histograms.

```{r}
#| label: fig-pred_hist_low
#| fig-cap: "Illustration of prediction accuracy" 
#| fig-cap-location: top

suppressWarnings(
predictionHistogram(data = aligned,
                    title = NA,
                    show = "actual",
                    cap = TRUE,
                    cuts = c(0,2.4,4),
                    hist_params = list(freq = TRUE, right = FALSE, labels = TRUE))
)

# note that this aligns with

# > table(aligned$GRADEGPA[aligned$pred < 2.4])

#  0 0.7   1 1.3 1.7   2 2.3 2.7   3 3.3 3.7   4 
#  69  30  27  21  32  41  32  43  66  50  35  61 

```



```{r}
#| label: fig-pred_hist_high
#| fig-cap: "Illustration of prediction accuracy" 
#| fig-cap-location: top

suppressWarnings(
predictionHistogram(data = aligned,
                    title = NA,
                    show = "predict",
                    cap = TRUE, 
                    cuts = c(0,2.4,4),
                    hist_params = list(freq = TRUE, right = FALSE, labels = TRUE))
)

```



This model also shows a high precision point at a predicted GPA of 0.9.  These candidates would predictably fail, and could be considered an admissions mistake.  However, there are only eight people so identified (out of 1,759), suggesting nearly all students admitted to the U have a strong expectation of passing initial math courses.  

```{r}
#| label: fig-fail_cm
#| fig-cap: "Confusion matrix for 'admission mistake' candidates" 
#| fig-cap-location: top

createCM(aligned, cuts = c(0,0.9,4)) |> drawCM(title = "Predictably failed with good precision")

```




## Variable importance  

```{r}
#| label: fig-importance
#| fig-cap: "Variable importance to the model"
#| fig-cap-location: top
#| fig-height: 5


print(plot(
varImp(model_list[["GRADEGPA"]]),
15,
main = "Relative variable importance"

))


```


```{r}

# Box plots of z-scores vs grades

zBox <- function(data, column, 
                 box_params = list(),
                 axis_params = list(),
                 title_params = list(),
                 title = NA
                 ){
  
  thePlot <- boxplot(get(column) ~ GRADEGPA,
                     data = data,
                     ylab = "",
                     xlab = "",
                     xaxt = "n",
                     yaxt = "n",
                     outline = FALSE
                     )
  
  rect(xleft = par("usr")[1],
       ybottom = par("usr")[3],
       xright = par("usr")[2],
       ytop = par("usr")[4],
       col = "gray95")

  grid(col = "gray90", lty =1, lwd = 2)
  
  default_box_params <- list( 
    formula = as.formula(paste(column, " ~ GRADEGPA")),
    data = data,
    ylab = "",
    xlab = "",
    outline = FALSE,
    col = "lavender",
    add = TRUE
    )
  
  box_params <- modifyList(default_box_params, box_params)
  do.call(boxplot, box_params)
  
  default_axis_params <- list(
    side = c(1,2),
    font = c(1,1),
    cex = c(1.2,1.2),
    line = c(2.32,2.31),
    text = c("Math grade", "Z-score")
  )

  axis_params <- modifyList(default_axis_params, axis_params)
  do.call(mtext, axis_params)
    
  default_title_params <- list(
        side = c(3,3),
    font = c(2,3),
    cex = c(2.05, 1.25),
    line = c(1.69, 0.5),
    text = c(title, "Z-score distance to prior median successful student")
  )
  
  title_params <- modifyList(default_title_params, title_params)
  do.call(mtext, title_params)

  
}

```


```{r}
#| label: fig-gpa
#| fig-cap: "#1 Variable: High school GPA"
#| fig-cap-location: top

zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "HSGPA",
     title = "High school GPA",
     axis_params = list(text = c("Math grade", "High school GPA")),
     box_params = list(col="lightsteelblue1", las=1),
     title_params = list(text = c("High school GPA","") ))

```

Put some words here describing the strong overlap


```{r}
#| label: fig-gpa_z
#| fig-cap: "#2 Variable: High school GPA z-score"
#| fig-cap-location: top

zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "HSGPA.z",
     title = "High school GPA",
     box_params = list(col="lightsteelblue2", las=1))

```


```{r}
#| label: fig-dist
#| fig-cap: "#3 Variable: Combined z-score distance"
#| fig-cap-location: top

zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "dist",
     title = "Combined distance",
     box_params = list(col="beige"))
     
     
```     

```{r}
#| label: fig-act_z
#| fig-cap: "#4 Variable: ACT Math z-score"
#| fig-cap-location: top

zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "ACTMATH.z",
     title = "ACT Math",
     box_params = list(col="pink2", las=1))


```

```{r}
#| label: fig-act
#| fig-cap: "#5 Variable: ACT Math"
#| fig-cap-location: top

zBox(data = data_list[["GRADEGPA"]][["training"]], 
     column = "ACTMATH",
     title = "ACT Math",
     axis_params = list(text = c("Math grade", "ACT Math")),
     box_params = list(col="pink1", las=1),
     title_params = list(text = c("ACT math", "")))


```


```{r}
#| label: fig-yr_diff
#| fig-cap: "#6 Variable: Years to initial math course"
#| fig-cap-location: top

# Density plot
xLim <- c(-0.5,4) #   range(aligned[,"yr_diff"])

success_ref <- aligned[aligned$cut_1.7 == "[0,1.7)","yr_diff"]
fail_ref <- aligned[aligned$cut_1.7 == "[1.7,4]","yr_diff"]

plot(density(success_ref), 
     col = "seagreen4", 
     lwd = 2, 
     xlim = xLim,
     main = "", 
     xlab = "Yrs to initial math course")
lines(density(fail_ref), 
      col = "mediumorchid4", 
      lwd = 2)
legend("topright", legend = c(">=1.7", "<1.7"), 
       col = c("seagreen4", "mediumorchid4"), lwd = 2)

mtext(text = "Years to initial math course",
      side = 3,
      line = 1.7,
      font = 2,
      cex = 2,
      )

mtext(text = "broken out by GPA",
      side = 3,
      line = 0.3,
      font = 1,
      cex = 1.3,
      )

```

There is a small tendency for people who fail to take the initial math course after the first year. (Most students (`r cut(aligned$yr_diff, breaks = c(-100,0,1, 100), right = FALSE ) |> table(useNA="always") |> proportions() |> round(2) |> (\(x){x*100})() |> (\(x){x["[0,1)"]})()`% in 2024) take their initial math course in the first year.)

# REFERENCES

```{r}
#| label: fig-guide
#| fig-cap: "How to interpret a confusion matrix"
#| fig-cap-location: top

drawGuide()

```

## Training data

```{r eval=TRUE}


skim(data_list[["GRADEGPA"]][["training"]])


```

