---
title: "Deeper dive into vH2 model"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vH2"
  target_classes: 
    - GRADEGPA
    - grade_quad
output:
  html_document:
    keep_md: true
---

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great!


source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

source(here::here("Sandbox", "Prediction map sandbox.R"))
source(here::here("Sandbox", "alignPrediction sandbox.R"))

```



```{r}

##########
## LOAD ##
##########

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", params$version, " ", params$target_classes, " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  params$target_classes
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", params$version, " ", params$target_classes, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  params$target_classes
)

```


```{r eval = FALSE}


# This will change based on the model prep.
# I might just want to delete this for a template "results"
# report.

## RAW DATA ##
ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))


##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

#
# identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
                          )
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
#  SATMATH, # leaving out SAT because it's a smaller proportion
#  SATVERBAL,
#  SATWRTG        
)

popCourses <- merge(popCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

popCourses <- merge(popCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
popCourses$cohort_year <- year(popCourses$COHORT_DT)

# Class year
popCourses$class_year <- year(popCourses$EOTDATE)

# Year to class
popCourses$yr_diff <- time_length(interval(popCourses$COHORT_DT, popCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
popCourses$season <- season_map[substr(popCourses$TERM, 4, 4)]


# print("Prep complete")

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

popCourses$cleanGrade <- clean_grade(popCourses$GRADE) 

popCourses$cleanGrade <- factor(popCourses$cleanGrade,
                                levels = c(
                                  "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                  "C", "C_minus", "D",   
                                  "E", "EU", "W", "I", "CR",  "NC", "V", 
                                  "missing", NA
                                )
)

# RARE AND UNUSUAL VALUES

rareGrades <- popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
# popCourses$wdraw_binary <- NA
# popCourses$wdraw_binary <- ifelse(popCourses$GRADE == "W","withdraw","not_withdraw")


# Combine grades
# BINARY
popCourses$grade_binary <- NA
hiFilter <- popCourses$GRADE %in% c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
loFilter <- popCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
popCourses$grade_binary[hiFilter] <- 1
popCourses$grade_binary[loFilter] <- 0

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
popCourses$grade_trinary <- NA
hiFilter <- popCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- popCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- popCourses$GRADE %in% c( "D+", "D", "D-", "E" ) 
popCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
popCourses$grade_trinary[medFilter] <- 1 # "med_grade"
popCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD (EXPANDED FAILURE)
popCourses$grade_quad <- NA
hiFilter <- popCourses$GRADE %in% c("A")
medPlusFilter <- popCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- popCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- popCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU", "W" ) 
popCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
popCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
popCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
popCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

# print("Transformations complete")


```

```{r}

## PREDICTIONS ##

thePredictions <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  print(x)
  print(table(missingFilter))
}
  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions) <- params$target_classes

```


### Models  

*MODEL 1: binary* 

  - Target buckets:   
    1: "A", "A-", "B+", "B", "B-", "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 
  
*MODEL 2: trinary*  

  - Target buckets:     
    2: "A", "A-", "B+", "B", "B-"  
    1: "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 

*MODEL 3: quad* 

  - Target buckets:   
    3: "A"  
    2: "A-", "B+", "B", "B-"  
    1: "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 

*MODEL 4: grade GPA* 

  - Targets: 
    4, 3.7, 3.3, 3, 2.7, 2.3, 2, 1.7, 1.3, 1, 0.7, 0

### Predictors  

`r colnames(data_list[[1]][["training"]])[-1]`. 

"load" is the credit load per student per term. 
"yr_diff" is the time (expressed as fractions of a year) between the cohort date and the end of term date for the course.  

The other variables are either self-explanatory or taken directly from OBIA tables.  

# Results

```{r}

theResults <- lapply(params$target_classes, function(x) {
  
  print(x)
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  print(x)
  print(table(missingFilter))
}
  
  postResample(pred = thePredictions[[x]],
                  obs = data_list[[x]][["testing"]][[x]][missingFilter]
  
               )
})
names(theResults) <- params$target_classes

# I'd like to normalize the errors with the range and standard deviation

# but it doesn't change that these aren't very good models

# normalized errors

normalized_metrics <- function(pred, obs) {
  rmse <- sqrt(mean((pred - obs)^2))
  mae  <- mean(abs(pred - obs))
  r2   <- cor(pred, obs)^2
  
  range_y <- diff(range(obs, na.rm = TRUE))
  sd_y    <- sd(obs, na.rm = TRUE)
  
  data.frame(
    RMSE = rmse,
    MAE  = mae,
    R2   = r2,
    NRMSE_range = rmse / range_y,
    NMAE_range  = mae  / range_y,
    NRMSE_sd    = rmse / sd_y,
    NMAE_sd     = mae  / sd_y
  )
}  
  
normResults <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  print(x)
  print(table(missingFilter))
}
  
  normalized_metrics(
    pred = thePredictions[[x]],
    obs = data_list[[x]][["testing"]][[x]][missingFilter]
  )
  
})  
names(normResults) <- params$target_classes

library(kableExtra)

normResults |>
  (\(x){do.call(rbind,x)})() |>
  (\(x){round(x,2)})()|>
    kbl(
    caption = "Normalized Regression Results Across Grading Buckets",
    align = "c",
    col.names = c("",
      "RMSE", "MAE", "R-sq", 
      "NRMSE (Range)", "NMAE (Range)", 
      "NRMSE (SD)", "NMAE (SD)"
    )
  ) |>
  kable_classic(full_width = FALSE, html_font = "Calibri") |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>  # header
  row_spec(1:length(params$target_classes), background = c("#ffffff", "#f9f9f9")) |> # alternating rows
  add_header_above(c(" " = 1, "Raw Metrics" = 3, "Normalized Metrics" = 4))

```


```{r}

regression_diagnostics <- function(pred, reference, plot_title = "Predicted vs Actual") {
  if (length(pred) != length(reference)) {
    stop("`pred` and `reference` must be the same length.")
  }
  
  incoming.par <- par(mfrow= c(2,1), mar = c(3,2.5,0,1), oma = c(1.5,1.5,2.25,0))
  
  on.exit(par(incoming.par))
  
  # Compute metrics
  rmse <- sqrt(mean((pred - reference)^2))
  mae  <- mean(abs(pred - reference))
  r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2) 
  
  # Print metrics
#  cat("Regression diagnostics:\n")
#  cat(sprintf("RMSE: %.4f\n", rmse))
#  cat(sprintf("MAE:  %.4f\n", mae))
#  cat(sprintf("R-squared (variance decomposition): %.4f\n", r2))

  
  # Scatter plot
  plot(reference, pred,
      # xlab = "Actual",
      # ylab = "Predicted",
       # main = plot_title,
       pch = 19, col = "steelblue",
       las = 1)
  abline(a = 0, b = 1, col = "red", lwd = 2)  # 45-degree reference line
  grid()
  
  # boxplot
  bp <- boxplot(pred ~ reference,
        # xlab = "Actual",
        # ylab = "Predicted",
         # main = plot_title,
         col = "lightblue",
         border = "steelblue",
         las=1)
  
    # Add a red reference line for each box
        for (i in seq_along(bp$names)) {
          actual_val <- as.numeric(bp$names[i])
          segments(x0 = i - 0.4, y0 = actual_val,
                   x1 = i + 0.4, y1 = actual_val,
                   col = "red", lwd = 2)
        }
  
mtext(side = 3, line =0.75, text = plot_title, cex = 1.5, font = 2, outer = TRUE)  
  
mtext(side = 2, line =0.25, text = "Predicted", cex = 1.25, font = 1, outer = TRUE)  

mtext(side = 1, line =0.25, text = "Actual", cex = 1.25, font = 1, outer = TRUE)  

  # Return metrics invisibly as a list
  invisible(list(RMSE = rmse, MAE = mae, R2 = r2))
}


invisible(lapply(params$target_classes, function(x){
  
      # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  
regression_diagnostics(pred = thePredictions[[x]],
                       reference = data_list[[x]][["testing"]][[x]][missingFilter],
                       plot_title = x)
})
)

```



```{r}

aligned <- alignPrediction(prediction_list = thePredictions, reference_list = data_list)

# ACTUAL vs PREDICTED HEATMAP
predictionMap(data = aligned[["GRADEGPA"]], 
              plot_title = "Predicted vs Actual for 2024, all courses")



```

You know, what 

# Variable importance

```{r fig.height=5}

invisible(
  lapply(params$target_classes, function(x){

  print(plot(
    varImp(model_list[[x]]),
    15,
    main = x
       )  
  )
})
)

```

```{r eval=FALSE}

for(x in params$target_classes) {
  
    print(plot(
    varImp(model_list[[x]]),
    15,
    main = x
       )  
    )
}

```

# Confusion matrix

```{r}

GPA_grade_levels <- unique(data_list[["GRADEGPA"]][["testing"]][["GRADEGPA"]]) |> 
  (\(x){ x[order(x)]})()
intermediate_steps <- GPA_grade_levels |> (\(x){ x[order(x)]})() |> diff() |> (\(x){x*0.5})()
GRADEGPA_thresholds <- GPA_grade_levels |> 
  (\(x){ x[order(x)]})() |>
  (\(x){x+intermediate_steps})()
GRADEGPA_thresholds[GRADEGPA_thresholds == max(GRADEGPA_thresholds)] <- 10 # upper limit is high so any high predictions are leveled back as "A"

GRADEGPA_thresholds <- c(-10, GRADEGPA_thresholds) # lower limit is low so any low prediction is leveled back to "F"

quad_grade_levels <- unique(data_list[["grade_quad"]][["testing"]][["grade_quad"]]) |> 
  (\(x){ x[order(x)]})()
intermediate_steps <- quad_grade_levels |> (\(x){ x[order(x)]})() |> diff() |> (\(x){x*0.5})()
grade_quad_thresholds <- quad_grade_levels |> 
  (\(x){ x[order(x)]})() |>
  (\(x){x+intermediate_steps})()
grade_quad_thresholds[grade_quad_thresholds == max(grade_quad_thresholds)] <- 10 # upper limit is high so any high predictions are leveled back as "A"

grade_quad_thresholds <- c(-10, grade_quad_thresholds)

missingInTraining <- function(target) {
  
  # Removing courses that are in testing but not in training 
  
    # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[target]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[target]][["testing"]][[courseColumn]],data_list[[target]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[target]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[target]][["testing"]][[courseColumn]] %in% missing
}
  return(missingFilter)
  
}

  
createConfusionMatrix <- function(pred, reference, thresholds = NA, ...){
  
  # set thresholds if not provided
  if(any(is.na(thresholds))) {
  grade_levels <- unique(reference)
  pred_levels <- sapply(pred, function(x){
    grade_levels[which.min(abs(x-grade_levels))]
  })
  
  }
  
  # convert to discrete values
  discrete_predictions <- cut(pred, thresholds, include.lowest = TRUE, right = FALSE)
  discrete_references <- cut(reference, thresholds, include.lowest=TRUE, right = FALSE)
  
  # create uniform factor levels

all_levels <- sort(unique(c(as.character(discrete_predictions), 
                            as.character(discrete_references))))
  
discrete_predictions <- factor(discrete_predictions, levels = all_levels)
discrete_references <- factor(discrete_references, levels = all_levels)

  # create confusion matrix
  
  cm <- confusionMatrix(discrete_predictions, discrete_references, mode = "everything", ... )
  
  return(cm)
  
}

gradeMissing <- missingInTraining("GRADEGPA")  
GRADEGPA_cm <- createConfusionMatrix(pred = thePredictions[["GRADEGPA"]],
                                      reference = data_list[["GRADEGPA"]][["testing"]][["GRADEGPA"]][gradeMissing],
                                      thresholds = GRADEGPA_thresholds
                                      )

drawCM(GRADEGPA_cm) # fixed!

quadMissing <- missingInTraining("grade_quad")  
quad_cm <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = grade_quad_thresholds
                                      )
drawCM(quad_cm, nameMatrix = cbind(original = c("[-10,0.5]", "(0.5,1.5]", "(1.5,2.5]", "(2.5,10]"),
                                display =  c("[-10,0.5]", "(0.5,1.5]", "(1.5,2.5]", "(2.5,10]"))
)


# let's turn that into a binary"
# to make this work, you'll need to modify the "cut" to "right = TRUE"

quad_cm_bin <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = c(-10,2,10),
                                     positive = "[-10,2]"
                                      )
drawCM(quad_cm_bin, nameMatrix = cbind(original = c("[-10,2]", "(2,10]"),
                                display =  c("[-10,2]", "(2,10]"))
)

# This one had an EXCELLENT kappa value-- like 0.6

# starting to look a lot better
# Maybe I use different thresholds for the prediction as the reference?
# Why not?

# and let's use better thresholds --- interested in flunking, here--

quad_cm_bin <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = c(-10,1,10),
                                     positive = "[-10,1)"
                                      )
drawCM(quad_cm_bin, nameMatrix = cbind(original = c("[-10,1)", "[1,10]"),
                                display =  c("[-10,1)", "[1,10]"))
)


quad_cm_bin <- createConfusionMatrix(pred = thePredictions[["grade_quad"]],
                                      reference = data_list[["grade_quad"]][["testing"]][["grade_quad"]][quadMissing],
                                      thresholds = c(-10,2,10),
                                     positive = "[-10,2)"
                                      )
drawCM(quad_cm_bin, nameMatrix = cbind(original = c("[-10,2)", "[2,10]"),
                                display =  c("[-10,2)", "[2,10)"))
)

# ...and I need to update drawCM again, from vC6 to include "grepl(any....)"


# I should cough up my best one, but unfortunately it's not the most useful.



``` 


```{r}

drawGuide()

```

# Training data

Presentation order: 

```{r}



lapply(params$target_classes, function(x) {
  
  print(x)

  skim(data_list[[x]][["training"]])
  
})


```

