---
title: "Evaluate Grade Decision Trees Over Time vJ0"
author: "Bill Prisbrey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vJ0"
  target_classes: 
    - GRADEGPA
    - grade_quad
output:
  html_document:
    keep_md: true
---



```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great!

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########

# Strikes me that I want one list per target class, with multiple years per

# Build file paths
testYears <- 2006:2025
grid <- expand.grid(params$version, testYears,  params$target_classes)

model_files <- here::here("Models", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " model.rds"))

# Read this into a list of lists by target class

model_list <- setNames(
  lapply(params$target_classes, function(tClass){
    files <- model_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  params$target_classes
)



# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(params$target_classes, function(tClass){
    files <- data_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  params$target_classes
)


```

```{r}

# Let's check a few things

lapply(data_list, function(tClass){
  lapply(tClass, function(tYr){
  
    print(paste("training", nrow(tYr[["training"]]), sep = ": "))
    print(paste("testing", nrow(tYr[["testing"]]), sep = ": "))

  })
  
})

# oops
# I'd better just re-do it and do it right

lapply(data_list, function(tClass){
  lapply(tClass, function(tYr){
  
    print(paste("training", levels(tYr[["training"]][["season"]]), sep = ": "))
    print(paste("testing", levels(tYr[["testing"]][["season"]] ), sep = ": "))

  })
  
})

``` 


```{r}

## PREDICTIONS ##

predictions_over_time <- lapply(params$target_classes, function(tClass){
  perClass <- lapply(as.character(testYears), function(tYr){
    
#  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
    
    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  
            
      
    # predict 
    predict(model_list[[tClass]][[tYr]],
            newdata = data_list[[tClass]][[tYr]][["testing"]][missing,]
            )
  })
  names(perClass) <- testYears
  return(perClass)
})
names(predictions_over_time) <- params$target_classes


```


### Models  

*MODEL 1: binary* 

  - Target buckets:   
    1: "A", "A-", "B+", "B", "B-", "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 
  
*MODEL 2: trinary*  

  - Target buckets:     
    2: "A", "A-", "B+", "B", "B-"  
    1: "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 

*MODEL 3: quad* 

  - Target buckets:   
    3: "A"  
    2: "A-", "B+", "B", "B-"  
    1: "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 

*MODEL 4: grade GPA* 

  - Targets: 
    4, 3.7, 3.3, 3, 2.7, 2.3, 2, 1.7, 1.3, 1, 0.7, 0

### Predictors  

`r colnames(data_list[[1]][["training"]])[-1]`. 

"load" is the credit load per student per term. 
"yr_diff" is the time (expressed as fractions of a year) between the cohort date and the end of term date for the course.  

The other variables are either self-explanatory or taken directly from OBIA tables.  

# Results


```{r}

theResults <- lapply(params$target_classes, function(tClass){
  
  perClass <- lapply(as.character(testYears), function(tYr){

        # Identify discrepancies in the courses between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][[courseColumn]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

       postResample(pred = predictions_over_time[[tClass]][[tYr]],
                  obs = data_list[[tClass]][[tYr]][["testing"]][[tClass]][missingCourseFilter]
               ) 
    
        
  })
  names(perClass) <- testYears
  return(perClass)
})
  
names(theResults) <- params$target_classes

```



```{r eval = FALSE}

# if I ever care to develop the normalized results

# normalized errors # I don't think I care

normalized_metrics <- function(pred, obs) {
  rmse <- sqrt(mean((pred - obs)^2))
  mae  <- mean(abs(pred - obs))
  r2   <- cor(pred, obs)^2
  
  range_y <- diff(range(obs, na.rm = TRUE))
  sd_y    <- sd(obs, na.rm = TRUE)
  
  data.frame(
    RMSE = rmse,
    MAE  = mae,
    R2   = r2,
    NRMSE_range = rmse / range_y,
    NMAE_range  = mae  / range_y,
    NRMSE_sd    = rmse / sd_y,
    NMAE_sd     = mae  / sd_y
  )
}  



  
normResults <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  print(x)
  print(table(missingCourseFilter))
}
  
  normalized_metrics(
    pred = thePredictions[[x]],
    obs = data_list[[x]][["testing"]][[x]][missingCourseFilter]
  )
  
})  
names(normResults) <- params$target_classes

library(kableExtra)

normResults |>
  (\(x){do.call(rbind,x)})() |>
  (\(x){round(x,2)})()|>
    kbl(
    caption = "Normalized Regression Results Across Grading Buckets",
    align = "c",
    col.names = c("",
      "RMSE", "MAE", "R-sq", 
      "NRMSE (Range)", "NMAE (Range)", 
      "NRMSE (SD)", "NMAE (SD)"
    )
  ) |>
  kable_classic(full_width = FALSE, html_font = "Calibri") |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>  # header
  row_spec(1:length(params$target_classes), background = c("#ffffff", "#f9f9f9")) |> # alternating rows
  add_header_above(c(" " = 1, "Raw Metrics" = 3, "Normalized Metrics" = 4))

```



```{r eval=FALSE}

theResults <- lapply(params$target_classes, function(x) {
  
  print(x)
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  print(x)
  print(table(missingCourseFilter))
}
  
  postResample(pred = thePredictions[[x]],
                  obs = data_list[[x]][["testing"]][[x]][missingCourseFilter]
  
               )
})
names(theResults) <- params$target_classes

# I'd like to normalize the errors with the range and standard deviation

# but it doesn't change that these aren't very good models

# normalized errors

normalized_metrics <- function(pred, obs) {
  rmse <- sqrt(mean((pred - obs)^2))
  mae  <- mean(abs(pred - obs))
  r2   <- cor(pred, obs)^2
  
  range_y <- diff(range(obs, na.rm = TRUE))
  sd_y    <- sd(obs, na.rm = TRUE)
  
  data.frame(
    RMSE = rmse,
    MAE  = mae,
    R2   = r2,
    NRMSE_range = rmse / range_y,
    NMAE_range  = mae  / range_y,
    NRMSE_sd    = rmse / sd_y,
    NMAE_sd     = mae  / sd_y
  )
}  
  
normResults <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  print(x)
  print(table(missingCourseFilter))
}
  
  normalized_metrics(
    pred = thePredictions[[x]],
    obs = data_list[[x]][["testing"]][[x]][missingCourseFilter]
  )
  
})  
names(normResults) <- params$target_classes

library(kableExtra)

normResults |>
  (\(x){do.call(rbind,x)})() |>
  (\(x){round(x,2)})()|>
    kbl(
    caption = "Normalized Regression Results Across Grading Buckets",
    align = "c",
    col.names = c("",
      "RMSE", "MAE", "R-sq", 
      "NRMSE (Range)", "NMAE (Range)", 
      "NRMSE (SD)", "NMAE (SD)"
    )
  ) |>
  kable_classic(full_width = FALSE, html_font = "Calibri") |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>  # header
  row_spec(1:length(params$target_classes), background = c("#ffffff", "#f9f9f9")) |> # alternating rows
  add_header_above(c(" " = 1, "Raw Metrics" = 3, "Normalized Metrics" = 4))

```


```{r eval=FALSE}

# boxplots per model if I ever care for them

regression_diagnostics <- function(pred, reference, plot_title = "Predicted vs Actual") {
  if (length(pred) != length(reference)) {
    stop("`pred` and `reference` must be the same length.")
  }
  
  incoming.par <- par(mfrow= c(2,1), mar = c(3,2.5,0,1), oma = c(1.5,1.5,2.25,0))
  
  on.exit(par(incoming.par))
  
  # Compute metrics
  rmse <- sqrt(mean((pred - reference)^2))
  mae  <- mean(abs(pred - reference))
  r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2) 
  
  # Print metrics
#  cat("Regression diagnostics:\n")
#  cat(sprintf("RMSE: %.4f\n", rmse))
#  cat(sprintf("MAE:  %.4f\n", mae))
#  cat(sprintf("R-squared (variance decomposition): %.4f\n", r2))

  
  # Scatter plot
  plot(reference, pred,
      # xlab = "Actual",
      # ylab = "Predicted",
       # main = plot_title,
       pch = 19, col = "steelblue",
       las = 1)
  abline(a = 0, b = 1, col = "red", lwd = 2)  # 45-degree reference line
  grid()
  
  # boxplot
  bp <- boxplot(pred ~ reference,
        # xlab = "Actual",
        # ylab = "Predicted",
         # main = plot_title,
         col = "lightblue",
         border = "steelblue",
         las=1)
  
    # Add a red reference line for each box
        for (i in seq_along(bp$names)) {
          actual_val <- as.numeric(bp$names[i])
          segments(x0 = i - 0.4, y0 = actual_val,
                   x1 = i + 0.4, y1 = actual_val,
                   col = "red", lwd = 2)
        }
  
mtext(side = 3, line =0.75, text = plot_title, cex = 1.5, font = 2, outer = TRUE)  
  
mtext(side = 2, line =0.25, text = "Predicted", cex = 1.25, font = 1, outer = TRUE)  

mtext(side = 1, line =0.25, text = "Actual", cex = 1.25, font = 1, outer = TRUE)  

  # Return metrics invisibly as a list
  invisible(list(RMSE = rmse, MAE = mae, R2 = r2))
}


invisible(lapply(params$target_classes, function(x){
  
      # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  
regression_diagnostics(pred = thePredictions[[x]],
                       reference = data_list[[x]][["testing"]][[x]][missingCourseFilter],
                       plot_title = x)
})
)

```

```{r}

# Success over time

resultsDF <- lapply(theResults, function(x){
  
  do.call(rbind, x)
  
})
names(resultsDF) <- names(theResults)


```

```{r}

# Plot success over time

plot(x = as.numeric(row.names(resultsDF[["GRADEGPA"]])),
     y = resultsDF[["GRADEGPA"]][,"Rsquared"],
     type = "b",
     lty = 2,
     col = "red")

# needs help

# but it's something


``` 



# Variable importance

```{r}

# extract variable importance
# re-shape into a data frame per target class with one column per year

importance_over_time <- lapply(model_list, function(tClass) {
  
     # Extract importance
      impList <- lapply(tClass, function(tYr){
        
        imp <- varImp(tYr)$importance
        imp$var <- row.names(imp)
        return(imp)
      })  
      
    # Re-name 'overall' column  
      
    impList <- lapply(names(impList), function(yearName){
      
      colnames(impList[[yearName]])[colnames(impList[[yearName]]) == "Overall"] <- yearName
      return(impList[[yearName]])
    })  
      
  # Collapse into one data frame
    
    impOverTime <- Reduce(function(m1,m2){merge(m1,m2, by = "var", all = TRUE)}, impList)
    impOverTime <- impOverTime[order(impOverTime[,2], decreasing = TRUE),] # descending order by the first year of data
    
      return(impOverTime)
})


```

```{r}

# Plot variable importance
# This needs some work

varSelection <- which(importance_over_time[["GRADEGPA"]][["var"]] %in% c("ACTMATH", "ACTENGL", "ACTCOMP", "HSGPA" ))

plot(50,
     ylim = range(importance_over_time[["GRADEGPA"]][varSelection,as.character(2018:2020)] ),
     xlim = c(1,3),
  type = "n"
)

invisible(
lapply(varSelection, function(x){
lines(
  as.numeric(importance_over_time[["GRADEGPA"]][x, as.character(2018:2020)])
      )
})
)


```





```{r fig.height=5, eval=FALSE}

# This shows each year --- I am preferring the line graph over time 

invisible(
  lapply(params$target_classes, function(tClass){
    
    lapply(testYears, function(tYear){
       print(plot(  
           varImp(model_list[[tClass]][[as.character(tYear)]]),
           15, main = paste(tClass, tYear, sep = ": ")
       ))
           })
})
)



```
