---
title: "Evaluate Course Decision Trees vB0"
author: "Bill Prisbrey"
date: "2025-10-31"
params: 
  version: "vB0"
  target_classes: 
    - course_target
output:
  html_document:
    keep_md: true
---

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great! # NOT great!  I just need to make this available on my personal Github, or find a differetn way to source it. 

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```




```{r}

##########
## LOAD ##
##########

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree Courses ", params$version, " ", params$target_classes, " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  params$target_classes
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree Courses ", params$version, " ", params$target_classes, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  params$target_classes
)

```


```{r}

## PREDICTIONS ##

thePredictions <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  print(x)
  print(table(missingFilter))
}
  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions) <- params$target_classes

```

### Models  

Only one model. I could delete this section 

### Predictors  

`r colnames(data_list[[1]][["training"]])[-1]`. 

"load" is the credit load per student per term. 
"yr_diff" is the time (expressed as fractions of a year) between the cohort date and the end of term date for the course.  

The other variables are either self-explanatory or taken directly from OBIA tables.  

# Results

Develop and report the confusion matrices 

```{r}

cm <- confusionMatrix(data =  thePredictions[[params$target_classes]], reference =data_list[["course_target"]][["testing"]][["course_target"]] , mode = "everything" )

drawCM(cm, nameMatrix = cbind("original" = c(
  "MATH_1010", "MATH_1030", "MATH_1050", "MATH_1070", "MATH_1080", "MATH_1090", "MATH_1210", "MATH_1220", "other" 
  
) ,"display" = c(
  "1010", "1030", "1050", "1070", "1080", "1090", "1210", "1220", "other"
  
) ))


drawCM(cm, 
       basis = "recall",
       percentage = TRUE,
       metricsClass = "1050",
       nameMatrix = cbind("original" = c(
  "MATH_1010", "MATH_1030", "MATH_1050", "MATH_1070", "MATH_1080", "MATH_1090", "MATH_1210", "MATH_1220", "other" 
  
) ,"display" = c(
  "1010", "1030", "1050", "1070", "1080", "1090", "1210", "1220", "other"
  
) ))

# precision color is DEFINITELY screwed up  

```


# Variable importance

```{r fig.height=5}

invisible(
  lapply(params$target_classes, function(x){

  print(plot(
    varImp(model_list[[x]]),
    15,
    main = x
       )  
  )
})
)

```


