---
title: "Evaluate Grade Decision Trees Over Time vJ0"
author: "Bill Prisbrey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vJ0"
  target_classes: 
    - GRADEGPA
    - grade_quad
output:
  html_document:
    keep_md: true
    toc: true
---



```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########

# Strikes me that I want one list per target class, with multiple years per

# Build file paths
testYears <- 2007:2025
grid <- expand.grid(params$version, testYears,  params$target_classes)

model_files <- here::here("Models", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " model.rds"))

# Read this into a list of lists by target class

model_list <- setNames(
  lapply(params$target_classes, function(tClass){
    files <- model_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  params$target_classes
)



# Build file paths
data_files <- here::here("Data", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(params$target_classes, function(tClass){
    files <- data_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  params$target_classes
)


```

```{r eval=FALSE}

# Let's check a few things

lapply(data_list, function(tClass){
  lapply(tClass, function(tYr){
  
    print(paste("training", nrow(tYr[["training"]]), sep = ": "))
    print(paste("testing", nrow(tYr[["testing"]]), sep = ": "))
invisible(NULL)
  })
})

# It did it--- not the format I was expecting
# And 2006 went wrong for some reason, but I can filter it out.  Oh, because I didn't over-write it from the first pass.  I could delete it.  

for(tClass in data_list){
  for(tYr in tClass){
       print(paste("training:", levels(factor(tYr[["training"]][["season"]]))))
    print(paste("testing:", levels(factor(tYr[["testing"]][["season"]]))), collapse = ", ")
  }
}

# no fall for 2025 yet

``` 


```{r}

## PREDICTIONS ##

predictions_over_time <- lapply(params$target_classes, function(tClass){
  perClass <- lapply(as.character(testYears), function(tYr){
    
#  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
    
    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  
            
      
    # predict 
    predict(model_list[[tClass]][[tYr]],
            newdata = data_list[[tClass]][[tYr]][["testing"]][missing,]
            )
  })
  names(perClass) <- testYears
  return(perClass)
})
names(predictions_over_time) <- params$target_classes

# predictions_over_time[["GRADEGPA"]][["2024"]] |> hist() # looks o.k.



```


```{r}

theResults <- lapply(params$target_classes, function(tClass){
  
  perClass <- lapply(as.character(testYears), function(tYr){

    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  

       postResample(pred = predictions_over_time[[tClass]][[tYr]],
                  obs = data_list[[tClass]][[tYr]][["testing"]][[tClass]][missingCourseFilter]
               ) 
    
        
  })
  names(perClass) <- testYears
  return(perClass)
})
  
names(theResults) <- params$target_classes

```



```{r}

# Success over time

resultsDF <- lapply(theResults, function(x){
  
  do.call(rbind, x)
  
})


```


### SUMMARY 

### METHODOLOGY 

All math courses taken by first time freshmen were identified. Each student's first math courses were identified (or multiple courses if taken in the same term.)   Grade performance for these math courses was queried.  These are called the initial math courses.   

Predictive models were trained with package "xgboost" (extreme gradient boosting) with five-fold cross validation and a grid search pattern (nrounds = (100,200); max_depth(3,6); eta(0.05,0.2); colsample_bytree(0.8); subsample(0.8).)    

An expanding window of historical data was used, where one model was trained per year for nineteen models.  Each model incrementally added an additional year's data and predicted on the subsequent year.  For example, the predicted course selection for the year 2010 was trained on data from the years 2005 to 2009.  The predicted course selection for the year 2024 was trained on data from the years 2005 to 2023.  

A value for math course was predicted.  

#### Variables 

The variables included: 

*Demographics:* 

  * Age (less than 17 yrs old/17/18/19 or 20/greater than 21 years old)
  * Sex (male or female)
  * First generation status
  * Ethnicity 
  
*High school information and performance:*  

  * High school GPA 
  * Number of AP credits
  * Whether the high school was public or private   
  * Honors achievement

*Test scores:*    

  * ACT Math
  * ACT English
  * ACT Science
  * ACT comprehensive

*University information:*  

  * Residence status  
  * Pell grant eligibility  
  * Cohort year (2005 to 2024)
  * Class year (2005 to 2025)
  * Years from the cohort date (always fall of the year) until the first term with math courses 
  * Season (spring, summer, or fall)  

#### Filters  

The following filters were applied: 

 * Math labs are filtered out.  
 * Courses higher than 3000 level were removed. 
 * Students younger than 11 yrs old were removed. 
 * Withdraw and "rare" grades of V, I, NC, and CR were removed.  
 * Students with high school GPA values of zero were removed. 
 * Courses from students with unusually high credits taken that semester (greater than 99th percentile) were removed.  
 * Courses taken more than half a year prior to the cohort date (where all cohort dates are in the fall) were removed   
 * All records with incomplete information were removed.  

#### Transformations    

The following transformations were applied: 

  * AP credit values of "NA" were converted to zero.  
  * Math courses with ordered enrollment above the 95th percentile of cumulative total were combined as "other"  
  
#### Target values    

Each model predicted a value for a math course out of the following possibilities:  

MATH_1010, MATH_1050, MATH_1210, MATH_1030, MATH_1090, MATH_1070, MATH_1220, MATH_1080, MATH_1060, MATH_1310, MATH_2210, MATH_990, MATH_1100, MATH_1250 or "other".
 
The following math courses were combined into "other":  

MATH_1040, MATH_1311, MATH_1260, MATH_950, MATH_1270, MATH_1320, MATH_1321, MATH_1215, MATH_1170, MATH_2270, MATH_2250, MATH_1035, MATH_2000, MATH_1280, MATH_2200, MATH_1105, MATH_2271, MATH_2015, MATH_2280, MATH_2160, UC_1000, MATH_1180.
  
### DISCUSSION  

### CONCLUSIONS  

### RESULTS 


```{r}

# I want three plotly graphics, one per RMSE, R-squared, and MAE
# Each plot has one line per GRADEGPA and grade_quad

# Extract years from row names
years <- as.numeric(rownames(resultsDF$GRADEGPA))[-19]

# Create RMSE plot
plot_rmse <- plot_ly() |>
  add_trace(x = years, 
            y = resultsDF[["GRADEGPA"]][-19,"RMSE"], 
            name = "GPA Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
  add_trace(x = years, 
            y = resultsDF[["grade_quad"]][-19,"RMSE"], 
            name = "Quad Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
  layout(title = "RMSE Over Years",
         xaxis = list(title = "Year"),
         yaxis = list(title = "RMSE"))

# Create R-squared plot
plot_rsq <- plot_ly() |>
  add_trace(x = years, 
            y = resultsDF[["GRADEGPA"]][-19,"Rsquared"], 
            name = "GPA Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
  add_trace(x = years, 
            y = resultsDF[["grade_quad"]][-19,"Rsquared"], 
            name = "Quad Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
  layout(title = "R-squared Over Years",
         xaxis = list(title = "Year"),
         yaxis = list(title = "R-squared"))

# Create MAE plot
plot_mae <- plot_ly() |>
  add_trace(x = years, 
            y = resultsDF[["GRADEGPA"]][-19,"MAE"], 
            name = "GPA Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
  add_trace(x = years, 
            y = resultsDF[["grade_quad"]][-19,"MAE"], 
            name = "Quad Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
  layout(title = "MAE Over Years",
         xaxis = list(title = "Year"),
         yaxis = list(title = "MAE"))

# Display the plots
plot_rsq
plot_mae
plot_rmse


```



```{r}

# extract variable importance
# re-shape into a data frame per target class with one column per year

importance_over_time <- lapply(model_list, function(tClass) {
  
     # Extract importance
      impList <- lapply(tClass, function(tYr){
        
        imp <- varImp(tYr)$importance
        imp$var <- row.names(imp)
        return(imp)
      })  
      
    # Re-name 'overall' column  
      
    impList <- lapply(names(impList), function(yearName){
      
      colnames(impList[[yearName]])[colnames(impList[[yearName]]) == "Overall"] <- yearName
      return(impList[[yearName]])
    })  
      
  # Collapse into one data frame
    
    impOverTime <- Reduce(function(m1,m2){merge(m1,m2, by = "var", all = TRUE)}, impList)
    impOverTime <- impOverTime[order(impOverTime[,2], decreasing = TRUE),] # descending order by the first year of data
    
      return(impOverTime)
})


```





```{r}

# plotly version of importance_over_time

varFilter <- apply(importance_over_time[["GRADEGPA"]][,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time[["GRADEGPA"]][["var"]][varFilter & !is.na(varFilter)]

# varSelection <- importance_over_time[["GRADEGPA"]][["var"]][order(importance_over_time[["GRADEGPA"]][["2024"]], decreasing = TRUE)] |>
#  (\(x){x[2:10]})() # top 8 predictors in 2024

# Filter to selected variables
plotData <- importance_over_time[["GRADEGPA"]][
  importance_over_time[["GRADEGPA"]][,"var"] %in% varSelection, 
]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p |>
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p |>
  layout(
    title = 'Variable Importance Over Time (all variables, GPA)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```



```{r}

# plotly version of importance_over_time

varFilter <- apply(importance_over_time[["grade_quad"]][,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time[["grade_quad"]][["var"]][varFilter & !is.na(varFilter)]


# Filter to selected variables
plotData <- importance_over_time[["grade_quad"]][
  importance_over_time[["grade_quad"]][,"var"] %in% varSelection, 
]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p |>
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p |>
  layout(
    title = 'Variable Importance Over Time (all variables, Quad)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```



```{r}

# plotly version of importance_over_time, low importance variables only

varFilter <- apply(importance_over_time[["GRADEGPA"]][,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time[["GRADEGPA"]][["var"]][varFilter & !is.na(varFilter)]

varSelection <- varSelection[-which(varSelection %in% "HSGPA")]

# varSelection <- importance_over_time[["GRADEGPA"]][["var"]][order(importance_over_time[["GRADEGPA"]][["2024"]], decreasing = TRUE)] |>
#  (\(x){x[2:10]})() # top 8 predictors in 2024

# Filter to selected variables
plotData <- importance_over_time[["GRADEGPA"]][
  importance_over_time[["GRADEGPA"]][,"var"] %in% varSelection, 
]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p |>
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p |>
  layout(
    title = 'Variable Importance Over Time (low importance variables, GPA)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```


Discretize and create a confusion matrix

```{r}

# Discretize and create a couple of confusion matrices
# Should I do the high accuracy, or preserve the C+/D- barrier? 





```



```{r}

# Extract accuracy metrics per course

courses <- levels(data_list[["GRADEGPA"]][["2024"]][["testing"]]$course)

courseResults <- lapply(params$target_classes, function(tClass){
  
  perClass <- lapply(as.character(testYears), function(tYr){

    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  

# Iterate through each course at a time
  
perCourse <- lapply(courses, function(crse){ 

         course_filter <- data_list[[tClass]][[tYr]][["testing"]][["course"]] == crse
  
       postResample(pred = predictions_over_time[[tClass]][[tYr]][course_filter],
                  obs = data_list[[tClass]][[tYr]][["testing"]][[tClass]][missingCourseFilter][course_filter]
               ) 
    
        
  })

names(perCourse) <- courses
return(perCourse)

})

  names(perClass) <- testYears
  return(perClass)
})
  
names(courseResults) <- params$target_classes

# I could have built this into the lapply, but I'll work with it

gradeGPA_by_course <- lapply(courseResults[["GRADEGPA"]], function(x){ 
  
  do.call(rbind, x)
  })

# ok, I want a list of three dataframes, courses as row names and years as colnames

accuracy_metrics <- c("RMSE", "Rsquared", "MAE")  

course_over_time <- lapply(accuracy_metrics, function(metric){
  
  metricExtract <- lapply(gradeGPA_by_course, function(x){ 
    anExtract <- x[,metric]
    return(anExtract)
    })
  
  course_per_year <- do.call(cbind, metricExtract)
  
  return(course_per_year)
})

names(course_over_time) <- accuracy_metrics

# if I want, I could do "grade_quad"

``` 


```{r eval=FALSE}

# Check 2024 values for RMSE

course_counts2024 <- table(data_list[["GRADEGPA"]][["2024"]][["testing"]][["course"]])

weighted_rmse2024 <- sqrt(sum((gradeGPA_by_course[["2024"]][,"RMSE"]^2 *
                                 course_counts2024[rownames(gradeGPA_by_course[["2024"]])]), na.rm = TRUE) / 
        sum(course_counts2024[rownames(gradeGPA_by_course[["2024"]])], na.rm = TRUE))

cat("Overall RMSE:", theResults[["GRADEGPA"]][["2024"]]["RMSE"], "\n")
# Overall RMSE: 1.012924 
cat("Weighted RMSE from courses:", weighted_rmse2024, "\n")
# Weighted RMSE from courses: 1.012924 

# Looks like we're good!

```

```{r}

# plot accuracy per course over time  

plot_course_metric <- function(metric_df, metric_name = NULL) {
  library(plotly)
  
  # Get metric name from list if not provided
  if(is.null(metric_name)) {
    metric_name <- "Metric"
  }
  
  # Get courses and years
  courses <- rownames(metric_df)
  years <- as.numeric(colnames(metric_df))
  
  # Initialize plot
  p <- plot_ly()
  
  # Add a line for each course
  for(course in courses) {
    p <- p %>%
      add_trace(x = years,
                y = as.numeric(metric_df[course, ]),
                name = course,
                type = 'scatter',
                mode = 'lines+markers')
  }
  
  # Add layout
  p <- p %>%
    layout(title = paste(metric_name, "by Course Over Years"),
           xaxis = list(title = "Year"),
           yaxis = list(title = metric_name),
           hovermode = "closest")
  
  return(p)
}

# Usage:
plot_course_metric(course_over_time[["RMSE"]][,-which(colnames(course_over_time[["RMSE"]]) %in% "2025")], "RMSE")
plot_course_metric(course_over_time[["Rsquared"]][,-which(colnames(course_over_time[["Rsquared"]]) %in% "2025")], "R-squared")
plot_course_metric(course_over_time[["MAE"]][,-which(colnames(course_over_time[["MAE"]]) %in% "2025")], "MAE")



```



# Old stuff

Eventually delete everything below this

### Thoughts

So what is that 2024/2025 discontinuity?  All predictors jump from model trained on 2024, and 2025 accuracy falls.

And the ACT test scores are increasing in popularity.  Declining, then increasing.  

I think I'll try to at least knit a version of this report, and then predict the courses in overnight training. 

Or try to visualize some of the changes over time --

- How many students per year in the training set  
- Total AP credits / median ap credit for students that have them
- Some pdp plots. I guess I can't do these over time -- but I could do pdp plots showing class year.  And why is class_year -AND- cohort year shown?
- How many students take the ACT and the spread or interquartile range

Should I do a PCA?  Always love me a PCA.

### Models  


  | Letter  |***GPA model***| ***Quad Model*** |
  |---------|---------------|------------------|
  | E, EU   | 0             | "Low"            |
  | D-      | 0.7           | "Low"            |
  | D       | 1.0           | "Low"            |
  | D+      | 1.3           | "Low"            |
  | C-      | 1.7           | "Med-Minus"      | 
  | C       | 2.0           | "Med-Minus"      |  
  | C+      | 2.3           | "Med-Minus"      |
  | B-      | 2.7           | "Med-Plus"       |
  | B       | 3.0           | "Med-Plus"       |
  | B+      | 3.3           | "Med-Plus"       |
  | A-      | 3.7           | "Med-Plus"       |
  | A       | 4.0           | "High"           |


*MODEL 1: binary* 

  - Target buckets:   
    1: "A", "A-", "B+", "B", "B-", "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 
  
*MODEL 2: trinary*  

  - Target buckets:     
    2: "A", "A-", "B+", "B", "B-"  
    1: "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 

*MODEL 3: quad* 

  - Target buckets:   
    3: "A"  
    2: "A-", "B+", "B", "B-"  
    1: "C+", "C", "C-"  
    0: "D+", "D", "D-", "E", "EU" 

*MODEL 4: grade GPA* 

  - Targets: 
    4, 3.7, 3.3, 3, 2.7, 2.3, 2, 1.7, 1.3, 1, 0.7, 0

### Predictors  

`r colnames(data_list[[1]][["training"]])[-1]`. 

"load" is the credit load per student per term. 
"yr_diff" is the time (expressed as fractions of a year) between the cohort date and the end of term date for the course.  

The other variables are either self-explanatory or taken directly from OBIA tables.  

Results



```{r eval = FALSE}

# if I ever care to develop the normalized results

# normalized errors # I don't think I care

normalized_metrics <- function(pred, obs) {
  rmse <- sqrt(mean((pred - obs)^2))
  mae  <- mean(abs(pred - obs))
  r2   <- cor(pred, obs)^2
  
  range_y <- diff(range(obs, na.rm = TRUE))
  sd_y    <- sd(obs, na.rm = TRUE)
  
  data.frame(
    RMSE = rmse,
    MAE  = mae,
    R2   = r2,
    NRMSE_range = rmse / range_y,
    NMAE_range  = mae  / range_y,
    NRMSE_sd    = rmse / sd_y,
    NMAE_sd     = mae  / sd_y
  )
}  



  
normResults <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  print(x)
  print(table(missingCourseFilter))
}
  
  normalized_metrics(
    pred = thePredictions[[x]],
    obs = data_list[[x]][["testing"]][[x]][missingCourseFilter]
  )
  
})  
names(normResults) <- params$target_classes

library(kableExtra)

normResults |>
  (\(x){do.call(rbind,x)})() |>
  (\(x){round(x,2)})()|>
    kbl(
    caption = "Normalized Regression Results Across Grading Buckets",
    align = "c",
    col.names = c("",
      "RMSE", "MAE", "R-sq", 
      "NRMSE (Range)", "NMAE (Range)", 
      "NRMSE (SD)", "NMAE (SD)"
    )
  ) |>
  kable_classic(full_width = FALSE, html_font = "Calibri") |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>  # header
  row_spec(1:length(params$target_classes), background = c("#ffffff", "#f9f9f9")) |> # alternating rows
  add_header_above(c(" " = 1, "Raw Metrics" = 3, "Normalized Metrics" = 4))

```



```{r eval=FALSE}

# I think I can delete this section

theResults <- lapply(params$target_classes, function(x) {
  
  print(x)
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  print(x)
  print(table(missingCourseFilter))
}
  
  postResample(pred = thePredictions[[x]],
                  obs = data_list[[x]][["testing"]][[x]][missingCourseFilter]
  
               )
})
names(theResults) <- params$target_classes

# I'd like to normalize the errors with the range and standard deviation

# but it doesn't change that these aren't very good models

# normalized errors

normalized_metrics <- function(pred, obs) {
  rmse <- sqrt(mean((pred - obs)^2))
  mae  <- mean(abs(pred - obs))
  r2   <- cor(pred, obs)^2
  
  range_y <- diff(range(obs, na.rm = TRUE))
  sd_y    <- sd(obs, na.rm = TRUE)
  
  data.frame(
    RMSE = rmse,
    MAE  = mae,
    R2   = r2,
    NRMSE_range = rmse / range_y,
    NMAE_range  = mae  / range_y,
    NRMSE_sd    = rmse / sd_y,
    NMAE_sd     = mae  / sd_y
  )
}  
  
normResults <- lapply(params$target_classes, function(x) {
  
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  print(x)
  print(table(missingCourseFilter))
}
  
  normalized_metrics(
    pred = thePredictions[[x]],
    obs = data_list[[x]][["testing"]][[x]][missingCourseFilter]
  )
  
})  
names(normResults) <- params$target_classes

library(kableExtra)

normResults |>
  (\(x){do.call(rbind,x)})() |>
  (\(x){round(x,2)})()|>
    kbl(
    caption = "Normalized Regression Results Across Grading Buckets",
    align = "c",
    col.names = c("",
      "RMSE", "MAE", "R-sq", 
      "NRMSE (Range)", "NMAE (Range)", 
      "NRMSE (SD)", "NMAE (SD)"
    )
  ) |>
  kable_classic(full_width = FALSE, html_font = "Calibri") |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>  # header
  row_spec(1:length(params$target_classes), background = c("#ffffff", "#f9f9f9")) |> # alternating rows
  add_header_above(c(" " = 1, "Raw Metrics" = 3, "Normalized Metrics" = 4))

```


```{r eval=FALSE}

# boxplots per model if I ever care for them

regression_diagnostics <- function(pred, reference, plot_title = "Predicted vs Actual") {
  if (length(pred) != length(reference)) {
    stop("`pred` and `reference` must be the same length.")
  }
  
  incoming.par <- par(mfrow= c(2,1), mar = c(3,2.5,0,1), oma = c(1.5,1.5,2.25,0))
  
  on.exit(par(incoming.par))
  
  # Compute metrics
  rmse <- sqrt(mean((pred - reference)^2))
  mae  <- mean(abs(pred - reference))
  r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2) 
  
  # Print metrics
#  cat("Regression diagnostics:\n")
#  cat(sprintf("RMSE: %.4f\n", rmse))
#  cat(sprintf("MAE:  %.4f\n", mae))
#  cat(sprintf("R-squared (variance decomposition): %.4f\n", r2))

  
  # Scatter plot
  plot(reference, pred,
      # xlab = "Actual",
      # ylab = "Predicted",
       # main = plot_title,
       pch = 19, col = "steelblue",
       las = 1)
  abline(a = 0, b = 1, col = "red", lwd = 2)  # 45-degree reference line
  grid()
  
  # boxplot
  bp <- boxplot(pred ~ reference,
        # xlab = "Actual",
        # ylab = "Predicted",
         # main = plot_title,
         col = "lightblue",
         border = "steelblue",
         las=1)
  
    # Add a red reference line for each box
        for (i in seq_along(bp$names)) {
          actual_val <- as.numeric(bp$names[i])
          segments(x0 = i - 0.4, y0 = actual_val,
                   x1 = i + 0.4, y1 = actual_val,
                   col = "red", lwd = 2)
        }
  
mtext(side = 3, line =0.75, text = plot_title, cex = 1.5, font = 2, outer = TRUE)  
  
mtext(side = 2, line =0.25, text = "Predicted", cex = 1.25, font = 1, outer = TRUE)  

mtext(side = 1, line =0.25, text = "Actual", cex = 1.25, font = 1, outer = TRUE)  

  # Return metrics invisibly as a list
  invisible(list(RMSE = rmse, MAE = mae, R2 = r2))
}


invisible(lapply(params$target_classes, function(x){
  
      # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][[courseColumn]],data_list[[x]][["training"]][[courseColumn]])
  
  missingCourseFilter <- !data_list[[x]][["testing"]][[courseColumn]] %in% missingCourse
  
  
regression_diagnostics(pred = thePredictions[[x]],
                       reference = data_list[[x]][["testing"]][[x]][missingCourseFilter],
                       plot_title = x)
})
)

```

```{r}

# Plot success over time, one line per target  

plot(x = as.numeric(row.names(resultsDF[["GRADEGPA"]])),
     y = resultsDF[["GRADEGPA"]][,"Rsquared"],
     type = "b",
     lty = 2,
     col = "red")

# it might be nice to scale these and plot them on the same graphic

# There are two anomalies -- 2009 and 2025.
# both are worth a closer look  

# there is no covid discontinuity
# And this could all be explained by more data

``` 



# Variable importance

```{r}

# Plot variable importance
# This needs some work
# I think I'm doing this all wrong

# It does work, but I think I shold turn this into plotly. Some interactivity would be nice. 

varSelection <- importance_over_time[["GRADEGPA"]][["var"]][order(importance_over_time[["GRADEGPA"]][["2024"]], decreasing = TRUE)] |>
  (\(x){x[2:10]})() # top 8 predictors in 2024

# varSelection <- which(importance_over_time[["GRADEGPA"]][["var"]] %in% c("ACTMATH", "ACTENGL", "ACTCOMP", "HSGPA" ))

plotLines <- which(importance_over_time[["GRADEGPA"]][,"var"] %in% varSelection)

plot(50,
     ylim = range(importance_over_time[["GRADEGPA"]][plotLines, as.character(testYears) ], na.rm = TRUE ),
     xlim = range(testYears),
  type = "n"
)

invisible(
lapply(plotLines, function(x){
lines(
  y= as.numeric(importance_over_time[["GRADEGPA"]][x, as.character(testYears)]),
  x = testYears
      )
})
)


```


```{r fig.height=5, eval=FALSE}

# This shows variable importance with a separate graph for each year.

# It's not a bad appendix item, though


invisible(
  lapply(params$target_classes, function(tClass){
    
    lapply(testYears, function(tYear){
       print(plot(  
           varImp(model_list[[tClass]][[as.character(tYear)]]),
           15, main = paste(tClass, tYear, sep = ": ")
       ))
           })
})
)



```
