---
title: "Evaluate Course Decision Trees Over Time vC0"
author: "Bill Prisbrey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vC0"
output:
  html_document:
    keep_md: true
---


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great!

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########


# Build file paths
testYears <- 2007:2025
grid <- expand.grid(params$version, testYears)

model_files <- here::here("Models", "Over Time", paste0("Decision Tree Courses ", grid$Var1, " ", grid$Var2, " model.rds"))

# Read this into a list

model_list <-
  lapply(testYears, function(tYr){
    files <- model_files[grep(tYr, model_files)]
    readRDS(files)
  })
names(model_list) <- testYears

# Build file paths
data_files <- here::here("Data", "Over Time", paste0("Decision Tree Courses ", grid$Var1, " ", grid$Var2, " Data.rds"))

# Read each data into a named list
data_list <-
  lapply(testYears, function(tYr){
    files <- data_files[grep(tYr, data_files)]
    readRDS(files)
  })
names(data_list) <- testYears


```


```{r}

## PREDICTIONS ##

thePredictions <- lapply(as.character(testYears), function(x) {
  
  # identify course values in testing that aren't in training
  
#  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][["course_target"]],data_list[[x]][["training"]][["course_target"]])
  
  if (length(missingCourse) == 0) {
  # no missing values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingCourseFilter <- !data_list[[x]][["testing"]][["course_target"]] %in% missingCourse
  
#  print(x)
#  print(table(missingFilter))
}
  
  
        missingEthnicity <- setdiff(data_list[[x]] [["testing"]][["ETHNICITY"]],data_list[[x]][["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[x]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missingFilter <- missingCourseFilter & missingEthnicityFilter  

  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions) <- as.character(testYears)

```

# Results

Develop and report the confusion matrices 

```{r eval=FALSE}

## CHECK ##

sapply(thePredictions, length)
sapply(data_list, function(x){ nrow(x[["testing"]])  })

# I should dig into those discrepancies a little more
# looks like we added math courses or ethnicity levels;
# I need to be consistent in filtering them out # DONE


```

```{r}

theResults <- lapply(as.character(testYears), function(x) {

    missingCourse <- setdiff(data_list[[x]][["testing"]][["course_target"]],data_list[[x]][["training"]][["course_target"]])
  
  if (length(missingCourse) == 0) {
  # no missing values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingCourseFilter <- !data_list[[x]][["testing"]][["course_target"]] %in% missingCourse
  
#  print(x)
#  print(table(missingFilter))
}
  
  
        missingEthnicity <- setdiff(data_list[[x]] [["testing"]][["ETHNICITY"]],data_list[[x]][["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[x]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missingFilter <- missingCourseFilter & missingEthnicityFilter  

  
  
cm <- confusionMatrix(data =  thePredictions[[x]], reference =data_list[[x]][["testing"]][["course_target"]][missingFilter] , mode = "everything" )

})
names(theResults) <- as.character(testYears)

nameMatrix <- cbind("original" = c(
  "MATH_1010", "MATH_1030", "MATH_1050", "MATH_1060", "MATH_1070", "MATH_1080", "MATH_1090", "MATH_1100", "MATH_1210", "MATH_1220", "MATH_1250", "MATH_1310", "MATH_2210", "MATH_990",  "other"
), "display" = c(
  "1010", "1030", "1050", "1060", "1070", "1080", "1090", "1100", "1210", "1220", "1250", "1310", "2210", "990", "other"
))

# I might need a way to make a name matrix automatically per confusion matrix

# not sure why this is throwing an error
drawCM(theResults[[3]], nameMatrix = nameMatrix, percentage = TRUE, metricsClass = "1210")
drawCM(theResults[[3]],                          percentage = TRUE, metricsClass = "MATH_1210")

# precision color is screwed up

```


```{r}

# extract Kappa

theKappa <- sapply(theResults, function(x) {x[["overall"]]["Kappa"]})
names(theKappa) <- names(theResults)  

# plot Kappa
plot(y = theKappa,
     x= as.numeric(names(theKappa)),
     type = "b",
     col = "firebrick",
     las = 2,
     ylab = "Kappa",
     xlab = ""
     )

```

```{r}

# Metrics per course (one dataframe per course)
target_math_courses <- sapply(thePredictions, unique) |> unlist() |> as.character()  |> unique()

perCourse <- lapply(target_math_courses, function(math_course){
  
  course_over_time_list <- lapply(theResults, function(result){ 
   result[["byClass"]][grepl(math_course, row.names(result[["byClass"]])),]
    })
  
  course_over_time <- do.call(rbind, course_over_time_list)
  row.names(course_over_time) <- names(theResults)
  return(course_over_time)
  
})
names(perCourse) <- target_math_courses 

```


```{r}

importance_over_time <- lapply(names(model_list), function(tYr){
  
  imp <- varImp(model_list[[tYr]])$importance
  imp$var <- row.names(imp)
  colnames(imp)[colnames(imp) %in% "Overall"] <- tYr # rename "Overall" column
  return(imp)
  
})

importance_over_time <- Reduce(function(m1,m2){merge(m1,m2, by = "var", all = TRUE)}, importance_over_time)
importance_over_time <- importance_over_time[order(importance_over_time[,2], decreasing = TRUE),] # descending order by the first year of data

```


```{r}

# plotly version courtesy of Chat

## BEAUTIFUL! ##

# library(plotly)

varFilter <- apply(importance_over_time[,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time$var[varFilter & !is.na(varFilter)]

# Filter to selected variables
plotData <- importance_over_time[importance_over_time[,"var"] %in% varSelection[-1], ]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```


```{r fig.height=5, eval=FALSE}

# This shows variable importance with a separate graph for each year.

# It's not a bad appendix item, though


invisible(

    lapply(testYears, function(tYear){
       print(plot(  
           varImp(model_list[[as.character(tYear)]]),
           15, main = paste(tYear, sep = ": ")
       ))
           })
)

# course is a lot harder.

# let's pull in what we have


```

So accuracy at predicting courses drops incredibly.
And I'm not sure I want to include "course_level", but why was it only super-important for a few years?  

I want to see the median grade per course over time.  What's my trend over time?

Similarly -- maybe I want to see the proportion of students achieving a particular grade level or grade chunk.  It's getting big, right?  Could that show grade inflation? 

And we're getting closer to counter-factuals. What grade would they get in a particular course?

I need to submit this to Git, and write up both reports, and get a couple of key visualizations over time. 


