---
title: "Evaluate Grade Decision Trees using vD0 models"
date: "2025-09-10"
output:
  html_document:
    keep_md: true
---

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)

source(here::here("Functions", "Draw Confusion Matrix vC5.R"))

```


```{r}

##########
## LOAD ##
##########

target_classes <- c("GRADEGPA", "wdraw_binary", "grade_binary", "grade_trinary", "grade_quad")

# Build file paths
model_files <- here::here("Models", paste0("Decision Tree vD0 ", target_classes, " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  target_classes
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree vD0 ", target_classes, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  target_classes
)


```


# SUMMARY

Five models were trained using extreme gradient boosting.  Each model predicted a course grade for one of ten math courses at the University of Utah.  Each model used complete cases of a small number of predictors, such as high school GPA and ACT test scores.

The data included all years from 2005 to 2025.  

Four models bucketed grades into groups of increasing granularity and used classification.  One model trained on numeric GPA values and used regression.  

All models had low accuracy. The highest classification model had a Kappa value of 0.207, and the regression model had an r-squared value of 0.27.

Due to low accuracy, it is not recommended that these few predictors can be used to generate strong recommendations or firm cut-offs or thresholds for the various courses.  Rather, it could be used to develop probabilistic guidance that a student or advisor could access.  

Some key findings include:  


  - The high school GPA was the most predictive variable for the regression model. 
  - Math 1050 [?] had the highest accuracy (R^2 value of 0.29) while Math xxxx had the lowest accuracy.
  - An inspection of accuracy found that predictions were the most accurate for students with a high school GPA greater than [[3.8]]. 
  - Math grades increased with high school GPA and ACT math test scores.  
  - Other important variables.  
  - Math grades have increased during the time period investigated. 
  - A guidance plot shows a heatmap of predicted grades varied across two leading indicators and adjustable to the remaining variables.  




# MODEL SUMMARY

**MODEL 1:** ***Withdraw binary***
  - 

**MODEL 2:** ***Grade binary***
  - 

**MODEL 3:** ***Trinary***
  -
  
**MODEL 4:** ***Quad***
  -   
  
**MODEL 5:** ***GPA***
  -


NEXT STEPS: 

  - Filter the data to first term math courses only:  all the math courses taken in the first term a student takes a math course and re-train the models.  
  - Include the concurrent semester credit load.    
    - Possibly broken out by college or topic.  
  - Convert everything to regression models.   
  - Attempt different methods to predict withdrawals (grades of "W").  
  - Develop counter-factual tables of predicted grades in alternative courses in order to identify optimum course selections based on the limited data.  
  - Develop guidance plots that show the optimum course selection per individual with respect to grade. 
  - Develop a plot that shows the expected grade and range per course per individual.  
  
  
  
  
  
  
  

```{r}

# predict(model_list[[1]], newdata = data_list[[1]][["testing"]][,"wdraw_binary"]) 

# predict(model_list[[ target_classes[[3]] ]], 
#         newdata =  data_list[[  target_classes[[3]]   ]][["testing"]]       
#          ) |> table()


thePredictions <- lapply(target_classes, function(x) {
  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]])
  
} )
names(thePredictions) <- target_classes

# check
# lapply(thePredictions[-1], table)
# hist(thePredictions[[1]]) # went a little high I guess


```


```{r}

theCMs <- lapply(target_classes[-1], function(x) {
  
  confusionMatrix(data = thePredictions[[x]],
                  reference = factor(data_list[[x]][["testing"]][,x]) ,
                  mode = "everything")
  
  
})
names(theCMs) <- target_classes[-1]

kappaValues <- sapply(theCMs, function(x){ 
  
  x[["overall"]]["Kappa"]
  
  })

# invisible(lapply(theCMs, drawCM))
# Need to modify the title displayed


```

# Classification model accuracy

Because no classification model produced a Kappa value greater than 0.21 (indicating performance too close to random), no classification model will be interpreted or evaluated further.  

Confusion matrices for each model are available in the appendix. 

```{r}

## PLOT KAPPA ##

par(mar = c(2,4,4,1), bg = "lightsteelblue1" )

plot(kappaValues, 
     type = "n",
     xaxt = "n",
     xlab = "",
     ylab = "Kappa",
     las = 2)
usr <- par("usr")

rect(xleft = usr[1], ybottom = usr[3],
     xright = usr[2], ytop = 0.1, col = "red", border = NA)

rect(xleft = usr[1], ybottom = 0.1,
     xright = usr[2], ytop = usr[4], col = "pink", border = NA)

points(kappaValues, pch = 19, col = "gray10")

text(x = 1:4,
     y = kappaValues,
     labels = gsub(".Kappa","", names(kappaValues)),
     col = c("lightyellow","lightyellow","khaki1", "khaki1"),
     pos = c(4,4,4,2),
     xpd = TRUE,
     font = 2
     )

mtext(side = 3,
      text = "Accuracy of classification models",
      font = 2,
      cex = 1.5,
      line = 1.5)


```


```{r eval=FALSE}

#########################
## VARIABLE IMPORTANCE ##
#########################

par(mar = c(0,3,5,0))
lapply(names(model_list), function(x){plot(varImp(model_list[[x]]), main = x)})

# I need to figure out how to terrace these
# Probably from plotting the underlying data

varImp(model_list[[1]]) |> (\(x){(x$importance)})() |> as.matrix() |> rev() |> barplot(horiz = TRUE, beside = TRUE)

# Getting there


```

# Regression model accuracy

```{r}

# Now for the regression

regression_diagnostics <- function(pred, reference, plot_title = "Predicted vs Actual") {
  if (length(pred) != length(reference)) {
    stop("`pred` and `reference` must be the same length.")
  }
  
  incoming.par <- par(mfrow= c(2,1), mar = c(3,2.5,0,1), oma = c(1.5,1.5,2.25,0))
  
  on.exit(par(incoming.par))
  
  # Compute metrics
  rmse <- sqrt(mean((pred - reference)^2))
  mae  <- mean(abs(pred - reference))
  r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2)
  
  # Print metrics
  cat("Regression diagnostics:\n")
  cat(sprintf("RMSE: %.4f\n", rmse))
  cat(sprintf("MAE:  %.4f\n", mae))
  cat(sprintf("R-squared: %.4f\n", r2))

  
  # Scatter plot
  plot(reference, pred,
      # xlab = "Actual",
      # ylab = "Predicted",
       # main = plot_title,
       pch = 19, col = "steelblue",
       las = 1)
  abline(a = 0, b = 1, col = "red", lwd = 2)  # 45-degree reference line
  grid()
  
  # boxplot
  bp <- boxplot(pred ~ reference,
        # xlab = "Actual",
        # ylab = "Predicted",
         # main = plot_title,
         col = "lightblue",
         border = "steelblue",
         las=1)
  
    # Add a red reference line for each box
        for (i in seq_along(bp$names)) {
          actual_val <- as.numeric(bp$names[i])
          segments(x0 = i - 0.4, y0 = actual_val,
                   x1 = i + 0.4, y1 = actual_val,
                   col = "red", lwd = 2)
        }
  
mtext(side = 3, line =0.75, text = plot_title, cex = 1.5, font = 2, outer = TRUE)  
  
mtext(side = 2, line =0.25, text = "Predicted", cex = 1.25, font = 1, outer = TRUE)  

mtext(side = 1, line =0.25, text = "Actual", cex = 1.25, font = 1, outer = TRUE)  

  # Return metrics invisibly as a list
  invisible(list(RMSE = rmse, MAE = mae, R2 = r2))
}

pred <- thePredictions[["GRADEGPA"]]
reference <- data_list[["GRADEGPA"]][["testing"]][,"GRADEGPA"]
# pred = your predicted values
# reference = your true grades

regression_diagnostics(pred = pred,
                       reference = reference)

```


```{r fig.height=7}

plot(varImp(model_list[["GRADEGPA"]]),
     main = "Important variables for \n 'grade GPA' regression")

```


```{r}

errors <- pred - reference

# Scatter plot of errors by true grade
plot(reference, errors,
     xlab = "True Grade",
     ylab = "Prediction Error (Predicted - True)",
     main = "Prediction Errors by True Grade",
     pch = 19, col = "steelblue")
abline(h = 0, col = "red", lwd = 2)  # reference line: perfect prediction
grid()

# Optional: add a boxplot per true grade for clarity
boxplot(errors ~ reference,
        xlab = "True Grade",
        ylab = "Prediction Error",
        main = "Error Distribution by True Grade",
        col = "lightblue")
abline(h = 0, col = "red", lwd = 2)


```


```{r eval=TRUE}

# data frame per course

results <- data.frame(
  actual = data_list[["GRADEGPA"]][["testing"]][,"GRADEGPA"],
  predicted = thePredictions[["GRADEGPA"]],
  course=data_list[["GRADEGPA"]][["testing"]][,"course"]
)

library(dplyr)
course_metrics <- results %>%
  group_by(course) %>%
  summarise(
    n=n(),
    rmse = sqrt(mean((actual-predicted)^2)),
    mae=mean(abs(actual-predicted)),
    r_squared = cor(actual, predicted)^2
  ) %>%
  arrange(desc(r_squared))

library(kableExtra)

course_metrics |>
  kbl(caption = "Course Metrics Summary",
      digits = 2,          # round numeric columns
      align = "c") |>      # center columns
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "center",
                font_size = 14) 

# > course_metrics
# A tibble: 10 × 5
#   course        n  rmse   mae r_squared
#   <chr>     <int> <dbl> <dbl>     <dbl>
# 1 MATH_1070   879 0.971 0.709     0.299
# 2 MATH_1210  1848 1.02  0.777     0.271
# 3 MATH_2250   972 0.996 0.740     0.253
# 4 MATH_1090   735 1.10  0.845     0.239
# 5 MATH_1010  2015 1.14  0.900     0.236
# 6 MATH_1050  2100 1.17  0.929     0.233
# 7 MATH_2210  1174 0.928 0.659     0.228
# 8 MATH_1220  1603 1.02  0.763     0.223
# 9 MATH_1060  1052 1.16  0.916     0.201
# 10 MATH_1030   914 1.03  0.784     0.198

```

```{r}

regression_diagnostics <- function(data, pred_col, ref_col, group_col = NULL) {
  # data: data frame
  # pred_col, ref_col: column names (strings)
  # group_col: optional column name (string)
  
  if (!all(c(pred_col, ref_col) %in% names(data))) {
    stop("Data must contain columns for prediction and reference.")
  }
  
  if (!is.null(group_col) && !group_col %in% names(data)) {
    stop("Grouping column not found in data.")
  }
  
  # --------------------
  # Grouped mode
  # --------------------
  if (!is.null(group_col)) {
    groups <- unique(data[[group_col]])
    n <- length(groups)
    
    # Set up plotting grid (square-ish layout)
    nrow <- ceiling(sqrt(n))
    ncol <- ceiling(n / nrow)
    oldpar <- par(mfrow = c(nrow, ncol))
    on.exit(par(oldpar))  # restore layout
    
    results <- lapply(groups, function(g) {
      subset <- data[data[[group_col]] == g, ]
      pred <- subset[[pred_col]]
      reference <- subset[[ref_col]]
      
      # Metrics
      rmse <- sqrt(mean((pred - reference)^2))
      mae  <- mean(abs(pred - reference))
      r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2)
      
      if (length(unique(reference)) < 20) {
        # Boxplot for discrete actuals
        bp <- boxplot(subset[[pred_col]] ~ factor(subset[[ref_col]]),
                      xlab = "Actual",
                      ylab = "Predicted",
                      main = paste("Predicted vs Actual -", g),
                      col = "lightblue")
        
        # Add a red reference line for each box
        for (i in seq_along(bp$names)) {
          actual_val <- as.numeric(bp$names[i])
          segments(x0 = i - 0.4, y0 = actual_val,
                   x1 = i + 0.4, y1 = actual_val,
                   col = "red", lwd = 2)
        }
        
      } else {
        # Scatterplot for continuous actuals
        plot(reference, pred,
             xlab = "Actual",
             ylab = "Predicted",
             main = paste("Predicted vs Actual -", g),
             pch = 19, col = "steelblue")
        abline(a = 0, b = 1, col = "red", lwd = 2)
        grid()
      }
      
      # Return metrics
      data.frame(Group = g, RMSE = rmse, MAE = mae, R2 = r2)
    })
    
    return(do.call(rbind, results))
    
  } else {
    # --------------------
    # Ungrouped mode
    # --------------------
    pred <- data[[pred_col]]
    reference <- data[[ref_col]]
    
    # Metrics
    rmse <- sqrt(mean((pred - reference)^2))
    mae  <- mean(abs(pred - reference))
    r2   <- 1 - sum((pred - reference)^2) / sum((reference - mean(reference))^2)
    
    if (length(unique(reference)) < 20) {
      # Boxplot for discrete actuals
      bp <- boxplot(data[[pred_col]] ~ factor(data[[ref_col]]),
                    xlab = "Actual",
                    ylab = "Predicted",
                    main = "Predicted vs Actual (boxplot)",
                    col = "lightblue")
      
      # Add a red reference line for each box
      for (i in seq_along(bp$names)) {
        actual_val <- as.numeric(bp$names[i])
        segments(x0 = i - 0.4, y0 = actual_val,
                 x1 = i + 0.4, y1 = actual_val,
                 col = "red", lwd = 2)
      }
      
    } else {
      # Scatterplot for continuous actuals
      plot(reference, pred,
           xlab = "Actual",
           ylab = "Predicted",
           main = "Predicted vs Actual",
           pch = 19, col = "steelblue")
      abline(a = 0, b = 1, col = "red", lwd = 2)
      grid()
    }
    
    return(data.frame(RMSE = rmse, MAE = mae, R2 = r2))
  }
}



```


```{r}
# let's see how that column works out

# regression_diagnostics(data = results,
#                       pred_col = "predicted",
#                       ref_col = "actual")
# matches earlier version

# regression_diagnostics(data = results,
#                       pred_col = "predicted",
#                       ref_col = "actual",
#                       group_col = "course")

# course filter

resultsFilter <-results$course %in% c("MATH_1070", "MATH_1210", "MATH_2250")

regression_diagnostics(data = results[resultsFilter,],
                       pred_col = "predicted",
                       ref_col = "actual",
                       group_col = "course")


results <- data.frame(
  actual = data_list[["GRADEGPA"]][["testing"]][,"GRADEGPA"],
  predicted = thePredictions[["GRADEGPA"]],
  year=data_list[["GRADEGPA"]][["testing"]][,"year"]
)


resultsFilter <-results$year %in% 2020:2023

regression_diagnostics(data = results[resultsFilter,],
                       pred_col = "predicted",
                       ref_col = "actual",
                       group_col = "year")




```


Is there a pocket of high accuracy?

I think I should make sure all-l of these visualizations work,
including a few more ---

partial dependence plot

hs.gpa vs act scores as axis, actual - predicted grade (connected by a line on a countour plot?)

should I do a variable importance plot, where I shuffle a variable?

correlate ?

```{r eval=FALSE}

data_list[["GRADEGPA"]][["testing"]][,"GRADEGPA"]

plot(x = data_list[["GRADEGPA"]][["testing"]][,"GRADEGPA"],
      y = data_list[["GRADEGPA"]][["testing"]][,"HSGPA"]
      )

boxplot(x = data_list[["GRADEGPA"]][["testing"]][,"GRADEGPA"],
      y = data_list[["GRADEGPA"]][["testing"]][,"HSGPA"]
      )

```



```{r}

boxplot(HSGPA ~ GRADEGPA,
              data = data_list[["GRADEGPA"]][["testing"]],
              xlab = "Math Grade",
              ylab = "High School GPA",
              main = "HS GPA vs Math GPA (boxplot)",
              col = "lightblue")

boxplot(GRADEGPA ~ year,
              data = data_list[["GRADEGPA"]][["testing"]],
              xlab = "year",
              ylab = "Math grade",
              main = "Math GPA vs year (boxplot)",
              col = "lightblue")

# That's some impressive grade inflation  

# aggregate(GRADEGPA ~ year, data = data_list[["GRADEGPA"]][["training"]], function(x) 
#   {c(median(x), length(x))})

# why is "year" in whole numbers,
# and why is 'n' going down so much?

boxplot(yr_diff ~ GRADEGPA,
              data = data_list[["GRADEGPA"]][["testing"]],
              xlab = "math grade",
              ylab = "Year difference",
              main = "Years at college vs Math GPA (boxplot)",
              col = "lightblue",
        outline = FALSE)

# that's not a bad graphic
# needs some causal inference, though, to pull conclusions from it.

# ok, let's trouble-shoot my graphics, add some...
# and wrap up today.


```


So I am looking at two things here:
- Writing up this report
- Comparing the data set

If the data set compares, then I need to move to the next iteration of training (esp only using the first math class and possibly adding the number of credits in the first semester)  (I think there's something to also looking at the GPA for all of the other classes that semester.  I wonder if causal inference allows that kind of thing?)

If the data set does NOT compare, then this becomes less interesting and I should re-run the training on this data set.

I'm thinking I should also run the binary as a regression model to take advantage of the ordinal nature of the data.

I'd also really like to know what the range of grades are for a prediction of, say, 2.7 or 3.0.  I guess that's my RMSE or MAE.  It's also rotating my boxplots on the side, or one box plot per prediction.

Then there's reviewing my Claude convo and adding those conversations.

And there's a couple of my other sandbox -- selecting classes via hierarchical clustering and the year_diff density plots (years to class)



I need to see how the prediction/actual changes as one predictor changes



I'd like to see something like an individual guidance chart, one per student, that gives the probabilities (contours?) of grades per course.

So it has all the information we have about a student, and somehow tells them what their likelihood is.  

In my mind (and I know this won't work) I am imagining ACT Math along one edge, and HS GPA along the other, and each cell is a color with the probability of getting greater than "X" score.  One graphic per course.

It would make a cool widget in Shiny, that's for sure, where the student moves their widget up or down to determine their own cut-off dates.  

The student could then add information to the widget, and get a personalized prediction.  They could add as much information as they'd like and with-hold others.  
It would display something like an "expected grade" with error bars.

I am imaginining ... I'm not sure what I'm imaginging.

Something like each square represents a grade and a probability?  At the upper right corner is the highest probability, and the grade that appears in that corner changes based on the personal prediction. 

Anyway, def leading towards "probabilistic guidance." 

And I might as well try a per-course model.

And (of course) restrict it to the first math class taken by the first-time-freshman


I gotta show what I've got, though -- I need some show'n'tell.

```{r}

#############
## CONTOUR ##
#############

testData <- data_list[["GRADEGPA"]][["testing"]]


# Turn this into a function where you can specify values

#for(course_name in unique(testData$course)) {
#  course_data <- testData[testData$course == course_name, ]

predictionContour <- function(data, 
                              course_name = "MATH_1010",
                              show_actual = FALSE,
  SEX = "M",
  FIRST_GEN_STATUS_CD = "N",
  ETHNICITY = "C",
  RESSTAT = "R",
  FA_PELL = 0,
  AGE = 18,
  APCREDIT = 0,
  HSPRIVATE = "N",
  HONORS = 0,
  ACTCOMP = 24,
  ACTENGL = 24,
  ACTSCI = 24,
  year = 2019,
  yr_diff = 0.75
  ) {
  
  # subset the data
  course_data <- testData[testData$course == course_name, ]
  
  # Create prediction grid
  grid <- expand.grid(
    HSGPA = seq(min(course_data$HSGPA), max(course_data$HSGPA), length.out = 50),
    ACTMATH = seq(min(course_data$ACTMATH), max(course_data$ACTMATH), length.out = 50)
  )
  
  # Add other predictors (replicate across rows)
  grid$SEX                 <- SEX
  grid$FIRST_GEN_STATUS_CD <- FIRST_GEN_STATUS_CD
  grid$ETHNICITY           <- ETHNICITY
  grid$RESSTAT             <- RESSTAT
  grid$FA_PELL             <- FA_PELL
  grid$AGE                 <- AGE
  grid$APCREDIT            <- APCREDIT
  grid$HSPRIVATE           <- HSPRIVATE
  grid$HONORS              <- HONORS
  grid$ACTCOMP             <- ACTCOMP
  grid$ACTENGL             <- ACTENGL
  grid$ACTSCI              <- ACTSCI
  grid$year                <- year
  grid$yr_diff             <- yr_diff
  grid$course              <- course_name  
  
#   grid$course <- course_name
    
  grid$predicted_grade <- predict(model_list[["GRADEGPA"]], newdata = grid)
  
  # plot
thePlot <- ggplot() +
    # Heatmap of predictions
    geom_tile(data = grid, aes(x=HSGPA, y=ACTMATH, fill = predicted_grade)) +

    scale_fill_gradient2(low = "purple4", mid = "gray90", high = "darkcyan",  midpoint = 2.5, name = "Predicted\nGrade") +

    labs(title = paste("Grade Predictions:", course_name), 
         x="High School GPA", y = "ACT Math Score") +
    theme_minimal()

  if(show_actual){
    thePlot <- thePlot +
          # Actual student outcomes as points
    geom_point(data = course_data, aes(x=HSGPA, y = ACTMATH, color = GRADEGPA), alpha = 0.6, size = 2) +
          scale_color_gradient2(low = "darkorange2", mid = "yellowgreen", high = "darkgreen", midpoint = 2.5, name = "Actual\nGrade")
    
  }
 
thePlot 
  
 }

predictionContour(data = testData, course_name  = 
                  "MATH_1010")


predictionContour(data = testData, 
                  course_name  = "MATH_1010",
                  show_actual = TRUE)

# anyway... I like it

# except for the colors.

```

```{r eval = FALSE}

# stuff used in building of contour map function

 # Complete other variables
  grid$SEX <- SEX
  grid$FIRST_GEN_STATUS_CD <- FIRST_GEN_STATUS_CD
  grid$ETHNICITY <- "C"
  grid$RESSTAT <- "R"
  grid$FA_PELL <- 0
  grid$AGE <- 18
  grid$APCREDIT <- 0
  grid$HSPRIVATE <- "N"
  grid$HONORS <- 0
  grid$ACTCOMP <- 24
  grid$ACTENGL <- 24
  grid$ACTSCI <- 24
  grid$year <- 2019
  grid$yr_diff <- 0.75 # could use per-course values

  
  # plot
  ggplot() +
    # Heatmap of predictions
    geom_tile(data = grid, aes(x=HSGPA, y=ACTMATH, fill = predicted_grade)) +
    # Actual student outcomes as points
    geom_point(data = course_data, aes(x=HSGPA, y = ACTMATH, color = GRADEGPA), alpha = 0.6, size = 2) +
    scale_fill_gradient2(low = "purple4", mid = "gray90", high = "darkcyan",  midpoint = 2.5, name = "Predicted\nGrade") +
    scale_color_gradient2(low = "darkorange3", mid = "gray90", high = "royalblue4", midpoint = 2.5, name = "Actual\nGrade") + 
    labs(title = paste("Grade Predictions:", course_name), 
         x="High School GPA", y = "ACT Math Score") +
    theme_minimal()
  

```


Why can't I simply subtract one from the prediction?  Won't that make it more accurate?


I like that contour map.  Looking pretty good.
...and it also shows how AWFUL it is!

I still haven't answered if I have pockets of high accuracy. 


```{r eval=FALSE} 

#################################
## BINNED ACTUAL VS PREDICTIVE ##
##            MEAN             ##
#################################

# This looks really good, actually, because the "mean" smooths out the actual spread

# don't forget to add the prediction to the test data

predictionBin <- function(data,
                          course_name
) {
  
  if(!"predicted" %in% colnames(data)) {
  stop("Add column named 'predicted' with grade predictions.")
  }
    
# Calculate bins
  
  course_data  <- data[testData$course == course_name, ] %>%
  mutate(
    gpa_bin = cut(HSGPA, breaks = quantile(HSGPA, probs=seq(0,1,0.25)),
                  include.lowest = TRUE),
    act_bin = cut(ACTMATH, breaks = quantile(ACTMATH, probs= seq(0,1,0.25)),
                  include.lowest = TRUE)
  )
  
# Calculate mean actual and prediction by bin

summary_data <- course_data %>%
  group_by(gpa_bin, act_bin) %>%
  summarise(
    mean_actual = mean(GRADEGPA),
    mean_predicted = mean(predicted),
    n=n(),
    .groups = 'drop'
  )  
  
# Plot

ggplot(summary_data, aes(x=gpa_bin, y=act_bin)) +
  geom_tile(aes(fill = mean_predicted), color="white") +
  geom_text(aes(label=sprintf("Pred: %.2f\nActual: %.2f\n(n=%d",
                              mean_predicted, mean_actual, n)),
            size=3) +
            
scale_fill_gradient2(low = "red", mid = "yellow", high = "green", midpoint=2.5) +
  labs(title=paste("Grade Predictions by Student Profile:", course_name)) +
  theme_minimal()  

}





testData$predicted <- thePredictions[["GRADEGPA"]]

course_data  <- testData[testData$course == "MATH_1010", ] %>%
  mutate(
    gpa_bin = cut(HSGPA, breaks = quantile(HSGPA, probs=seq(0,1,0.25)),
                  include.lowest = TRUE),
    act_bin = cut(ACTMATH, breaks = quantile(ACTMATH, probs= seq(0,1,0.25)),
                  include.lowest = TRUE)
  )

# Calculate mean actual and prediction by bin

summary_data <- course_data %>%
  group_by(gpa_bin, act_bin) %>%
  summarise(
    mean_actual = mean(GRADEGPA),
    mean_predicted = mean(predicted),
    n=n(),
    .groups = 'drop'
  )

# Plot

ggplot(summary_data, aes(x=gpa_bin, y=act_bin)) +
  geom_tile(aes(fill = mean_predicted), color="white") +
  geom_text(aes(label=sprintf("Pred: %.2f\nActual: %.2f\n(n=%d",
                              mean_predicted, mean_actual, n)),
            size=3) +
            
scale_fill_gradient2(low = "red", mid = "yellow", high = "green", midpoint=2.5) +
  labs(title=paste("Grade Predictions by Student Profile:", course_name)) +
  theme_minimal()




```  


```{r eval=FALSE}

predictionBin2 <- function(data, course_name) {
  
  if(!"predicted" %in% colnames(data)) {
    stop("Add column named 'predicted' with grade predictions.")
  }
  
  # Subset for course
  course_data <- data[data$course == course_name, ] %>%
    filter(!is.na(HSGPA), !is.na(ACTMATH), !is.na(GRADEGPA), !is.na(predicted))
  
  # Create bins
  course_data <- course_data %>%
    mutate(
      gpa_bin = cut(HSGPA, breaks = quantile(HSGPA, probs = seq(0, 1, 0.25), na.rm = TRUE),
                    include.lowest = TRUE),
      act_bin = cut(ACTMATH, breaks = quantile(ACTMATH, probs = seq(0, 1, 0.25), na.rm = TRUE),
                    include.lowest = TRUE)
    )
  
  # Summarise per bin
  summary_data <- course_data %>%
    group_by(gpa_bin, act_bin) %>%
    summarise(
      mean_actual    = mean(GRADEGPA),
      mean_predicted = mean(predicted),
      sd_predicted   = sd(predicted),
      n              = n(),
      .groups = 'drop'
    )
  
  # Heatmap with spread overlay
  ggplot(summary_data, aes(x = gpa_bin, y = act_bin)) +
    geom_tile(aes(fill = mean_predicted), color = "white") +
    geom_text(aes(label = sprintf("Pred: %.2f\nActual: %.2f\n(n=%d)", 
                                  mean_predicted, mean_actual, n)),
              size = 3) +
    geom_point(data = course_data,
               aes(x = gpa_bin, y = act_bin, color = predicted),
               position = position_jitter(width = 0.2, height = 0.2),
               alpha = 0.5, size = 1.5) +
    scale_fill_gradient2(low = "red", mid = "yellow", high = "green", midpoint = 2.5,
                         name = "Mean Predicted") +
    scale_color_gradient2(low = "darkgreen", mid = "yellowgreen", high = "darkorange",
                          midpoint = 2.5, name = "Predicted") +
    labs(title = paste("Grade Predictions by Student Profile:", course_name),
         x = "HSGPA Bin", y = "ACT Math Bin") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}



```

```{r eval=FALSE}

library(tidyr)
predictionBin3 <- function(data, course_name) {
  
  if(!"predicted" %in% colnames(data)) {
    stop("Add column named 'predicted' with grade predictions.")
  }
  
  # Subset for course
  course_data <- data[data$course == course_name, ] %>%
    filter(!is.na(HSGPA), !is.na(ACTMATH), !is.na(GRADEGPA), !is.na(predicted))
  
  # Create bins
  course_data <- course_data %>%
    mutate(
      gpa_bin = cut(HSGPA, breaks = quantile(HSGPA, probs = seq(0, 1, 0.25), na.rm = TRUE),
                    include.lowest = TRUE),
      act_bin = cut(ACTMATH, breaks = quantile(ACTMATH, probs = seq(0, 1, 0.25), na.rm = TRUE),
                    include.lowest = TRUE)
    )
  
  # Reshape to long format for boxplots
  long_data <- course_data %>%
    select(gpa_bin, act_bin, predicted, GRADEGPA) %>%
    pivot_longer(cols = c(predicted, GRADEGPA),
                 names_to = "type",
                 values_to = "value")
  
  # Heatmap of bin counts (optional)
  bin_counts <- course_data %>%
    group_by(gpa_bin, act_bin) %>%
    summarise(n = n(), .groups = 'drop')
  
  ggplot() +
    geom_tile(data = bin_counts, aes(x = gpa_bin, y = act_bin, fill = n), color = "white") +
    geom_boxplot(
      data = long_data,
      aes(x = gpa_bin, y = value, fill = type),
      position = position_dodge(width = 0.8),
      width = 0.35,
      outlier.size = 0.5
    ) +
    scale_fill_manual(values = c(predicted = "lightblue", GRADEGPA = "orange"),
                      name = "Type",
                      labels = c("Predicted", "Actual")) +
    labs(title = paste("Predicted vs Actual by Student Profile:", course_name),
         x = "HSGPA Bin",
         y = "Grade") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

predictionBin3(data = testData, course_name = "MATH_1050")

```

```{r eval=FALSE}

predictionBin4 <- function(data, course_name) {
  library(dplyr)
  library(ggplot2)
  
  if(!"predicted" %in% colnames(data)) {
    stop("Add column named 'predicted' with grade predictions.")
  }
  
  # Subset for course
  course_data <- data %>%
    filter(course == course_name,
           !is.na(HSGPA), !is.na(ACTMATH), 
           !is.na(GRADEGPA), !is.na(predicted))
  
  # Create bins
  course_data <- course_data %>%
    mutate(
      gpa_bin = cut(HSGPA, 
                    breaks = quantile(HSGPA, probs = seq(0,1,0.25), na.rm = TRUE), 
                    include.lowest = TRUE),
      act_bin = cut(ACTMATH, 
                    breaks = quantile(ACTMATH, probs = seq(0,1,0.25), na.rm = TRUE),
                    include.lowest = TRUE)
    )
  
  # Prepare long-format data for boxplots (predicted vs actual)
  long_data <- course_data %>%
    select(gpa_bin, predicted, GRADEGPA) %>%
    tidyr::pivot_longer(cols = c(predicted, GRADEGPA),
                        names_to = "type",
                        values_to = "value")
  
  # Plot: two boxplots per GPA bin
  ggplot(long_data, aes(x = gpa_bin, y = value, fill = type)) +
    geom_boxplot(position = position_dodge(width = 0.8), width = 0.35, outlier.size = 0.5) +
    scale_fill_manual(values = c(predicted = "lightblue", GRADEGPA = "orange"),
                      name = "Type",
                      labels = c("Predicted", "Actual")) +
    labs(title = paste("Predicted vs Actual by Student Profile:", course_name),
         x = "HSGPA Bin",
         y = "Grade") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

predictionBin4(data = testData,
               course_name = "MATH_1050"
               )
# I want to verify this.  I'm not so sure

```

This bin4 looks terrible.  I don't quite believe it.

Instead of four pairs, I wonder if I can have the four y-bins as well (sub-bins for the grade).

Next, I need to do a pdp plot.


```{r}

library(pdp)

importance <- varImp(model_list[["GRADEGPA"]])
top_vars <- rownames(importance$importance)[order(importance$importance$Overall, decreasing = TRUE)[1:3]]
top_vars

# For a single variable
pdp_obj <- partial(object = model_list[["GRADEGPA"]], 
                   pred.var = "HSGPA",
                   train = model_list[["GRADEGPA"]]$trainingData,
                   type = "regression")

# Plot
plot(pdp_obj, main = "Partial Dependence of HSGPA on Predicted Grade")


```

```{r}

## PARTIAL DEPENDENCE PER COURSE ##

library(pdp)
library(ggplot2)

# Courses to loop over
courses <- unique(model_list[["GRADEGPA"]]$trainingData$course)

# Function to make PDP for one course
get_pdp_course <- function(course_name, pred_var = "HSGPA") {
  model <- model_list[["GRADEGPA"]]
  train_df <- model$trainingData[model$trainingData$course == course_name, ]
  
  # remove .outcome column
#  train_df <- train_df[, setdiff(names(train_df), ".outcome"), drop = FALSE]
  
  # Create a "fixed" grid with all other predictors at their median / most common value
fixed_vals <- train_df[1, ]  # take first row as default

# Override course to the one you want
fixed_vals$course <- course_name

# Drop HSGPA because we want it to vary
fixed_vals$HSGPA <- NULL
  
  # Partial dependence for HSGPA, fixing course
  pdp_obj <- partial(
    object = model,
    pred.var = pred_var,
    train = train_df,
    type = "regression"
  )
  
  pdp_obj$course <- course_name
  pdp_obj
  
}

# Loop through courses
all_pdp <- lapply(courses, get_pdp_course, pred_var = "HSGPA")
all_pdp_df <- do.call(rbind, all_pdp)

# Plot faceted by course
ggplot(all_pdp_df, aes(x = HSGPA, y = yhat)) +
  geom_line(color = "steelblue") +
  facet_wrap(~course) +
  labs(title = "PDP of HSGPA by Course",
       x = "High School GPA", y = "Predicted Grade GPA") +
  theme_minimal()

# I'd rather see one line per course

ggplot(all_pdp_df, aes(x = HSGPA, y = yhat, color = course)) +
  geom_line(size = 1) +
  labs(title = "PDP of HSGPA by Course",
       x = "High School GPA", y = "Predicted Grade GPA",
       color = "Course") +
  theme_minimal()

```

```{r}

## PARTIAL DEPENDENCE PER COURSE AND ACTMATH ##

# Loop through courses
pdp_ACTMATH <- lapply(courses, get_pdp_course, pred_var = "ACTMATH")
pdp_ACTMATH_df <- do.call(rbind, pdp_ACTMATH)


ggplot(pdp_ACTMATH_df, aes(x = ACTMATH, y = yhat, color = course)) +
  geom_line(size = 1) +
  labs(title = "PDP of ACTMATH by Course",
       x = "ACT Math Score", y = "Predicted Grade GPA",
       color = "Course") +
  theme_minimal()

```


```{r eval=FALSE}

# two variables

pdp_obj <- partial(model, pred.var = c("HSGPA", "ACTMATH"),
                   train = train_df, type = "regression")

ggplot(pdp_obj, aes(x = HSGPA, y = ACTMATH, fill = yhat)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "Partial Dependence on HSGPA and ACT Math",
       x = "HSGPA", y = "ACT Math") +
  theme_minimal()

# This is cool.
# I kinda already did it with an over-lay of actual plots, though

``` 



```{r eval=FALSE}

## ACCURATE LEAVES ## 

# Strategy 1:  Mine the XGBoost trees directly

model_dump <- xgb.model.dt.tree(model = model_list[["GRADEGPA"]]$finalModel)

leaves <- model_dump[Feature ==  "Leaf"]

# Get leaf assignments

leaf_ids <- predict(model_list[["GRADEGPA"]]$finalModel, newdata =  data_list[["GRADEGPA"]][["training"]][,-1] , predleaf = TRUE )

leaf_ids <- predict(model_list[["GRADEGPA"]]$finalModel, newdata =  model_list[["GRADEGPA"]]$trainingData , predleaf = TRUE )


leaf_analysis <- data.frame(
  leaf_id = leaf_ids[,1], #use first tree for simplicity or iterate through all
  actual_grade = 
  
)

# hitting some walls with this one.
# I'll look at it in the morning


```


```{r}

# STRATEGY 3: RECURSIVE PARTITIONING TO FIND POCKETS  

library(rpart)

trainData <- data_list[["GRADEGPA"]][["training"]]

trainData$pred <-  predict(model_list[["GRADEGPA"]], newdata =  trainData)  

trainData$absolute_error <- abs(trainData$GRADEGPA-trainData$pred)

startTime <- Sys.time()
accuracy_tree <- rpart(
  absolute_error ~ .,
  data = trainData[,colnames(trainData[setdiff(colnames(trainData), c("GRADEGPA", "pred"))])],
  control = rpart.control(
   minsplit = 100,
   maxdepth = 4,
   cp = 0.01
  )
)
  
endTime <- Sys.time() # 0.75 sec # Wow, fast!

# plot(accuracy_tree) 
# text(accuracy_tree, cex = 0.8)

# I can almost see it

trainData$accuracy_leaf <- predict(accuracy_tree, newdata = trainData)

leaf_summary <- trainData %>%
  mutate(leaf_id = as.character(accuracy_leaf)) %>%
  group_by(leaf_id) %>%
  summarise(
    n=n(),
    mean_error = mean(absolute_error),
    median_error = median(absolute_error),
    pct_within_half_grade = mean(absolute_error<0.5)*100,
    mean_hs_gpa = mean(HSGPA),
    mean_act = mean(ACTMATH)
  ) %>%
  arrange(mean_error)

# print(leaf_summary)

# I am struggling to interpret this.

# here's more from Chat

library(rpart.plot)

# Aggregate mean error per leaf (for coloring)
leaf_errors <- trainData %>%
  group_by(accuracy_leaf) %>%
  summarise(mean_error = mean(absolute_error), .groups = "drop")

# Map leaf predictions to colors
error_palette <- colorRampPalette(c("darkgreen", "yellow", "red"))  # low error = green, high = red
n_leaves <- nrow(leaf_errors)
leaf_colors <- setNames(error_palette(n_leaves), leaf_errors$accuracy_leaf)

# Plot tree with rpart.plot
rpart.plot(
  accuracy_tree,
  type = 3,               # type 3: label all nodes
  extra = 101,            # display mean response and n in each node
  box.palette = leaf_colors, # color nodes by mean error
  shadow.col = "gray", 
  fallen.leaves = TRUE,
  tweak = 1.2
)

# well isnt that interesting.
# pretty incredible, all told

# I guess the hyper-achievers are pretty careful?
# But it doesn't tell us which course to put them into


# I will look at this in the morning, I guess

  
```

A plot I want to see is ---
It's individual for a single person
x-axis has all of the courses
y-axis has the predicted grade, and the estimated spread of GPA.
It would be nice if I could show the spread of actual GPA for the people with that estimate


# APPENDIX

### Selecting high volume courses using hierarchical clustering

```{r eval = FALSE}

child_path <- here::here("Reports and Analysis", 
                         "Evaluate Decision Trees", 
                         "Children", 
                         "Selecting high volume math courses child.Rmd")

knitr::knit_child(child_path)


```

### Classification model results  

```{r child = here::here("Reports and Analysis", "Evaluate Decision Trees", "Children", "Confusion matrices for decision tree grades vD0 models child"), eval = FALSE}
```





