---
title: "Evaluate Course Decision Trees Over Time vD0"
author: "Bill Prisbrey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vD0"
output:
  html_document:
    keep_md: true
    toc: true
---





```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(stringr)
library(purrr)

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########


# Build file paths
testYears <- 2007:2024
grid <- expand.grid(params$version, testYears)

model_files <- here::here("Models", "Over Time", paste0("Decision Tree Courses ", grid$Var1, " ", grid$Var2, " model.rds"))

# Read this into a list

model_list <-
  lapply(testYears, function(tYr){
    files <- model_files[grep(tYr, model_files)]
    readRDS(files)
  })
names(model_list) <- testYears

# Build file paths
data_files <- here::here("Data", "Over Time", paste0("Decision Tree Courses ", grid$Var1, " ", grid$Var2, " Data.rds"))

# Read each data into a named list
data_list <-
  lapply(testYears, function(tYr){
    files <- data_files[grep(tYr, data_files)]
    readRDS(files)
  })
names(data_list) <- testYears


```


```{r}

# Extract all training data
# (Recombine the training data)

# trainData <- do.call(rbind, unique(map(data_list, "training")))

trainData <- data_list[["2024"]][["training"]]

trainData$course_option <- droplevels(trainData$course_option) # oops, thought I did that already

# optionsData_top90[optionsData_top90$enroll == "enroll", c("class_year","course_option")] |> table() |> as.data.frame.matrix() |> apply(MARGIN = 1, proportions)

# looks like the decline in accuracy falls off a cliff when the Math 1010 proportion declines



courseProportion <- trainData[trainData$enroll == "enroll", c("class_year","course_option")] |> table() |> as.data.frame.matrix() |> apply(MARGIN = 1, proportions) 

```


```{r eval=FALSE}

# plotting the course proportion

plot(x = mean(courseProportion),
     ylim = range(courseProportion),
     xlim = c(0,ncol(courseProportion)),
     ylab = "",
     xlab = "",
     type = "n")

cols <- RColorBrewer::brewer.pal(9, "Set1")

lapply(1:nrow(courseProportion), function(ln){ 
  
  lines(courseProportion[ln,], col = cols[ln]  )
  
  })

# I don't know why I'm doing this.  It's in "predictors over time"
# I guess it might be sharper in this filtered data set ?

```


```{r}

## PREDICTIONS ##

thePredictions <- lapply(as.character(testYears), function(x) {
  
  # identify course values in testing that aren't in training
  
#  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][["enroll"]],data_list[[x]][["training"]][["enroll"]])
  
  if (length(missingCourse) == 0) {
  # no missing values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingCourseFilter <- !data_list[[x]][["testing"]][["enroll"]] %in% missingCourse
  
#  print(x)
#  print(table(missingFilter))
}
  
  
        missingEthnicity <- setdiff(data_list[[x]] [["testing"]][["ETHNICITY"]],data_list[[x]][["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[x]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missingFilter <- missingCourseFilter & missingEthnicityFilter  

  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions) <- as.character(testYears)

```



```{r eval=FALSE}

## CHECK ##

sapply(thePredictions, length)
sapply(data_list, function(x){ nrow(x[["testing"]])  })

# I should dig into those discrepancies a little more
# looks like we added math courses or ethnicity levels;
# I need to be consistent in filtering them out # DONE


```

```{r}

theResults <- lapply(as.character(testYears), function(x) {

#    missingCourse <- setdiff(data_list[[x]][["testing"]][["enroll"]],data_list[[x]][["training"]][["enroll"]])
  
#  if (length(missingCourse) == 0) {
#  # no missing values — keep everything
#  missingCourseFilter <- rep(TRUE,  nrow(data_list[[x]][["testing"]]))
# } else {
#  # normal case — exclude the missing courses
#  missingCourseFilter <- !data_list[[x]][["testing"]][["enroll"]] %in% missingCourse
  
#  print(x)
#  print(table(missingFilter))
# }
  
  
        missingEthnicity <- setdiff(data_list[[x]] [["testing"]][["ETHNICITY"]],data_list[[x]][["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[x]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missingFilter <- missingEthnicityFilter # missingCourseFilter &   

  
  
cm <- confusionMatrix(data =  thePredictions[[x]], reference =data_list[[x]][["testing"]][["enroll"]][missingFilter] , mode = "everything" )

})
names(theResults) <- as.character(testYears)

nameMatrix <- cbind("original" = c(
  "MATH_1010", "MATH_1030", "MATH_1050", "MATH_1060", "MATH_1070", "MATH_1080", "MATH_1090", "MATH_1100", "MATH_1210", "MATH_1220", "MATH_1250", "MATH_1310", "MATH_2210", "MATH_990",  "other"
), "display" = c(
  "1010", "1030", "1050", "1060", "1070", "1080", "1090", "1100", "1210", "1220", "1250", "1310", "2210", "990", "other"
))




```


```{r}

########################
## RESULTS PER COURSE ##
########################

theCourses <- levels(data_list[["2024"]][["testing"]][["course_option"]])

# check
# theCourses <- sapply(data_list, function(dl){
  
#  levels(dl[["testing"]][["course_option"]])
  
#})
# all(apply(theCourses, 2, identical, theCourses[,1])) # TRUE
# all identical, as expected

courseResults <- lapply(theCourses, function(loop_course){ 

  # print(loop_course)
  
  
year_results <- lapply(as.character(testYears), function(loop_year) {

 #  print(loop_year)
  
  
          missingEthnicity <- setdiff(data_list[[loop_year]] [["testing"]][["ETHNICITY"]],data_list[[loop_year]][["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[loop_year]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[loop_year]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missingFilter <- missingEthnicityFilter # missingCourseFilter &   

  results_frame <- data.frame(pred = thePredictions[[loop_year]],
                              ref = data_list[[loop_year]][["testing"]][["enroll"]][missingFilter],
                              course_option = data_list[[loop_year]][["testing"]][["course_option"]][missingFilter] )
  
courseCM <- 
  
  confusionMatrix(
    data =  results_frame$pred[results_frame$course_option == loop_course],
    reference = results_frame$ref[results_frame$course_option == loop_course], 
    mode = "everything" )
  
return(courseCM)

})
names(year_results) <- testYears
return(year_results)

})
names(courseResults) <- theCourses




```


### SUMMARY

In order to examine trends in initial math course selection by first time freshmen over time, a series of predictive models were trained that used minimal and readily available information about first time freshmen. 

Eighteen predictive models were trained using extreme gradient boosting and an expanding window, where each year a consecutive year's worth of data was added and a new subsequent year's course selection was predicted. 

The predictive model found a moderate relationship between math course selection and predictors early on (Kappa of ~0.3 to 0.45), but that relationship eroded to a non-existent relationship by 2024 (when the Kappa value declined to 0.06).   

Kappa and precision started to abruptly declined in 2017, when enrollment in Math 1010 began declining and enrollment in other courses began increasing. 

The most important predictor of course selection in every year was the ACT test scores, followed by the combined standardized distance to the prior year's median successful student.  Distantly, high school GPA and the high school GPA z-score were next highest in model importance.  

Some courses such as Math 1220 increased in accuracy, peaking at a Kappa of 0.24 in 2020, while 1210 fluctuated between Kappa values of 0.06 (2017) and 0.39 (2010.)

The alignment between model accuracy and the decline in Math 1010 enrollment suggests that Math 1010 may have had stronger guidelines in directing students to select it, or was otherwise preferred at the expense of the other courses.    

Several conclusions are reasonable:

  * Until 2020, pre-University data (especially ACT math test scores) had a moderate to weak relationship with course selection.  
  * After 2020, pre-University data had a weak to no relationship with course selection.  
  * That the models are unable to identify and account for trends in the changing course selection mix, suggesting a cause that is not captured in the data.   
  * That some courses are too equivalent, or similar, or overlap with respect to the predictors that the model can't distinguish them.     

The overall decline in the model's ability to predict course selection suggests a weakening relationship between predictors like test scores and high school GPA, and math course selection.  Some possible explanations include: 

  * A new free-for-all in course selection absent of historical guidelines.    
  * The introduction of new guidelines that do not align with historical practices.  
  * The possibility of additional data that might influence course selection after admission, such as the choice of major influencing the choice of math course. 
  
### METHODOLOGY 

Data was prepared as described in the separate report, "Source Data, Cleaning, and Preparation."

Data was additionally filtered to the top 90th cumulative percentile in course enrollment over the full time period. The courses selected were Math 1010, 1030, 1050, 1060, 1070, 1080, 1090, 1210, and 1220.

Predictive models were trained with the package "xgboost." 

An expanding window of historical data was used, where one model was trained per year from 2007 to 2024.  Each model incrementally added an additional year's data and predicted on the subsequent year.  For example, the predicted course selection for the year 2020 was trained on data from the years 2005 to 2019.

Each row consisted of one of nine possible course selections, the student's data duplicated nine times, and the z-scores and combined distance per student per course calculated (relative to the prior year's median 'successful' or 'on track' student.)  The target column consisted of the values "enroll" or "not enroll", with one value of "enroll" per student identifying the course actually enrolled in.     

The models were evaluated with Kappa, precision, and recall. 


### DISCUSSION  

The relationship between early information and course selection was moderately predictive for many years, with Kappa values above 0.20 until 2019, and as high as 0.46 in 2009.  It began steeply declining in 2017 when it rapidly eroded to a Kappa value of 0.06 in 2024, or essentially no predictive ability or no relationship between the predictors and course selection.  

Kappa values also dipped in 2011 and 2012, suggesting a change in practices or guidelines or expectations for course selection.

The most important predictor over all years was the test scores for ACT math, followed by the combined standardized distance per course in ACT math and high school GPA.  This suggests the test score may have been explicitly used as a guideline in the past and remained an influential variable of course selection until 2023, when the model accuracy dipped below a Kappa value of 0.1 and found no relationship between the predictors and course selection.   

Some notable courses include: 

  - *Math 1010:*  The model was able to correctly identify participants in Math 1010 until the year 2020 with relatively high precision and recall scores (except for the 2011 dip), but declined steeply to Kappa and recall scores near zero in 2024.  
  - *Math 1210:*  The model correctly identified selection of Math 1210 with the second highest Kappa, ranging between a low of 0.06 in 2017 and a high of 0.39 in 2010.  In 2024, it had the third highest Kappa value of 0.12
  - *Math 1220:*  The course bucked the trend and started to increased in accuracy starting in 2015 and peaking in 2020 with a Kappa of 0.24.  In 2024, it had the second highest Kappa value of 0.16.   
  
Taken together, the overall decline in model accuracy (Kappa, precision, and recall) over the years suggests several possibilities:  

  (1) A collapse of guidelines and a free-for-all in selecting courses
  (2) The introduction of new guidelines or mechanisms that aligned weakly with prior practices. (At the same time, some courses with improving accuracy (such as Math 1220 with improving precision and recall) suggest (slightly) more rigid course selection.)  
  (3) The possibility of additional data that might influence course selection after admission, such as the choice of major influencing the choice of math course.
    


### CONCLUSIONS  

Initial math course selection for first time freshmen is decreasingly determined by variables known at the time of enrollment, until it is completely determined by variables other than pre-University information in 2024.

Predictive models had Kappa values as high as 0.46, indicating a moderate relationship between early information (such as high school performance and test scores) and course selection.  However, that relationship eroded rapidly starting in 2017 until Kappa values were as low as 0.06 by 2024, indicating an essentially random or no relationship between prior information and course selection. 

ACT math scores were the most important predictor of course selection, followed by the standardized distance in ACT math scores and high school GPA, until 2020 when the model declined to find a relationship.    

In summary:

  * Until 2020, pre-University data (especially ACT math test scores) had a moderate to weak relationship with course selection.  
  * After 2020, pre-University data had a weak to no relationship with course selection.  
  * The models are unable to identify and account for trends in the changing course selection mix, suggesting a cause that is not captured in the data.   
  * Some courses are possibly too equivalent, or similar, or overlap with respect to the predictors enough that the model can't distinguish them.

### POSSIBLE NEXT STEPS 

If the model can't distinguish between highly similar courses (in terms of incoming student qualifications), then the model might be able to distinguish between clusters of similar courses.  Repeating this analysis but with course clusters usefulness of data available at the time of admittance.   

A closer inspection of the confusion matrix for Math 1010, investigating which courses the model confused for Math 1010, could inform knowledge of the similarities between courses with respect to the model variables.  

A conversation with math instructors familiar with departmental history could inform the experiences suggested by these models.  

### RESULTS 


```{r}

# extract overall Kappa, precision, and recall

theKappas <- sapply(theResults, function(x) {x[["overall"]]["Kappa"]})
names(theKappas) <- names(theResults)  

thePrecisions <- lapply(theResults, function(results) {
    results[["byClass"]]["Precision"]
})
thePrecisions <- do.call(cbind, thePrecisions)  
  
theRecalls <- lapply(theResults, function(results) {
    results[["byClass"]]["Recall"]
})
theRecalls <- do.call(cbind, theRecalls)    



```

```{r}

# extract Kappa, precision, and recall by course

courseKappas <- lapply(courseResults, function(loop_course) {
  
  sapply(loop_course, function(x){
  
    x[["overall"]]["Kappa"]  
    
  })

  })

coursePrecision <- lapply(courseResults, function(loop_course) {
  
  sapply(loop_course, function(x){
  
    x[["byClass"]]["Precision"]  
    
  })

  })

courseRecall <- lapply(courseResults, function(loop_course) {
  
  sapply(loop_course, function(x){
  
    x[["byClass"]]["Recall"]  
    
  })

  })


```



```{r}

# plot overall Kappa

# plot Kappa
# plot(y = theKappas,
#     x= as.numeric(names(theKappas)),
#     type = "b",
#     col = "firebrick",
#     las = 2,
#     ylab = "Kappa",
#     xlab = ""
#     )

# This would make a good plotly

# library(plotly)

years <- as.numeric(names(theKappas))
kappa_values <- as.numeric(round(theKappas,3))

plot_ly(x = years, #[-length(years)], 
        y = kappa_values, #[-length(kappa_values)], 
        type = 'scatter', 
        mode = 'lines+markers',
        line = list(color = 'firebrick', width = 2),
        marker = list(color = 'firebrick', size = 8)) %>%
  layout(title = 'Model Performance (Kappa) Over Time',
         xaxis = list(title = '', dtick = 1, tickangle = -45),
         yaxis = list(title = 'Kappa'),
         hovermode = 'closest')

```




```{r}

# plot overall precision

plotData <- thePrecisions

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = gsub("Class: MATH_","", rownames(plotData)[i]), #     plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", gsub("Class: MATH_","", rownames(thePrecisions)[i]), "</b><br>",
        "Year: %{x}<br>",
        "Precision: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Precision Over Time',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Precision"),
    hovermode = "closest",
    legend = list(title = list(text = "Course"))
  )

p

```


```{r}

# plot overall recall

plotData <- theRecalls

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = gsub("Class: MATH_","", rownames(theRecalls)[i]), #     plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", gsub("Class: MATH_","", rownames(theRecalls)[i]), "</b><br>",
        "Year: %{x}<br>",
        "Recall: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Recall Over Time',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Recall"),
    hovermode = "closest",
    legend = list(title = list(text = "Course"))
  )

p


```

```{r}

# Plot kappa by course  

plotData <- do.call(rbind, courseKappas)
colnames(plotData) <- gsub(".Kappa", "", colnames(plotData))

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = gsub("Class: MATH_","", rownames(plotData)[i]), #     plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", gsub("Class: MATH_","", rownames(plotData)[i]), "</b><br>",
        "Year: %{x}<br>",
        "Kappa: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Kappa by course over time',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Kappa"),
    hovermode = "closest",
    legend = list(title = list(text = "Course"))
  )

p




```



```{r}

# Plot Recall by course  

plotData <- do.call(rbind, courseRecall)
colnames(plotData) <- gsub(".Recall", "", colnames(plotData))

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = gsub("Class: MATH_","", rownames(plotData)[i]), #     plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", gsub("Class: MATH_","", rownames(plotData)[i]), "</b><br>",
        "Year: %{x}<br>",
        "Recall: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Recall by course over time',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Recall"),
    hovermode = "closest",
    legend = list(title = list(text = "Course"))
  )

p




```

```{r}

# Plot Precision by course  

plotData <- do.call(rbind, coursePrecision)
colnames(plotData) <- gsub(".Precision", "", colnames(plotData))

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = gsub("Class: MATH_","", rownames(plotData)[i]), #     plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", gsub("Class: MATH_","", rownames(plotData)[i]), "</b><br>",
        "Year: %{x}<br>",
        "Precision: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Precision by course over time',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Precision"),
    hovermode = "closest",
    legend = list(title = list(text = "Course"))
  )

p




```

```{r}

# Variable importance over time

importance_over_time <- lapply(names(model_list), function(tYr){
  
  imp <- varImp(model_list[[tYr]])$importance
  imp$var <- row.names(imp)
  colnames(imp)[colnames(imp) %in% "Overall"] <- tYr # rename "Overall" column
  return(imp)
  
})

importance_over_time <- Reduce(function(m1,m2){merge(m1,m2, by = "var", all = TRUE)}, importance_over_time)
importance_over_time <- importance_over_time[order(importance_over_time[,2], decreasing = TRUE),] 

# why do I have NA values for ACTMATH.z ? 
# I guess it was just collinear with ACTMATH and dropped
# But wouldn't HSGPA mean the same ?



```



```{r eval=FALSE}

# Apparently I don't need this

# Metrics per course (one dataframe per course)
target_math_courses <- sapply(thePredictions, unique) |> unlist() |> as.character()  |> unique()

perCourse <- lapply(target_math_courses, function(math_course){
  
  course_over_time_list <- lapply(theResults, function(result){ 
   result[["byClass"]][grepl(math_course, row.names(result[["byClass"]])),]
    })
  
  course_over_time <- do.call(rbind, course_over_time_list)
  row.names(course_over_time) <- names(theResults)
  return(course_over_time)
  
})
names(perCourse) <- target_math_courses 

```





```{r}

# Plotly graphic of all predictors that have an importance value over 10 in any model

varFilter <- apply(importance_over_time[,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time$var[varFilter & !is.na(varFilter)]

# Filter to selected variables
plotData <- importance_over_time[importance_over_time[,"var"] %in% varSelection, ]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Variable Importance Over Time (all variables)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```



```{r eval=FALSE}

# Plotly graphic of all predictors that have an importance value less than fifty and over 10 in any model


varFilter <- apply(importance_over_time[,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time$var[varFilter & !is.na(varFilter)]

# Filter to selected variables
plotData <- importance_over_time[importance_over_time[,"var"] %in% varSelection[!varSelection %in% c("ACTMATH", "course_level")], ]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Variable Importance Over Time (low importance variables)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```


```{r fig.height=5, eval=FALSE}

# This shows variable importance with a separate graph for each year.

# It's not a bad appendix item, though


invisible(

    lapply(testYears, function(tYear){
       print(plot(  
           varImp(model_list[[as.character(tYear)]]),
           15, main = paste(tYear, sep = ": ")
       ))
           })
)

# course is a lot harder.

# let's pull in what we have


```


```{r}

# An example confusion matrix

drawCM(theResults[["2024"]], title = "2024 Results")

#,nameMatrix = nameMatrix, metricsClass = "2210", title = "2024 Results") 

# precision color is screwed up

```  

```{r}

skim(data_list[["2024"]][["training"]])

```

