---
title: "Evaluate Course Decision Trees Over Time vC0"
author: "Bill Prisbrey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vC0"
output:
  html_document:
    keep_md: true
---



```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(stringr)

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########


# Build file paths
testYears <- 2007:2025
grid <- expand.grid(params$version, testYears)

model_files <- here::here("Models", "Over Time", paste0("Decision Tree Courses ", grid$Var1, " ", grid$Var2, " model.rds"))

# Read this into a list

model_list <-
  lapply(testYears, function(tYr){
    files <- model_files[grep(tYr, model_files)]
    readRDS(files)
  })
names(model_list) <- testYears

# Build file paths
data_files <- here::here("Data", "Over Time", paste0("Decision Tree Courses ", grid$Var1, " ", grid$Var2, " Data.rds"))

# Read each data into a named list
data_list <-
  lapply(testYears, function(tYr){
    files <- data_files[grep(tYr, data_files)]
    readRDS(files)
  })
names(data_list) <- testYears


```


```{r}

## PREDICTIONS ##

thePredictions <- lapply(as.character(testYears), function(x) {
  
  # identify course values in testing that aren't in training
  
#  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missingCourse <- setdiff(data_list[[x]][["testing"]][["course_target"]],data_list[[x]][["training"]][["course_target"]])
  
  if (length(missingCourse) == 0) {
  # no missing values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingCourseFilter <- !data_list[[x]][["testing"]][["course_target"]] %in% missingCourse
  
#  print(x)
#  print(table(missingFilter))
}
  
  
        missingEthnicity <- setdiff(data_list[[x]] [["testing"]][["ETHNICITY"]],data_list[[x]][["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[x]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missingFilter <- missingCourseFilter & missingEthnicityFilter  

  
  predict(model_list[[x]],
          newdata = data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions) <- as.character(testYears)

```

# Results

Develop and report the confusion matrices 

```{r eval=FALSE}

## CHECK ##

sapply(thePredictions, length)
sapply(data_list, function(x){ nrow(x[["testing"]])  })

# I should dig into those discrepancies a little more
# looks like we added math courses or ethnicity levels;
# I need to be consistent in filtering them out # DONE


```

```{r}

theResults <- lapply(as.character(testYears), function(x) {

    missingCourse <- setdiff(data_list[[x]][["testing"]][["course_target"]],data_list[[x]][["training"]][["course_target"]])
  
  if (length(missingCourse) == 0) {
  # no missing values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingCourseFilter <- !data_list[[x]][["testing"]][["course_target"]] %in% missingCourse
  
#  print(x)
#  print(table(missingFilter))
}
  
  
        missingEthnicity <- setdiff(data_list[[x]] [["testing"]][["ETHNICITY"]],data_list[[x]][["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[x]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missingFilter <- missingCourseFilter & missingEthnicityFilter  

  
  
cm <- confusionMatrix(data =  thePredictions[[x]], reference =data_list[[x]][["testing"]][["course_target"]][missingFilter] , mode = "everything" )

})
names(theResults) <- as.character(testYears)

nameMatrix <- cbind("original" = c(
  "MATH_1010", "MATH_1030", "MATH_1050", "MATH_1060", "MATH_1070", "MATH_1080", "MATH_1090", "MATH_1100", "MATH_1210", "MATH_1220", "MATH_1250", "MATH_1310", "MATH_2210", "MATH_990",  "other"
), "display" = c(
  "1010", "1030", "1050", "1060", "1070", "1080", "1090", "1100", "1210", "1220", "1250", "1310", "2210", "990", "other"
))




```


```{r}

# extract Kappa

theKappa <- sapply(theResults, function(x) {x[["overall"]]["Kappa"]})
names(theKappa) <- names(theResults)  

# plot Kappa
# plot(y = theKappa,
#     x= as.numeric(names(theKappa)),
#     type = "b",
#     col = "firebrick",
#     las = 2,
#     ylab = "Kappa",
#     xlab = ""
#     )

# This would make a good plotly

# library(plotly)

years <- as.numeric(names(theKappa))
kappa_values <- as.numeric(round(theKappa,3))

plot_ly(x = years, 
        y = kappa_values, 
        type = 'scatter', 
        mode = 'lines+markers',
        line = list(color = 'firebrick', width = 2),
        marker = list(color = 'firebrick', size = 8)) %>%
  layout(title = 'Model Performance (Kappa) Over Time',
         xaxis = list(title = '', dtick = 1, tickangle = -45),
         yaxis = list(title = 'Kappa'),
         hovermode = 'closest')

```

```{r}

# Metrics per course (one dataframe per course)
target_math_courses <- sapply(thePredictions, unique) |> unlist() |> as.character()  |> unique()

perCourse <- lapply(target_math_courses, function(math_course){
  
  course_over_time_list <- lapply(theResults, function(result){ 
   result[["byClass"]][grepl(math_course, row.names(result[["byClass"]])),]
    })
  
  course_over_time <- do.call(rbind, course_over_time_list)
  row.names(course_over_time) <- names(theResults)
  return(course_over_time)
  
})
names(perCourse) <- target_math_courses 

```


```{r}

importance_over_time <- lapply(names(model_list), function(tYr){
  
  imp <- varImp(model_list[[tYr]])$importance
  imp$var <- row.names(imp)
  colnames(imp)[colnames(imp) %in% "Overall"] <- tYr # rename "Overall" column
  return(imp)
  
})

importance_over_time <- Reduce(function(m1,m2){merge(m1,m2, by = "var", all = TRUE)}, importance_over_time)
importance_over_time <- importance_over_time[order(importance_over_time[,2], decreasing = TRUE),] # descending order by the first year of data

```



```{r}

# Plotly graphic of all predictors that have an importance value over 10 in any model

varFilter <- apply(importance_over_time[,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time$var[varFilter & !is.na(varFilter)]

# Filter to selected variables
plotData <- importance_over_time[importance_over_time[,"var"] %in% varSelection, ]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Variable Importance Over Time (all variables)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```



```{r}

# Plotly graphic of all predictors that have an importance value less than fifty and over 10 in any model


varFilter <- apply(importance_over_time[,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time$var[varFilter & !is.na(varFilter)]

# Filter to selected variables
plotData <- importance_over_time[importance_over_time[,"var"] %in% varSelection[!varSelection %in% c("ACTMATH", "course_level")], ]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Variable Importance Over Time (low importance variables)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```


```{r fig.height=5, eval=FALSE}

# This shows variable importance with a separate graph for each year.

# It's not a bad appendix item, though


invisible(

    lapply(testYears, function(tYear){
       print(plot(  
           varImp(model_list[[as.character(tYear)]]),
           15, main = paste(tYear, sep = ": ")
       ))
           })
)

# course is a lot harder.

# let's pull in what we have


```


Should I put a heatmap?  Really would tell the story well, don't you think?

I should put in an example confusion matrix.  


Possible next steps:
- explain NA values
- explain course_level strong showing


Show precision/recall over time

```{r}

# An example confusion matrix

drawCM(theResults[["2024"]],nameMatrix = nameMatrix, metricsClass = "1050") # eh.

# precision color is screwed up

```  



```{r}

# extract precision and recall

thePrecisions <- lapply(theResults, function(results) {
    results[["byClass"]][,"Precision"]
})
thePrecisions <- do.call(cbind, thePrecisions)  
  
theRecalls <- lapply(theResults, function(results) {
    results[["byClass"]][,"Recall"]
})
theRecalls <- do.call(cbind, theRecalls)    



```


```{r}

# Precision and recall per course over time

plotData <- thePrecisions

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = gsub("Class: MATH_","", rownames(thePrecisions)[i]), #     plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", gsub("Class: MATH_","", rownames(thePrecisions)[i]), "</b><br>",
        "Year: %{x}<br>",
        "Precision: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Precision Over Time',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Precision"),
    hovermode = "closest",
    legend = list(title = list(text = "Course"))
  )

p


plotData <- theRecalls

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p %>%
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = gsub("Class: MATH_","", rownames(theRecalls)[i]), #     plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", gsub("Class: MATH_","", rownames(theRecalls)[i]), "</b><br>",
        "Year: %{x}<br>",
        "Recall: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p %>%
  layout(
    title = 'Recall Over Time',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Recall"),
    hovermode = "closest",
    legend = list(title = list(text = "Course"))
  )

p


```


Where is my "cleaning" and "prep" boilerplate?


So accuracy at predicting courses drops incredibly.
And I'm not sure I want to include "course_level", but why was it only super-important for a few years?  

I want to see the median grade per course over time.  What's my trend over time?

Similarly -- maybe I want to see the proportion of students achieving a particular grade level or grade chunk.  It's getting big, right?  Could that show grade inflation? 

And we're getting closer to counter-factuals. What grade would they get in a particular course?

I need to submit this to Git, and write up both reports, and get a couple of key visualizations over time. 

I'd like to see "precision" for the top three courses over time.  And "recall", sure.

I need to make the confusion matrices work--  

  - name matrix
  - whatever that error it's throwing 
  
I think I'll make a separate "over time" report that explores the predictors for both grades and course over time.  

I'll work on all three reports in parallel -- 

  - Prediction results  (course and grade)  
  - Predictors over time
  

### SUMMARY

### METHODOLOGY 

All math courses taken by first time freshmen were identified.  The term per student with that student's first math class was identified, and grade performance in all math courses taken during that term were queried.  These are called the initial math courses.   
Math labs were filtered out.

The early information variables included: 

*Demographics:* 

  * Age (less than 17 yrs old/17/18/19 or 20/greater than 21 years old)
  * Sex (male or female)
  * First generation status
  * Ethnicity 
  
*High school information and performance:*  

  * High school GPA 
  * Number of AP credits
  * Whether the high school was public or private   
  * Honors achievement

*Test scores:*    

  * ACT Math
  * ACT English
  * ACT Science
  * ACT comprehensive

*University information:*  

  * Residence status  
  * Pell grant eligibility  
  * Cohort year (2005 to 2024)
  * Class year (2005 to 2025)
  * Years from the cohort date (always fall of the year) until the first term with math courses 
  * Season (spring, summer, or fall)  


The following filters were applied: 

 * Courses higher than 3000 level were removed. 
 * Students younger than 11 yrs old were removed. 
 * Withdraw and "rare" grades of V, I, NC, and CR were removed.  
 * Students with high school GPA values of zero were removed. 
 * Courses from students with unusually high credits taken that semester (greater than 99th percentile) were removed.  
 * Courses taken more than half a year prior to the cohort date (where all cohort dates are in the fall) were removed   
 * All records with incomplete information were removed.  

The following transformations were applied: 

  * AP credit values of "NA" were converted to zero.  
  * Math courses with ordered enrollment above the 95th percentile of cumulative total were combined as "other"  
  
A predictive model was trained with package "xgboost" (extreme gradient boosting) with five-fold cross validation and a grid search pattern (nrounds = (100,200); max_depth(3,6); eta(0.05,0.2);colsample_bytree(0.8); subsample(0.8)).  

An expanding window of historical data was used, where one model was trained per year, incrementally adding an additional year's data and predicting on the subsequent year.  For example, the course selection for the year 2010 was trained on data from the years 2005 to 2009.  The course selection for the year 2024 was trained on data from the years 2005 to 2023.  Nineteen models were trained. 

Each model predicted a value for a math course out of the following possibilities:  

    "MATH_1010", "MATH_1050", "MATH_1210", "MATH_1030", "MATH_1090", "MATH_1070",   "MATH_1220", "MATH_1080", "MATH_1060", "MATH_1310", "MATH_2210", "MATH_990",  "MATH_1100", "MATH_1250" or "other".
 
The following math courses were combined into "other":  

    "MATH_1040, MATH_1311, MATH_1260, MATH_950, MATH_1270, MATH_1320, MATH_1321, MATH_1215, MATH_1170, MATH_2270, MATH_2250, MATH_1035, MATH_2000, MATH_1280, MATH_2200, MATH_1105, MATH_2271, MATH_2015, MATH_2280, MATH_2160, UC_1000, MATH_1180, MATH_1010, MATH_1030, MATH_1050, MATH_1060, MATH_1070, MATH_1080, MATH_1090, MATH_1100, MATH_1210, MATH_1220, MATH_1250, MATH_1310, MATH_2210, MATH_990"
  
*this belongs in the other report, not here!*
In one model, the actual GPA performance was targeted with the following values:  

  |         |***GPA model***| ***Quad Model*** |
  | Letter  | GPA           |  Grouping        |
  | E, EU   | 0             | "Low"            |
  | D-      | 0.7           | "Low"            |
  | D       | 1.0           | "Low"            |
  | D+      | 1.3           | "Low"            |
  | C-      | 1.7           | "Med-Minus"      | 
  | C       | 2.0           | "Med-Minus"      |  
  | C+      | 2.3           | "Med-Minus"      |
  | B-      | 2.7           | "Med-Plus"       |
  | B       | 3.0           | "Med-Plus"       |
  | B+      | 3.3           | "Med-Plus"       |
  | A-      | 3.7           | "Med-Plus"       |
  | A       | 4.0           | "High"           |


### DISCUSSION  

The relationship between early information and course selection was moderate for many years, until it started eroding in 2016 and approached random (or a very weak relationship) after covid (2021 onwards).  

The most important predictor over all years, by far, was test scores for ACT math.  This suggests the test score may have been explicitly used as a guideline in the past and remains the most important variable in predicting course selection.  

The model was able to correctly identify participants in Math 1010 until the year 2022 with a reasonably high recall score of 0.74, after which recall declined to 0.08 in 2024.  In 2024, the model's greatest success was in identifying Math 1210 with a recall score of 0.52 (which course had also been gently declining over time.) 

Historically, the model could most accurately predict 2210 with a precision of 0.93 (2014) and 1010 with a precision of 0.83 (2016.)  In 2024, the most precise predictions were 2210 (0.44) and 1090 (0.39).

Taken together, the decline in model accuracy (Kappa, precision, and recall) over the years suggests a collapse of guidelines and either a free-for-all in selecting courses, or the use of some other guideline or mechanism that aligns weakly with prior knowledge.


### CONCLUSIONS  

Initial math course selection for first time freshmen is decreasingly determined by variables known at the time of enrollment, until it is almost completely determined by variables other than early available information in 2024.

Predictive models had Kappa values as high as 0.4, indicating a moderate degree of determinism between early information (such as high school performance and test scores) and course selection.  However, that relationship eroded rapidly starting in 2016 until 2024 Kappa values were as low as 0.13, indicating a near-random relationship between prior information and course selection. 

All models in all years used ACTMATH scores as the most important predictor of course selection.  APCREDIT increased in importance in predicting math course selection, as did information about the class year and cohort year.  




By 2024, models that use a handful of variables are almost completeley unable to predict initial course selection.  Or, in other words, course selection is almost completely determined by factors other than these variables.  This can be compared to 2013 to 2016, 


### RESULTS 

Put graphics down here. 
