---
title: "Evaluate Grade Decision Trees Over Time `r params$version`"
author: "Bill Prisbrey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  version: "vQ0"
  target_classes: 
    - GRADEGPA
output:
  html_document:
    keep_md: true
    toc: true
---



```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########

# Strikes me that I want one list per target class, with multiple years per

# Build file paths
testYears <- 2007:2025
grid <- expand.grid(params$version, testYears,  params$target_classes)

model_files <- here::here("Models", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " model.rds"))

# Read this into a list of lists by target class

model_list <- setNames(
  lapply(params$target_classes, function(tClass){
    files <- model_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  params$target_classes
)



# Build file paths
data_files <- here::here("Data", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(params$target_classes, function(tClass){
    files <- data_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  params$target_classes
)


```

```{r eval=FALSE}

# Let's check a few things

lapply(data_list, function(tClass){
  lapply(tClass, function(tYr){
  
    print(paste("training", nrow(tYr[["training"]]), sep = ": "))
    print(paste("testing", nrow(tYr[["testing"]]), sep = ": "))
invisible(NULL)
  })
})

# It did it--- not the format I was expecting
# And 2006 went wrong for some reason, but I can filter it out.  Oh, because I didn't over-write it from the first pass.  I could delete it.  

for(tClass in data_list){
  for(tYr in tClass){
       print(paste("training:", levels(factor(tYr[["training"]][["season"]]))))
    print(paste("testing:", levels(factor(tYr[["testing"]][["season"]]))), collapse = ", ")
  }
}

# no fall for 2025 yet

``` 


```{r}

## PREDICTIONS ##

predictions_over_time <- lapply(params$target_classes, function(tClass){
  perClass <- lapply(as.character(testYears), function(tYr){
    
#  courseColumn <- ifelse("course" %in% colnames(data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
    
    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  
            
      
    # predict 
    predict(model_list[[tClass]][[tYr]],
            newdata = data_list[[tClass]][[tYr]][["testing"]][missing,]
            )
  })
  names(perClass) <- testYears
  return(perClass)
})
names(predictions_over_time) <- params$target_classes

# predictions_over_time[["GRADEGPA"]][["2024"]] |> hist() # looks o.k.



```


```{r}

theResults <- lapply(params$target_classes, function(tClass){
  
  perClass <- lapply(as.character(testYears), function(tYr){

    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  

       postResample(pred = predictions_over_time[[tClass]][[tYr]],
                  obs = data_list[[tClass]][[tYr]][["testing"]][[tClass]][missingCourseFilter]
               ) 
    
        
  })
  names(perClass) <- testYears
  return(perClass)
})
  
names(theResults) <- params$target_classes

```



```{r}

# Success over time

# resultsDF <- lapply(theResults[["GRADEGPA"]], function(x){
  
#  do.call(rbind, x)
  
# })

resultsDF <- do.call(rbind, theResults[["GRADEGPA"]])

```


### SUMMARY 

### METHODOLOGY 

Data was prepared as described in the separate report, "Source Data, Cleaning, and Preparation".  

An additional filter was applied to include only the top 90th cumulative percentile of courses in attendance over all years.  

Predictive models were trained with package "xgboost".

An expanding window of historical data was used, where one model was trained per year from 2007 to 2024.  Each model incrementally added an additional year's data and predicted on the subsequent year.  For example, the predicted grades for the year 2020 was trained on data from the years 2005 to 2019.  

A value for math grade was predicted, and evaluated with root mean square error, mean absolute error, and r-squared values.

#### Variables 

The following variables were included: 

*Demographics:* 

  * Age (bucketed as less than 17 yrs old/17/18/19 or 20/greater than 21 years old)
  * Sex (male or female)
  * First generation status
  * Ethnicity 
  
*High school:*  

  * High school GPA 
  * Number of AP credits
  * Whether the high school was public or private   

*Test scores:*    

  * ACT Math
  * ACT English
  * ACT Science
  * ACT comprehensive

*University information:*  

  * Math course taken 
  * Course level (as 1000 or 2000)
  * Residence status  
  * Pell grant eligibility  
  * Cohort year (2005 to 2024)
  * Class year (2005 to 2025)
  * Years from the cohort date (always fall of the year) until the first term with math courses (as yr_diff)
  * Season (spring, summer, or fall)  
  * Honors status

*Derived features:* 

  * The high school GPA converted to a z-score distance from the prior year's course median successful student  
  * The ACT math test scores converted to a z-score distance from the prior year's course median successful student  
  * The combined Euclidean distance of these z-scores 

### DISCUSSION  

### CONCLUSIONS  

### RESULTS 


```{r}

# I want three plotly graphics, one per RMSE, R-squared, and MAE
# Each plot has one line per GRADEGPA and grade_quad

# Extract years from row names
years <- as.numeric(rownames(resultsDF))[-19]

# Create RMSE plot
plot_rmse <- plot_ly() |>
  add_trace(x = years, 
            y = resultsDF[-19,"RMSE"], 
            name = "GPA Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
#  add_trace(x = years, 
#            y = resultsDF[["grade_quad"]][-19,"RMSE"], 
#            name = "Quad Model",
#            type = 'scatter', 
#            mode = 'lines+markers') |>
  layout(title = "RMSE Over Years",
         xaxis = list(title = "Year"),
         yaxis = list(title = "RMSE"))

# Create R-squared plot
plot_rsq <- plot_ly() |>
  add_trace(x = years, 
            y = resultsDF[-19,"Rsquared"], 
            name = "GPA Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
#  add_trace(x = years, 
#            y = resultsDF[["grade_quad"]][-19,"Rsquared"], 
#            name = "Quad Model",
#            type = 'scatter', 
#            mode = 'lines+markers') |>
  layout(title = "R-squared Over Years",
         xaxis = list(title = "Year"),
         yaxis = list(title = "R-squared"))

# Create MAE plot
plot_mae <- plot_ly() |>
  add_trace(x = years, 
            y = resultsDF[-19,"MAE"], 
            name = "GPA Model",
            type = 'scatter', 
            mode = 'lines+markers') |>
#  add_trace(x = years, 
#            y = resultsDF[["grade_quad"]][-19,"MAE"], 
#            name = "Quad Model",
#            type = 'scatter', 
#            mode = 'lines+markers') |>
  layout(title = "MAE Over Years",
         xaxis = list(title = "Year"),
         yaxis = list(title = "MAE"))

# Display the plots
plot_rsq
plot_mae
plot_rmse


```



```{r}

# extract variable importance
# re-shape into a data frame per target class with one column per year

importance_over_time <- lapply(model_list, function(tClass) {
  
     # Extract importance
      impList <- lapply(tClass, function(tYr){
        
        imp <- varImp(tYr)$importance
        imp$var <- row.names(imp)
        return(imp)
      })  
      
    # Re-name 'overall' column  
      
    impList <- lapply(names(impList), function(yearName){
      
      colnames(impList[[yearName]])[colnames(impList[[yearName]]) == "Overall"] <- yearName
      return(impList[[yearName]])
    })  
      
  # Collapse into one data frame
    
    impOverTime <- Reduce(function(m1,m2){merge(m1,m2, by = "var", all = TRUE)}, impList)
    impOverTime <- impOverTime[order(impOverTime[,2], decreasing = TRUE),] # descending order by the first year of data
    
      return(impOverTime)
})


```





```{r}

# plotly version of importance_over_time

# eliminate year 2025

lastCol <- ncol(importance_over_time[["GRADEGPA"]])

testYears <- testYears[-which(testYears %in% "2025")]

varFilter <- apply(importance_over_time[["GRADEGPA"]][,-1*c(1, lastCol )],1, function(x){any(x>10)}) 

varSelection <- importance_over_time[["GRADEGPA"]][["var"]][varFilter & !is.na(varFilter)]

# varSelection <- importance_over_time[["GRADEGPA"]][["var"]][order(importance_over_time[["GRADEGPA"]][["2024"]], decreasing = TRUE)] |>
#  (\(x){x[2:10]})() # top 8 predictors in 2024

# Filter to selected variables
plotData <- importance_over_time[["GRADEGPA"]][
  importance_over_time[["GRADEGPA"]][,"var"] %in% varSelection,-lastCol 
]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p |>
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p |>
  layout(
    title = 'Variable Importance Over Time (all variables, GPA)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```



```{r eval=FALSE}

# plotly version of importance_over_time

varFilter <- apply(importance_over_time[["grade_quad"]][,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time[["grade_quad"]][["var"]][varFilter & !is.na(varFilter)]


# Filter to selected variables
plotData <- importance_over_time[["grade_quad"]][
  importance_over_time[["grade_quad"]][,"var"] %in% varSelection, 
]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p |>
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p |>
  layout(
    title = 'Variable Importance Over Time (all variables, Quad)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```



```{r eval=FALSE}

# I just don't care

# plotly version of importance_over_time, low importance variables only

varFilter <- apply(importance_over_time[["GRADEGPA"]][,-1],1, function(x){any(x>10)}) 

varSelection <- importance_over_time[["GRADEGPA"]][["var"]][varFilter & !is.na(varFilter)]

varSelection <- varSelection[-which(varSelection %in% "HSGPA")]

# varSelection <- importance_over_time[["GRADEGPA"]][["var"]][order(importance_over_time[["GRADEGPA"]][["2024"]], decreasing = TRUE)] |>
#  (\(x){x[2:10]})() # top 8 predictors in 2024

# Filter to selected variables
plotData <- importance_over_time[["GRADEGPA"]][
  importance_over_time[["GRADEGPA"]][,"var"] %in% varSelection, 
]

# Initialize plot
p <- plot_ly()

# Add a line for each variable
for(i in 1:nrow(plotData)) {
  p <- p |>
    add_lines(
      x = testYears,
      y = as.numeric(plotData[i, as.character(testYears)]),
      name = plotData[i, "var"],
      hovertemplate = paste0(
        "<b>", plotData[i, "var"], "</b><br>",
        "Year: %{x}<br>",
        "Importance: %{y:.2f}<br>",
        "<extra></extra>"
      )
    )
}

# Customize layout
p <- p |>
  layout(
    title = 'Variable Importance Over Time (low importance variables, GPA)',
    xaxis = list(title = "Test Year"),
    yaxis = list(title = "Variable Importance"),
    hovermode = "closest",
    legend = list(title = list(text = "Variable"))
  )

p


```


Discretize and create a confusion matrix

```{r}

# Discretize and create a couple of confusion matrices
# Should I do the high accuracy, or preserve the C+/B- barrier? 





```



```{r}

# Extract accuracy metrics per course

courses <- levels(data_list[["GRADEGPA"]][["2024"]][["testing"]]$course)

courseResults <- lapply(params$target_classes, function(tClass){
  
  perClass <- lapply(as.character(testYears), function(tYr){

    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["course"]],data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  

# Iterate through each course at a time
  
perCourse <- lapply(courses, function(crse){ 

         course_filter <- data_list[[tClass]][[tYr]][["testing"]][["course"]] == crse
  
       postResample(pred = predictions_over_time[[tClass]][[tYr]][course_filter],
                  obs = data_list[[tClass]][[tYr]][["testing"]][[tClass]][missingCourseFilter][course_filter]
               ) 
    
        
  })

names(perCourse) <- courses
return(perCourse)

})

  names(perClass) <- testYears
  return(perClass)
})
  
names(courseResults) <- params$target_classes

# I could have built this into the lapply, but I'll work with it

gradeGPA_by_course <- lapply(courseResults[["GRADEGPA"]], function(x){ 
  
  do.call(rbind, x)
  })

# ok, I want a list of three dataframes, courses as row names and years as colnames

accuracy_metrics <- c("RMSE", "Rsquared", "MAE")  

course_over_time <- lapply(accuracy_metrics, function(metric){
  
  metricExtract <- lapply(gradeGPA_by_course, function(x){ 
    anExtract <- x[,metric]
    return(anExtract)
    })
  
  course_per_year <- do.call(cbind, metricExtract)
  
  return(course_per_year)
})

names(course_over_time) <- accuracy_metrics

# if I want, I could do "grade_quad"

``` 


```{r eval=FALSE}

# Check 2024 values for RMSE

course_counts2024 <- table(data_list[["GRADEGPA"]][["2024"]][["testing"]][["course"]])

weighted_rmse2024 <- sqrt(sum((gradeGPA_by_course[["2024"]][,"RMSE"]^2 *
                                 course_counts2024[rownames(gradeGPA_by_course[["2024"]])]), na.rm = TRUE) / 
        sum(course_counts2024[rownames(gradeGPA_by_course[["2024"]])], na.rm = TRUE))

cat("Overall RMSE:", theResults[["GRADEGPA"]][["2024"]]["RMSE"], "\n")
# Overall RMSE: 1.012924 
cat("Weighted RMSE from courses:", weighted_rmse2024, "\n")
# Weighted RMSE from courses: 1.012924 

# Looks like we're good!

```

```{r}

# This is probably worth fixing

# plot accuracy per course over time  

plot_course_metric <- function(metric_df, metric_name = NULL) {
  # library(plotly) # already called for this report
  
  # Get metric name from list if not provided
  if(is.null(metric_name)) {
    metric_name <- "Metric"
  }
  
  # Get courses and years
  courses <- rownames(metric_df)
  years <- as.numeric(colnames(metric_df))
  
  # Initialize plot
  p <- plot_ly()
  
  # Add a line for each course
  for(course in courses) {
    p <- p %>%
      add_trace(x = years,
                y = as.numeric(metric_df[course, ]),
                name = course,
                type = 'scatter',
                mode = 'lines+markers')
  }
  
  # Add layout
  p <- p %>%
    layout(title = paste(metric_name, "by Course Over Years"),
           xaxis = list(title = "Year"),
           yaxis = list(title = metric_name),
           hovermode = "closest")
  
  return(p)
}

# Usage:
plot_course_metric(course_over_time[["Rsquared"]], "R-squared")
plot_course_metric(course_over_time[["MAE"]], "MAE")
plot_course_metric(course_over_time[["RMSE"]], "RMSE")





```

# Representative data from the year 2024

```{r}

skim(data_list[["GRADEGPA"]][["2024"]][["testing"]])

```


