---
title: "Summary of findings from predicting math success"
author: "Bill Prisbrey"
date: "2025-11-07"
output: html_document
---

```{r}

knitr::opts_chunk$set(echo = TRUE)

source(here::here("Functions", "Predict Math Success Functions.R"))

library(extrafont)
loadfonts()

```


Highlights  (work, brain, work) 

Brain, brain, brain

Strongest predictors of grade performance is HSGPA and ACTMATH. 
Strongest predictors of course is ACTMATH.

Ability to predict course is declining over time, almost gone.

Ability to predict grade is about plus-or-minus 0.6 GPA points (R-sq of 0.3) consistently 

Course seems irrelevant to predicting the grade, although developing a counterfactual will shed more light into that.  

2020 --- discontinuity, and 2024 is esp weird.

I probably want to put a couple of highlights here, but otherwise call the child reports.  Or they are separate reports. 

Looks like ACTMATH is declining in importance.  Isn't that interesting.  

Let's trim off 2025 from all of these graphics, then work on the predictor graphic.  





```{r fig.height=5, fig.width=10}

displayPerformance()

```

```{r fig.height=5, fig.width=11}

displayPipeline()

```