---
title: "Data reconciliation"
author: "Bill Prisbrey"
date: "2025-10-30"
output:
  html_document:
    keep_md: true
---

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(lubridate)
library(tidyverse)
library(readxl)
library(skimr)

```




```{r}

##########
## LOAD ##
##########


ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

whData <- read_excel(here::here("Data", "WH Dataset - Math Placement Project - First Time Students from Fall 2021 - 2024 Cohorts and Their First Math Course 2025.09.30.xlsx"))



```


```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

# add unique identifier key here 
mathCourses$key <- paste(mathCourses$EMPLID, mathCourses$CATNBR, mathCourses$TERM, sep = "_")
                   
## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", "COHORTTERM", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course[recent_filter]) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course[recent_filter]) |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.9]) })()

courseFilter <- firstCourses$course %in% higherVolCourses

# > higherVolCourses
# [1] "MATH_1010" "MATH_1050" "MATH_1210" "MATH_1030" "MATH_1090" "MATH_1060"
# [7] "MATH_1070" "MATH_1220" "MATH_1080"

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

# print("Prep complete")

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# Unique Identifier
# firstCourses$key <- paste(firstCourses$EMPLID, firstCourses$CATNBR, firstCourses$TERM, sep = "_")

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)

cleanData$course <- factor(cleanData$course)



```


```{r}

##########
## PREP ##
##########

whData <- whData[!is.na(whData$FIRST_MATH_COURSE_CATNBR),]

whData$key <- paste(whData$EMPLID, whData$FIRST_MATH_COURSE_CATNBR, whData$FIRST_MATH_TERM, sep = "_")


```


```{r}

############################
## IDENTIFY DISCREPANCIES ##
############################

whPerTerm <- lapply(unique(whData$COHORTTERM), function(x) {
  emplidPerTerm <- unique(whData$EMPLID[whData$COHORTTERM == x]) 
  keyPerTerm <- unique(whData$key[whData$COHORTTERM == x])
  return(list(emplid = emplidPerTerm, key = keyPerTerm))
  })
names(whPerTerm) <- unique(whData$COHORTTERM)

cleanPerTerm <- lapply(unique(cleanData$COHORTTERM), function(x) {
  emplidPerTerm <- unique(cleanData$EMPLID[cleanData$COHORTTERM == x]) 
  keyPerTerm <- unique(cleanData$key[cleanData$COHORTTERM == x])
  return(list(emplid = emplidPerTerm, key = keyPerTerm))
})
names(cleanPerTerm) <- unique(cleanData$COHORTTERM)


comboPerTerm <- lapply(unique(whData$COHORTTERM), function(x){ 
  x <- as.character(x)
  EMPLIDs <- merge(data.frame(emplid = whPerTerm[[x]][["emplid"]], wh = TRUE), data.frame(emplid = cleanPerTerm[[x]][["emplid"]], clean = TRUE), by = "emplid", all = TRUE  )
  EMPLIDs[is.na(EMPLIDs)] <- FALSE
  keys <- merge(data.frame(key = whPerTerm[[x]][["key"]], wh = TRUE), data.frame(key = cleanPerTerm[[x]][["key"]], clean = TRUE), by = "key", all = TRUE  )
  keys[is.na(keys)] <- FALSE
  return(list(emplid = EMPLIDs, key = keys))
  })
names(comboPerTerm) <- unique(whData$COHORTTERM)

missingCleanKeys <- unlist(lapply(names(comboPerTerm), function(x){
  
  comboPerTerm[[x]][["key"]][["key"]][comboPerTerm[[x]][["key"]]["clean"] == FALSE]
  
}))

missingWHKeys <- unlist(lapply(names(comboPerTerm), function(x){
  
  # comboPerTerm[[x]][["key"]][["key"]][comboPerTerm[[x]][["key"]]["clean"] == FALSE]
  comboPerTerm[[x]][["key"]][["key"]][comboPerTerm[[x]][["key"]]["wh"] == FALSE]
  
  
}))

```


# Executive summary

**PURPOSE:** The purpose of this report is to compare the data sets derived by Bill Prisbrey and Whitney Holt and explain any discrepancies.  

**SITUATION:** Bill Prisbrey and Whitney Holt used different methods to derive data sets used to examine grades in first math courses.  Prisbrey (who is new to this data) derived the data set in R, while Holt used SQL.  Prisbrey extracted all available years, while Holt extracted after Fall 2021. 

**COMPLICATION:** Prisbrey's data set was missing `r length(missingCleanKeys) |> format(big.mark = ",")` rows that were found in Holt's baseline data set for the terms that they had in common.  

**RESOLUTION:** There are no unexplained differences.  An investigation of the filters used by Prisbrey explained all discrepancies with Holt's base data extraction.  

# Filters used by Prisbrey

  |Filter          |Removes                                                                   |
  |----------------|--------------------------------------------------------------------------|
  |High School GPA | Students without a high school GPA                                       |
  |Withdraw        | 'W' grade values                                                         |
  |Pre-math        | Students taking math more than half a year before their cohort date      |
  |Volume          | Courses that are greater than 90% of the cumulative total enrollment     |
  |Rare            | 'V', 'I', 'NC', 'CR' grade values                                        |
  |High load       | Students taking more than the 99th percentile in total credits that term |
  |Course level    | Courses higher than 3000 level                                           |
  |Recent          | Courses that were not after before Spring 2021                           |
  |First Term      | Courses taken in a term after the first term a math course is taken      |
  |Age             | Students younger than 12 years old                                       |
  |Clean           | 'NA' grade values                                                        |
  |Math Lab        | Math labs                                                                |
  

```{r}

filterList <- list(
 # baseline = mathCourses$key,
  math_lab =  mathCourses$key[!mathLabFilter],
  first_term = mathCourses$key[firstTermFilter],
  course_level = firstCourses$key[course_levelFilter],
  recent = firstCourses$key[recent_filter],
  volume = firstCourses$key[courseFilter],
  age = firstCourses$key[ageFilter],
  clean = firstCourses$key[!is.na(firstCourses$cleanGrade)],
  rare = firstCourses$key[!rareGrades],
  withdraw = firstCourses$key[withdrawFilter],
  high_load = firstCourses$key[!highLoad_filter],
  pre_math = firstCourses$key[!preMath_filter],
  HS_GPA = firstCourses$key[!zeroHSGPA_filter]
#  combined = cleanData$key[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter]
)

par(mar = c(7.1,4.1,4.1,2.1))

sapply(filterList, function(x) length(missingCleanKeys) - sum(missingCleanKeys %in% x)) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  barplot(
    ylab = "rows explained",
    las = 2,
    main = "Filters that (when combined) explain all discrepancies\nwith W.H. base data set",
    col=viridis::viridis(length(filterList)))
abline(h = length(missingCleanKeys), col = "red", lwd = 2, lty = 2 )

```
