---
title: "Act Math by High School GPA"
output: html_document
date: "2025-10-16"
---

**PURPOSE:**  This report contains visualizations describing math performance for first time freshman in their first math courses.  It includes heat maps of enrollment and performance by ACT math test scores and high school GPA.  It does not contain the results of any predictive model.   


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=10)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(ggplot2)
library(patchwork)

source(here::here("Functions", "Draw Confusion Matrix vC5.R"))
source(here::here("Functions", "Predict Math Success Functions.R"))

```




```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

```

```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course[recent_filter]) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

# print("Prep complete")


```

```{r}

###########################
## GRADE TRANSFORMATIONS ##
###########################

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
 firstCourses$wdraw_binary <- NA
 firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")


# Combine grades
# BINARY
firstCourses$grade_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_binary[hiFilter] <- 1
firstCourses$grade_binary[loFilter] <- 0

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")

courseClusters_high <- names(courseClusters[courseClusters == 1])

courseClusters_low <- names(courseClusters[courseClusters == 2])

``` 


```{r}

###########
## CLEAN ##
###########

target_classes <- c("GRADEGPA", "grade_quad")

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)

cleanData$course <- factor(cleanData$course)

#clean_and_hi_filter
#clean_and_low_filter

# Before and after Fall 2021
cleanData$covid <- ifelse(cleanData$cohort_year < 2021, "before", "after")
cleanData$covid <- factor(cleanData$covid, levels = c("before", "after"))


```


### Data

This displays all years of available data (Fall 2005 to Fall 2024.)

The first term that a first time freshman took a math course was determined.  Then, all math courses in that term were extracted. 

Next, the following filters are applied:  

  - High school GPA values of zero are removed. 
  - Rare, missing, and withdraw grades ("missing", "V", "I", "NC", "CR", "W", NA) are removed.  
  - Students with more than the 99th percentile in course load (19 credits) are removed. 
  - Math courses taken before half a year before the cohort date (yr_diff < -0.5) are removed.  This allows for taking math classes the summer before the fall cohort date.  
  - Students aged less than 12 are removed.    
  - All courses greater than 3000 are removed.  
  - Classes that were not offered after 2021 are removed.  

### Enrollment by course

```{r}

par(mar=c(6.1,4.1,3.1,2.1))

table(cleanData$course) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  (\(x){
    barplot(x, 
            main = "Count of students per course",
          col = ifelse(names(x) %in% names(courseClusters)[courseClusters ==1], "aliceblue", "lemonchiffon" ),
          las = 2,
          ylab = "")})()

legend("topright",
       legend = c("High volume", "Low volume"),
       pch = 15,
       pt.cex=3,
       col = c("aliceblue", "lemonchiffon"))

```


### ACT math scores appear to have guided course selection

```{r}

par(mar=c(6.1,4.1,3.1,2.1))

# c(5.1,4.1,4.1,2.1) # default

boxplot(ACTMATH ~ course, data = cleanData[cleanData$vol_cluster == "hi_vol",], 
        outline = FALSE,
        col = "aliceblue",
        las = 2,
        xlab = "",
        main = "High volume courses")

boxplot(ACTMATH ~ course, data = cleanData[cleanData$vol_cluster == "low_vol",], outline = FALSE,
      col = "lemonchiffon",
      las = 2,
      xlab = "",
      main = "Low volume courses")

```

### ACT math scores per course over time

```{r fig.height=7}

# I need to do this covid boxplot per course

boxplot(ACTMATH ~ covid,  data = cleanData[cleanData$vol_cluster == "hi_vol",],
        col = "aliceblue",
        main = "ACT math scores before and after Fall 2021",
        xlab = "")
mtext(text = "Fall 2021 is included in 'after'.", font=3, side = 1, line = 3)

par(mfrow=c(5,1), mar = c(0,3,3,1), oma = c(4,2,0,0))

for(crse in names(courseClusters[courseClusters == 1]) ){
daBox <- boxplot(ACTMATH ~ course + cohort_year, data = droplevels(cleanData[cleanData$course == "MATH_1010",]), 
         outline = FALSE,
         col = "aliceblue",
         las = 2,
         xlab = "",
         main = crse,
        xaxt = "n")
}

axis(side = 1,
     at = 1:length(2005:2024),
     labels = 2005:2024,
     outer = TRUE,
     las = 2,
     font = 2,
     cex.axis = 1.2)

mtext(text = "ACT math score", side = 2, line = 0.25, outer = TRUE)

# let's see that as lines, one line per course

median_ACT <- aggregate(ACTMATH ~ course + cohort_year, data = cleanData[cleanData$vol_cluster == "hi_vol",], median, na.rm = TRUE)

plot(median(median_ACT$ACTMATH),
     xlim = range(median_ACT$cohort_year),
     ylim = range(median_ACT$ACTMATH), 
     xlab = "",
     ylab = "",
     xaxt = "n",
    # yaxt = "n",
     type = "n",
    las = 2,
    main = "Median ACT math score per course over time"
)

colorMap <- c("lightsteelblue", "powderblue", "beige", "moccasin", "rosybrown")
names(colorMap) <- names(courseClusters[courseClusters == 1])

lapply(names(courseClusters[courseClusters == 1]), function(crse){
  
  filter <- median_ACT$course == crse
  
  lines(y = median_ACT$ACTMATH[filter],
        x = median_ACT$cohort_year[filter],
        col = colorMap[crse],
        lwd = 3
        )
  
})


axis(side = 1,
     at = 2005:2024,
     labels = 2005:2024,
     las = 2,
     font = 2,
     cex.axis = 1.2
     )

legend("top",
       legend = names(colorMap),
       lty = 1,
       lwd = 3,
       col = colorMap)

# I need to off-set the legend to the right

``` 




### Grade distribution per course

```{r}

par(mar=c(6.1,4.1,3.1,2.1))

# c(5.1,4.1,4.1,2.1) # default

boxplot(GRADEGPA ~ course, data = cleanData[cleanData$vol_cluster == "hi_vol",], 
        outline = FALSE,
        col = "aliceblue",
        las = 2,
        xlab = "",
        main = "High volume courses")

boxplot(GRADEGPA ~ course, data = cleanData[cleanData$vol_cluster == "low_vol",], outline = FALSE,
      col = "lemonchiffon",
      las = 2,
      xlab = "",
      main = "Low volume courses")

```

### High school GPA and math grades

```{r}

par(mar=c(4.1,4.1,3.1,2.1))

daBox <- boxplot(HSGPA ~ GRADEGPA, data = cleanData[cleanData$vol_cluster == "hi_vol",], 
        outline = FALSE,
        col = "aliceblue",
        las = 2,
        xlab = "Math grade",
        main = "High volume courses")

boxplot(HSGPA ~ GRADEGPA, data = cleanData[cleanData$vol_cluster == "low_vol",], outline = FALSE,
      col = "lemonchiffon",
      las = 2,
      xlab = "Math grade",
      main = "Low volume courses")

```

### Course load and math grades


```{r}

par(mar=c(4.1,4.1,3.1,2.1))

boxplot(load ~ GRADEGPA, data = cleanData[cleanData$vol_cluster == "hi_vol",], 
        outline = FALSE,
        col = "aliceblue",
        las = 2,
        xlab = "Math grade",
        main = "High volume courses")

boxplot(load ~ GRADEGPA, data = cleanData[cleanData$vol_cluster == "low_vol",], outline = FALSE,
      col = "lemonchiffon",
      las = 2,
      xlab = "Math grade",
      main = "Low volume courses")


```

### Year to math (yr_diff) and math grades

```{r}

par(mar=c(4.1,4.1,3.1,2.1))

boxplot(yr_diff ~ GRADEGPA, data = cleanData[cleanData$vol_cluster == "hi_vol",], 
        outline = FALSE,
        col = "aliceblue",
        las = 2,
        xlab = "Math grade",
        main = "High volume courses")

boxplot(yr_diff ~ GRADEGPA, data = cleanData[cleanData$vol_cluster == "low_vol",], outline = FALSE,
      col = "lemonchiffon",
      las = 2,
      xlab = "Math grade",
      main = "Low volume courses")


```

### Heatmaps 

```{r}

plotHex(data = cleanData, title = "Full data (All courses, all years, all filters")

plotHex(data = cleanData[cleanData$class_year <= 2020,], title = "<=2020\n(All courses, year 2020 or less, all filters")

plotHex(data = cleanData[cleanData$class_year > 2020,], title = ">=2021\n(All courses, year 2021 or greater, all filters")

plotHex(data = cleanData[cleanData$class_year == 2024,], title = "2024\n(All courses, year 2024, all filters")


plotHex(data = cleanData[cleanData$class_year == 2024 & cleanData$vol_cluster == "hi_vol",], title = "2024\n(High vol courses, year 2024, all filters")

plotHex(data = cleanData[cleanData$class_year == 2024 & cleanData$vol_cluster == "low_vol",], title = "2024\n(Low vol courses, year 2024, all filters")


for(crse in names(courseClusters[courseClusters == 1])){

  
  thePlot <- plotHex(data = cleanData[cleanData$course == crse,],
          title = paste(crse, "\n(All years, after all filters)")
  )

   beforePlot <- plotHex(data = cleanData[cleanData$course == crse & cleanData$class_year <=2020,],
          title = paste(crse, "\n(<=2020, after all filters)")
  )  
  
  plot2024 <- plotHex(data = cleanData[cleanData$course == crse & cleanData$class_year ==2024,],
          title = paste(crse, "\n(2024, after all filters)")
  )
  
 print(thePlot)  
 print(beforePlot)
 print(plot2024)  

}

# I might need to make an app so I can really look at these

```

```{r eval=FALSE}

# Too many; kinda cluttered

for(crse in names(courseClusters[courseClusters == 2])){


#lapply(courseClusters[courseClusters == 1], function(crse) {
  
  thePlot <- plotHex(data = cleanData[cleanData$course == crse,],
          title = paste(crse, "\nafter all filters")
  )
  
 print(thePlot)
  
#  })

}

```


```{r eval=TRUE}

skim(cleanData)

```