---
title: "Predictors over time"
author: "Bill Prisbrey"
date: "2026-02-11"
output:
  html_document:
    keep_md: true
    toc: true
---


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=11)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(stringr)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great!

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))
source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Sandbox", "Predictors over time with plotly sandbox.R"))
source(here::here("Sandbox", "Rate over time sandbox.R"))

```

```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

# print("You made it past loading")

```



```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course) |> # [recent_filter]
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course) |> # [recent_filter] 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.95]) })()

courseFilter <- firstCourses$course %in% higherVolCourses


# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

# print("Prep complete")

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")


# Combine grades
# BINARY
firstCourses$grade_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_binary[hiFilter] <- 1
firstCourses$grade_binary[loFilter] <- 0

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")

#################
## PRE-PROCESS ##
#################


# Select columns

# colnames(popCourses)

keepColumns <- c(
  #   "EMPLID"                      ,
  #   "TERM"                        ,
  # "course"                      ,
  #   "class"                       ,
  #   "CATNBR"                      ,
  #   "TITLE"                       ,
  #   "SECTION"                     ,
  #   "UNITS"                       ,
  #   "GRADE"                       ,
  #   "GRADEGPA"                    ,
  #   "INSTEMPLID"                  ,
  #   "INSTNAME"                    ,
  #   "EOTDATE"                     ,
  #   "ACADYR"                      ,
  #   "TERMEXTRACT"                 ,
  #   "MAX_SNAP"                    ,
  #   "CENSUSDATE"                  ,
  #   "ALT_ID"                      ,
  #   "FULLNAME"                    ,
  #   "STUDENTCAREER"               ,
  #   "SUBJECT_CD"                  ,
  #   "SUBJECT_NAME"                ,
  #   "SUBJECT_LONG"                ,
  #   "SUBJECT_ACAD_ORG_CD"         ,
  #   "CATNBR2"                     ,
  #   "CLASSNBR"                    ,
  #   "OFFERINGNBR"                 ,
  #   "SESSIONCODE"                 ,
  #   "COURSECAREER"                ,
  #   "SCHEDFLAG"                   ,
  #   "WAUTOENRL"                   ,
  #   "AUTOENROLL"                  ,
  #   "COMPONENT"                   ,
  #   "GENED"                       ,
  #   "VARCREDIT"                   ,
  #   "CONTRACT"                    ,
  #   "DIRECTPAY"                   ,
  #   "CORRESPONDENCE"              ,
  #   "ONLINECOURSE"                ,
  #   "IVC"                         ,
  #   "IVC_HYBRID"                  ,
  #   "COURSE_MODALITY"             ,
  #   "INSTRUCTION_MODE"            ,
  #   "TELECOURSE"                  ,
  #   "STUDYABROAD"                 ,
  #   "EDNET"                       ,
  #   "HYBRIDCOURSE"                ,
  #   "COURSE_LEVEL"                ,
  #   "USHE_COURSE_LEVEL"           ,
  #   "FISCAL_YEAR_OF_STARTDT"      ,
  #   "VP"                          ,
  #   "VP_SHORT"                    ,
  #   "VP_FORMAL"                   ,
  #   "ACAD_COLLEGE_CD"             ,
  #   "COLLEGE"                     ,
  #   "COLLEGE_SHORT"               ,
  #   "COLLEGE_FORMAL"              ,
  #   "ACAD_COLLEGE_REG_SUPP"       ,
  #   "ACAD_COLLEGE_TYPE"           ,
  #   "ACAD_COLLEGE_CIP_CD"         ,
  #   "ACAD_DEPARTMENT_CD"          ,
  #   "DEPARTMENT"                  ,
  #   "DEPARTMENT_SHORT"            ,
  #   "DEPARTMENT_FORMAL"           ,
  #   "ACAD_DEPARTMENT_REG_SUPP"    ,
  #   "ACAD_DEPARTMENT_TYPE"        ,
  #   "ACAD_DEPARTMENT_CIP_CD"      ,
  #   "ACAD_DIVISION_CD"            ,
  #   "DIVISION"                    ,
  #   "DIVISION_SHORT"              ,
  #   "DIVISION_FORMAL"             ,
  #   "ACAD_DIVISION_REG_SUPP"      ,
  #   "ACAD_DIVISION_TYPE"          ,
  #   "ACAD_DIVISION_CIP_CD"        ,
  #   "VP_CD"                       ,
  #   "COLLEGE_CD"                  ,
  #   "DEPARTMENT_CD"               ,
  #   "DIVISION_CD"                 ,
  #   "ROLLUP_SORT_ORDER"           ,
  #   "PS_ACAD_ORG"                 ,
  #   "PS_ACAD_GROUP"               ,
  #   "CAMPUS"                      ,
  #   "COURSE_CAMPUS"               ,
  #   "USHE_SITE_TYPE_CD"           ,
  #   "USHE_SITE_TYPE"              ,
  #   "ONOFFCAMPUS"                 ,
  #   "COURSELOCATION"              ,
  #   "CONTACTMINUTES"              ,
  #   "TEAMTAUGHT"                  ,
  #   "XLIST"                       ,
  #   "BEGTIME1"                    ,
  #   "BEGTIME2"                    ,
  #   "BEGTIME3"                    ,
  #   "DAYS1"                       ,
  #   "DAYS2"                       ,
  #   "DAYS3"                       ,
  #   "ENDTIME1"                    ,
  #   "ENDTIME2"                    ,
  #   "ENDTIME3"                    ,
  #   "CLASSLOC1"                   ,
  #   "CLASSLOC2"                   ,
  #   "CLASSLOC3"                   ,
  #   "CLASSLOCBUILDNAME1"          ,
  #   "CLASSLOCBUILDROOM1"          ,
  #   "CLASSLOCBUILDNAME2"          ,
  #   "CLASSLOCBUILDROOM2"          ,
  #   "CLASSLOCBUILDNAME3"          ,
  #   "CLASSLOCBUILDROOM3"          ,
  #   "STARTDT"                     ,
  #   "ENDDT"                       ,
  #   "BUDGETCODE"                  ,
  #   "LINEITEM"                    ,
  #   "SERVICELEARNING"             ,
  #   "XLIST_ID"                    ,
  #   "COMBINEDID"                  ,
  #   "USHE_ACADYR"                 ,
  #   "USHE_TERM"                   ,
  #   "TERM2"                       ,
  #   "ORG_EFFDT"                   ,
  #   "CLASSENROLLMENTCAPACITY"     ,
  #   "ROOM_MAX_1"                  ,
  #   "ROOM_MAX_2"                  ,
  #   "ROOM_MAX_3"                  ,
  #   "TERM_NBR"                    ,
  #   "CLASS_ATTR_LIST"             ,
  #   "EXCLUDE_BUDGET_SCH"          ,
  #   "SPR_CORRECTION_NOT_USHE_FLAG",
  #   "Section Divider: OLD"        ,
  #   "SUBJECTCOLL"                 ,
  #   "SUBJECT"                     ,
  #   "class.1"                     ,
  #   "COHORT_DT"                   ,
  "SEX"                         ,
  "FIRST_GEN_STATUS_CD"         ,
  "ETHNICITY"                   ,
  "RESSTAT"                     ,
  "FA_PELL"                     ,
  # "AGE"                         ,
  "APCREDIT"                    ,
  "HSGPA"                       ,
  "HSPRIVATE"                   ,
  "HONORS"                      ,
  "ACTCOMP"                     ,
  "ACTENGL"                     ,
  "ACTMATH"                     ,
  "ACTSCI"                      ,
  #  "load"                        ,
  "cohort_year"                 ,
  "class_year"                  ,
  "yr_diff"                     ,
  "season"                      ,
#  "course_level"                ,
#   "vol_cluster"                 ,
  "age_cut"                     
  # "cleanGrade"                  
  #   "wdraw_binary"                ,
  #   "grade_binary"                ,
  #   "grade_trinary"               ,
  #   "grade_quad" 
)

# target_classes <- c("GRADEGPA",  "grade_binary", "grade_trinary", "grade_quad") # "wdraw_binary",
target_classes <- c("GRADEGPA", "grade_quad")

# stop("Work on keep columns")

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter  & ageFilter & withdrawFilter,] # & recent_filter
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)

cleanData$course <- factor(cleanData$course)
cleanData$ETHNICITY <- factor(cleanData$ETHNICITY)

# Combine math courses
topCourses <- table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){x[x<0.95]})() |>
  names()

# I'm only adding three more to my volume clustering.
# I guess I'll comment that column out as a ringer

cleanData$course_target <- ifelse(cleanData$course %in% topCourses, as.character(cleanData$course), "other")
cleanData$course_target <- factor(cleanData$course_target)

# stop("Next step is the training")

#############################################  
## DISTANCE FROM MEDIAN SUCCESSFUL STUDENT ##  
#############################################  

hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()  

# stop("Good to here")

# Merge to cleanData 
# ADJUSTMENT -- I need to shift this by one year! 

hiGrades$class_year_shift <- hiGrades$class_year + 1

# saveRDS(hiGrades, here::here("Data", "hiGrades from vH4.rds"))

cleanData <- merge(cleanData, hiGrades[,c("course", "class_year", "class_year_shift", "ACTMATH.median", "ACTMATH.stdev", "HSGPA.median", "HSGPA.stdev", "HSGPA.count")], by.x = c("course", "class_year"), by.y = c("course","class_year_shift"),  all.x=TRUE)

# Calculate distance as a z-score

# I need to take care to shift this in my training data
# 

cleanData$ACTMATH.z <- (cleanData$ACTMATH - cleanData$ACTMATH.median)/cleanData$ACTMATH.stdev
cleanData$HSGPA.z <- (cleanData$HSGPA - cleanData$HSGPA.median)/cleanData$HSGPA.stdev

  

cleanData$dist <- sqrt(
  ((cleanData$ACTMATH - cleanData$ACTMATH.median)/cleanData$ACTMATH.stdev)^2 +
    ((cleanData$HSGPA - cleanData$HSGPA.median)/cleanData$HSGPA.stdev)^2
)

# Check
# View(cleanData[cleanData$HSGPA.z < -20 & !is.na(cleanData$HSGPA.z) & !is.na(cleanData$ACTMATH), c(keepColumns,"HSGPA.z", "HSGPA.median", "HSGPA.stdev", "HSGPA.count", "class_year.y" )])

# recall that the aggregate used in hiGrades uses complete cases for both columns 

# Calcualte the distance  
# courseProximity$dist <- sqrt(
#  ((courseProximity$ACTMATH - courseProximity$ACTMATH.median)/courseProximity$ACTMATH.stdev)^2 +
#    ((courseProximity$HSGPA - courseProximity$HSGPA.median)/courseProximity$HSGPA.stdev)^2
# )


# stop("Next step is the training loop")



```

# Executive Summary

### **PURPOSE:**  *This report examines predictors of math course selection and academic performance over time.*  

### **EXECUTIVE SUMMARY:**  *This report finds that math grades awarded have increased; that ACT math test scores has drastically decreased; and that the mix of courses selected has changed.  The students who submit test scores are increasingly elite, with better high school GPA's, test scores, and grade performance in initial math courses.  It also finds that courses and grades are consistently mildly distinguished by high school GPA and ACT math test scores over time; and that transforming these qualifications into z-scores per course improves stability.*

# Discussion  

This report finds many changes over time.  Most notably, grades have increased; test score submission rates have decreased;  and the mix of courses selected has changed. 

***Grades have increased.***  This report finds that the proportion of students earning an 4.0 ("A") has steadily increased from ~20% to ~30% since 2005.  Conversely, the proportion of students achieving a 2.3 ("C+") or worse has declined from ~40% to ~30% in the same time span.  

***Test score submission rates have decreased.***  This report finds that a decreasing number of students submit ACT test scores.  Only one-third of students in high volume courses submitted math test scores in 2024, compared to about 90% submission until 2016.  These students are increasingly elite, with better high school GPA's, test scores, and grade performance in initial math courses.

***Mix of courses selected has changed.***  This report finds that overall enrollment has steadily increased, and that enrollment has increased more in high volume courses (1010, 1050, 1210, 1030, and 1090.)  Among these, Math 1010 and Math 1090 have shifted positions as highest and lowest enrolled classes.  Where Math 1010 used to be the highest enrolled course by a large margin, it is now the lowest enrolled course by a large margin. 

Awarded grades are also higher in the low-volume courses than the high volume courses, with a median grade of 3.7 in the low-volume courses compared to 3.0 for the high-volume courses in 2024.  It is also observed that Math 1090 had a rising median course GPA since 2015, peaking with a median grade of 4.0 in 2020.  The median grade for 1090 has since fallen to be more in alignment with the other high volume courses in 2024.

**Courses and grades are at least mildly distinguished by ACT math test scores and high school GPA.**  Low-volume classes have higher test scores and high school GPA.  Among the high volume classes, Math 1010 has students with the lowest median test scores and high school GPA, and Math 1210 has the highest.

Transforming these incoming qualifications of ACT math test scores and high school GPA into z-scores introduces stability.  The median z-scores and combined distance for on track students is mostly stable.  The median z-scores and combined distance for at risk students shows a strong trend in the last few years, however, particularly in the combined distance for the median at risk student.  This aligns with the the grade compression already noted.  (Here, "on track" is defined as achieving a grade of 2.7 ("B-") or better, and "at risk" is defined as achieving a grade of 2.3 ("C+") or worse.  This cut-off was identified as the optimum in the predictive model described in the accompanying report.)
  
# Enrollment  

### *Courses are separated into high volume or low volume courses.  Courses with less than five percent of the cumulative enrollment are combined together as 'other'.* 


Five math courses have distinctly high enrollment.  These courses are categorized together as "high volume" and designated as "hi_vol" in legends in this course.  These courses are Math 1010, 1050, 1210, 1030, and 1090.  

An additional nine courses are categorized as "low volume."  They are Math 1070, 1220, 1080, 1060, 1310, 2210, 990, 1100, and 1250.

Many small courses that consist of less than 5% of the total enrollment combined are lumped together as "other."



```{r}

# Show enrollment and the top 95th percentile (with a red line, or a second cumulative graphic.) 
# Highlight high volume courses

# table(cleanData$course, useNA = "always") |> 
#  (\(x){x[order(x, decreasing = TRUE)]})() |> 
#  proportions() |> 
#  cumsum() |>
#  (\(x){names(x[x<0.95]) })()
  # plot()

#  eh let's just do the barplot

# high volume CATNBR per cluster  
hiVolCatNbr <- names(courseClusters)[courseClusters == 1] |>
(\(x){gsub("MATH_", "", x) })()


theBar <- table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  barplot(plot = FALSE)

table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> # put in order
  (\(x){names(x) <- gsub("MATH_", "", names(x)) ; return(x) })() |> # remove preceding "MATH_"
  (\(x){
  barplot(x,
          plot = TRUE,
          col = ifelse(names(x) %in% hiVolCatNbr, "darkorange2","dodgerblue3"),
          las = 2)
    })()

table(cleanData$course, useNA = "always") |> 
    (\(x){x[order(x, decreasing = TRUE)]})() |> 
    proportions() |> 
    cumsum() |>
    (\(x){which(x<0.95)})() |>
    max() |>
  (\(x){abline(v=mean(theBar[c(x,x+1)]), lwd = 2, col = "blue")})()

# I'd like to mess with graphics
# but let's put a text on here at least

text(x = theBar[14],
     y = 8000,
     label = "Keep",
     adj = c(1,0))

text(x = theBar[15],
     y = 8000,
     label = "Lump as 'Other'",
     adj = c(0,0))

# needs arrows

# Add a block arrow to your plot
arrows(x0 = theBar[14.5],   # starting x position
       y0 = 10000,   # starting y position
       x1 = theBar[11],   # ending x position
       y1 = 10000,   # ending y position
       col = "blue",
       lwd = 3,
       length = 0.15,  # arrowhead size
       angle = 20,     # arrowhead angle
       code = 2)       # code=2 puts arrow at the end


arrows(x0 = theBar[15.5],   # starting x position
       y0 = 10000,   # starting y position
       x1 = theBar[18],   # ending x position
       y1 = 10000,   # ending y position
       col = "blue",
       lwd = 3,
       length = 0.15,  # arrowhead size
       angle = 20,     # arrowhead angle
       code = 2)       # code=1 puts arrow at the beginning


mtext(side = 1, text = "Course", cex = 1.2, line = 4)
mtext(side = 2, text = "Enrollment", cex = 1.2, line = 3)
mtext(side = 3, text = "Enrollment by math course", cex = 1.75, font = 2, line = 2)

legend("topright",
       col = c("darkorange2","dodgerblue3"),
       pch = 19,
       pt.cex =2,
       legend = c("High","Low"))


```



```{r eval=FALSE}

# This section looks at enrollment in recent courses.  I decided it didn't add enough to the report to include.  I was thinking of floating an asterisk above some bars instead of creating a couple of new bar graphs, but this section would need some work to include it with the rest of the report.

# These courses are more prominent recently but have been included in the 'other' category.

droppedRecentCourses <- recentCourses[!recentCourses %in% topCourses]
droppedRecentCoursesFilter <- cleanData$course %in% droppedRecentCourses

  
theBar <- table(droplevels(cleanData$course[droppedRecentCoursesFilter]), useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  barplot(plot = FALSE)

table(droplevels(cleanData$course[droppedRecentCoursesFilter]), useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> # put in order
  (\(x){names(x) <- gsub("MATH_", "", names(x)) ; return(x) })() |> # remove preceding "MATH_"
  (\(x){
  barplot(x,
          plot = TRUE,
          col = ifelse(names(x) %in% hiVolCatNbr, "darkorange2","dodgerblue3"),
          las = 2)
    })()


recent_filter_clean <- cleanData$course %in% recentCourses

theBar <- table(droplevels(cleanData$course[recent_filter_clean]), useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  barplot(plot = FALSE)

table(droplevels(cleanData$course[recent_filter_clean]), useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> # put in order
  (\(x){names(x) <- gsub("MATH_", "", names(x)) ; return(x) })() |> # remove preceding "MATH_"
  (\(x){
  barplot(x,
          plot = TRUE,
          col = ifelse(names(x) %in% hiVolCatNbr, "darkorange2","dodgerblue3"),
          las = 2)
    })()


```


```{r}

# Additional column creation and filters

completeCases <- complete.cases(cleanData[, keepColumns])
# View(cleanData[!completeCases, keepColumns]) # at a quick glance it's all missing ACT test submissions  

# Enrollment

remove2025 <- cleanData$class_year < 2025

cleanData$all <- rep(1, nrow(cleanData))

cleanData$submitted <- ifelse(is.na(cleanData$ACTMATH), "not_submitted","submitted")

volumeMap <- data.frame(
      color = c("darkorange2", "dodgerblue3"),
      by = c("hi_vol", "low_vol"),
      lty = rep(1,2) ,
      lwd = rep(3,2)
    )

submitMap <- data.frame(
      color = c( "forestgreen", "mediumpurple3"),
      by = c("submitted", "not_submitted"),
      lty = rep(1,2) ,
      lwd = rep(3,2)
    )

courseMap <- data.frame(color = c( c("lightsteelblue", "powderblue", "beige", "moccasin", "rosybrown"), c( 
      "darkseagreen3",   
      "lightgoldenrod3", 
      "tan3",            
      "thistle3",        
      "lightskyblue3",   
      "paleturquoise3", #"wheat3",          
      "plum3",           
      "darkseagreen4",   
      "burlywood3"
    )),
    lty = c(rep(1,5), rep(2:5,3)[1:9]),
    course = c(
      "MATH_1010", "MATH_1050", "MATH_1210", "MATH_1030", "MATH_1090", "MATH_1070", "MATH_1220", "MATH_1080", "MATH_1060", "MATH_1310",
      "MATH_2210", "MATH_990",  "MATH_1100", "MATH_1250"
    ),
    lwd = c(rep(3,5), rep(2,9))
    )

yCex <- 1.5
subCex <- 1.1
mainCex <- 1.7

```


### *Although enrollment has steadily increased over time, the course mixture has changed as Math 1010 has declined to be replaced by Math 1090.* 

```{r}

# I probably need a ylab value of "Count of students" 

overTime(data= cleanData[remove2025,],
         variable = "EMPLID", agg = length, by = "all",
         title = "",
         legend_params_list = list(plot=FALSE)
         )
mtext(side = 3, 
      text = "Enrollment over time",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Enrollment has steadily increased.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Count of students",
      line = 3,
      cex = yCex
      )

overTime(data= cleanData[remove2025,],
         variable = "EMPLID", agg = length, by = "vol_cluster",
         title = "",
         featureMap = volumeMap)
mtext(side = 3, 
      text = "Enrollment over time broken by course cluster",
      line = 2.8,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Students enrolled in high volume courses have increased more.", 
      line = 1.2, 
      font = subCex)
mtext(side = 2,
      text = "Count of students",
      line = 3,
      cex = yCex
      )

overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,],
         variable = "EMPLID", agg = length,
         title = "")
mtext(side = 3, 
      text = "High enrollment volume courses over time",
      line = 2.8,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Enrollment in 1010 has declined while others, especially 1090, have increased.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Count of students",
      line = 3,
      cex = yCex
      )



# decline to show this
# overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,],
#         variable = "EMPLID", agg = length)

# COMPLETE.CASES enrollment

# overTime(data= cleanData[remove2025 & completeCases,],
#         variable = "EMPLID", agg = length, by = "vol_cluster")


# overTime(data= cleanData[remove2025,],
#         variable = "EMPLID", agg = length, by = "all")

```

# Grade distribution over time

### *The proportion of students receiving an 'A' has increased.*  

```{r}

# Proportion of grades over time

grades_per_year <- aggregate(GRADEGPA ~ class_year,  data = cleanData, table) |>
  (\(x){ do.call(data.frame,x) })()

grade_proportions_per_year <- apply(grades_per_year[,-1], 1, proportions)
colnames(grade_proportions_per_year) <- as.character(grades_per_year[,1])


gradeValues <- unique(cleanData$GRADEGPA[!is.na(cleanData$GRADEGPA)]) |>
  (\(x){x[order(x,decreasing = FALSE)]})() # |>
#  (\(x){x+0.1})()

# plot grade proportions per year

plot(x = 5,
     y = 0.5,
     type = "n",
     xlim = c(-2,ncol(grade_proportions_per_year)-1),
     ylim = c(0,0.4),
     xlab = "",
     ylab = "",
     xaxt = "n",
     las = 1)

rect(par("usr")[1],
     par("usr")[3],
     par("usr")[2],
     par("usr")[4],
     col = "gray93"
     )

grid(col = "gray87", lty = 2, lwd = 3)

# number of grades
grade_value_count <- nrow(grade_proportions_per_year)

invisible(
lapply(1:grade_value_count, function(plot_row){
  
  lines(grade_proportions_per_year[plot_row, -which(colnames(grade_proportions_per_year) %in% "2025")  ], col = viridis::viridis(n = grade_value_count , option = "C", end = 0.7)[plot_row])
  
})
)

axis(side = 1, at = 1:(ncol(grade_proportions_per_year) - 1) , labels = colnames(grade_proportions_per_year[,-which(colnames(grade_proportions_per_year) %in% "2025")]))

legend("topleft",
       legend = rev(gradeValues),
       lwd = 2,
      # type = "line",
       lty = 1,
       col = viridis::viridis(n = grade_value_count , option = "C", end = 0.7, direction = -1)
      )

mtext("Proportion of students per grade over time",
      side = 3,
      line = 2,
      font = 2,
      cex = 2)

mtext("An increasing proportion of students earn an 'A'",
      line = 0.3,
      side = 3,
      font = 3,
      cex = 1.3)

mtext("Proportion of grades awarded",
      side = 2,
      line = 2.6,
      cex = 1.2)

``` 

```{r}

# Establishing where to draw the binary value 

# Surely there's a better way than this nested loop, but it works 

annual_binaries <- lapply(2005:2024, function(loop_year){ 

binaries <- lapply(gradeValues + 0.1,
       
       function(x){ 
         
         cut(cleanData$GRADEGPA[cleanData$class_year == loop_year],
             breaks = c(0,x,4),
             include.lowest = TRUE,
             right = FALSE) |>
           table() |>
           proportions()
         
         })

 names(binaries) <- as.character(gradeValues + 0.1)
 return(binaries)  
  
})
names(annual_binaries) <- as.character(2005:2024)

byMids <- lapply(as.character(gradeValues + 0.1), function(mid){ 
  
  midRow <- sapply(annual_binaries, function(x){x[[mid]][1]})
  return(midRow)
  
})
names(byMids) <- as.character(gradeValues + 0.1)

plot(x = 5,
     y = 1,
     type = "n",
     xlim = c(-2,ncol(grade_proportions_per_year)+2),
     ylim = c(0,.8),
     xlab = "",
     ylab = "",
     xaxt = "n",
     las = 1)

rect(par("usr")[1],
     par("usr")[3],
     par("usr")[2],
     par("usr")[4],
     col = "gray93"
     )

grid(col = "gray87", lty = 2, lwd = 3)


invisible(
  
  lapply(1:length(byMids), function(mid){
    
    lines(byMids[[mid]],
          
          col = viridis::viridis(n = length(byMids), option = "C", end = 0.7)[mid]
          
          )
    
    
  })
  
)

axis(side = 1, at = 1:(ncol(grade_proportions_per_year) - 1) , labels = colnames(grade_proportions_per_year[,-which(colnames(grade_proportions_per_year) %in% "2025")]))

text(x = c(19.9),
     y = c(byMids[[6]][[20]], byMids[[9]][[20]]), 
     labels = c("2.0", "3.0"),
     pos = 4
       )


legend("topleft",
       legend = c(paste( "<", rev( as.character(gradeValues[12])), sep = "   " ) , paste( "<=", rev( as.character(gradeValues[-12])) )),
       lwd = 3,
      # type = "line",
       lty = 1,
       col = viridis::viridis(n = grade_value_count , option = "C", end = 0.7, direction = -1)
      )

mtext("Proportion of students receiving\nless than various grade thresholds",
      side = 3,
      line = 0.75,
      font = 2,
      cex = 1.9)

# mtext("Large gaps between thresholds increases model ability",
#      line = 0.3,
#      side = 3,
#      font = 3,
#      cex = 1.3)

mtext("Proportion less than grade threshold",
      side = 2,
      line = 2.6,
      cex = 1.2)

lines(x = c(19, 22),
      y = rep(mean(c(byMids[[7]][[20]], byMids[[8]][[20]])),2),
      lty = 3,
      lwd = 4,
      col = "gray20"
      )

# text(x=20.9,
#     y = mean(c(byMids[[7]][[20]], byMids[[8]][[20]])),
#     pos = 4,
#     labels = "On track\nAt risk",
#     xpd = TRUE
# )

text(x=21.9,
     y = mean(c(byMids[[7]][[20]], byMids[[8]][[20]])),
     pos = 3,
     labels = "On track",
     xpd = TRUE
)

text(x=21.9,
     y = mean(c(byMids[[7]][[20]], byMids[[8]][[20]])),
     pos = 1,
     labels = "At risk",
     xpd = TRUE
)

```



# ACT test scores
### *Submission of ACT test scores has steeply declined, especially for high volume courses.  Only a third of students in some courses (like Math 1090) have submitted ACT test scores.*


```{r}


overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "all",
         title = "Submission of ACT math test scores has declined")

```



```{r}

ratesOverTime(data= cleanData[remove2025, ], variable ="ACTMATH", by = "vol_cluster", featureMap = transform(volumeMap, lty = 3))
mtext(side = 2,
      text = "                 Fraction of students",
      line = 2,
      cex = 1.0
      )


ratesOverTime(data= cleanData[remove2025 & cleanData$vol_cluster == "hi_vol", ], variable ="ACTMATH", by = "course")
mtext(side = 2,
      text = "Fraction of students",
      line = 2,
      cex = 1.0,
      xpd = TRUE
      )



```

```{r eval = FALSE}

# FRACTION SUBMITTING TEST SCORES DEVELOPED

enrollment_by_vol_cluster <- aggregate(EMPLID ~ vol_cluster + class_year, data = cleanData, length)
actmath_by_vol_cluster <- aggregate(ACTMATH ~ vol_cluster + class_year, data = cleanData, length)

score_submission <- merge(enrollment_by_vol_cluster, actmath_by_vol_cluster, by = c("vol_cluster","class_year"), all = TRUE)


score_submission$submit_rate <- score_submission$ACTMATH/score_submission$EMPLID

plot(x = score_submission$class_year,
     y = score_submission$submit_rate,
     ylab = "",
     xlab = "",
     type = "n",
     las = 2)

invisible(
lapply(unique(score_submission$vol_cluster), function(clust){
  
  lineFilter <- score_submission$vol_cluster == clust
  lines(x = score_submission$class[lineFilter],
        y = score_submission$submit_rate[lineFilter],
     type = "l",
     lwd = 3,
     lty = 1,
     col = c("darkorange2", "dodgerblue3")[factor(score_submission$vol_cluster)[lineFilter]]
     )
})
)

legend(x="topright",
       legend = c("hi_vol", "low_vol"),
       title = "Course enrollment",
       col = c("darkorange2", "dodgerblue3"),
       pch = 19,
       pt.cex = 3)

mtext(side = 2, text = "Fraction of enrollment", line=3, cex = 1.2)
mtext(side =3, 
      text = "Fraction of students submitting ACT math scores",
      line = 2.8, 
      cex = 1.5, 
      font = 2)
mtext(side = 3, 
      text = "Submission of test scores has greatly declined", 
      line = 1.2, 
      font = 3,
      cex = 1.1
        )

``` 


```{r eval=FALSE}

# Turned this into it's own function
# almost want to expand the "overTime" function to caclulate and display "rate" 


submission_by_cluster <- overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "vol_cluster",
         title = "Test submission has especially declined for high volume math courses",
         featureMap = volumeMap)

enrollment_by_cluster <- overTime(data= cleanData[remove2025,],
         variable = "EMPLID", agg = length, by = "vol_cluster",
         title = "Enrollment has increased",
         featureMap = volumeMap)


submit_rate <- merge(submission_by_cluster, enrollment_by_cluster[,-which(colnames(enrollment_by_cluster) %in% c("color","lty","lwd"))], by = c("vol_cluster","class_year"))

submit_rate$rate <- submit_rate$ACTMATH/submit_rate$EMPLID



```


```{r eval = FALSE}

# FRACTION SUBMITTING TEST SCORES SANDBOX DEUX

enrollment_by_vol_cluster <- aggregate(EMPLID ~ vol_cluster + class_year, data = cleanData, length)
actmath_by_vol_cluster <- aggregate(ACTMATH ~ vol_cluster + class_year, data = cleanData, length)

score_submission <- merge(enrollment_by_vol_cluster, actmath_by_vol_cluster, by = c("vol_cluster","class_year"), all = TRUE)

plot(y = score_submission$ACTMATH/score_submission$EMPLID, 
     x = score_submission$class_year,
     col = c("red","green")[factor(score_submission$vol_cluster)] )

# wow, ok that's interesting.
# I wonder if I can make my "overTime" cough that up?

score_submission <- aggregate(cbind(enrollment = EMPLID, actmath = ACTMATH) ~ vol_cluster + class_year, 
                              data = cleanData, 
                              FUN = length)

score_submission <- transform(
  aggregate(cbind(enrollment = EMPLID, actmath = ACTMATH) ~ vol_cluster + class_year, 
            data = cleanData, 
            FUN = length),
  submission_rate = actmath / enrollment
)

# I think it would take a bit of work to convert this to my overTime function,...
# but why not? Instead of calculating an agg, I'd just pass a completed one to it.  That preserves colors, formatting, and style  


```


```{r eval=FALSE}

# REPLACED
# test score submission

overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "all",
         title = "Submission of ACT math test scores has declined")


overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "vol_cluster",
         title = "Test submission has especially declined for low volume math courses")


overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "ACTMATH", agg = length,
         title = "Students in 1010 have essentially stopped submitting test scores")

overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,], variable = "ACTMATH", agg = length,
         title = "Students in 1080 have essentially stopped submitting test scores")




```

### *Students who submit test scores are a self-selected elite, with better high school GPAs and more AP credits, a rising median ACT score, and consequent better math performance.*   

```{r}

medHSGPA <- overTime(data=cleanData[remove2025,], variable = "HSGPA", by = "submitted", agg = "median", title = "Median high school GPA by test submission over time", featureMap = submitMap)

overTime(data=cleanData[remove2025,], variable = "HSGPA", by = "submitted", agg = "sd", title = "Math grade standard deviation by test submission over time", featureMap = submitMap)

```

On average, test submitters' high school GPA is `r aggregate(HSGPA ~ submitted, data = medHSGPA, mean) |> (\(x){ diff(x[,"HSGPA"])})() |> round(2) ` higher than non-test submitters. 

```{r}

medAP <- overTime(data=cleanData[remove2025,], variable = "APCREDIT", by = "submitted", agg = "median", title = "Median AP credits by test submission over time", featureMap = submitMap)

overTime(data=cleanData[remove2025,], variable = "APCREDIT", by = "submitted", agg = "sd", title = "AP credit standard deviation by test submission over time", featureMap = submitMap)


```

Test score submitters' median AP credit was three credits higher than non-test submitters' median.


```{r}

# cleanData$all <- rep(1, nrow(cleanData))

overTime(data=cleanData[remove2025,], variable = "ACTMATH", by = "all", agg = "median", title = "Median ACT math score submitted over time")

overTime(data=cleanData[remove2025,], variable = "ACTMATH", by = "all", agg = "mean", title = "Mean ACT math score submitted over time")

```

```{r}

# COMPARING TEST SUBMITTERS AND NON-SUBMITTERS 

math_grade <- overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = "submitted", agg = "median", title = "Median math grade by score submission over time", featureMap = submitMap)

overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = "submitted", agg = "sd", title = "Math grade standard deviation by score submission over time", featureMap = submitMap)

```

Math grades have diverged until test score submitters have a median math grade 0.7 points higher than non-submitters in 2024 (3.7 compared to 3.0.) 


```{r eval = FALSE}

# Modify to enable a list of two "by" values
# Can I get two "by" values?
overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = list( "submitted", "vol_cluster") , agg = "median") 
# hold -up right now is the feature map.

# this will take a little work
overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = list( "submitted", "vol_cluster") , agg = "median", featureMap = data.frame(color= c("red","blue","green","gray"),
           by = c("submitted","not_submitted","hi_vol","low_vol"),
           lty = 1:4,
           lwd = 1:4
          )) 

# wow, really diverging
overTime(data=cleanData[remove2025 & cleanData$vol_cluster == "hi_vol",], variable = "GRADEGPA", by = "submitted", agg = "median") 

overTime(data=cleanData[remove2025 & cleanData$vol_cluster == "low_vol",], variable = "GRADEGPA", by = "submitted", agg = "median") 


```

###  *Courses are distinguished by ACT math scores with heavy overlap.*

```{r}

# Boxplot of 2024 ACT test scores per class.  

boxplotSubset <- cleanData[cleanData$vol_cluster == "hi_vol" & cleanData$class_year == 2024,] 

course_levels <- levels(droplevels(boxplotSubset$course))

# Align to color map
boxplotColors <- courseMap$color[match(course_levels, courseMap$course)]

boxplot(ACTMATH ~ droplevels(course),
        data = boxplotSubset,
        col = boxplotColors,
        xlab = "",
        ylab = "")

mtext(side = 3,
      text = "ACT Math test scores per course in 2024",
      line = 2.8,
      font = 2,
      cex = mainCex
      )
# mtext(side = 3, 
#      text = "Courses are distinguished by ACT math scores with heavy overlap.", 
#      line = 1.2, 
#      font = 3,
#      cex = subCex)
mtext(side = 2,
      text = "ACT Math",
      line = 2,
      cex = yCex
      )







```

```{r}


overTime(data = cleanData[remove2025,],
         variable = "ACTMATH",
         agg = median,
         by = "vol_cluster")


overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,],
         variable = "ACTMATH",
         agg = median,
         by = "course")


```


```{r fig.height=9, eval=FALSE}

### ACT math scores per course over time

# This graphic is simply wrong

par(mfrow=c(5,1), mar = c(0,3,3,1), oma = c(4,2,0,0))

for(crse in names(courseClusters[courseClusters == 1]) ){
  
daBox <- boxplot(ACTMATH ~ course + cohort_year, data = droplevels(cleanData[cleanData$course == "MATH_1010",]), 
         outline = FALSE,
         col = courseMap$color[courseMap$course == crse], #  "aliceblue",
         las = 2,
         xlab = "",
         main = crse,
        xaxt = "n")
}

axis(side = 1,
     at = 1:length(2005:2024),
     labels = 2005:2024,
     outer = TRUE,
     las = 2,
     font = 2,
     cex.axis = 1.2)

mtext(text = "ACT math score", side = 2, line = 0.25, outer = TRUE)

```

# High school GPA
### *Median high school GPA has increased over the years, with the courses mildly distinguished by high school GPA with heavy overlap.* 

```{r}

overTime(data = cleanData[remove2025,],
         variable = "HSGPA",
         agg = median,
         by = "all",
         title = "Median high school GPA for all math courses")

overTime(data = cleanData[remove2025,],
         variable = "HSGPA",
         agg = median,
         by = "vol_cluster",
         title = "Median high school GPA by volume cluster")

overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,],
         variable = "HSGPA",
         agg = median,
         by = "course",
         title = "Median high school GPA by course")

overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,],
         variable = "HSGPA",
         agg = median,
         by = "course",
         title = "Median high school GPA by course")

```


```{r}

# Boxplot of 2024 high school GPA per class.  

boxplotSubset <- cleanData[cleanData$vol_cluster == "hi_vol" & cleanData$class_year == 2024,] 

course_levels <- levels(droplevels(boxplotSubset$course))

# Align to color map
boxplotColors <- courseMap$color[match(course_levels, courseMap$course)]

boxplot(HSGPA ~ droplevels(course),
        data = boxplotSubset,
        col = boxplotColors,
        xlab = "",
        ylab = "")

mtext(side = 3,
      text = "High school GPA per course in 2024",
      line = 2.8,
      font = 2,
      cex = mainCex
      )
# mtext(side = 3, 
#      text = "Courses are distinguished by ACT math scores with heavy overlap.", 
#      line = 1.2, 
#      font = 3,
#      cex = subCex)
mtext(side = 2,
      text = "High school GPA",
      line = 2,
      cex = yCex
      )


```



# Math grades 
### *The median math grade fluctuates over time without a clear trend, while above-median grades shift higher:  the 70th percentile shifts from a B+ to an A in 2016.*  

```{r}


overTime(data = cleanData[remove2025,],
         variable = "GRADEGPA",
         agg = function(x) {quantile(x,0.5)},
         by = "all",
         title = "Median math grade for all math courses")

overTime(data = cleanData[remove2025,],
         variable = "GRADEGPA",
         agg = function(x) {quantile(x,0.7)},
         by = "all",
         title = "70th percentile math grade for all math courses")

```

### *Low-volume classes have maintained a higher median GPA than high-volume classes, and seem to have recently stabilized at their twenty-year high.* 

```{r}

overTime(data = cleanData[remove2025,],
         variable = "GRADEGPA",
         agg = function(x) {quantile(x,0.5)},
         by = "vol_cluster",
         title = "Median math grade by volume cluster")

# overTime(data = cleanData[remove2025,],
#         variable = "GRADEGPA",
#         agg = function(x) {quantile(x,0.6)},
#         by = "vol_cluster",
#         title = "60th percentile math grade by volume cluster")

```


```{r}

overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,],
         variable = "GRADEGPA",
         agg = median,
         by = "course")


overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,],
         variable = "GRADEGPA",
         agg = median,
         by = "course")

```

Math 1090 peaked with a median grade of 4.0 in 2020.
 
### *Courses are not distinguished by grade distribution.*   


```{r}

# Boxplot of 2024 ACT test scores per class.  

boxplotSubset <- cleanData[cleanData$vol_cluster == "hi_vol" & cleanData$class_year == 2024,] 

course_levels <- levels(droplevels(boxplotSubset$course))

# Align to color map
boxplotColors <- courseMap$color[match(course_levels, courseMap$course)]

boxplot(GRADEGPA ~ droplevels(course),
        data = boxplotSubset,
        col = boxplotColors,
        xlab = "",
        ylab = "")

mtext(side = 3,
      text = "Grade distribution per course in 2024",
      line = 2.8,
      font = 2,
      cex = mainCex
      )
# mtext(side = 3, 
#      text = "Courses are not distinguished by grade distribution.", 
#      line = 1.2, 
#      font = 3,
#      cex = subCex)
mtext(side = 2,
      text = "GPA",
      line = 2,
      cex = yCex
      )


```




# Z-scores

### *Successfully on track students are closer to the prior year's median successful student.*  

```{r}

# Z-score distance prep

successMap <- data.frame(color = c("deeppink3","darkslategray4"), by = c("At Risk", "On Track"), lty = c(1,1), lwd = c(3,3) )

cleanData$success_cut <- cut(cleanData$GRADEGPA, breaks = c(0,2.4,4), include.lowest = TRUE,
                             right = FALSE) |>
  (\(x){ifelse(as.character(x) == "[0,2.4)","At Risk","On Track" )})()


```


```{r}

# ACT MATH z-scores

boxplot(ACTMATH.z ~ success_cut, 
        data = cleanData[remove2025,], 
        col = successMap$color, 
        outline = FALSE,
        ylab = "",
        xlab = "")


mtext(side = 3, 
      text = "ACT math converted to z-scores per course",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Students at risk of failing are below the prior year's class median.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Median-centered z-score",
      line = 3,
      cex = yCex
      )

mtext(side = 1,
      text = "Grade outcome",
      line = 2.6,
      cex = yCex
      )

mtext(side = 1,
      text = "where 'at risk' is <=2.3",
      line = 4,
      cex = yCex,
      font = 3
      )


overTime(data= cleanData[remove2025,],
         variable = "ACTMATH.z", agg = median, by = "success_cut",
         title = "",
         featureMap = successMap)

mtext(side = 3, 
      text = "Median ACT math z-score per course over time",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Students at risk of failing have ACT math test scores below the prior year's course median.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Median-centered z-score",
      line = 3,
      cex = yCex
      )



```




```{r}

# HSGPA z-scores

boxplot(HSGPA.z ~ success_cut, 
        data = cleanData[remove2025,], 
        col = successMap$color, 
        outline = FALSE,
        ylab = "",
        xlab = "")


mtext(side = 3, 
      text = "High school GPA converted to z-scores",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Students at risk of failing are below the prior year's class median.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Median-centered z-score",
      line = 3,
      cex = yCex
      )

mtext(side = 1,
      text = "Grade outcome",
      line = 2.6,
      cex = yCex
      )

mtext(side = 1,
      text = "where 'at risk' is <=2.3",
      line = 4,
      cex = yCex,
      font = 3
      )


overTime(data= cleanData[remove2025,],
         variable = "HSGPA.z", agg = median, by = "success_cut",
         title = "",
         featureMap = successMap)

mtext(side = 3, 
      text = "Median high school GPA z-score per course over time",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Students at risk of failing have a high school GPA below the prior year's course median.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Median-centered z-score",
      line = 3,
      cex = yCex
      )


```



```{r}

# DISTANCE OVER TIME

# Maybe I want to break these graphs further into hi/low volume courses (one graph per volume cluster), or look at the high volume courses 

boxplot(dist ~ success_cut, 
        data = cleanData[remove2025,], 
        col = successMap$color, 
        outline = FALSE,
        ylab = "",
        xlab = "")


mtext(side = 3, 
      text = "Combined distance to prior median student qualifications",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Successful students are closer to the prior year's median successful student.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Standardized distance",
      line = 3,
      cex = yCex
      )

mtext(side = 1,
      text = "Grade outcome",
      line = 2.8,
      cex = yCex
      )

mtext(side = 1,
      text = "where 'at risk' is <=2.3",
      line = 4,
      cex = yCex,
      font = 3
      )


# Distance over time


overTime(data= cleanData[remove2025,],
         variable = "dist", agg = median, by = "success_cut",
         title = "",
         featureMap = successMap)


mtext(side = 3, 
      text = "Median combined distance to prior median student qualifications",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Distance to prior median successful student is stable.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Standardized distance",
      line = 3,
      cex = yCex
      )



```





# Additional comparison graphics  

```{r fig.width = 4.5, fig.height = 9}

ggplot(cleanData[cleanData$vol_cluster == "hi_vol" & 
                 cleanData$class_year %in% c(2014, 2024),], 
       aes(x = HSGPA, y = ACTMATH)) +
  geom_hex(bins = 30) +
  facet_grid(course ~ class_year) +  # rows ~ columns
  scale_fill_gradientn(colors = c("lightblue", "yellow", "orange", "red"),
                       name = "Count") +
  theme_minimal() +
  labs(x = "HSGPA", y = "ACTMATH", title = "Entrance changes for high volume courses")

``` 




```{r}

# Grade density, all courses

ggplot(cleanData[cleanData$class_year %in% c(2014, 2024),],
       aes(x = GRADEGPA, color = factor(class_year), fill = factor(class_year))) +
    geom_density(alpha = 0.3) +  # alpha makes the fill transparent
    theme_minimal() +
    labs(x = "Grade GPA", 
         y = "Density",
         color = "Class Year",
         fill = "Class Year",
         title = "Grade distribution shift from 2014 to 2024") +
    scale_color_manual(values = c("blue", "red")) +
    scale_fill_manual(values = c("blue", "red"))



```

```{r}

# grade density by volume cluster

ggplot(cleanData[cleanData$class_year %in% c(2014, 2024),],
       aes(x = GRADEGPA, color = factor(class_year), fill = factor(class_year))) +
    geom_density(alpha = 0.3) +
    facet_wrap(~vol_cluster, ncol = 1) +  # one column, stacked rows
    theme_minimal() +
    labs(x = "Grade GPA", 
         y = "Density",
         color = "Class Year",
         fill = "Class Year",
         title = "Grade distribution shift from 2014 to 2024 by enrollment volume") +
    scale_color_manual(values = c("blue", "red")) +
    scale_fill_manual(values = c("blue", "red"))



```

```{r  fig.width=3, fig.height=9}

# Grade density, high volume courses

ggplot(cleanData[cleanData$vol_cluster == "hi_vol" & 
                 cleanData$class_year %in% c(2014, 2024),],
       aes(x = GRADEGPA, color = factor(class_year), fill = factor(class_year))) +
  geom_density(alpha = 0.3) +
  facet_wrap(~course, ncol = 1) +  # one column, stacked rows
  theme_minimal() +
  labs(x = "Grade GPA", 
       y = "Density",
       color = "Class Year",
       fill = "Class Year",
       title = "Grade distribution shift from\n2014 to 2024 for\nhigh volume courses") +
  scale_color_manual(values = c("blue", "red")) +
  scale_fill_manual(values = c("blue", "red"))


```

