---
title: "Predictors over time"
author: "Bill Prisbrey"
date: "2025-11-05"
output: html_document
output:
  html_document:
    keep_md: true
---

Where is that document I was working on?
I was looking at predictors over time, and it was a report of descriptive things. 

It was "Math grade by ACT math score and high school GPA.Rmd"

Seems like I c/should break that into two reports, and one report is just the heatmaps. Or, more accurately, a child I guess.  And it should be "admittance" on left and "performance" on right.  

I'm also pretty sure I haven't knitted the latest version of it.


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(stringr)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great!

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```

```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

# print("You made it past loading")

```


```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course) |> # [recent_filter]
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course) |> # [recent_filter] 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.95]) })()

courseFilter <- firstCourses$course %in% higherVolCourses


# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

# print("Prep complete")

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")


# Combine grades
# BINARY
firstCourses$grade_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_binary[hiFilter] <- 1
firstCourses$grade_binary[loFilter] <- 0

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")

#################
## PRE-PROCESS ##
#################


# Select columns

# colnames(popCourses)

keepColumns <- c(
  #   "EMPLID"                      ,
  #   "TERM"                        ,
  # "course"                      ,
  #   "class"                       ,
  #   "CATNBR"                      ,
  #   "TITLE"                       ,
  #   "SECTION"                     ,
  #   "UNITS"                       ,
  #   "GRADE"                       ,
  #   "GRADEGPA"                    ,
  #   "INSTEMPLID"                  ,
  #   "INSTNAME"                    ,
  #   "EOTDATE"                     ,
  #   "ACADYR"                      ,
  #   "TERMEXTRACT"                 ,
  #   "MAX_SNAP"                    ,
  #   "CENSUSDATE"                  ,
  #   "ALT_ID"                      ,
  #   "FULLNAME"                    ,
  #   "STUDENTCAREER"               ,
  #   "SUBJECT_CD"                  ,
  #   "SUBJECT_NAME"                ,
  #   "SUBJECT_LONG"                ,
  #   "SUBJECT_ACAD_ORG_CD"         ,
  #   "CATNBR2"                     ,
  #   "CLASSNBR"                    ,
  #   "OFFERINGNBR"                 ,
  #   "SESSIONCODE"                 ,
  #   "COURSECAREER"                ,
  #   "SCHEDFLAG"                   ,
  #   "WAUTOENRL"                   ,
  #   "AUTOENROLL"                  ,
  #   "COMPONENT"                   ,
  #   "GENED"                       ,
  #   "VARCREDIT"                   ,
  #   "CONTRACT"                    ,
  #   "DIRECTPAY"                   ,
  #   "CORRESPONDENCE"              ,
  #   "ONLINECOURSE"                ,
  #   "IVC"                         ,
  #   "IVC_HYBRID"                  ,
  #   "COURSE_MODALITY"             ,
  #   "INSTRUCTION_MODE"            ,
  #   "TELECOURSE"                  ,
  #   "STUDYABROAD"                 ,
  #   "EDNET"                       ,
  #   "HYBRIDCOURSE"                ,
  #   "COURSE_LEVEL"                ,
  #   "USHE_COURSE_LEVEL"           ,
  #   "FISCAL_YEAR_OF_STARTDT"      ,
  #   "VP"                          ,
  #   "VP_SHORT"                    ,
  #   "VP_FORMAL"                   ,
  #   "ACAD_COLLEGE_CD"             ,
  #   "COLLEGE"                     ,
  #   "COLLEGE_SHORT"               ,
  #   "COLLEGE_FORMAL"              ,
  #   "ACAD_COLLEGE_REG_SUPP"       ,
  #   "ACAD_COLLEGE_TYPE"           ,
  #   "ACAD_COLLEGE_CIP_CD"         ,
  #   "ACAD_DEPARTMENT_CD"          ,
  #   "DEPARTMENT"                  ,
  #   "DEPARTMENT_SHORT"            ,
  #   "DEPARTMENT_FORMAL"           ,
  #   "ACAD_DEPARTMENT_REG_SUPP"    ,
  #   "ACAD_DEPARTMENT_TYPE"        ,
  #   "ACAD_DEPARTMENT_CIP_CD"      ,
  #   "ACAD_DIVISION_CD"            ,
  #   "DIVISION"                    ,
  #   "DIVISION_SHORT"              ,
  #   "DIVISION_FORMAL"             ,
  #   "ACAD_DIVISION_REG_SUPP"      ,
  #   "ACAD_DIVISION_TYPE"          ,
  #   "ACAD_DIVISION_CIP_CD"        ,
  #   "VP_CD"                       ,
  #   "COLLEGE_CD"                  ,
  #   "DEPARTMENT_CD"               ,
  #   "DIVISION_CD"                 ,
  #   "ROLLUP_SORT_ORDER"           ,
  #   "PS_ACAD_ORG"                 ,
  #   "PS_ACAD_GROUP"               ,
  #   "CAMPUS"                      ,
  #   "COURSE_CAMPUS"               ,
  #   "USHE_SITE_TYPE_CD"           ,
  #   "USHE_SITE_TYPE"              ,
  #   "ONOFFCAMPUS"                 ,
  #   "COURSELOCATION"              ,
  #   "CONTACTMINUTES"              ,
  #   "TEAMTAUGHT"                  ,
  #   "XLIST"                       ,
  #   "BEGTIME1"                    ,
  #   "BEGTIME2"                    ,
  #   "BEGTIME3"                    ,
  #   "DAYS1"                       ,
  #   "DAYS2"                       ,
  #   "DAYS3"                       ,
  #   "ENDTIME1"                    ,
  #   "ENDTIME2"                    ,
  #   "ENDTIME3"                    ,
  #   "CLASSLOC1"                   ,
  #   "CLASSLOC2"                   ,
  #   "CLASSLOC3"                   ,
  #   "CLASSLOCBUILDNAME1"          ,
  #   "CLASSLOCBUILDROOM1"          ,
  #   "CLASSLOCBUILDNAME2"          ,
  #   "CLASSLOCBUILDROOM2"          ,
  #   "CLASSLOCBUILDNAME3"          ,
  #   "CLASSLOCBUILDROOM3"          ,
  #   "STARTDT"                     ,
  #   "ENDDT"                       ,
  #   "BUDGETCODE"                  ,
  #   "LINEITEM"                    ,
  #   "SERVICELEARNING"             ,
  #   "XLIST_ID"                    ,
  #   "COMBINEDID"                  ,
  #   "USHE_ACADYR"                 ,
  #   "USHE_TERM"                   ,
  #   "TERM2"                       ,
  #   "ORG_EFFDT"                   ,
  #   "CLASSENROLLMENTCAPACITY"     ,
  #   "ROOM_MAX_1"                  ,
  #   "ROOM_MAX_2"                  ,
  #   "ROOM_MAX_3"                  ,
  #   "TERM_NBR"                    ,
  #   "CLASS_ATTR_LIST"             ,
  #   "EXCLUDE_BUDGET_SCH"          ,
  #   "SPR_CORRECTION_NOT_USHE_FLAG",
  #   "Section Divider: OLD"        ,
  #   "SUBJECTCOLL"                 ,
  #   "SUBJECT"                     ,
  #   "class.1"                     ,
  #   "COHORT_DT"                   ,
  "SEX"                         ,
  "FIRST_GEN_STATUS_CD"         ,
  "ETHNICITY"                   ,
  "RESSTAT"                     ,
  "FA_PELL"                     ,
  # "AGE"                         ,
  "APCREDIT"                    ,
  "HSGPA"                       ,
  "HSPRIVATE"                   ,
  "HONORS"                      ,
  "ACTCOMP"                     ,
  "ACTENGL"                     ,
  "ACTMATH"                     ,
  "ACTSCI"                      ,
  #  "load"                        ,
  "cohort_year"                 ,
  "class_year"                  ,
  "yr_diff"                     ,
  "season"                      ,
  "course_level"                ,
#   "vol_cluster"                 ,
  "age_cut"                     
  # "cleanGrade"                  
  #   "wdraw_binary"                ,
  #   "grade_binary"                ,
  #   "grade_trinary"               ,
  #   "grade_quad" 
)

# target_classes <- c("GRADEGPA",  "grade_binary", "grade_trinary", "grade_quad") # "wdraw_binary",
target_classes <- c("GRADEGPA", "grade_quad")

# stop("Work on keep columns")

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter  & ageFilter & withdrawFilter,] # & recent_filter
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)

cleanData$course <- factor(cleanData$course)

# Combine math courses
topCourses <- table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){x[x<0.95]})() |>
  names()

# I'm only adding three more to my volume clustering.
# I guess I'll comment that column out as a ringer

cleanData$course_target <- ifelse(cleanData$course %in% topCourses, as.character(cleanData$course), "other")
cleanData$course_target <- factor(cleanData$course_target)

# stop("Next step is the training")



```

I need to create an appendix that shows all of these cleaning steps. All of these filters and complete cases. 

But first, let's show our math classes by enrollment

STARTING DESCRIPTIVES


```{r}

# Show enrollment and the top 95th percentile (with a red line, or a second cumulative graphic.) 
# Highlight high volume courses

table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.95]) })()
  # plot()

#  eh let's just do the barplot

# high volume CATNBR per cluster  
hiVolCatNbr <- names(courseClusters)[courseClusters == 1] |>
(\(x){gsub("MATH_", "", x) })()


theBar <- table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  barplot(plot = FALSE)

table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> # put in order
  (\(x){names(x) <- gsub("MATH_", "", names(x)) ; return(x) })() |> # remove preceding "MATH_"
  (\(x){
  barplot(x,
          plot = TRUE,
          col = ifelse(names(x) %in% hiVolCatNbr, "red","green"),
          las = 2)
    })()

table(cleanData$course, useNA = "always") |> 
    (\(x){x[order(x, decreasing = TRUE)]})() |> 
    proportions() |> 
    cumsum() |>
    (\(x){which(x<0.95)})() |>
    max() |>
  (\(x){abline(v=mean(theBar[c(x,x+1)]), lwd = 2, col = "blue")})()

# I'd like to mess with graphics
# but let's put a text on here at least

text(x = theBar[14],
     y = 8000,
     label = "Keep",
     adj = c(1,0))

text(x = theBar[15],
     y = 8000,
     label = "Lump as 'Other'",
     adj = c(0,0))

# needs arrows

# Add a block arrow to your plot
arrows(x0 = theBar[14.5],   # starting x position
       y0 = 10000,   # starting y position
       x1 = theBar[11],   # ending x position
       y1 = 10000,   # ending y position
       col = "blue",
       lwd = 3,
       length = 0.15,  # arrowhead size
       angle = 20,     # arrowhead angle
       code = 2)       # code=2 puts arrow at the end


arrows(x0 = theBar[15.5],   # starting x position
       y0 = 10000,   # starting y position
       x1 = theBar[18],   # ending x position
       y1 = 10000,   # ending y position
       col = "blue",
       lwd = 3,
       length = 0.15,  # arrowhead size
       angle = 20,     # arrowhead angle
       code = 2)       # code=1 puts arrow at the beginning


mtext(side = 1, text = "Course", cex = 1.2, line = 4)
mtext(side = 2, text = "Enrollment", cex = 1.2, line = 3)
mtext(side = 3, text = "Enrollment by math course", cex = 1.75, font = 2, line = 2)

legend("topright",
       col = c("red","green"),
       pch = 19,
       pt.cex =2,
       legend = c("High","Regular"))

# mess with colors another day

```


### ACT math scores per course over time

```{r fig.height=9}

par(mfrow=c(5,1), mar = c(0,3,3,1), oma = c(4,2,0,0))

for(crse in names(courseClusters[courseClusters == 1]) ){
daBox <- boxplot(ACTMATH ~ course + cohort_year, data = droplevels(cleanData[cleanData$course == "MATH_1010",]), 
         outline = FALSE,
         col = "aliceblue",
         las = 2,
         xlab = "",
         main = crse,
        xaxt = "n")
}

axis(side = 1,
     at = 1:length(2005:2024),
     labels = 2005:2024,
     outer = TRUE,
     las = 2,
     font = 2,
     cex.axis = 1.2)

mtext(text = "ACT math score", side = 2, line = 0.25, outer = TRUE)

```


```{r}

# let's see that as lines, one line per course

par(mfrow=c(1,1), mar = c(0,3,3,7), oma = c(4,2,0,0))

median_ACT <- aggregate(ACTMATH ~ course + class_year, data = cleanData[cleanData$vol_cluster == "hi_vol",], median, na.rm = TRUE)

plot(median(median_ACT$ACTMATH),
     xlim = range(median_ACT$class_year),
     ylim = range(median_ACT$ACTMATH), 
     xlab = "",
     ylab = "",
     xaxt = "n",
    # yaxt = "n",
     type = "n",
    las = 2,
    main = "Median ACT math score per course over time"
)

colorMap <- c("lightsteelblue", "powderblue", "beige", "moccasin", "rosybrown")
names(colorMap) <- names(courseClusters[courseClusters == 1]) 

lapply(names(courseClusters[courseClusters == 1]), function(crse){
  
  filter <- median_ACT$course == crse
  
  lines(y = median_ACT$ACTMATH[filter],
        x = median_ACT$class_year[filter],
        col = colorMap[crse],
        lwd = 3
        )
  
})


axis(side = 1,
     at = 2005:2025,
     labels = 2005:2025,
     las = 2,
     font = 1,
     cex.axis = 1.2
     )

legend("topright",
       legend = names(colorMap) |> (\(x){  gsub("MATH_","",x)})(),
       lty = 1,
       lwd = 3,
       col = colorMap,
       xpd = TRUE,
       inset = c(-0.25,0))


``` 

And this would look good as a plotly.

I need to see if ACT has gone down over time  
I need to see how the math grade and the HSGPA have changed over time 

```{r}

median_overall_ACT <- aggregate(ACTMATH ~ class_year, data = cleanData[cleanData$vol_cluster == "hi_vol",], median, na.rm = TRUE)


boxplot(ACTMATH ~ class_year, data = cleanData[cleanData$vol_cluster == "hi_vol",])

# So that's stable over time for the high volume courses 

# And it's really stable for all-l the courses
boxplot(ACTMATH ~ class_year, data = cleanData) 

# now let's see high school grade

boxplot(HSGPA ~ class_year, 
        data = cleanData,
        outline = FALSE)

# now actual math grade

boxplot(GRADEGPA ~ class_year, 
        data = cleanData,
        outline = FALSE)

``` 

```{r eval=FALSE}

# Hunting for changes in grade over time

library(ggridges)
library(ggplot2)

ggplot(your_data, aes(x = HSGPA, y = as.factor(year), fill = as.factor(year))) +
  geom_density_ridges(alpha = 0.7) +
  scale_x_continuous(limits = c(0, 4)) +
  labs(y = "Year", x = "HSGPA") +
  theme_ridges() +
  theme(legend.position = "none")

# doesn't show the subtlety

```


```{r}

# Time to turn this into a function


overTime <- function(variable = "ACTMATH", 
                     agg = median, 
                     data = cleanData[cleanData$vol_cluster == "hi_vol",], title = NA, 
                     featureMap = NA,
                     plot_params_list = list(),
                     legend_params_list = list(),
                     ...){

  if(is.na(title)){
    
    agg_name <- deparse(substitute(agg))
    title <- paste(str_to_title(agg_name), variable, "per course over time", sep=" ")
    
  }
  
  if(is.na(featureMap)){
    
    featureMap <- data.frame(color = c( c("lightsteelblue", "powderblue", "beige", "moccasin", "rosybrown"), c( 
        "darkseagreen3",   
  "lightgoldenrod3", 
  "tan3",            
  "thistle3",        
  "lightskyblue3",   
  "wheat3",          
  "plum3",           
  "darkseagreen4",   
  "burlywood3"
      )),
  lty = c(rep(1,5), rep(2:5,3)[1:9]),
  course = c(
     "MATH_1010", "MATH_1050", "MATH_1210", "MATH_1030", "MATH_1090", "MATH_1070", "MATH_1220", "MATH_1080", "MATH_1060", "MATH_1310",
"MATH_2210", "MATH_990",  "MATH_1100", "MATH_1250"
  ),
lwd = c(rep(3,5), rep(2,9))
                             )
  }


plot_params_default <- list(mfrow=c(1,1), mar = c(0,3,3,7), oma = c(4,2,0,0))  
plot_params_list <- modifyList(plot_params_default, plot_params_list)
do.call(par, plot_params_list)

# par(mfrow=c(1,1), mar = c(0,3,3,7), oma = c(4,2,0,0))

theAgg <- aggregate(get(variable) ~ course + class_year, data = data, FUN = agg, ...)

# merge 

theAgg <- merge(theAgg, featureMap, by = "course", all.x = TRUE)

theAgg <- theAgg[order(theAgg$class_year),]

plot(median(theAgg[,3]),
     xlim = range(theAgg[,2]),
     ylim = range(theAgg[,3]), 
     xlab = "",
     ylab = "",
     xaxt = "n",
    # yaxt = "n",
     type = "n",
    las = 2,
    main = title
)


lapply(unique(theAgg$course), function(crse){
  
  filter <- theAgg[,1] == crse
  
  lines(y = theAgg[,3][filter],
        x = theAgg[,2][filter],
        col = theAgg[filter,"color"],
        lwd = theAgg[filter,"lwd"],
        lty = theAgg[filter,"lty"],
        )
  
})


axis(side = 1,
     at = 2005:2025,
     labels = 2005:2025,
     las = 2,
     font = 1,
     cex.axis = 1.2
     )

forLegend <- unique(theAgg[,c("course", "color", "lwd", "lty")])
legend_params_default <- list(x = "center",
                              legend = gsub("MATH_","",forLegend[,"course"]),
                              lty = forLegend[,"lty"],
                              lwd = forLegend[,"lwd"],
                              col = forLegend[,"color"],
                              xpd = TRUE,
                              inset = c(-0.15,0)
                                )

legend_params_list <- modifyList(legend_params_default, legend_params_list)
do.call(legend, legend_params_list)



#legend("topright",
#       legend = forLegend[,"course"] |> (\(x){  gsub("MATH_","",x)})(),
#       lty = forLegend[,"lty"],
#       lwd = forLegend[,"lwd"],
#       col = forLegend[,"color"],
#       xpd = TRUE,
#       inset = c(-0.25,0))

#legend("bottomleft",
#       legend = theAgg[,3] |> unique() |> (\(x){  gsub("MATH_","",x)})(),
#       lty = unique(theAgg[,"lty"]),
#       lwd = unique(theAgg[,"lwd"]),
#       col = unique(theAgg[,"color"]),
#       xpd = TRUE,
#       inset = c(-0.25,0))

}

# time to mess with the legend
# Getting closer


``` 

I want to see course attendandance over time.
Is "other" getting more popular?


