---
title: "Predictors over time"
author: "Bill Prisbrey"
date: "2025-11-05"
output:
  html_document:
    keep_md: true
    toc: true
---


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=11)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(stringr)

# source("https://raw.githubusercontent.com/u0693875-billprisbrey/Draw-Confusion-Matrix/refs/heads/main/Draw%20Confusion%20Matrix.R?token=GHSAT0AAAAAADL3K4CPCVTG2DS7Y2JOIWDK2IAEEWQ") # great!

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))
source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Sandbox", "Predictors over time with plotly sandbox.R"))
source(here::here("Sandbox", "Rate over time sandbox.R"))

```

```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

# print("You made it past loading")

```



```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course) |> # [recent_filter]
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course) |> # [recent_filter] 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.95]) })()

courseFilter <- firstCourses$course %in% higherVolCourses


# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

# print("Prep complete")

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")


# Combine grades
# BINARY
firstCourses$grade_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_binary[hiFilter] <- 1
firstCourses$grade_binary[loFilter] <- 0

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")

#################
## PRE-PROCESS ##
#################


# Select columns

# colnames(popCourses)

keepColumns <- c(
  #   "EMPLID"                      ,
  #   "TERM"                        ,
  # "course"                      ,
  #   "class"                       ,
  #   "CATNBR"                      ,
  #   "TITLE"                       ,
  #   "SECTION"                     ,
  #   "UNITS"                       ,
  #   "GRADE"                       ,
  #   "GRADEGPA"                    ,
  #   "INSTEMPLID"                  ,
  #   "INSTNAME"                    ,
  #   "EOTDATE"                     ,
  #   "ACADYR"                      ,
  #   "TERMEXTRACT"                 ,
  #   "MAX_SNAP"                    ,
  #   "CENSUSDATE"                  ,
  #   "ALT_ID"                      ,
  #   "FULLNAME"                    ,
  #   "STUDENTCAREER"               ,
  #   "SUBJECT_CD"                  ,
  #   "SUBJECT_NAME"                ,
  #   "SUBJECT_LONG"                ,
  #   "SUBJECT_ACAD_ORG_CD"         ,
  #   "CATNBR2"                     ,
  #   "CLASSNBR"                    ,
  #   "OFFERINGNBR"                 ,
  #   "SESSIONCODE"                 ,
  #   "COURSECAREER"                ,
  #   "SCHEDFLAG"                   ,
  #   "WAUTOENRL"                   ,
  #   "AUTOENROLL"                  ,
  #   "COMPONENT"                   ,
  #   "GENED"                       ,
  #   "VARCREDIT"                   ,
  #   "CONTRACT"                    ,
  #   "DIRECTPAY"                   ,
  #   "CORRESPONDENCE"              ,
  #   "ONLINECOURSE"                ,
  #   "IVC"                         ,
  #   "IVC_HYBRID"                  ,
  #   "COURSE_MODALITY"             ,
  #   "INSTRUCTION_MODE"            ,
  #   "TELECOURSE"                  ,
  #   "STUDYABROAD"                 ,
  #   "EDNET"                       ,
  #   "HYBRIDCOURSE"                ,
  #   "COURSE_LEVEL"                ,
  #   "USHE_COURSE_LEVEL"           ,
  #   "FISCAL_YEAR_OF_STARTDT"      ,
  #   "VP"                          ,
  #   "VP_SHORT"                    ,
  #   "VP_FORMAL"                   ,
  #   "ACAD_COLLEGE_CD"             ,
  #   "COLLEGE"                     ,
  #   "COLLEGE_SHORT"               ,
  #   "COLLEGE_FORMAL"              ,
  #   "ACAD_COLLEGE_REG_SUPP"       ,
  #   "ACAD_COLLEGE_TYPE"           ,
  #   "ACAD_COLLEGE_CIP_CD"         ,
  #   "ACAD_DEPARTMENT_CD"          ,
  #   "DEPARTMENT"                  ,
  #   "DEPARTMENT_SHORT"            ,
  #   "DEPARTMENT_FORMAL"           ,
  #   "ACAD_DEPARTMENT_REG_SUPP"    ,
  #   "ACAD_DEPARTMENT_TYPE"        ,
  #   "ACAD_DEPARTMENT_CIP_CD"      ,
  #   "ACAD_DIVISION_CD"            ,
  #   "DIVISION"                    ,
  #   "DIVISION_SHORT"              ,
  #   "DIVISION_FORMAL"             ,
  #   "ACAD_DIVISION_REG_SUPP"      ,
  #   "ACAD_DIVISION_TYPE"          ,
  #   "ACAD_DIVISION_CIP_CD"        ,
  #   "VP_CD"                       ,
  #   "COLLEGE_CD"                  ,
  #   "DEPARTMENT_CD"               ,
  #   "DIVISION_CD"                 ,
  #   "ROLLUP_SORT_ORDER"           ,
  #   "PS_ACAD_ORG"                 ,
  #   "PS_ACAD_GROUP"               ,
  #   "CAMPUS"                      ,
  #   "COURSE_CAMPUS"               ,
  #   "USHE_SITE_TYPE_CD"           ,
  #   "USHE_SITE_TYPE"              ,
  #   "ONOFFCAMPUS"                 ,
  #   "COURSELOCATION"              ,
  #   "CONTACTMINUTES"              ,
  #   "TEAMTAUGHT"                  ,
  #   "XLIST"                       ,
  #   "BEGTIME1"                    ,
  #   "BEGTIME2"                    ,
  #   "BEGTIME3"                    ,
  #   "DAYS1"                       ,
  #   "DAYS2"                       ,
  #   "DAYS3"                       ,
  #   "ENDTIME1"                    ,
  #   "ENDTIME2"                    ,
  #   "ENDTIME3"                    ,
  #   "CLASSLOC1"                   ,
  #   "CLASSLOC2"                   ,
  #   "CLASSLOC3"                   ,
  #   "CLASSLOCBUILDNAME1"          ,
  #   "CLASSLOCBUILDROOM1"          ,
  #   "CLASSLOCBUILDNAME2"          ,
  #   "CLASSLOCBUILDROOM2"          ,
  #   "CLASSLOCBUILDNAME3"          ,
  #   "CLASSLOCBUILDROOM3"          ,
  #   "STARTDT"                     ,
  #   "ENDDT"                       ,
  #   "BUDGETCODE"                  ,
  #   "LINEITEM"                    ,
  #   "SERVICELEARNING"             ,
  #   "XLIST_ID"                    ,
  #   "COMBINEDID"                  ,
  #   "USHE_ACADYR"                 ,
  #   "USHE_TERM"                   ,
  #   "TERM2"                       ,
  #   "ORG_EFFDT"                   ,
  #   "CLASSENROLLMENTCAPACITY"     ,
  #   "ROOM_MAX_1"                  ,
  #   "ROOM_MAX_2"                  ,
  #   "ROOM_MAX_3"                  ,
  #   "TERM_NBR"                    ,
  #   "CLASS_ATTR_LIST"             ,
  #   "EXCLUDE_BUDGET_SCH"          ,
  #   "SPR_CORRECTION_NOT_USHE_FLAG",
  #   "Section Divider: OLD"        ,
  #   "SUBJECTCOLL"                 ,
  #   "SUBJECT"                     ,
  #   "class.1"                     ,
  #   "COHORT_DT"                   ,
  "SEX"                         ,
  "FIRST_GEN_STATUS_CD"         ,
  "ETHNICITY"                   ,
  "RESSTAT"                     ,
  "FA_PELL"                     ,
  # "AGE"                         ,
  "APCREDIT"                    ,
  "HSGPA"                       ,
  "HSPRIVATE"                   ,
  "HONORS"                      ,
  "ACTCOMP"                     ,
  "ACTENGL"                     ,
  "ACTMATH"                     ,
  "ACTSCI"                      ,
  #  "load"                        ,
  "cohort_year"                 ,
  "class_year"                  ,
  "yr_diff"                     ,
  "season"                      ,
#  "course_level"                ,
#   "vol_cluster"                 ,
  "age_cut"                     
  # "cleanGrade"                  
  #   "wdraw_binary"                ,
  #   "grade_binary"                ,
  #   "grade_trinary"               ,
  #   "grade_quad" 
)

# target_classes <- c("GRADEGPA",  "grade_binary", "grade_trinary", "grade_quad") # "wdraw_binary",
target_classes <- c("GRADEGPA", "grade_quad")

# stop("Work on keep columns")

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter  & ageFilter & withdrawFilter,] # & recent_filter
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)

cleanData$course <- factor(cleanData$course)

# Combine math courses
topCourses <- table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){x[x<0.95]})() |>
  names()

# I'm only adding three more to my volume clustering.
# I guess I'll comment that column out as a ringer

cleanData$course_target <- ifelse(cleanData$course %in% topCourses, as.character(cleanData$course), "other")
cleanData$course_target <- factor(cleanData$course_target)

# stop("Next step is the training")



```

**PURPOSE:**  This report examines predictors of math course selection and academic performance over time.  

**EXECUTIVE SUMMARY:**  This report finds that a decreasing number of students submit ACT test scores.  Theses students are increasingly elite, with better high school GPA's, test scores, and grade performance in initial math courses.

# DISCUSSION



# Enrollment  

### *Courses are separated into high volume or low volume courses.  Courses with less than five percent of the cumulative enrollment are combined together as 'other'.* 


Five math courses have distinctly high enrollment.  These courses are categorized together as "high volume" and designated as "hi_vol" in legends in this course.  These courses are Math 1010, 1050, 1210, 1030, and 1090.  

An additional nine courses are categorized as "low volume."  They are Math 1070, 1220, 1080, 1060, 1310, 2210, 990, 1100, and 1250.

Many small courses that consist of less than 5% of the total enrollment combined are lumped together as "other."



```{r}

# Show enrollment and the top 95th percentile (with a red line, or a second cumulative graphic.) 
# Highlight high volume courses

# table(cleanData$course, useNA = "always") |> 
#  (\(x){x[order(x, decreasing = TRUE)]})() |> 
#  proportions() |> 
#  cumsum() |>
#  (\(x){names(x[x<0.95]) })()
  # plot()

#  eh let's just do the barplot

# high volume CATNBR per cluster  
hiVolCatNbr <- names(courseClusters)[courseClusters == 1] |>
(\(x){gsub("MATH_", "", x) })()


theBar <- table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  barplot(plot = FALSE)

table(cleanData$course, useNA = "always") |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> # put in order
  (\(x){names(x) <- gsub("MATH_", "", names(x)) ; return(x) })() |> # remove preceding "MATH_"
  (\(x){
  barplot(x,
          plot = TRUE,
          col = ifelse(names(x) %in% hiVolCatNbr, "darkorange2","dodgerblue3"),
          las = 2)
    })()

table(cleanData$course, useNA = "always") |> 
    (\(x){x[order(x, decreasing = TRUE)]})() |> 
    proportions() |> 
    cumsum() |>
    (\(x){which(x<0.95)})() |>
    max() |>
  (\(x){abline(v=mean(theBar[c(x,x+1)]), lwd = 2, col = "blue")})()

# I'd like to mess with graphics
# but let's put a text on here at least

text(x = theBar[14],
     y = 8000,
     label = "Keep",
     adj = c(1,0))

text(x = theBar[15],
     y = 8000,
     label = "Lump as 'Other'",
     adj = c(0,0))

# needs arrows

# Add a block arrow to your plot
arrows(x0 = theBar[14.5],   # starting x position
       y0 = 10000,   # starting y position
       x1 = theBar[11],   # ending x position
       y1 = 10000,   # ending y position
       col = "blue",
       lwd = 3,
       length = 0.15,  # arrowhead size
       angle = 20,     # arrowhead angle
       code = 2)       # code=2 puts arrow at the end


arrows(x0 = theBar[15.5],   # starting x position
       y0 = 10000,   # starting y position
       x1 = theBar[18],   # ending x position
       y1 = 10000,   # ending y position
       col = "blue",
       lwd = 3,
       length = 0.15,  # arrowhead size
       angle = 20,     # arrowhead angle
       code = 2)       # code=1 puts arrow at the beginning


mtext(side = 1, text = "Course", cex = 1.2, line = 4)
mtext(side = 2, text = "Enrollment", cex = 1.2, line = 3)
mtext(side = 3, text = "Enrollment by math course", cex = 1.75, font = 2, line = 2)

legend("topright",
       col = c("darkorange2","dodgerblue3"),
       pch = 19,
       pt.cex =2,
       legend = c("High","Low"))


```


```{r}

# Additional column creation and filters

completeCases <- complete.cases(cleanData[, keepColumns])
# View(cleanData[!completeCases, keepColumns]) # at a quick glance it's all missing ACT test submissions  

# Enrollment

remove2025 <- cleanData$class_year < 2025

cleanData$all <- rep(1, nrow(cleanData))

cleanData$submitted <- ifelse(is.na(cleanData$ACTMATH), "not_submitted","submitted")

volumeMap <- data.frame(
      color = c("darkorange2", "dodgerblue3"),
      by = c("hi_vol", "low_vol"),
      lty = rep(1,2) ,
      lwd = rep(3,2)
    )

submitMap <- data.frame(
      color = c("mediumpurple3", "forestgreen"),
      by = c("not_submitted", "submitted"),
      lty = rep(1,2) ,
      lwd = rep(3,2)
    )

courseMap <- data.frame(color = c( c("lightsteelblue", "powderblue", "beige", "moccasin", "rosybrown"), c( 
      "darkseagreen3",   
      "lightgoldenrod3", 
      "tan3",            
      "thistle3",        
      "lightskyblue3",   
      "paleturquoise3", #"wheat3",          
      "plum3",           
      "darkseagreen4",   
      "burlywood3"
    )),
    lty = c(rep(1,5), rep(2:5,3)[1:9]),
    course = c(
      "MATH_1010", "MATH_1050", "MATH_1210", "MATH_1030", "MATH_1090", "MATH_1070", "MATH_1220", "MATH_1080", "MATH_1060", "MATH_1310",
      "MATH_2210", "MATH_990",  "MATH_1100", "MATH_1250"
    ),
    lwd = c(rep(3,5), rep(2,9))
    )

yCex <- 1.5
subCex <- 1.1
mainCex <- 1.7

```


### *Although enrollment has steadily increased over time, the course mixture has changed as Math 1010 has declined to be replaced by Math 1090.* 

```{r}

# I probably need a ylab value of "Count of students" 

overTime(data= cleanData[remove2025,],
         variable = "EMPLID", agg = length, by = "all",
         title = "",
         legend_params_list = list(plot=FALSE)
         )
mtext(side = 3, 
      text = "Enrollment over time",
      line = 2.6,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Enrollment has steadily increased.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Count of students",
      line = 3,
      cex = yCex
      )

overTime(data= cleanData[remove2025,],
         variable = "EMPLID", agg = length, by = "vol_cluster",
         title = "",
         featureMap = volumeMap)
mtext(side = 3, 
      text = "Enrollment over time broken by course cluster",
      line = 2.8,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Students enrolled in high volume courses have increased more.", 
      line = 1.2, 
      font = subCex)
mtext(side = 2,
      text = "Count of students",
      line = 3,
      cex = yCex
      )

overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,],
         variable = "EMPLID", agg = length,
         title = "")
mtext(side = 3, 
      text = "High enrollment courses over time",
      line = 2.8,
      font = 2, 
      cex=mainCex)
mtext(side = 3, 
      text = "Enrollment in 1010 has declined while others, especially 1090, have increased.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "Count of students",
      line = 3,
      cex = yCex
      )



# decline to show this
# overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,],
#         variable = "EMPLID", agg = length)

# COMPLETE.CASES enrollment

# overTime(data= cleanData[remove2025 & completeCases,],
#         variable = "EMPLID", agg = length, by = "vol_cluster")


# overTime(data= cleanData[remove2025,],
#         variable = "EMPLID", agg = length, by = "all")

```

# ACT test scores
### *Submission of ACT test scores has steeply declined, especially for high volume courses.  Only a third of students in some courses (like Math 1090) have submitted ACT test scores.*


```{r}


overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "all",
         title = "Submission of ACT math test scores has declined")

```



```{r}

ratesOverTime(data= cleanData[remove2025, ], variable ="ACTMATH", by = "vol_cluster", featureMap = transform(volumeMap, lty = 3))
mtext(side = 2,
      text = "                 Fraction of students",
      line = 2,
      cex = 1.0
      )


ratesOverTime(data= cleanData[remove2025 & cleanData$vol_cluster == "hi_vol", ], variable ="ACTMATH", by = "course")
mtext(side = 2,
      text = "Fraction of students",
      line = 2,
      cex = 1.0,
      xpd = TRUE
      )



```

```{r eval = FALSE}

# FRACTION SUBMITTING TEST SCORES DEVELOPED

enrollment_by_vol_cluster <- aggregate(EMPLID ~ vol_cluster + class_year, data = cleanData, length)
actmath_by_vol_cluster <- aggregate(ACTMATH ~ vol_cluster + class_year, data = cleanData, length)

score_submission <- merge(enrollment_by_vol_cluster, actmath_by_vol_cluster, by = c("vol_cluster","class_year"), all = TRUE)


score_submission$submit_rate <- score_submission$ACTMATH/score_submission$EMPLID

plot(x = score_submission$class_year,
     y = score_submission$submit_rate,
     ylab = "",
     xlab = "",
     type = "n",
     las = 2)

invisible(
lapply(unique(score_submission$vol_cluster), function(clust){
  
  lineFilter <- score_submission$vol_cluster == clust
  lines(x = score_submission$class[lineFilter],
        y = score_submission$submit_rate[lineFilter],
     type = "l",
     lwd = 3,
     lty = 1,
     col = c("darkorange2", "dodgerblue3")[factor(score_submission$vol_cluster)[lineFilter]]
     )
})
)

legend(x="topright",
       legend = c("hi_vol", "low_vol"),
       title = "Course enrollment",
       col = c("darkorange2", "dodgerblue3"),
       pch = 19,
       pt.cex = 3)

mtext(side = 2, text = "Fraction of enrollment", line=3, cex = 1.2)
mtext(side =3, 
      text = "Fraction of students submitting ACT math scores",
      line = 2.8, 
      cex = 1.5, 
      font = 2)
mtext(side = 3, 
      text = "Submission of test scores has greatly declined", 
      line = 1.2, 
      font = 3,
      cex = 1.1
        )

``` 


```{r eval=FALSE}

# Turned this into it's own function
# almost want to expand the "overTime" function to caclulate and display "rate" 


submission_by_cluster <- overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "vol_cluster",
         title = "Test submission has especially declined for high volume math courses",
         featureMap = volumeMap)

enrollment_by_cluster <- overTime(data= cleanData[remove2025,],
         variable = "EMPLID", agg = length, by = "vol_cluster",
         title = "Enrollment has increased",
         featureMap = volumeMap)


submit_rate <- merge(submission_by_cluster, enrollment_by_cluster[,-which(colnames(enrollment_by_cluster) %in% c("color","lty","lwd"))], by = c("vol_cluster","class_year"))

submit_rate$rate <- submit_rate$ACTMATH/submit_rate$EMPLID



```


```{r eval = FALSE}

# FRACTION SUBMITTING TEST SCORES SANDBOX DEUX

enrollment_by_vol_cluster <- aggregate(EMPLID ~ vol_cluster + class_year, data = cleanData, length)
actmath_by_vol_cluster <- aggregate(ACTMATH ~ vol_cluster + class_year, data = cleanData, length)

score_submission <- merge(enrollment_by_vol_cluster, actmath_by_vol_cluster, by = c("vol_cluster","class_year"), all = TRUE)

plot(y = score_submission$ACTMATH/score_submission$EMPLID, 
     x = score_submission$class_year,
     col = c("red","green")[factor(score_submission$vol_cluster)] )

# wow, ok that's interesting.
# I wonder if I can make my "overTime" cough that up?

score_submission <- aggregate(cbind(enrollment = EMPLID, actmath = ACTMATH) ~ vol_cluster + class_year, 
                              data = cleanData, 
                              FUN = length)

score_submission <- transform(
  aggregate(cbind(enrollment = EMPLID, actmath = ACTMATH) ~ vol_cluster + class_year, 
            data = cleanData, 
            FUN = length),
  submission_rate = actmath / enrollment
)

# I think it would take a bit of work to convert this to my overTime function,...
# but why not? Instead of calculating an agg, I'd just pass a completed one to it.  That preserves colors, formatting, and style  


```


```{r eval=FALSE}

# REPLACED
# test score submission

overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "all",
         title = "Submission of ACT math test scores has declined")


overTime(data= cleanData[remove2025,],
         variable = "ACTMATH", agg = length, by = "vol_cluster",
         title = "Test submission has especially declined for low volume math courses")


overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "ACTMATH", agg = length,
         title = "Students in 1010 have essentially stopped submitting test scores")

overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,], variable = "ACTMATH", agg = length,
         title = "Students in 1080 have essentially stopped submitting test scores")




```

### *Students who submit test scores are a self-selected elite, with better high school GPAs and more AP credits, a rising median ACT score, and consequent better math performance.*   

```{r}

medHSGPA <- overTime(data=cleanData[remove2025,], variable = "HSGPA", by = "submitted", agg = "median", title = "Median high school GPA by test submission over time", featureMap = submitMap)

overTime(data=cleanData[remove2025,], variable = "HSGPA", by = "submitted", agg = "sd", title = "Math grade standard deviation by test submission over time", featureMap = submitMap)

```

On average, test submitters' high school GPA is `r aggregate(HSGPA ~ submitted, data = medHSGPA, mean) |> (\(x){ diff(x[,"HSGPA"])})() |> round(2) ` higher than non-test submitters. 

```{r}

medAP <- overTime(data=cleanData[remove2025,], variable = "APCREDIT", by = "submitted", agg = "median", title = "Median AP credits by test submission over time", featureMap = submitMap)

overTime(data=cleanData[remove2025,], variable = "APCREDIT", by = "submitted", agg = "sd", title = "AP credit standard deviation by test submission over time", featureMap = submitMap)


```

Test score submitters' median AP credit was three credits higher than non-test submitters' median.


```{r}

# cleanData$all <- rep(1, nrow(cleanData))

overTime(data=cleanData[remove2025,], variable = "ACTMATH", by = "all", agg = "median", title = "Median ACT math score submitted over time")

overTime(data=cleanData[remove2025,], variable = "ACTMATH", by = "all", agg = "mean", title = "Mean ACT math score submitted over time")

```

```{r}

# COMPARING TEST SUBMITTERS AND NON-SUBMITTERS 

math_grade <- overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = "submitted", agg = "median", title = "Median math grade by score submission over time", featureMap = submitMap)

overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = "submitted", agg = "sd", title = "Math grade standard deviation by score submission over time", featureMap = submitMap)

```

Math grades have diverged until test score submitters have a median math grade 0.7 points higher than non-submitters in 2024 (3.7 compared to 3.0.) 




```{r eval = FALSE}

# Modify to enable a list of two "by" values
# Can I get two "by" values?
overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = list( "submitted", "vol_cluster") , agg = "median") 
# hold -up right now is the feature map.

# this will take a little work
overTime(data=cleanData[remove2025,], variable = "GRADEGPA", by = list( "submitted", "vol_cluster") , agg = "median", featureMap = data.frame(color= c("red","blue","green","gray"),
           by = c("submitted","not_submitted","hi_vol","low_vol"),
           lty = 1:4,
           lwd = 1:4
          )) 

# wow, really diverging
overTime(data=cleanData[remove2025 & cleanData$vol_cluster == "hi_vol",], variable = "GRADEGPA", by = "submitted", agg = "median") 

overTime(data=cleanData[remove2025 & cleanData$vol_cluster == "low_vol",], variable = "GRADEGPA", by = "submitted", agg = "median") 


```

###  *Courses are distinguished by ACT math scores with heavy overlap.*

```{r}

# Boxplot of 2024 ACT test scores per class.  

boxplotSubset <- cleanData[cleanData$vol_cluster == "hi_vol" & cleanData$class_year == 2024,] 

course_levels <- levels(droplevels(boxplotSubset$course))

# Align to color map
boxplotColors <- courseMap$color[match(course_levels, courseMap$course)]

boxplot(ACTMATH ~ droplevels(course),
        data = boxplotSubset,
        col = boxplotColors,
        xlab = "",
        ylab = "")

mtext(side = 3,
      text = "ACT Math test scores per course in 2024",
      line = 2.8,
      font = 2,
      cex = mainCex
      )
mtext(side = 3, 
      text = "Courses are distinguished by ACT math scores with heavy overlap.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "ACT Math",
      line = 2,
      cex = yCex
      )







```

```{r}


overTime(data = cleanData[remove2025,],
         variable = "ACTMATH",
         agg = median,
         by = "vol_cluster")


overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,],
         variable = "ACTMATH",
         agg = median,
         by = "course")


```



```{r}

# Boxplot of 2024 ACT test scores per class.  

boxplotSubset <- cleanData[cleanData$vol_cluster == "hi_vol" & cleanData$class_year == 2024,] 

course_levels <- levels(droplevels(boxplotSubset$course))

# Align to color map
boxplotColors <- courseMap$color[match(course_levels, courseMap$course)]

boxplot(GRADEGPA ~ droplevels(course),
        data = boxplotSubset,
        col = boxplotColors,
        xlab = "",
        ylab = "")

mtext(side = 3,
      text = "Grade distribution per course in 2024",
      line = 2.8,
      font = 2,
      cex = mainCex
      )
mtext(side = 3, 
      text = "Courses are not distinguished by grade distribution.", 
      line = 1.2, 
      font = 3,
      cex = subCex)
mtext(side = 2,
      text = "ACT Math",
      line = 2,
      cex = yCex
      )


```



# Nothing happens after here

Grade inflation

Looks like 1010 is tough and 1090 is easy (....?)

```{r eval=FALSE}

# GRADE INFLATION
overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "GRADEGPA", agg = function(x){quantile(x,0.75)}) # getting an A is easier?

# overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "GRADEGPA", agg = function(x){quantile(x,0.25)}) # getting harder to flunk?



```


Check out hsgpa, and gpa, and actmath scores by submitted and un-submitted over time.
Looks like submitters might simply be an increasingly elite group of students, so it "looks like" grade inflation but I'm just getting the more elite students to self-identify.












This report needs a complete overhaul.

What I need to do: 
 - Keep your first graphic that explains enrollment numbers and lumping into "other"
 - Repeat function "overTime" for the top handful of predictors
 - Double-check the use of "length" to mean the number of people enrolling or submitting test scores.  It seems too low 
    - I want to find out what percentage of students submit ACTMATH scores. 
 - Pull in some of the graphics from "Math grade by ACT and test score" into here, and leave that for just the heat maps
 - Compare pre-2020 heatmaps vs the heatmaps of 2024

==

Where is that document I was working on?
I was looking at predictors over time, and it was a report of descriptive things. 

It was "Math grade by ACT math score and high school GPA.Rmd"

Seems like I c/should break that into two reports, and one report is just the heatmaps. Or, more accurately, a child I guess.  And it should be "admittance" on left and "performance" on right.  

I'm also pretty sure I haven't knitted the latest version of it.


I need to create an appendix that shows all of the cleaning steps. All of these filters and complete cases. 

But first, let's show our math classes by enrollment

STARTING DESCRIPTIVES


### ACT math scores per course over time

```{r fig.height=9}

# Improve this  graphic by:
# Using the matched color per course
# Placing a grid behind the boxes ---
#  ... or 
# ....comparing all the courses in year 2024 at the end

par(mfrow=c(5,1), mar = c(0,3,3,1), oma = c(4,2,0,0))

for(crse in names(courseClusters[courseClusters == 1]) ){
daBox <- boxplot(ACTMATH ~ course + cohort_year, data = droplevels(cleanData[cleanData$course == "MATH_1010",]), 
         outline = FALSE,
         col = "aliceblue",
         las = 2,
         xlab = "",
         main = crse,
        xaxt = "n")
}

axis(side = 1,
     at = 1:length(2005:2024),
     labels = 2005:2024,
     outer = TRUE,
     las = 2,
     font = 2,
     cex.axis = 1.2)

mtext(text = "ACT math score", side = 2, line = 0.25, outer = TRUE)

```


```{r eval=FALSE}

# let's see that as lines, one line per course

par(mfrow=c(1,1), mar = c(0,3,3,7), oma = c(4,2,0,0))

median_ACT <- aggregate(ACTMATH ~ course + class_year, data = cleanData[cleanData$vol_cluster == "hi_vol",], median, na.rm = TRUE)

plot(median(median_ACT$ACTMATH),
     xlim = range(median_ACT$class_year),
     ylim = range(median_ACT$ACTMATH), 
     xlab = "",
     ylab = "",
     xaxt = "n",
    # yaxt = "n",
     type = "n",
    las = 2,
    main = "Median ACT math score per course over time"
)

colorMap <- c("lightsteelblue", "powderblue", "beige", "moccasin", "rosybrown")
names(colorMap) <- names(courseClusters[courseClusters == 1]) 

lapply(names(courseClusters[courseClusters == 1]), function(crse){
  
  filter <- median_ACT$course == crse
  
  lines(y = median_ACT$ACTMATH[filter],
        x = median_ACT$class_year[filter],
        col = colorMap[crse],
        lwd = 3
        )
  
})


axis(side = 1,
     at = 2005:2025,
     labels = 2005:2025,
     las = 2,
     font = 1,
     cex.axis = 1.2
     )

legend("topright",
       legend = names(colorMap) |> (\(x){  gsub("MATH_","",x)})(),
       lty = 1,
       lwd = 3,
       col = colorMap,
       xpd = TRUE,
       inset = c(-0.25,0))


``` 

And this would look good as a plotly.

I need to see if ACT has gone down over time  
I need to see how the math grade and the HSGPA have changed over time 

```{r eval=FALSE}

median_overall_ACT <- aggregate(ACTMATH ~ class_year, data = cleanData[cleanData$vol_cluster == "hi_vol",], median, na.rm = TRUE)


boxplot(ACTMATH ~ class_year, data = cleanData[cleanData$vol_cluster == "hi_vol",])

# So that's stable over time for the high volume courses 

# And it's really stable for all-l the courses
boxplot(ACTMATH ~ class_year, data = cleanData) 

# now let's see high school grade

boxplot(HSGPA ~ class_year, 
        data = cleanData,
        outline = FALSE)

# now actual math grade

boxplot(GRADEGPA ~ class_year, 
        data = cleanData,
        outline = FALSE)

``` 

```{r eval=FALSE}

# Hunting for changes in grade over time

library(ggridges)
library(ggplot2)

ggplot(your_data, aes(x = HSGPA, y = as.factor(year), fill = as.factor(year))) +
  geom_density_ridges(alpha = 0.7) +
  scale_x_continuous(limits = c(0, 4)) +
  labs(y = "Year", x = "HSGPA") +
  theme_ridges() +
  theme(legend.position = "none")

# doesn't show the subtlety

```

```{r eval=FALSE}

remove2025 <- cleanData$class_year < 2025

# GRADE INFLATION
overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "GRADEGPA", agg = function(x){quantile(x,0.75)}) # getting an A is easier?
overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "GRADEGPA", agg = function(x){quantile(x,0.15)}) # getting harder to flunk?

# TEST SCORE SUBMISSION
overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "ACTMATH", agg = length)
overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,], variable = "ACTMATH", agg = length)

# ENROLLMENT
overTime(data = cleanData[cleanData$vol_cluster == "hi_vol" & remove2025,], variable = "GRADEGPA", agg = length)
overTime(data = cleanData[cleanData$vol_cluster == "low_vol" & remove2025,], variable = "GRADEGPA", agg = length)

# am I positive my colors are consistent across graphs?
# Also, I should truncate and not show 2025
# Also, I need an "other" category --> replace select "course" names with "other"
#     --> Maybe make this part of my plot ability?
#     --> Maybe make this a second function that transforms data before feeding it in?

# but honestly, this is a great look across time.

```

```{r eval=FALSE}

# seems I can dynamically re-name "course" values to aggregate as I want
# Is this a whole 'nuther function?

overTime(data = cleanData, by = "vol_cluster")


```

Seems like I want a report that is "overview of first time freshman" and describes a few factoids about first time freshmen over time --

-- Like, how many, median HSGPA and ACT MATH scores, how many submit the scores 
-- Focus on the predictors that I'm using.

Then, I want an "overview of first math courses for first time freshmen" ... and then describe like I'm doing here.  What fraction of first time freshmen take a math class at the U? What fraction take them in Semesters 1 thru 4?  What fraction only take one math class ... and they are done?   I'd want to compare this to the top ten math courses, and include a nod to the number of instructors. 

Then, I can show "predicting course selection and academic performance" ... and that's my predictive models.  

Then, using causal inference methods, I can write a "recommending course".

A couple of explanatory graphics includes --
- PERFORMANCE = MOTIVATION x ABILITY and subset ability to "Math"  and "Study" Ability

- Time experience  (Prior knowledge --> course selection --> course performance)

There's a lot that I don't have explained  --

Like, so --- so the grade prediction remains high even though the course performance is low.  How irrelevant is course selection?  Isn't that interesting? 

If I told a student, "You'll get a B plus-or-minus one letter grade," I mean, is that what I'm telling him?  

I have a lot of good stuff here.  I just gotta, like, write it up.

I still want to see heatmaps at different instances in time.  

It's almost like I need to write several reports.  One, the massive appendix, is a pile of technical reports that can be referenced.  Then, on top of that, is the "summary" for the lay-reader with the main take-aways and points. 

O.k., let's get three reports polished and moved to the U drive-- Course over time, Grades over time, and predictors over time. 


