---
title: "Exploring Principal Components Analysis"
author: "Bill Prisbrey"
date: "2026-01-29"
output:
  word_document: default
  html_document:
    keep_md: true
---

**PURPOSE:** This report describes using principal components analysis in training predictive models (xgboost) to estimate math grades.  


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
#library(ggplot2)
#library(patchwork)
#library(kableExtra)
library(factoextra)
library(FactoMineR)

source(here::here("Functions", "Predict Math Success Functions.R"))

```

```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

# print("You made it past loading")


```

```{r}


##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# prepare FTF data

# Remove students with multiple values for COHORT_DT in FTF data
# problemIDs <- c("00517160", "00572252", "00280932", "00643029", "00548206")

theDupes <- table(ftfData$EMPLID) |>
  (\(x){x[x>1]})() |>
  names()

duplicateFilter <- ftfData$EMPLID %in% theDupes

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

# stop("Prior to merge with difficult filter")

firstCourses <- merge(firstCourses, ftfData[!duplicateFilter, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course[recent_filter]) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course[recent_filter]) |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.9]) })()

courseFilter <- firstCourses$course %in% higherVolCourses

# > higherVolCourses
# [1] "MATH_1010" "MATH_1050" "MATH_1210" "MATH_1030" "MATH_1090" "MATH_1060"
# [7] "MATH_1070" "MATH_1220" "MATH_1080"

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

# print("Prep complete")


```

```{r}

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")


# Combine grades
# BINARY
firstCourses$grade_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_binary[hiFilter] <- 1
firstCourses$grade_binary[loFilter] <- 0

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")

```

```{r}

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)

cleanData$course <- factor(cleanData$course)
cleanData$ETHNICITY <- factor(cleanData$ETHNICITY)

# Add row_id
cleanData$rowid <- paste("ID",cleanData$TERM, cleanData$EMPLID, cleanData$CATNBR, sep = "_" )

# stop("Stop after cleaning")
# print("Clean complete")

```

```{r}
#############################################  
## DISTANCE FROM MEDIAN SUCCESSFUL STUDENT ##  
#############################################  

hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()  

# stop("Good to here")

# Merge to cleanData 
# ADJUSTMENT -- I need to shift this by one year! 

hiGrades$class_year_shift <- hiGrades$class_year + 1

# saveRDS(hiGrades, here::here("Data", "hiGrades from vH3.rds"))

cleanData_orig <- cleanData
cleanData <- merge(cleanData, hiGrades[,c("course", "class_year", "class_year_shift", "ACTMATH.median", "ACTMATH.stdev", "HSGPA.median", "HSGPA.stdev", "HSGPA.count")], by.x = c("course", "class_year"), by.y = c("course","class_year_shift"),  all.x=TRUE)

# Calculate distance as a z-score

# I need to take care to shift this in my training data
# 

cleanData$ACTMATH.z <- (cleanData$ACTMATH - cleanData$ACTMATH.median)/cleanData$ACTMATH.stdev
cleanData$HSGPA.z <- (cleanData$HSGPA - cleanData$HSGPA.median)/cleanData$HSGPA.stdev



cleanData$dist <- sqrt(
  ((cleanData$ACTMATH - cleanData$ACTMATH.median)/cleanData$ACTMATH.stdev)^2 +
    ((cleanData$HSGPA - cleanData$HSGPA.median)/cleanData$HSGPA.stdev)^2
)



# Check
# View(cleanData[cleanData$HSGPA.z < -20 & !is.na(cleanData$HSGPA.z) & !is.na(cleanData$ACTMATH), c(keepColumns,"HSGPA.z", "HSGPA.median", "HSGPA.stdev", "HSGPA.count", "class_year.y" )])

# recall that the aggregate used in hiGrades uses complete cases for both columns 

# Calcualte the distance  
# courseProximity$dist <- sqrt(
#  ((courseProximity$ACTMATH - courseProximity$ACTMATH.median)/courseProximity$ACTMATH.stdev)^2 +
#    ((courseProximity$HSGPA - courseProximity$HSGPA.median)/courseProximity$HSGPA.stdev)^2
# )

# print("Z-scores added")
# stop("Starting principal components")

```

```{r}

##########################  
## PRINCIPAL COMPONENTS ##  
##########################  

# Filters

actFilter <- !is.na(cleanData$ACTMATH)
hsgpaFilter <- !is.na(cleanData$HSGPA) & cleanData$HSGPA > 0

# Convert rowid to row names
rownames(cleanData) <- cleanData$rowid

set.seed(123) # to ensure consistent sign application

no_of_years <- 1
initial_year <- min(cleanData$class_year)+no_of_years - 1
final_year <- max(cleanData$class_year)

rolling_pca <- lapply(initial_year:final_year, function(loop_year){
  
  # Set filters
  target_yr_filter <- (cleanData$class_year %in% (loop_year + 1)) & !is.na(cleanData$class_year)
  loop_yr_filter <- cleanData$class_year %in% (loop_year - no_of_years +1 ):(loop_year) & !is.na(cleanData$class_year)
  courses_in_loop_yr <- unique(cleanData$course[loop_yr_filter & actFilter & hsgpaFilter])
  courses_in_target_yr <- unique(cleanData$course[target_yr_filter & actFilter & hsgpaFilter]) 
  common_courses <- intersect(courses_in_loop_yr, courses_in_target_yr) |> as.character()
  
  per_course_pca <- lapply(common_courses, function(loop_course){ 
    
    # print(loop_course)
    
    courseFilter <- cleanData$course == loop_course & !is.na(cleanData$course)  
    
    course_data <- cleanData[
      loop_yr_filter & courseFilter & actFilter
      , c("ACTMATH","HSGPA", "GRADEGPA")] |> droplevels()
    
    #  course_data_check <- cleanData[
    #    loop_yr_filter & courseFilter & actFilter
    #    , c("ACTMATH","HSGPA", "GRADEGPA", "class_year","course")] |> droplevels()
    
    course_data$PassFail <- ifelse(course_data$GRADEGPA >= 2.4, "Pass", "Fail")
    
    if(nrow(course_data) < 3){ return(NA) } else {
      
      course_pca <- PCA(course_data, 
                        quanti.sup = 3,
                        quali.sup = 4,
                        graph = FALSE) # |> 
      # flipPCA(vars_to_check = c("HSGPA", "ACTMATH")) # to ensure consistent sign application
      
    }
    
  })
  names(per_course_pca) <- common_courses
  return(per_course_pca)
})
names(rolling_pca) <- initial_year:final_year

no_of_years <- 3
initial_year <- min(cleanData$class_year)+no_of_years - 1
final_year <- max(cleanData$class_year)

rolling_pca_3 <- lapply(initial_year:final_year, function(loop_year){
  
  # Set filters
  target_yr_filter <- (cleanData$class_year %in% (loop_year + 1)) & !is.na(cleanData$class_year)
  loop_yr_filter <- cleanData$class_year %in% (loop_year - no_of_years +1 ):(loop_year) & !is.na(cleanData$class_year)
  courses_in_loop_yr <- unique(cleanData$course[loop_yr_filter & actFilter & hsgpaFilter])
  courses_in_target_yr <- unique(cleanData$course[target_yr_filter & actFilter & hsgpaFilter]) 
  common_courses <- intersect(courses_in_loop_yr, courses_in_target_yr) |> as.character()
  
  per_course_pca <- lapply(common_courses, function(loop_course){ 
    
    # print(loop_course)
    
    courseFilter <- cleanData$course == loop_course & !is.na(cleanData$course)  
    
    course_data <- cleanData[
      loop_yr_filter & courseFilter & actFilter
      , c("ACTMATH","HSGPA", "GRADEGPA")] |> droplevels()
    
    #  course_data_check <- cleanData[
    #    loop_yr_filter & courseFilter & actFilter
    #    , c("ACTMATH","HSGPA", "GRADEGPA", "class_year","course")] |> droplevels()
    
    course_data$PassFail <- ifelse(course_data$GRADEGPA >= 2.4, "Pass", "Fail")
    
    if(nrow(course_data) < 3){ return(NA) } else {
      
      course_pca <- PCA(course_data, 
                        quanti.sup = 3,
                        quali.sup = 4,
                        graph = FALSE) # |> 
      # flipPCA(vars_to_check = c("HSGPA", "ACTMATH")) # to ensure consistent sign application
      
    }
    
  })
  names(per_course_pca) <- common_courses
  return(per_course_pca)
})
names(rolling_pca_3) <- initial_year:final_year

# print("PCA complete")

```

```{r}

## CENTROID DIFF ##

centroidDiff <- lapply(rolling_pca, function(loop_yr_list){
  
  lapply(loop_yr_list, function(loop_course){ 
    
    if(any(is.na(loop_course))) {return(NA)}
    if(nrow(loop_course$quali.sup$coord) < 2 ) {return(NA)}
    
    # centroid_diff <- sqrt(sum((loop_course$quali.sup$coord["Pass",] - loop_course$quali.sup$coord["Fail",])^2))
    
    # for binary, the centroids are always collinear with 0, so I can use
    centroid_diff <- sum(loop_course$quali.sup$dist)
    
    return(centroid_diff)
    
  })
  
})

centroidDiff_3 <- lapply(rolling_pca_3, function(loop_yr_list){
  
  lapply(loop_yr_list, function(loop_course){ 
    
    if(any(is.na(loop_course))) {return(NA)}
    if(nrow(loop_course$quali.sup$coord) < 2 ) {return(NA)}
    
    # centroid_diff <- sqrt(sum((loop_course$quali.sup$coord["Pass",] - loop_course$quali.sup$coord["Fail",])^2))
    
    # for binary, the centroids are always collinear with 0, so I can use
    centroid_diff <- sum(loop_course$quali.sup$dist)
    
    return(centroid_diff)
    
  })
  
})

# print("CENTROID DIFF COMPLETE")

```

```{r}

## PREDICTIONS ## 

# Pick the PCA list to use
pca_list <- rolling_pca

no_of_years <- 1
initial_year <- min(cleanData$class_year)+no_of_years - 1
final_year <- max(cleanData$class_year)

rolling_pred <- lapply(initial_year:(final_year-1), function(loop_year){
  
  # Set filters
  target_yr_filter <- (cleanData$class_year %in% (loop_year + 1)) & !is.na(cleanData$class_year)
  loop_yr_filter <- cleanData$class_year %in% (loop_year - no_of_years +1 ):(loop_year) & !is.na(cleanData$class_year)
  courses_in_loop_yr <- unique(cleanData$course[loop_yr_filter & actFilter & hsgpaFilter])
  courses_in_target_yr <- unique(cleanData$course[target_yr_filter & actFilter & hsgpaFilter]) 
  common_courses <- intersect(courses_in_loop_yr, courses_in_target_yr) |> as.character()
  
  per_course_pred <- lapply(common_courses, function(loop_course){ 
    
    # print(loop_course)
    
    courseFilter <- cleanData$course == loop_course & !is.na(cleanData$course)  
    
    course_data <- cleanData[
      target_yr_filter & courseFilter & actFilter
      , c("ACTMATH","HSGPA")] |> droplevels()
    
    #  course_data_check <- cleanData[
    #    loop_yr_filter & courseFilter & actFilter
    #    , c("ACTMATH","HSGPA", "GRADEGPA", "class_year","course")] |> droplevels()
    
    #  course_data$PassFail <- ifelse(course_data$GRADEGPA >= 2.4, "Pass", "Fail")
    
    if(nrow(course_data) < 1){ return(NA) }
    if(any(is.na(pca_list[[as.character(loop_year)]][[loop_course]])) | any(is.null(pca_list[[as.character(loop_year)]][[loop_course]]))  ) {return(NA)}
    
    # predict(perCourse_pca[[math_course]], cleanData[yrFilter_2024 & loop_course_filter, c("HSGPA","ACTMATH") ])
    
    course_pred <- predict(pca_list[[as.character(loop_year)]][[loop_course]],
                           course_data) # |> 
    # flipPCA(vars_to_check = c("HSGPA", "ACTMATH")) # to ensure consistent sign application
    
    # CALCULATE THE DISTANCE      
    
    prior_coord <- cbind(pca_list[[as.character(loop_year)]][[loop_course]]$ind$coord, pca_list[[as.character(loop_year)]][[loop_course]]$ind$dist, pca_list[[as.character(loop_year)]][[loop_course]]$call$X)
    
    dim1_med <- aggregate(Dim.1 ~ PassFail, data = prior_coord, FUN = median)
    dim2_med <- aggregate(Dim.2 ~ PassFail, data = prior_coord, FUN = median)
    
    dist_fail <- sqrt( (course_pred$coord[,"Dim.1"] - dim1_med[1,2])^2 +
                         (course_pred$coord[,"Dim.2"] - dim2_med[1,2])^2
    )
    
    dist_pass <- sqrt(
      (course_pred$coord[,"Dim.1"] - dim1_med[2,2])^2 + 
        (course_pred$coord[,"Dim.2"] - dim2_med[2,2])^2
    )
    
    
    signed_dist <- dist_fail - dist_pass 
    
    
    
    
    return(
      list(course_pred = course_pred,
           distance = cbind(dist_fail, dist_pass, signed_dist),
           prior_median = c(dim1_med, dim2_med),
           target_yr = unique(cleanData$class_year[target_yr_filter]),
           course = loop_course,
           prior_coord = prior_coord
      )
    )
    
  })
  names(per_course_pred) <- common_courses
  return(per_course_pred)
})
names(rolling_pred) <- (initial_year+1):final_year

# print(rep("PREDICTION 1 COMPLETE\n",3))

pca_list <- rolling_pca_3

no_of_years <- 3
initial_year <- min(cleanData$class_year)+no_of_years - 1
final_year <- max(cleanData$class_year)

rolling_pred_3 <- lapply(initial_year:(final_year-1), function(loop_year){
  
  # Set filters
  target_yr_filter <- (cleanData$class_year %in% (loop_year + 1)) & !is.na(cleanData$class_year)
  loop_yr_filter <- cleanData$class_year %in% (loop_year - no_of_years +1 ):(loop_year) & !is.na(cleanData$class_year)
  courses_in_loop_yr <- unique(cleanData$course[loop_yr_filter & actFilter & hsgpaFilter])
  courses_in_target_yr <- unique(cleanData$course[target_yr_filter & actFilter & hsgpaFilter]) 
  common_courses <- intersect(courses_in_loop_yr, courses_in_target_yr) |> as.character()
  
  per_course_pred <- lapply(common_courses, function(loop_course){ 
    
    # print(loop_course)
    
    courseFilter <- cleanData$course == loop_course & !is.na(cleanData$course)  
    
    course_data <- cleanData[
      target_yr_filter & courseFilter & actFilter
      , c("ACTMATH","HSGPA")] |> droplevels()
    
    #  course_data_check <- cleanData[
    #    loop_yr_filter & courseFilter & actFilter
    #    , c("ACTMATH","HSGPA", "GRADEGPA", "class_year","course")] |> droplevels()
    
    #  course_data$PassFail <- ifelse(course_data$GRADEGPA >= 2.4, "Pass", "Fail")
    
    if(nrow(course_data) < 1){ return(NA) }
    if(any(is.na(pca_list[[as.character(loop_year)]][[loop_course]])) | any(is.null(pca_list[[as.character(loop_year)]][[loop_course]]))  ) {return(NA)}
    
    # predict(perCourse_pca[[math_course]], cleanData[yrFilter_2024 & loop_course_filter, c("HSGPA","ACTMATH") ])
    
    course_pred <- predict(pca_list[[as.character(loop_year)]][[loop_course]],
                           course_data) # |> 
    # flipPCA(vars_to_check = c("HSGPA", "ACTMATH")) # to ensure consistent sign application
    
    # CALCULATE THE DISTANCE      
    
    prior_coord <- cbind(pca_list[[as.character(loop_year)]][[loop_course]]$ind$coord, pca_list[[as.character(loop_year)]][[loop_course]]$ind$dist, pca_list[[as.character(loop_year)]][[loop_course]]$call$X)
    
    dim1_med <- aggregate(Dim.1 ~ PassFail, data = prior_coord, FUN = median)
    dim2_med <- aggregate(Dim.2 ~ PassFail, data = prior_coord, FUN = median)
    
    dist_fail <- sqrt( (course_pred$coord[,"Dim.1"] - dim1_med[1,2])^2 +
                         (course_pred$coord[,"Dim.2"] - dim2_med[1,2])^2
    )
    
    dist_pass <- sqrt(
      (course_pred$coord[,"Dim.1"] - dim1_med[2,2])^2 + 
        (course_pred$coord[,"Dim.2"] - dim2_med[2,2])^2
    )
    
    
    signed_dist <- dist_fail - dist_pass 
    
    
    
    
    return(
      list(course_pred = course_pred,
           distance = cbind(dist_fail, dist_pass, signed_dist),
           prior_median = c(dim1_med, dim2_med),
           target_yr = unique(cleanData$class_year[target_yr_filter]),
           course = loop_course,
           prior_coord = prior_coord
      )
    )
    
  })
  names(per_course_pred) <- common_courses
  return(per_course_pred)
})
names(rolling_pred_3) <- (initial_year+1):final_year


# print("PCA predictions complete") 

```

```{r}

#####################################
## EXTRACTING AND MERGING DISTANCE ##
#####################################

year_distance <- list()
yr_inc <- 0
for(year_name in names(rolling_pred)){ 
  yr_inc <- yr_inc + 1
  
  course_distances <- list()
  course_inc <- 0
  for(course_name in names(rolling_pred[[year_name]]) ) {
    
    course_inc <- course_inc + 1
    
    #print(paste(i, inc))
    
    if( all(is.na(rolling_pred[[year_name]][[course_inc]]) ) ) { 
      
      course_distances[[course_inc]] <- NA } else {
        
        course_distances[[course_inc]] <- rolling_pred[[year_name]][[course_inc]][["distance"]] 
        
      }
    
    names(course_distances)[[course_inc]] <- course_name
    
  }
  
  year_distance[[yr_inc]] <- course_distances
  names(year_distance)[[yr_inc]] <- year_name
}

# ok, looks fine.  FINALLY!


distance_frame <- lapply(year_distance, function(x) {
  
  annual_frame <- do.call(rbind, x)
  
}) |>
  (\(x){ do.call(rbind,x)})() |>
  as.data.frame()

# print("DISTANCE EXTRACTED")

# stop("Distance extracted")

# Merge back into cleanData
# cleanData_orig <- cleanData

cleanData <- merge(cleanData, as.data.frame(distance_frame), by = "row.names",all.x = TRUE)
row.names(cleanData) <- cleanData$Row.names
cleanData$Row.names <- NULL

```

# EXECUTIVE SUMMARY 

In order to improve predictive accuracy, transforming the data and engineering new features using principal components was attempted.  This was expected to improve on the previous best model due to de-correlating the main variables.

It worked by extracting the principal components in prior years, projecting the target year data onto the prior principal components, and then calculating several distance metrics.

The approach did not improve model accuracy.  Using distances calculated from z-scores resulted in an R-squared value of 0.26, and using distances calculated from principal components also resulted in an R-squared value of 0.26.  Due to the increased computational complexity of extracting principal components for no improvement, the approach is abandoned.

However, the graphical representation from principal components may be instructive.  As well, a comparison in model accuracy between calculating principal components from just one prior year, or from three prior years, found that one prior year was more predictive. This suggests a trend in the data or gradual change over time as opposed to stability.  


# APPROACH  

The two leading predictors for math grades are the high school GPA and the ACT math test scores.  Previously, these scores were first filtered to just the values for "successful" students (a grade greater than 2.4), next standardized to derive a z-score centered on that filtered median, and finally a Euclidean distance was calculated.  The median-centered z-score was calculated for the prior year, and the distance was calculated for the upcoming year.

However, although the two scores are correlated, the z-scores were calculated independently.  By both de-correlating the variables, and constructing additional features, an improvement on the predictive model accuracy was anticipated. 

The following steps were observed:

  * An additional cleaning step was added (where a few dozen students were removed due to having multiple cohort dates).
  * Principal components were derived per course from ACT Math test scores and high school GPA.
    - Components were derived on a rolling basis of one or three years  
  * Component medians for passing or failing were calculated, where passing was a math grade of 2.4 or higher  
  * Data for the following year were projected onto the principal components derived from prior data. 
  * The Euclidean distance for values in that following year and the prior component medians was calculated.
  * The difference in the distances was calculated.
  
This resulted in three new variables: dist_pass (the distance to the component medians of prior successful students), dist_fail (the distance to the component medians of prior failed students), and the difference (signed_dist, calculated as dist_fail - dist_pass).

These new variables were then passed into a training model (extreme gradient boosting) where data from years less than the year 2024 were used for training, and data from the year 2024 were used for testing. R-squared and mean absolute error were compared. 

# DESCRIPTION 

### Original values compared to principal components

```{r}

plotCourseFilter <- cleanData$course == "MATH_1090" & !is.na(cleanData$course )
plotYearFilter <- cleanData$class_year == 2023 & !is.na(cleanData$class_year)
actPlotFilter <- !is.na(cleanData$ACTMATH)
  
plotFilter <- plotCourseFilter & plotYearFilter & actPlotFilter

plot(y=cleanData$ACTMATH[plotFilter],
     x = cleanData$HSGPA[plotFilter],
     pch = ifelse(cleanData$GRADEGPA[plotFilter] > 2.4, 17, 16 ),
     col = ifelse(cleanData$GRADEGPA[plotFilter] > 2.4, "darkturquoise", "firebrick" ),
     main = "Raw Data\nMath 1090 from 2023 (where 'Pass' is >2.4 GPA)",
     ylab = "ACT MATH",
     xlab = "High school GPA"
     )
legend("topleft",
       legend = c("Pass", "Fail"),
       pch = c(17,16),
       col = c("darkturquoise","firebrick"))


p <- fviz_pca_biplot(rolling_pca[["2023"]][["MATH_1090"]], 
                label = "var",
                habillage = 4,
              #  addEllipses = TRUE,
                mean.point = TRUE,
                mean.point.size = 5,
                title = "Principal Components\nMath 1090 from 2023 (where 'Pass' is >2.4 GPA)")

p$layers[[5]] <- NULL
p

```
   
### 2024 data projected onto the prior principal components   
   
```{r}

plot(x = rolling_pred[["2024"]][["MATH_1090"]][["course_pred"]] [["coord"]][,"Dim.1"],
     y = rolling_pred[["2024"]][["MATH_1090"]][["course_pred"]][["coord"]][,"Dim.2"],
col = "gray35",
 pch = 16,
ylab = "Dim.2",
xlab = "Dim.1",
main = "2024 data projected onto prior components for Math 1090")

library(RColorBrewer)
vals <- rolling_pred[["2024"]][["MATH_1090"]][["distance"]][,"signed_dist"]
cols <- colorRampPalette((brewer.pal(11, "RdYlBu")))(100)[
  cut(vals, breaks = quantile(vals, probs = seq(0, 1, length.out = 101)), 
      labels = FALSE, include.lowest = TRUE)
]

#cols <- colorRampPalette(c("firebrick", "darkturquoise"))(100)[
#  cut(rolling_pred[["2024"]][["MATH_1090"]][["distance"]][,"signed_dist"],
#      breaks = 100, labels = FALSE)
#]

plot(x = rolling_pred[["2024"]][["MATH_1090"]][["course_pred"]] [["coord"]][,"Dim.1"],
     y = rolling_pred[["2024"]][["MATH_1090"]][["course_pred"]][["coord"]][,"Dim.2"],
col = cols,
 pch = 16,
ylab = "Dim.2",
xlab = "Dim.1",
main = "2024 data projected onto prior components for Math 1090\ncolored by signed distance")


# It would be nice to plot the "actual" now, with maybe a square or circle around each point that is bright red or bright green

# that would take a little data wrangling though



```
    
### Components from different rolling periods 

```{r}

plot(x = rolling_pred[["2024"]][["MATH_1090"]][["course_pred"]] [["coord"]][,"Dim.1"],
     y = rolling_pred[["2024"]][["MATH_1090"]][["course_pred"]][["coord"]][,"Dim.2"],
col = "darkorange3",
 pch = 16,
ylab = "Dim.2",
xlab = "Dim.1",
main = "2024 data projected onto prior components for Math 1090\nwith different rolling periods")

points(
  x = rolling_pred_3[["2024"]][["MATH_1090"]][["course_pred"]] [["coord"]][,"Dim.1"],
     y = rolling_pred_3[["2024"]][["MATH_1090"]][["course_pred"]][["coord"]][,"Dim.2"],
col = "purple",
 pch = 1,
  
)

legend("topright",
       legend = c("Rolling 1 yr period", "Rolling 3 yr period"),
       pch = c(16,1),
       col = c("darkorange3", "purple"),
       pt.cex = 2)

legend("bottomright",
       legend = NA,
       bty = "n",
       title = "Projections are slightly different.",
       title.font = 3)

```


### Correlations

```{r}

cor_matrix <- cor(cleanData[actFilter,c("dist_fail","dist_pass","signed_dist","dist","HSGPA.z", "ACTMATH.z", "HSGPA", "ACTMATH")], use = "pairwise.complete.obs")
high_cors <- which(abs(cor_matrix) > 0.7 & cor_matrix < 1, arr.ind = TRUE)
heatmap(cor_matrix, main = "Correlation between distance variables")

# cor_matrix |> 
#  round(2) |>
#  kbl()

```



```{r}

plot(y = cleanData$dist_pass, 
     x= cleanData$dist,
     col = "gray40",
     ylab = "Distance to 'pass' medoid using Principal component 1",
     xlab = "Distance to 'pass' medoid using z-scores",
     main = "As expected, 'successful' distance calculated from principal components\nis highly correlated with distance calculated from z-scores",
     xlim = c(0,20))

corVal <- cor(cleanData$dist_pass, cleanData$dist, use = "complete.obs")

text(y = 8,
     x = 30,
     labels = paste("Correlation of ", 100*round(corVal,2), "%", sep = ""),
     font = 3)

legend("topright",
       title = paste("Correlation of ", 100*round(corVal,2), "%", sep = ""),
       legend = c(NA),
       bty = "n",
       title.font = 4,
       title.cex = 1.1)

```


# MODEL TRAINING RESULTS 

Model | Variable Set | Rolling Period | R-squared | Mean Absolute Error
------|--------------|----------------|-----------|--------------------
vH3 | Z-scores | 1 yr | 0.26 | 0.81
vM2 | Both | 3 yrs | 0.24  | 0.81
vM3 | PCA | 3 yrs | 0.23 | 0.84 
vM4 | PCA | 1 yr | 0.26 | 0.82


Due to an inability to improve on predictive accuracy using z-scores, deriving a "distance" measure from principal components is abandoned.

