---
title: "Course guide"
author: "Bill Prisbrey"
date: "2025-12-12"
output: html_document
---

**PURPOSE:**  This report is a guide to recommended math courses for first time freshmen.  It is based on the incoming qualifications of the median successful student per course in the prior year. 

Add a "year" parameter

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)

source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```


```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

source(here::here("Functions", "Predict Math Success Functions.R"))

```


```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course[recent_filter]) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course[recent_filter]) |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.9]) })()

courseFilter <- firstCourses$course %in% higherVolCourses

# > higherVolCourses
# [1] "MATH_1010" "MATH_1050" "MATH_1210" "MATH_1030" "MATH_1090" "MATH_1060"
# [7] "MATH_1070" "MATH_1220" "MATH_1080"

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

```


```{r}


## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")

# zeroBinary
firstCourses$zero_binary <- NA
firstCourses$zero_binary <- ifelse(firstCourses$GRADE %in% c("E","EU"),"zero","not_zero")

# Combine grades
# ADJUSTED BINARY
firstCourses$D_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-",  "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-") 
firstCourses$D_binary[hiFilter] <- "hi"
firstCourses$D_binary[loFilter] <- "low"

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")


```  



```{r}

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# zeroFilter
zeroFilter <- firstCourses$zero_binary != "zero"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)
cleanData$course <- factor(cleanData$course)

# stop("Prep complete")

``` 


```{r}

############################################
## STATISTICS AND MEDIANS PER COURSE/YEAR ##
############################################


hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()


lowGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                       data = cleanData[cleanData$GRADEGPA <= 2.4,], function(x){
                         c(median = median(x),
                           IQR = IQR(x),
                           stdev = sd(x),
                           Q = quantile(x)
                         ) 
                       }) |>
  (\(x){
    do.call(data.frame,x)
  })()


```


What I am thinking is that I develop a flow chart--

By drawing the rectangles, I am seeing that:  

ACTMATH <= 26 is Cluster 1 course
24 < ACTMATH <= 31 is Cluster 2 courses (note the overlap)
ACTMATH > 30 is Cluster 3 courses

And I could probably do a second column of HSGPA to the right of that, with the overlaps as well.

Now that my scatter is perfectly designed to handle courses . . . 
 . . . I want to plot nice smooth artistic ellipses on my scatterplot using the clusters!
 
...but I like this use of clusters.

And I wonder what the population is per cluster?  Is it 90% in Cluster 2?

```{r}

#####################
## ASSIGN CLUSTERS ##
#####################

yearFilter <- hiGrades$class_year == 2023

dist_methods <- c("euclidean", "maximum", "manhattan", "canberra", "binary", "minkowski")
clust_methods <- c("ward.D", "ward.D2", "single","complete","average","mcquitty","median","centroid")
clust_combos <- expand.grid(dist = dist_methods, clust = clust_methods)

row_id <- 7 # euclidean, ward.D2
garbo <- courseScatter(hiGrades[yearFilter,], 
              showCourse = FALSE,
              # clusterFilter = c(1,2),
              # ellipseFilter = hiGrades$course == "MATH_1090" & yearFilter,
              dist_params = list(method = clust_combos[ row_id,"dist"]),
              hclust_params = list(method = clust_combos[ row_id,"clust"]),
              cutree_params = list(k=3), 
              # ellipse_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.5,4), ylim = c(10,36)),
              mtext_params = list(side = 3, font = 2, cex = 1.3, line = 0.25, text = paste(row_id, unlist(clust_combos[row_id,]), collapse = ", "))
)

# what happened to my rectangles?
# I need to merge to get there


```


```{r}

# experiment with clustering the detail

actDetail$clust1 <- actDetail |>
  (\(x){row.names(x) <- x$course; 
  x$course <- NULL; 
  return(x[,-which(colnames(x) %in% "ACTMATH.count" )])  })() |>
  dist(method ="euclidean") |>
  hclust(method = "ward.D2")  |>
  #vplot()
 cutree(k=4) #|> table()

  
courseScatter(data = hiGrades[yearFilter & hiGrades$course %in% actDetail$course[actDetail$clust1 ==2],],
  cutree_params=list(k=1),
  plot_params = list(ylim = c(15,36), xlim = c(3.5,4.0))
  )

# I think I like these four groupings

# To Do:
#   Figure out how to plot the courses colored by these clusters
#   Get me a plotly
#   Do I bother adding HSGPA to the clustering criteria? --> No
#   Figure out how to plot the four clusters by themselves in a very visually appealing way
#   Make the table defining the cut-offs per cluster  
#      - including ACT and GPA  

hiGrades |>
  (\(x){merge(hiGrades[yearFilter,], actDetail[,c("course","clust1")], by = "course")})() |>
  courseScatter(
    
    cutree_params = list(k=1)
  )

cluster1_maps <- list(clusterMap = data.frame(color = c("skyblue", "tan4","maroon4", "forestgreen" ), clust = 1:4 ))


hiGrades |>
  (\(x){merge(hiGrades[yearFilter,], actDetail[,c("course","clust1")], by = "course")})() |>
  courseScatter(
    featureMap = cluster1_maps,  
    cutree_params = list(k=1)
  )

# ugh


  
courseScatter(data = hiGrades[yearFilter,],
    ellipse_params = list(plot = FALSE),
    point_params = list(pch = NA_integer_),  
    cutree_params = list(k=1),
    legend_params = list(plot = FALSE),
    plot_params = list(ylim = c(15,36), xlim = c(3.5,4.0))

  )

hiGrades |>
  (\(x){merge(hiGrades[yearFilter,], actDetail, by = "course")})() |>
  
aggregate(ACTMATH.50. ~ clust1,  data = actDetail, mean) |>
  (\(plot){
  points(x=c(3.6,3.7,3.8,3.9),
         y =  c(15,20,25,30),
           # plot$ACTMATH.50.,
         col = c("skyblue", "tan4","maroon4", "forestgreen" ),
         pch = 0:3,
         )
})()


aggregate(ACTMATH.50. ~ clust1,  data = actDetail, mean) |>
  (\(plot){

courseScatter(
    data = plot,
    ellipse_params = list(plot = "n"),
    point_params = list(
      x = c(3.6,3.6,3.6,3.6),
      y = plot$ACTMATH.50.,
      pch = 0:3,
      col = c("skyblue", "tan4","maroon4", "forestgreen" )
      ),  
    cutree_params = list(k=1),
    # legend_params = list(plot = FALSE),
    plot_params = list(ylim = c(15,36), xlim = c(3.5,4.0))
  )
})()


## WHAT IS GOING ON HERE? ##
## FIxED! ##

courseScatter(data = hiGrades[yearFilter,],
    ellipse_params = list(plot = FALSE),
    point_params = list(pch = NA_integer_),  
    cutree_params = list(k=1),
    legend_params = list(plot = FALSE),
    plot_params = list(ylim = c(15,36), xlim = c(3.5,4.0))

  )


points(x=3.75,
       y = 25,
       pch = 7,
       col = "green",
       cex = 1
       )

courseScatter(data = hiGrades[yearFilter,],
    ellipse_params = list(plot = FALSE),
    #point_params = list(pch = NA_integer_),  
    cutree_params = list(k=1),
    legend_params = list(plot = FALSE),
    plot_params = list(ylim = c(15,36), xlim = c(3.5,4.0))
  )

merge(actDetail, hsDetail, by = "course") |>
(\(x){aggregate(cbind(ACTMATH.50., HSGPA.50.) ~ clust1,  data = x, mean) })()|>

# aggregate(ACTMATH.50. ~ clust1,  data = actDetail, mean) |>
  
  (\(plot){
  points(x=plot$HSGPA.50.,
         y = plot$ACTMATH.50.,
         col = c("skyblue", "tan4","maroon4", "forestgreen" ),
         pch = 0:3,
         cex = 3
         )
})()
  
# Man.... almost there but still oddly off !
# My cluster mean-of-the-medians looks too high
# Is it the wrong year ?
  
```



```{r}

############################# 
## PER COURSE DETAIL TABLE ## 
############################# 

actDetail <- aggregate(ACTMATH ~ course, data = cleanData[cleanData$class_year == 2023 & cleanData$GRADEGPA > 2.4, ], function(x){c(quantile(x,probs = c(0.1,0.25,0.5,0.75,0.9), na.rm=TRUE),count = length(x))}) |>
  (\(x){
    do.call(data.frame,x)
  })() 


# cockney <- merge(actDetail, garbo[,c("course","clust")], by = "course")

# actDetail |>
#  kbl()

hsDetail <- aggregate(HSGPA ~ course, data = cleanData[cleanData$class_year == 2023 & cleanData$GRADEGPA > 2.4, ], function(x){c(quantile(x,probs = c(0.1,0.25,0.5,0.75,0.9), na.rm=TRUE ), count = length(x))}) |>
    (\(x){
    do.call(data.frame,x)
  })()

# hsDetail |>
#  kbl()

``` 




```{r}

# ABANDON
# Let's get this from start to stop 

actDetail$clust1 <- actDetail |>
  (\(x){row.names(x) <- x$course; 
  x$course <- NULL; 
  return(x[,-which(colnames(x) %in% "ACTMATH.count" )])  })() |>
  dist(method ="euclidean") |>
  hclust(method = "ward.D2")  |>
  #vplot()
 cutree(k=4)

# check

all(actDetail$course %in% hsDetail$course) # TRUE
all(hiGrades$course %in% actDetail$course) # FALSE # oh
all(hiGrades$course[hiGrades$class_year == 2023] %in% actDetail$course) # TRUE

theDetail <- merge(actDetail, hsDetail, by = "course")

hiGrades_2023 <- merge(hiGrades[hiGrades$class_year == 2023,], theDetail, by = "course")

```

# START HERE 

```{r} 

################  
## START HERE ##  
################  

hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()


hiGrades$clust1 <- hiGrades |>
  (\(x){ x[,colnames(x)[grepl("ACTMATH\\.Q\\.", colnames(x))]]  })() |>
  dist(method ="euclidean") |>
  hclust(method = "ward.D2")  |>
  #vplot()
  cutree(k=4)



lowGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA <= 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()



``` 


# Ok, this is FINALLY what I want

```{r}

###########################
## SHOW CLUSTER ELLIPSES ##
###########################

clusterColors <- c("darkorange", "steelblue", "purple3", "turquoise3")
# clusterColors <- c("tomato","forestgreen","purple3","orange2")

# c("darkorange", "steelblue",   "turquoise4", "yellow2")

  
  
courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              legend_params = list(plot = FALSE),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              )
              )

clusterPoints <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], mean)

par(xpd=NA)
DrawEllipse(x=clusterPoints$HSGPA.Q.50.[1:3],
       y = clusterPoints$ACTMATH.Q.50.[1:3],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[1:3],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.75.-clusterPoints$ACTMATH.Q.25.)[1:3], col = adjustcolor(clusterColors, alpha.f = 0.35)[1:3],
       border = clusterColors[1:3]
)

DrawEllipse(x=clusterPoints$HSGPA.Q.50.[4],
       y = clusterPoints$ACTMATH.Q.50.[4],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[4],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.90.-clusterPoints$ACTMATH.Q.10.)[4], col = adjustcolor(clusterColors, alpha.f = 0.35)[4],
       border = clusterColors[4]
)

legend("topright",
       bg = "snow",
       legend = c("basic", "intermediate","advanced","elite"),
       pt.cex = 3,
       col = clusterColors,
       pch = 19,
       xpd = TRUE,
       inset = c(-0.155,0)
       )

```


# Now I put in a table, I guess?

Or do I show cluster-by-cluster the courses as a zoom-in, uinsg different pch?

```{r}

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 1,
              par_params = list(xpd=FALSE),
              #ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.5,4),ylim = c(15,30)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(15:18,0,1)),
              legend_params = list(inset = c(-0.155,0), title = "Basic", pch = c(15:18,0,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[1],
      text = "BASIC",
      line = -1.5)


courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 2,
              par_params = list(xpd=FALSE),
              # ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4),ylim = c(20,30)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(0:5,1)),
              legend_params = list(inset = c(-0.155,0), title = "Intermediate", pch = c(0:5,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[2],
      text = "INTERMEDIATE",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 3,
              #ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.75,4),ylim = c(20,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,1)),
              legend_params = list(inset = c(-0.155,0), title = "Advanced", pch = c(8,11,12,14,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[3],
      text = "ADVANCED",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              # showCourse = TRUE,
              clusterFilter = 4,
              # ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.9,4),ylim = c(25,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,2,3,1)),
              legend_params = list(inset = c(-0.155,0), title = "Elite", pch = c(8,11,12,14,2,3,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[4],
      text = "ELITE",
      line = -1.5)

```

```{r}

#####################
## GUIDELINE TABLE ##
#####################

clusterTable <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], function(x) {c(min = min(x), max=max(x))} ) |>
  (\(x){
    do.call(data.frame,x)
  })()

# o.k. this will be fun
# first column is name
# second column is ACT range (as text)
# third column is HSGPA range (as text)

data.frame(
  name = c("Basic", "Intermediate", "Advanced", "Elite"),
  min.ACT = floor(clusterTable$ACTMATH.Q.25..min),
  max.ACT = ceiling(clusterTable$ACTMATH.Q.75..max),
  min.HSGPA = round(clusterTable$HSGPA.Q.25..min, 2),
  max.HSGPA = round(clusterTable$HSGPA.Q.75..max, 2)
) |>
  kbl(col.names = c("Level", "Min", "Max", "Min", "Max"),
      align = c("l", "c", "c", "c", "c")) |>
  add_header_above(c(" " = 1, "ACT Math" = 2, "High School GPA" = 2)) |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```


