---
title: "Course guide"
author: "Bill Prisbrey"
date: "2025-12-12"
output: html_document
---

**PURPOSE:**  This report is a guide to recommended math courses for first time freshmen.  It is based on the incoming qualifications of the median successful student per course in the prior year. 

Add a "year" parameter

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```


```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

source(here::here("Functions", "Predict Math Success Functions.R"))

```


```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course[recent_filter]) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course[recent_filter]) |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.9]) })()

courseFilter <- firstCourses$course %in% higherVolCourses

# > higherVolCourses
# [1] "MATH_1010" "MATH_1050" "MATH_1210" "MATH_1030" "MATH_1090" "MATH_1060"
# [7] "MATH_1070" "MATH_1220" "MATH_1080"

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

```


```{r}


## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")

# zeroBinary
firstCourses$zero_binary <- NA
firstCourses$zero_binary <- ifelse(firstCourses$GRADE %in% c("E","EU"),"zero","not_zero")

# Combine grades
# ADJUSTED BINARY
firstCourses$D_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-",  "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-") 
firstCourses$D_binary[hiFilter] <- "hi"
firstCourses$D_binary[loFilter] <- "low"

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")


```  



```{r}

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# zeroFilter
zeroFilter <- firstCourses$zero_binary != "zero"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)
cleanData$course <- factor(cleanData$course)

# stop("Prep complete")

``` 


What I am thinking is that I develop a flow chart--

By drawing the rectangles, I am seeing that:  

ACTMATH <= 26 is Cluster 1 course
24 < ACTMATH <= 31 is Cluster 2 courses (note the overlap)
ACTMATH > 30 is Cluster 3 courses

And I could probably do a second column of HSGPA to the right of that, with the overlaps as well.

Now that my scatter is perfectly designed to handle courses . . . 
 . . . I want to plot nice smooth artistic ellipses on my scatterplot using the clusters!
 
...but I like this use of clusters.

And I wonder what the population is per cluster?  Is it 90% in Cluster 2?

```{r eval=FALSE}

#####################
## ASSIGN CLUSTERS ##
#####################

yearFilter <- hiGrades$class_year == 2023

dist_methods <- c("euclidean", "maximum", "manhattan", "canberra", "binary", "minkowski")
clust_methods <- c("ward.D", "ward.D2", "single","complete","average","mcquitty","median","centroid")
clust_combos <- expand.grid(dist = dist_methods, clust = clust_methods)

row_id <- 7 # euclidean, ward.D2


# I think I like these four groupings

# To Do:
#   Figure out how to plot the courses colored by these clusters
#   Get me a plotly
#   Do I bother adding HSGPA to the clustering criteria? --> No
#   Figure out how to plot the four clusters by themselves in a very visually appealing way
#   Make the table defining the cut-offs per cluster  
#      - including ACT and GPA  

```


# START HERE 

```{r} 

################  
## START HERE ##  
################  

hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()


hiGrades$clust1 <- hiGrades |>
  (\(x){ x[,colnames(x)[grepl("ACTMATH\\.Q\\.", colnames(x))]]  })() |>
  dist(method ="euclidean") |>
  hclust(method = "ward.D2")  |>
  #vplot()
  cutree(k=4)



lowGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA <= 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()



``` 


# Ok, this is FINALLY what I want

```{r}

###########################
## SHOW CLUSTER ELLIPSES ##
###########################

clusterColors <- c("darkorange", "steelblue", "purple3", "turquoise3")
# clusterColors <- c("tomato","forestgreen","purple3","orange2")

# c("darkorange", "steelblue",   "turquoise4", "yellow2")

  
  
courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              legend_params = list(plot = FALSE),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              )
              )

clusterPoints <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], mean)

par(xpd=NA)
DrawEllipse(x=clusterPoints$HSGPA.Q.50.[1:3],
       y = clusterPoints$ACTMATH.Q.50.[1:3],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[1:3],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.75.-clusterPoints$ACTMATH.Q.25.)[1:3], col = adjustcolor(clusterColors, alpha.f = 0.35)[1:3],
       border = clusterColors[1:3]
)

DrawEllipse(x=clusterPoints$HSGPA.Q.50.[4],
       y = clusterPoints$ACTMATH.Q.50.[4],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[4],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.90.-clusterPoints$ACTMATH.Q.10.)[4], col = adjustcolor(clusterColors, alpha.f = 0.35)[4],
       border = clusterColors[4]
)

legend("topright",
       bg = "snow",
       legend = c("basic", "intermediate","advanced","elite"),
       pt.cex = 3,
       col = clusterColors,
       pch = 19,
       xpd = TRUE,
       inset = c(-0.155,0)
       )

```


```{r}

#####################
## GUIDELINE TABLE ##
#####################

clusterTable <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], function(x) {c(min = min(x), max=max(x))} ) |>
  (\(x){
    do.call(data.frame,x)
  })()

# o.k. this will be fun
# first column is name
# second column is ACT range (as text)
# third column is HSGPA range (as text)

data.frame(
  name = c("Basic", "Intermediate", "Advanced", "Elite"),
  min.ACT = floor(clusterTable$ACTMATH.Q.25..min),
  max.ACT = ceiling(clusterTable$ACTMATH.Q.75..max),
  min.HSGPA = round(clusterTable$HSGPA.Q.25..min, 2),
  max.HSGPA = round(clusterTable$HSGPA.Q.75..max, 2)
) |>
  kbl(col.names = c("Level", "Min", "Max", "Min", "Max"),
      align = c("l", "c", "c", "c", "c")) |>
  add_header_above(c(" " = 1, "ACT Math" = 2, "High School GPA" = 2)) |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```



# Now I put in a table, I guess?

Or do I show cluster-by-cluster the courses as a zoom-in, uinsg different pch?

```{r}

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 1,
              par_params = list(xpd=FALSE),
              #ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.5,4),ylim = c(15,30)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(15:18,0,1)),
              legend_params = list(inset = c(-0.155,0), title = "Basic", pch = c(15:18,0,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[1],
      text = "BASIC",
      line = -1.5)


courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 2,
              par_params = list(xpd=FALSE),
              # ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4),ylim = c(20,30)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(0:5,1)),
              legend_params = list(inset = c(-0.155,0), title = "Intermediate", pch = c(0:5,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[2],
      text = "INTERMEDIATE",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 3,
              #ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.75,4),ylim = c(20,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,1)),
              legend_params = list(inset = c(-0.155,0), title = "Advanced", pch = c(8,11,12,14,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[3],
      text = "ADVANCED",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              # showCourse = TRUE,
              clusterFilter = 4,
              # ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.9,4),ylim = c(25,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("HSGPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,2,3,1)),
              legend_params = list(inset = c(-0.155,0), title = "Elite", pch = c(8,11,12,14,2,3,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[4],
      text = "ELITE",
      line = -1.5)

```

