---
title: "Course guide"
author: "Bill Prisbrey"
date: "2025-12-12"
output: html_document
---

**PURPOSE:**  This report is a guide to recommended math courses for first time freshmen.  It is based on the incoming qualifications of successful students (who achieve a minimum grade of B-) in the prior year. 

*Report improvements* 

  - Add explanatory text    
  - Turn "year" into a parameter  
  - Turn plots into interactive "plotly" plots (allowing mouse-over and hover)  
  - Add number lines next to the table for ACT math test scores and high school GPA with the percentile markers 
  - Add descriptive information like the count or % of population per cluster or course
  - Add sections: 
      - Distance to course / top three closest courses (interactive)  
      - Grade prediction per course / top three grades (interactive)    
      - Explanation of grade prediction (SHAP values) 
      
*Additional work* 

  - Investigate median unsuccessful student and compare distances 
  - Investigate whether higher HSGPA and ACTMATH than course median predicts lower grades (suggesting that students should be bumped up if they are >75th or >90th percentiles for a course) ***DONE***   
  - Train a new predictive model that incorporates z-score distance.  ***DONE***

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=11)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```


```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

source(here::here("Functions", "Predict Math Success Functions.R"))




```


```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course[recent_filter]) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course[recent_filter]) |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.9]) })()

courseFilter <- firstCourses$course %in% higherVolCourses

# > higherVolCourses
# [1] "MATH_1010" "MATH_1050" "MATH_1210" "MATH_1030" "MATH_1090" "MATH_1060"
# [7] "MATH_1070" "MATH_1220" "MATH_1080"

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

```


```{r}


## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")

# zeroBinary
firstCourses$zero_binary <- NA
firstCourses$zero_binary <- ifelse(firstCourses$GRADE %in% c("E","EU"),"zero","not_zero")

# Combine grades
# ADJUSTED BINARY
firstCourses$D_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-",  "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-") 
firstCourses$D_binary[hiFilter] <- "hi"
firstCourses$D_binary[loFilter] <- "low"

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")


```  



```{r}

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# zeroFilter
zeroFilter <- firstCourses$zero_binary != "zero"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)
cleanData$course <- factor(cleanData$course)

# stop("Prep complete")

``` 



```{r eval=FALSE}

#####################
## ASSIGN CLUSTERS ##
#####################

yearFilter <- hiGrades$class_year == 2023

dist_methods <- c("euclidean", "maximum", "manhattan", "canberra", "binary", "minkowski")
clust_methods <- c("ward.D", "ward.D2", "single","complete","average","mcquitty","median","centroid")
clust_combos <- expand.grid(dist = dist_methods, clust = clust_methods)

row_id <- 7 # euclidean, ward.D2


# I think I like these four groupings

# To Do:
#   Figure out how to plot the courses colored by these clusters
#   Get me a plotly
#   Do I bother adding HSGPA to the clustering criteria? --> No
#   Figure out how to plot the four clusters by themselves in a very visually appealing way
#   Make the table defining the cut-offs per cluster  
#      - including ACT and GPA  

```


# Cluster placement

```{r} 

################  
## START HERE ##  
################  

hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

# hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
#                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
#                        c(median = median(x),
#                          IQR = IQR(x),
#                          stdev = sd(x),
#                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
#                          count = length(x)
#                        ) 
#                      }) |>
#  (\(x){
#    do.call(data.frame,x)
#  })()


hiGrades$clust1 <- hiGrades |>
  (\(x){ x[,colnames(x)[grepl("ACTMATH\\.Q\\.", colnames(x))]]  })() |>
  dist(method ="euclidean") |>
  hclust(method = "ward.D2")  |>
  #vplot()
  cutree(k=4)



lowGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA <= 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()



``` 



```{r}

###########################
## SHOW CLUSTER ELLIPSES ##
###########################

clusterColors <- c("darkorange", "steelblue", "purple3", "turquoise3")
# clusterColors <- c("tomato","forestgreen","purple3","orange2")

# c("darkorange", "steelblue",   "turquoise4", "yellow2")

  
  
courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4.0), ylim = c(17,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              )
              )

clusterPoints <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], mean)

par(xpd=TRUE)
DrawEllipse(x=clusterPoints$HSGPA.Q.50.[1:3],
       y = clusterPoints$ACTMATH.Q.50.[1:3],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[1:3],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.75.-clusterPoints$ACTMATH.Q.25.)[1:3], col = adjustcolor(clusterColors, alpha.f = 0.35)[1:3],
       border = clusterColors[1:3]
)

par(xpd=TRUE)
DrawEllipse(x=clusterPoints$HSGPA.Q.50.[4],
       y = clusterPoints$ACTMATH.Q.50.[4],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[4],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.90.-clusterPoints$ACTMATH.Q.10.)[4], col = adjustcolor(clusterColors, alpha.f = 0.35)[4],
       border = clusterColors[4]
)

par(xpd=TRUE)
legend("topright",
       bg = "snow",
       legend = rev(c("Basic", "Intermediate","Advanced","Elite")),
       pt.cex = 3,
       col = rev(clusterColors),
       pch = 19,
       xpd = TRUE,
       inset = c(-0.155,0)
       )

```



```{r}

#####################
## GUIDELINE TABLE ##
#####################

clusterTable <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], function(x) {c(min = min(x), max=max(x))} ) |>
  (\(x){
    do.call(data.frame,x)
  })()

# o.k. this will be fun
# first column is name
# second column is ACT range (as text)
# third column is HSGPA range (as text)

data.frame(
  name = c("Basic", "Intermediate", "Advanced", "Elite"),
  min.ACT = floor(clusterTable$ACTMATH.Q.25..min),
  max.ACT = ceiling(clusterTable$ACTMATH.Q.75..max),
  min.HSGPA = round(clusterTable$HSGPA.Q.25..min, 2),
  max.HSGPA = round(clusterTable$HSGPA.Q.75..max, 2)
) |>
  kbl(col.names = c("Level", "Low", "High", "Low", "High"),
      align = c("l", "c", "c", "c", "c")) |>
  add_header_above(c(" " = 1, "ACT Math" = 2, "High School GPA" = 2)) |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```


# Cluster by cluster

```{r}

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 1,
              par_params = list(xpd=FALSE),
              #ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.5,4),ylim = c(15,30)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(15:18,0,1)),
              legend_params = list(inset = c(-0.155,0), title = "Basic", pch = c(15:18,0,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[1],
      text = "BASIC",
      line = -1.5)


courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 2,
              par_params = list(xpd=FALSE),
              # ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4),ylim = c(22,32)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(0:5,1)),
              legend_params = list(inset = c(-0.155,0), title = "Intermediate", pch = c(0:5,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[2],
      text = "INTERMEDIATE",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 3,
              #ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.75,4),ylim = c(24,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,1)),
              legend_params = list(inset = c(-0.155,0), title = "Advanced", pch = c(8,11,12,14,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[3],
      text = "ADVANCED",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              # showCourse = TRUE,
              clusterFilter = 4,
              # ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.9,4),ylim = c(25,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,2,3,1)),
              legend_params = list(inset = c(-0.155,0), title = "Elite", pch = c(8,11,12,14,2,3,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[4],
      text = "ELITE",
      line = -1.5)

```

# Course detail 

## Incoming ACT math test score percentiles of successful students by course  

*where successful students achieved a grade of B-minus or better.*

```{r}

hiGrades[hiGrades$class_year == 2023, c("course", colnames(hiGrades)[grepl("ACTMATH\\.Q", colnames(hiGrades))])] |>
  (\(x){x[order(x$ACTMATH.Q.25),] })() |>
    kbl(caption = "Incoming ACT Math test score percentiles of students achieving B-minus or better by course", col.names = c("course", "10th", "25th", "50th", "75th","90th"),
        row.names = FALSE,
      align = c("l", "c", "c", "c", "c", "c")) |>
#  add_header_above("Incoming qualifications of successful students by course") |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```


## Incoming high school GPA percentiles of successful students by course

*where successful students achieved a grade of B-minus or better.*

```{r}

hiGrades[hiGrades$class_year == 2023,c("course", colnames(hiGrades)[grepl("HSGPA\\.Q", colnames(hiGrades))])] |>
  (\(x){x[order(x$HSGPA.Q.25),] })() |>
  (\(x){x[,2:6] <- round(x[,2:6],2); return(x)})() |>
    kbl(caption = "Incoming high school GPA percentiles of students achieving B-minus or better by course", col.names = c("course", "10th", "25th", "50th", "75th","90th"),
        row.names = FALSE,
      align = c("l", "c", "c", "c", "c", "c")) |>
#  add_header_above("Incoming qualifications of successful students by course") |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```

# Demonstration: Closest courses

Given a student's identification number, the incoming qualifications can be retrieved and the three closest courses by z-score can be returned. 


```{r}


courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4.0), ylim = c(17,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math", "Plotting individual students against incoming qualifications", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              )
              )

set.seed(123)
intermediateConditionFilter <-  cleanData$HSGPA > 3.8 & !is.na(cleanData$HSGPA) &
                 cleanData$HSGPA < 3.9 & !is.na(cleanData$HSGPA) &
                 cleanData$ACTMATH > 25 & !is.na(cleanData$ACTMATH) &
                 cleanData$ACTMATH < 30 & !is.na(cleanData$ACTMATH)

yearFilter <- cleanData$class_year == 2024

interEmplid <- sample(cleanData$EMPLID[intermediateConditionFilter & yearFilter] , 10)

intermediateEmplidFilter <- cleanData$EMPLID == interEmplid[1]

points(x = cleanData$HSGPA[intermediateEmplidFilter] ,
       y = cleanData$ACTMATH[intermediateEmplidFilter],
       cex = 3,
       pch = 13,
       col = "gray10")

text(x = cleanData$HSGPA[intermediateEmplidFilter] ,
     y = cleanData$ACTMATH[intermediateEmplidFilter],
     labels = interEmplid[1],
     pos = 3
     )


```



This could emphasize the three closest courses and identify them in the legend.

A table unique to the student could also be produced with the distance to all courses.  

# Demonstration:  Predicted grades 

Given a student's identification number, 

  - High-risk course selections can be identified (with [XX]% risk of failing) 
  - Grades of close courses can be estimated  

These estimates have a wide margin of error. Many students with high predictions fail and vice versa.   
```{r}  

#########################
## LOAD MODEL AND DATA ##
#########################

# hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  list("GRADEGPA")
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  list("GRADEGPA")
)

# Re-name 

vH3_data_list <- data_list  # test score
vH3_model_list <- model_list # test score

# Align

vH3_data_list[["GRADEGPA"]][["training"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["trainID"]]

vH3_data_list[["GRADEGPA"]][["testing"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["testID"]]

aligned_vH3 <- vH3_data_list[["GRADEGPA"]][["testing"]]


```

```{r}

## DEVELOP COuNTERFACTUALS ## 

testingGround <- aligned_vH3  

courseGrid <- expand.grid(unique(testingGround$EMPLID), unique(testingGround$course))

testingGround <- merge(testingGround, courseGrid, by.x = "EMPLID", by.y="Var1", all = TRUE)
names(testingGround)[names(testingGround) == "course"] <- "course_orig"
names(testingGround)[names(testingGround) == "Var2"] <- "course"

## CALCULATE DISTANCES ## 

tg2 <- merge(testingGround[,-which(colnames(testingGround) %in% c("HSGPA.z", "ACTMATH.z","dist") )],
                        hiGrades[,c("course", "class_year", "class_year_shift", "ACTMATH.median", "ACTMATH.stdev", "HSGPA.median", "HSGPA.stdev", "HSGPA.count")], by.x = c("course", "class_year"), by.y = c("course","class_year_shift"),  all.x=TRUE 
                        )

# Calculate the distance as a z-score 

tg2$ACTMATH.z <- (tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev
tg2$HSGPA.z <- (tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev

# Calculate distance 

tg2$dist <- sqrt(
  ((tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev)^2 +
    ((tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev)^2
)

## COUNTERFACTUAL PREDICTIONS ## 

trainColumns <- vH3_model_list[["GRADEGPA"]]$trainingData |> colnames() |> (\(x){x[-which(x %in% ".outcome")]})()

completeCasesFilter <- tg2[,trainColumns] |>
    complete.cases()

tg2$pred.vH3 <- NA
tg2$pred.vH3[completeCasesFilter] <- predict(vH3_model_list[["GRADEGPA"]], tg2[completeCasesFilter,]               
                        ) 
testingGround$pred.vH3 <- predict(vH3_model_list[["GRADEGPA"]], testingGround) 

```


```{r}

displayCourseAlternatives(
  data = tg2,
  dist_cutoff = 2,
  unid = interEmplid[1]
)


```


```{r eval = FALSE}

Currently, these predictions: 

  - Have a wide error (approximately plus or minus 0.85 grade points) 
  - May be more useful as a rank order of classes (to be determined) 
  - May decrease in accuracy with the distance to the class (needs to be investigated) 
  - Many students who are predicted to have high grades achieve grades of zero 
  - Maybe should just be used to predict high/low (B- or better vs C+ or less)

```

# Demonstration: An explanation of the predicted grades

Per the ethical guidelines [CITE], a student may be entitled to an explanation of machine learning that informs their academic career.  This section uses SHAP values to identify the components that led to the grade prediction per course.   

An interactive app that allowed toggling the inputs and showing the difference in the grade prediction could be useful.  

```{r}  



``` 

