---
title: "Course guide"
author: "Bill Prisbrey"
date: "2025-12-12"
output: html_document
---

**PURPOSE:**  This report is a guide to recommended math courses for first time freshmen.  It is based on the incoming qualifications of successful students (who achieve a minimum grade of B-) in the prior year. 

*Report improvements* 

  - Add explanatory text    
  - Turn "year" into a parameter  
  - Turn plots into interactive "plotly" plots (allowing mouse-over and hover)  
  - Add number lines next to the table for ACT math test scores and high school GPA with the percentile markers 
  - Add descriptive information like the count or % of population per cluster or course
  - Add sections: 
      - Distance to course / top three closest courses (interactive)  
      - Grade prediction per course / top three grades (interactive)    
      - Explanation of grade prediction (SHAP values) 
      
*Additional work* 

  - Investigate median unsuccessful student and compare distances 
  - Investigate whether higher HSGPA and ACTMATH than course median predicts lower grades (suggesting that students should be bumped up if they are >75th or >90th percentiles for a course) ***DONE***   
  - Train a new predictive model that incorporates z-score distance.  ***DONE***

```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=11)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))
source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```


```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

source(here::here("Functions", "Predict Math Success Functions.R"))




```


```{r}

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Course clusters by volume
courseClusters <- table(firstCourses$course[recent_filter]) |>
  (\(x){x[order(x, decreasing = TRUE)]})() |>
  dist() |>
  hclust() |>
  cutree(k=2)

firstCourses$vol_cluster <- ifelse(firstCourses$course %in% names(courseClusters)[courseClusters == 1], "hi_vol","low_vol")

# Course filter
higherVolCourses <- table(firstCourses$course[recent_filter]) |> 
  (\(x){x[order(x, decreasing = TRUE)]})() |> 
  proportions() |> 
  cumsum() |>
  (\(x){names(x[x<0.9]) })()

courseFilter <- firstCourses$course %in% higherVolCourses

# > higherVolCourses
# [1] "MATH_1010" "MATH_1050" "MATH_1210" "MATH_1030" "MATH_1090" "MATH_1060"
# [7] "MATH_1070" "MATH_1220" "MATH_1080"

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

```


```{r}


## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")

# zeroBinary
firstCourses$zero_binary <- NA
firstCourses$zero_binary <- ifelse(firstCourses$GRADE %in% c("E","EU"),"zero","not_zero")

# Combine grades
# ADJUSTED BINARY
firstCourses$D_binary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-",  "C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-") 
firstCourses$D_binary[hiFilter] <- "hi"
firstCourses$D_binary[loFilter] <- "low"

#popCourses$grade_binary <- factor(popCourses$grade_binary,
#                                  levels = c("hi_grade", "low_grade"))


# TRINARY
firstCourses$grade_trinary <- NA
hiFilter <- firstCourses$GRADE %in% c("A", "A-", "B+", "B", "B-")
medFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_trinary[hiFilter] <-  2 #"hi_grade"
firstCourses$grade_trinary[medFilter] <- 1 # "med_grade"
firstCourses$grade_trinary[loFilter] <-  0 #"low_grade"


# popCourses$grade_trinary <- factor(popCourses$grade_trinary,
#                                  levels = c("hi_grade", "med_grade", "low_grade"))

# QUAD
firstCourses$grade_quad <- NA
hiFilter <- firstCourses$GRADE %in% c("A")
medPlusFilter <- firstCourses$GRADE %in% c("A-", "B+", "B", "B-")
medMinusFilter <- firstCourses$GRADE %in% c("C+", "C", "C-")
loFilter <- firstCourses$GRADE %in% c( "D+", "D", "D-", "E", "EU" ) 
firstCourses$grade_quad[hiFilter] <- 3 # "hi_grade"
firstCourses$grade_quad[medPlusFilter] <- 2 # "medPlus_grade"
firstCourses$grade_quad[medMinusFilter] <- 1 #"medMinus_grade"
firstCourses$grade_quad[loFilter] <-0 # "low_grade"

# popCourses$grade_quad <- factor(popCourses$grade_quad,
#                                  levels = c("hi_grade", 
#                                             "medPlus_grade",
#                                             "medMinus_grade",
#                                             "low_grade"))

# popCourses$cleanGrade[popCourses$cleanGrade %in% c("D_plus", "D", "D_minus") ] <- "D"

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")


```  



```{r}

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to includ the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# zeroFilter
zeroFilter <- firstCourses$zero_binary != "zero"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)
cleanData$course <- factor(cleanData$course)

# stop("Prep complete")

``` 



```{r eval=FALSE}

#####################
## ASSIGN CLUSTERS ##
#####################

yearFilter <- hiGrades$class_year == 2023

dist_methods <- c("euclidean", "maximum", "manhattan", "canberra", "binary", "minkowski")
clust_methods <- c("ward.D", "ward.D2", "single","complete","average","mcquitty","median","centroid")
clust_combos <- expand.grid(dist = dist_methods, clust = clust_methods)

row_id <- 7 # euclidean, ward.D2


# I think I like these four groupings

# To Do:
#   Figure out how to plot the courses colored by these clusters
#   Get me a plotly
#   Do I bother adding HSGPA to the clustering criteria? --> No
#   Figure out how to plot the four clusters by themselves in a very visually appealing way
#   Make the table defining the cut-offs per cluster  
#      - including ACT and GPA  

```


# Cluster placement

```{r} 

################  
## START HERE ##  
################  

hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

# hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
#                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
#                        c(median = median(x),
#                          IQR = IQR(x),
#                          stdev = sd(x),
#                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
#                          count = length(x)
#                        ) 
#                      }) |>
#  (\(x){
#    do.call(data.frame,x)
#  })()


hiGrades$clust1 <- hiGrades |>
  (\(x){ x[,colnames(x)[grepl("ACTMATH\\.Q\\.", colnames(x))]]  })() |>
  dist(method ="euclidean") |>
  hclust(method = "ward.D2")  |>
  #vplot()
  cutree(k=4)



lowGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA <= 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()



``` 



```{r}

###########################
## SHOW CLUSTER ELLIPSES ##
###########################

clusterColors <- c("darkorange", "steelblue", "purple3", "turquoise3")
# clusterColors <- c("tomato","forestgreen","purple3","orange2")

# c("darkorange", "steelblue",   "turquoise4", "yellow2")

  
  
courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4.0), ylim = c(17,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              )
              )

clusterPoints <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], mean)

par(xpd=TRUE)
DrawEllipse(x=clusterPoints$HSGPA.Q.50.[1:3],
       y = clusterPoints$ACTMATH.Q.50.[1:3],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[1:3],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.75.-clusterPoints$ACTMATH.Q.25.)[1:3], col = adjustcolor(clusterColors, alpha.f = 0.35)[1:3],
       border = clusterColors[1:3]
)

par(xpd=TRUE)
DrawEllipse(x=clusterPoints$HSGPA.Q.50.[4],
       y = clusterPoints$ACTMATH.Q.50.[4],
       radius.x = 0.5*(clusterPoints$HSGPA.Q.75.-clusterPoints$HSGPA.Q.25.)[4],
       radius.y = 0.5*(clusterPoints$ACTMATH.Q.90.-clusterPoints$ACTMATH.Q.10.)[4], col = adjustcolor(clusterColors, alpha.f = 0.35)[4],
       border = clusterColors[4]
)

par(xpd=TRUE)
legend("topright",
       bg = "snow",
       legend = rev(c("Basic", "Intermediate","Advanced","Elite")),
       pt.cex = 3,
       col = rev(clusterColors),
       pch = 19,
       xpd = TRUE,
       inset = c(-0.155,0)
       )

```



```{r}

#####################
## GUIDELINE TABLE ##
#####################

clusterTable <- aggregate(.~clust1, data = hiGrades[hiGrades$class_year == 2023,], function(x) {c(min = min(x), max=max(x))} ) |>
  (\(x){
    do.call(data.frame,x)
  })()

# o.k. this will be fun
# first column is name
# second column is ACT range (as text)
# third column is HSGPA range (as text)

data.frame(
  name = c("Basic", "Intermediate", "Advanced", "Elite"),
  min.ACT = floor(clusterTable$ACTMATH.Q.25..min),
  max.ACT = ceiling(clusterTable$ACTMATH.Q.75..max),
  min.HSGPA = round(clusterTable$HSGPA.Q.25..min, 2),
  max.HSGPA = round(clusterTable$HSGPA.Q.75..max, 2)
) |>
  kbl(col.names = c("Level", "Low", "High", "Low", "High"),
      align = c("l", "c", "c", "c", "c")) |>
  add_header_above(c(" " = 1, "ACT Math" = 2, "High School GPA" = 2)) |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```


# Cluster by cluster

```{r}

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 1,
              par_params = list(xpd=FALSE),
              #ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.5,4),ylim = c(15,30)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(15:18,0,1)),
              legend_params = list(inset = c(-0.155,0), title = "Basic", pch = c(15:18,0,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[1],
      text = "BASIC",
      line = -1.5)


courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 2,
              par_params = list(xpd=FALSE),
              # ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4),ylim = c(22,32)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(0:5,1)),
              legend_params = list(inset = c(-0.155,0), title = "Intermediate", pch = c(0:5,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[2],
      text = "INTERMEDIATE",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              clusterFilter = 3,
              #ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.75,4),ylim = c(24,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,1)),
              legend_params = list(inset = c(-0.155,0), title = "Advanced", pch = c(8,11,12,14,1))
              )

mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[3],
      text = "ADVANCED",
      line = -1.5)

courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              # showCourse = TRUE,
              clusterFilter = 4,
              # ellipse_params = list(plot = FALSE),
              par_params = list(xpd=FALSE),
              qrect_params = list(border = NA),
              #legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.9,4),ylim = c(25,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              ),
              point_params = list(pch = c(8,11,12,14,2,3,1)),
              legend_params = list(inset = c(-0.155,0), title = "Elite", pch = c(8,11,12,14,2,3,1))
              )


mtext(side = 3,
      font = 4,
      cex = 3,
      col = clusterColors[4],
      text = "ELITE",
      line = -1.5)

```

# Course detail 

## Incoming ACT math test score percentiles of successful students by course  

*where successful students achieved a grade of B-minus or better.*

```{r}

hiGrades[hiGrades$class_year == 2023, c("course", colnames(hiGrades)[grepl("ACTMATH\\.Q", colnames(hiGrades))])] |>
  (\(x){x[order(x$ACTMATH.Q.25),] })() |>
    kbl(caption = "Incoming ACT Math test score percentiles of students achieving B-minus or better by course", col.names = c("course", "10th", "25th", "50th", "75th","90th"),
        row.names = FALSE,
      align = c("l", "c", "c", "c", "c", "c")) |>
#  add_header_above("Incoming qualifications of successful students by course") |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```


## Incoming high school GPA percentiles of successful students by course

*where successful students achieved a grade of B-minus or better.*

```{r}

hiGrades[hiGrades$class_year == 2023,c("course", colnames(hiGrades)[grepl("HSGPA\\.Q", colnames(hiGrades))])] |>
  (\(x){x[order(x$HSGPA.Q.25),] })() |>
  (\(x){x[,2:6] <- round(x[,2:6],2); return(x)})() |>
    kbl(caption = "Incoming high school GPA percentiles of students achieving B-minus or better by course", col.names = c("course", "10th", "25th", "50th", "75th","90th"),
        row.names = FALSE,
      align = c("l", "c", "c", "c", "c", "c")) |>
#  add_header_above("Incoming qualifications of successful students by course") |>
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

```

# Predictive Modeling  

A predictive model for math grades is developed that uses information readily available upon enrollment. 

The model primarily relies on the student's high school GPA and ACT math test scores.  It also uses the difference between the student's qualifications and the prior year's median successful student's qualifications. 

Generally, the model finds that math grades improve with higher ACT math test scores and high school GPA, and decline with distance to the prior year's successful student.   

## *Identify high and low risk groups:*

A predictive model can identify a group of students with a high rate of failure.  Roughly one-third of these "high risk" students will fail (achieving a grade less than 1.7, or D+ or less.)  This compares to the "low risk" students, with only a 5% failure rate (one in twenty.)

## *Compare predicted grades across courses:* 

The model can produce predicted grades per course per student and can be used to compare courses and select courses where the expected grade is higher. 

These predicted grades align to the actual grades with an r-squared value of 0.26, meaning that the incoming qualifications account for only about a quarter of the variation in grade performance. Qualities and factors that presumably have an impact but are not included in the model are individual motivation, student-teacher alignment, course schedule, difficulty and number of concurrent courses, and other individual circumstances.  

The predicted grades also have a mean absolute error of 0.81, putting almost an entire grade point margin of error onto any individual prediction.  This margin of error suggests that incoming qualifications of high school GPA and ACT math test scores are not deterministic and that high values in either category do not guarantee success.  

It also suggests that the predictive model should only be used as a loose guideline.  Examples of individual recommendations accompany this guide. 


```{r}  

#########################
## LOAD MODEL AND DATA ##
#########################

# hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  list("GRADEGPA")
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  list("GRADEGPA")
)

# Re-name 

vH3_data_list <- data_list  # test score
vH3_model_list <- model_list # test score

# Align

vH3_data_list[["GRADEGPA"]][["training"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["trainID"]]

vH3_data_list[["GRADEGPA"]][["testing"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["testID"]]

aligned_vH3 <- vH3_data_list[["GRADEGPA"]][["testing"]]

# Predict

aligned_vH3$pred <- predict(vH3_model_list[["GRADEGPA"]], aligned_vH3)


```



```{r}

# Predicted vs actual hexplot

predictionMap(data = aligned_vH3, 
              plot_title = "Predicted vs Actual for 2024, all courses")


```



```{r}

# Risk table

aligned_vH3[["pred_cut_2.4"]] <- cut(aligned_vH3[["pred"]], breaks = c(0,2.4,4), include.lowest = TRUE)

aligned_vH3[["cut_1.7"]] <- cut(aligned_vH3[["GRADEGPA"]], c(0,1.7,4), include.lowest = TRUE, right = FALSE )


# Create confusion matrix with proportions
conf_table <- table(aligned_vH3[,c("pred_cut_2.4", "cut_1.7")])

# Convert to proportions and percentages
prop_table <- prop.table(conf_table) * 100

# Combine counts and percentages
combined <- matrix(
  paste0(conf_table, "\n(", round(prop_table, 1), "%)"),
  nrow = nrow(conf_table)
)

dimnames(combined) <- dimnames(conf_table)
rownames(combined) <- c("High risk", "Low risk")

combined |>
  kbl(caption = "Performance by risk group",
      col.names = c("Fail (<1.7)", "Pass (≥1.7)"),
      #row.names = c("High_risk", "Low_risk"),
      align = "c",
      escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                font_size = 18) |>
  # add_header_above(c("Group" = 1, "Performance" = 2),
  #                 bold = TRUE, font_size = 20) |>
  row_spec(0, bold = TRUE, font_size = 20, background = "#e6f2ff")

recall <- (conf_table |> as.matrix() |> (\(x){x[1,1]})()  / conf_table |> as.matrix() |> (\(x){sum(x[,1])})()) |> round(2) |> (\(x){x*100})()

precision <- (conf_table |> as.matrix() |> (\(x){x[1,1]})()  / conf_table |> as.matrix() |> (\(x){sum(x[1,])})()) |> round(2) |> (\(x){x*100})()

```   


