---
title: "Individual Math Course Recommendations"
author: "Bill Prisbrey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params: 
  emplid: "01495399"
  class_yr: 2023  
  dist_cutoff: 2 
  closest_cutoff: 6
output:
  html_document:
    keep_md: true
---

**EXPLANATION:** 

This document is a demonstration of a report that could be used to help the incoming freshman select a math course.   It shows the actual performance of students from the year 2024 compared to their grade projections.  

It provides the statistical distance to the qualifications of successful students in the prior year's courses (2023).  Academic performance roughly decreases with distance and courses with a distance measure larger than three are not recommended.  

It also provides individual grade expectations per course based on limited available information about the student and the department's historical records.  

These projected grades have a large margin of error and many students who are projected to perform well end up performing poorly, and vice versa.  Individual effort is important.  Qualities like motivation and factors like concurrent course load, course schedule, and individual circumstances are not captured in the projections. 


**IMPORTANT NOTE:** 

***This document shows historical projections and actual performance for students in 2024.*** 
  
```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=11)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))
# source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```

```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
# mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
# creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

source(here::here("Functions", "Predict Math Success Functions.R"))


```


```{r}  

#########################
## LOAD MODEL AND DATA ##
#########################

# hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  list("GRADEGPA")
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  list("GRADEGPA")
)

# Re-name 

vH3_data_list <- data_list  # test score
vH3_model_list <- model_list # test score

# Align

vH3_data_list[["GRADEGPA"]][["training"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["trainID"]]

vH3_data_list[["GRADEGPA"]][["testing"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["testID"]]

aligned_vH3 <- vH3_data_list[["GRADEGPA"]][["testing"]]


```


```{r}

## DEVELOP COuNTERFACTUALS ## 

testingGround <- aligned_vH3  

courseGrid <- expand.grid(unique(testingGround$EMPLID), unique(testingGround$course))

testingGround <- merge(testingGround, courseGrid, by.x = "EMPLID", by.y="Var1", all = TRUE)
names(testingGround)[names(testingGround) == "course"] <- "course_orig"
names(testingGround)[names(testingGround) == "Var2"] <- "course"

## CALCULATE DISTANCES ## 

tg2 <- merge(testingGround[,-which(colnames(testingGround) %in% c("HSGPA.z", "ACTMATH.z","dist") )],
                        hiGrades[,c("course", "class_year", "class_year_shift", "ACTMATH.median", "ACTMATH.stdev", "HSGPA.median", "HSGPA.stdev", "HSGPA.count")], by.x = c("course", "class_year"), by.y = c("course","class_year_shift"),  all.x=TRUE 
                        )

# Calculate the distance as a z-score 

tg2$ACTMATH.z <- (tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev
tg2$HSGPA.z <- (tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev

# Calculate distance 

tg2$dist <- sqrt(
  ((tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev)^2 +
    ((tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev)^2
)

## COUNTERFACTUAL PREDICTIONS ## 

trainColumns <- vH3_model_list[["GRADEGPA"]]$trainingData |> colnames() |> (\(x){x[-which(x %in% ".outcome")]})()

completeCasesFilter <- tg2[,trainColumns] |>
    complete.cases()

tg2$pred.vH3 <- NA
tg2$pred.vH3[completeCasesFilter] <- predict(vH3_model_list[["GRADEGPA"]], tg2[completeCasesFilter,]               
                        ) 
testingGround$pred.vH3 <- predict(vH3_model_list[["GRADEGPA"]], testingGround) 

```



```{r}

## CALCULATE COURSE CLUSTERS ## 

hiGrades$clust1 <- hiGrades |>
  (\(x){ x[,colnames(x)[grepl("ACTMATH\\.Q\\.", colnames(x))]]  })() |>
  dist(method ="euclidean") |>
  hclust(method = "ward.D2")  |>
  #vplot()
  cutree(k=4)

```

```{r eval=FALSE}

# This enables the selection of example unid's that are called in a separate documnet

criteria <- list("elite" = !is.na(aligned_vH3$ACTMATH) & aligned_vH3$ACTMATH > 30 & aligned_vH3$HSGPA > 3.93 & !is.na(aligned_vH3$HSGPA),
                 "advanced" = !is.na(aligned_vH3$ACTMATH) &  aligned_vH3$ACTMATH > 25 & aligned_vH3$ACTMATH <= 30 & aligned_vH3$HSGPA > 3.9 & aligned_vH3$HSGPA <= 3.93 & !is.na(aligned_vH3$HSGPA),
                 "intermediate" = !is.na(aligned_vH3$ACTMATH) & aligned_vH3$ACTMATH > 23 & aligned_vH3$ACTMATH <= 25 & aligned_vH3$HSGPA > 3.8 & aligned_vH3$HSGPA <= 3.9 & !is.na(aligned_vH3$HSGPA),
                 "basic" = !is.na(aligned_vH3$ACTMATH) & aligned_vH3$ACTMATH > 10 & aligned_vH3$ACTMATH <= 23 & aligned_vH3$HSGPA > 3.7 & aligned_vH3$HSGPA <= 3.8 & !is.na(aligned_vH3$HSGPA),
                 "high" = !is.na(aligned_vH3$ACTMATH) & aligned_vH3$ACTMATH > 30 & aligned_vH3$ACTMATH <= 36 & aligned_vH3$HSGPA > 3.00 & aligned_vH3$HSGPA <= 3.8 & !is.na(aligned_vH3$HSGPA),
                 "low"= !is.na(aligned_vH3$ACTMATH) & aligned_vH3$ACTMATH > 10 & aligned_vH3$ACTMATH <= 20 & aligned_vH3$HSGPA > 3.00 & aligned_vH3$HSGPA <= 3.8 & !is.na(aligned_vH3$HSGPA)
                 )

set.seed(42)
candidates <- lapply(criteria, function(x) {
  sample(aligned_vH3$EMPLID[x],1)
  
})

# candidates


```


# `r paste("Student ID:", params$emplid, sep=" ")`


```{r}

high_risk_qualification <- any(tg2$pred.vH3[tg2$EMPLID == params$emplid & !is.na(tg2$dist) & tg2$dist <= params$dist_cutoff] <= 2.4)

low_risk_qualification <- all(tg2$pred.vH3[tg2$EMPLID == params$emplid & !is.na(tg2$dist) & tg2$dist <= params$dist_cutoff] > 2.4)


```


```{r eval=high_risk_qualification, fig.height=2, fig.width=11}

par(oma = c(0,0,0,0), mar = c(0,0,0,0))

plot(x=0.5,
     y=0.5,
     xlim = c(0,1),
     ylim = c(0,1),
     xaxt = "n",
     yaxt = "n",
     xlab = "",
     ylab = "",
     type = "n" 
     )

box(col = "red", lwd = 3)

rect(xleft = par("usr")[1],
     ybottom = par("usr")[3],
     xright = par("usr")[2],
     ytop = par("usr")[4],
     col = "pink")

text(x=0.5, y = 0.619,
     labels = "HIGH RISK",
     font = 2,
     cex = 3.619,
     col = "red")

text(x=0.5,
     y = 0.3,
     labels = "This student is at high risk of failing one or more math course options.",
     font = 3,
     cex = 1.9)

```



```{r eval=low_risk_qualification, fig.height=2, fig.width=11}

par(oma = c(0,0,0,0), mar = c(0,0,0,0))

plot(x=0.5,
     y=0.5,
     xlim = c(0,1),
     ylim = c(0,1),
     xaxt = "n",
     yaxt = "n",
     xlab = "",
     ylab = "",
     type = "n" 
     )

box(col = "forestgreen", lwd = 3)

rect(xleft = par("usr")[1],
     ybottom = par("usr")[3],
     xright = par("usr")[2],
     ytop = par("usr")[4],
     col = "palegreen")

text(x=0.5, y = 0.619,
     labels = "LOW RISK",
     font = 2,
     cex = 3.619,
     col = "forestgreen")

text(x=0.5,
     y = 0.3,
     labels = "This student is at low risk of failing.",
     font = 3,
     cex = 1.9)

```



## Distance  

```{r}

emplidFilter <- ftfData$EMPLID == params$emplid

xLim <- range(c(ftfData$HSGPA[emplidFilter], 3.6, 4.0))
yLim <- range(c(ftfData$ACTMATH[emplidFilter]-1, 17, 36))

courseScatter(hiGrades[hiGrades$class_year == params$class_yr,], 
              clusterColumn = "clust1",
              ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              legend_params = list(plot = FALSE),
              plot_params = list(xlim = xLim, ylim = yLim),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACT Math", "Plotting individual students against incoming qualifications", "Median student achieving at least B- per course"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              )
              )



points(x = ftfData$HSGPA[emplidFilter] ,
       y = ftfData$ACTMATH[emplidFilter],
       cex = 3,
       pch = 13,
       col = "gray10")

text(x = ftfData$HSGPA[emplidFilter] ,
     y = ftfData$ACTMATH[emplidFilter],
     labels = params$emplid,
     pos = 3
     )


```

## Projections 

```{r}

if(high_risk_qualification) { 

displayCourseAlternatives(
  data = tg2,
  dist_cutoff = params$dist_cutoff,
  closest_cutoff = params$closest_cutoff,
  unid = params$emplid,
  legend_params = list(inset = c(-0.225,0)),
  risk_legend_params = list(inset = c(-0.225,0)),
  cutoff_text_params = list(line = c(2.7,3.7),
  at = par("usr")[2] + 0.25 * (par("usr")[2] - par("usr")[1]) )
)
  
} else {
  
  displayCourseAlternatives(
  data = tg2,
  dist_cutoff = params$dist_cutoff,
  closest_cutoff = params$closest_cutoff,
  unid = params$emplid,
  legend_params = list(inset = c(-0.225,0)),
  risk_legend_params = list(inset = c(-0.225,0)),
  cutoff_text_params = list(line = c(2.5),
  at = par("usr")[2] + 0.25 * (par("usr")[2] - par("usr")[1]) )
)
  
}

```

## Table 

```{r}  

# Student qualifications  

ftfData |>
  (\(x){x[emplidFilter, c("HSGPA","ACTMATH")]})() |>
  unique() |>
  round(2)  |>
  kbl(caption = paste("Student ", params$emplid),row.names = FALSE,
      col.names = c("HS GPA", "ACT Math"),  # nicer column names
      align = "c") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "left") 

```


```{r}

# Table of course distances

# Adjust cut off to use closest_cutoff instead

row_count <- tg2 |>
  (\(x){x[x$EMPLID == params$emplid, ]})() |>
  nrow()

row_check <- tg2 |>
  (\(x){x[x$EMPLID == params$emplid, ]})() |>
  (\(x){x[x$dist <= params$dist_cutoff & !is.na(x$dist), ]})() |>
  nrow()

if(row_check < params$closest_cutoff) {
  
  tableFilter <- tg2 |>
  (\(x){x[x$EMPLID == params$emplid, ]})() |>
  (\(x){x[order(x$dist),]})() |> # descending order  
  (\(x){c(rep(TRUE, params$closest_cutoff), rep(FALSE, row_count - params$closest_cutoff) ) })()
  
  
} else {
  
    tableFilter <- tg2 |>
  (\(x){x[x$EMPLID == params$emplid, ]})() |>
  (\(x){x[order(x$dist),]})() |> # descending order
  (\(x){x$dist <= params$dist_cutoff & !is.na(x$dist)})()
  
}

tg2 |>
  (\(x){x[x$EMPLID == params$emplid, ]})() |> # filter to emplid
  (\(x){x[order(x$dist),]})() |> # descending order
  (\(x){x[tableFilter, ]})() |> # filter to distance cut-off 
  (\(x){row.names(x) <- x[,"course"]; return(x)})() |>
  (\(x){x[,c("dist", 
             "pred.vH3",
          #   "ACTMATH",
          #   "HSGPA",
             "ACTMATH.median", 
             "HSGPA.median")]})() |>
  round(2) |>
  kbl(caption = "Distance to median successful student and predicted grades",
      col.names = c("Distance", "Predicted Grade", "Median ACT Math", "Median HS GPA"),
      align = "c") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

```


## Projection explanation 

The grade projection is based on [list of factors].  A separate logistic regression analysis found that [factors] are statistically significant predictors (p < .05).  

Further explanation of any particular projection is available upon request. 

A full explanation of the methodology is available.  

 



