---
title: "Counterfactuals Explained (vQ0)"
author: "Bill Prisbrey"
date: "2026-01-12"
format: 
  html:
    keep-md: true
    page-layout: article
    toc: true 
    toc-location: left 
    fig-cap-location: top 
execute: 
  echo: false
  warning: false
  message: false
  error: false
fig-height: 5
fig-width: 8 
editor: source
---

**PURPOSE:** This report describes using counterfactuals and distance in recommending a course.

# What are counterfactuals?

Counterfactuals are a "what if" question. In this case, it is an attempt to answer: *What if the student took a different course?* It uses the model that was trained on the actual courses taken, swaps out the course for the other options, and produces an expected grade.

Counterfactuals have a few problems. Chief among them is that there is no way to evaluate their accuracy: there is no way to know what would have actually happened if a student took a different course. It is taking a model grounded in reality and applying it to an imaginary landscape, exploring an alternate reality where people made different choices.

One way to improve the reasonableness of a counterfactual estimate is to stay within the bounds of the training data. That is to say, the model would produce a poor estimate of a grade for a course that had completely dissimilar students. For example, take Math 2250, where the minimum incoming qualifications for students were an ACT math test score of 23 and a high school GPA of 3.11. The model's predicted grade for that course for a student with much lower qualifications (say, an ACT math test score of 17 with a high school GPA of 3.0) might not be believably realistic.

In order to get the best counterfactual estimates, the predictions should be bounded to be close to what happened in reality.

In this case, it is recommended to limit interpreting counterfactuals to courses that are less than three distance units away. This limit is explored below. (A more thorough approach would be to use the "MatchIt" package to better match the fictitious counterfactual estimates with similar actual students.)

# Using distance to limit counterfactuals

Two variables are the most relevant towards predicting the grade for initial math courses for first time freshmen: high school GPA and ACT math test scores.

Another variable can be derived from these two, called "distance". It is the distance to the median qualifications of the successful student from the prior year. (Please see the section "Distance calculation explained" below for more detail.)

In the first place, the GPA gently declines as the distance increases, as shown in @fig-GPA. This suggests that a student should prefer courses that are closer.

In the second place, prediction accuracy also gently declines (or the prediction error mildly increases) with distance (as shown in @fig-errors.) Actually, if you look at the graph closely, you'll see that the prediction error increases up until a distance of about four, and then starts declining again. This increase in accuracy is due to the fact above: that the model easily predicts low grades for distant courses. Accuracy, but not desirability, is improved for the very distant courses.

Students have historically been capable of sorting themselves (or were guided) into close courses: 91% of the students took a course within three of these distance units (@fig-cuml-perc).

Therefore, in order to improve desirability of a course and preserve a reasonable interpretation of a counterfactual, the author recommends limiting interpreting counterfactuals to courses that are less than three units away.

***Distance calculation explained***

For each course, how students who earned a B- or better performed on ACT math and high school GPA is calculated. An incoming student's distance from that successful group is measured in standardized units across both dimensions.

More specifically, the median and standard deviation are calculated per course per year for the ACT math test scores and the high school GPA for students who achieved a grade of at least a B-minus.  

Next, a standardized score is calculated for any given student.  It is the difference for each of the student's ACT math test score and high school GPA and the prior year's median, divided by the standard deviation.  (Note that this uses the median rather than the mean, making it more robust to outliers.) 

Finally, the distance is calculated in Euclidean standardized space as the square root of the scores squared and summed. 



```{r}
#| include: false

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=11)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(ggplot2)
library(patchwork)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))

```

```{r}

##########
## LOAD ##
##########

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))
creditLoad <- readRDS(here::here("Data", "Freshman_career_credit_load.rds"  ))

##########
## PREP ##
##########

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)

# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

# filter out courses at 3000+ level
course_levelFilter <- firstCourses$course_level < 3

# filter out courses that aren't recent
course_by_year <- table(firstCourses[course_levelFilter, c("course","class_year")  ]) |> as.data.frame()
recentCourses <- unique(course_by_year$course[course_by_year$class_year %in% c(2021:2025) & course_by_year$Freq > 0])
recent_filter <- firstCourses$course %in% recentCourses

# Age buckets
firstCourses$age_cut <- cut(firstCourses$AGE, breaks = c(0,17,18,19,21,100))

# Age filter
ageFilter <- firstCourses$AGE > 11

# print("Prep complete")

## GRADE TRANSFORMATIONS ## 

# convert grade value names to values compatible with R naming

clean_grade <- function(x) {
  x <- as.character(x)
  x <- gsub("-", "_minus", x)   # replace '-' with '_minus'
  x <- gsub("\\+", "_plus", x)  # replace '+' with '_plus'
  x <- gsub(" ", "missing", x)        # replace spaces with '_'
  
  return(x)
}

firstCourses$cleanGrade <- clean_grade(firstCourses$GRADE) 

firstCourses$cleanGrade <- factor(firstCourses$cleanGrade,
                                  levels = c(
                                    "A", "A_minus", "B_plus", "B", "B_minus", "C_plus",
                                    "C", "C_minus", "D_plus", "D", "D_minus",   
                                    "E", "EU", "W", "I", "CR",  "NC", "V", 
                                    "missing", NA
                                  )
)

# RARE AND UNUSUAL VALUES

rareGrades <- firstCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") # , "EU" convert EU to 0

# WITHDRAW
firstCourses$wdraw_binary <- NA
firstCourses$wdraw_binary <- ifelse(firstCourses$GRADE == "W","withdraw","not_withdraw")

## CONVERT APCREDIT NA VALUES to ZERO ##  

firstCourses$APCREDIT[is.na(firstCourses$APCREDIT)] <- 0

# print("Transformations complete")

#################
## PRE-PROCESS ##
#################


# Select columns

# colnames(popCourses)

keepColumns <- c(
  "EMPLID"                      ,
  "course"                      ,
  "SEX"                         ,
  "FIRST_GEN_STATUS_CD"         ,
  "ETHNICITY"                   ,
  "RESSTAT"                     ,
  "FA_PELL"                     ,
  "APCREDIT"                    ,
  "HSGPA"                       ,
  "HSPRIVATE"                   ,
  "HONORS"                      ,
  "ACTCOMP"                     ,
  "ACTENGL"                     ,
  "ACTMATH"                     ,
  "ACTSCI"                      ,
  "cohort_year"                 ,
  "class_year"                  ,
  "yr_diff"                     ,
  "season"                      ,
  "course_level"                ,
  "age_cut"                     ,
  "cleanGrade"                  ,
  "HSGPA.z"                     ,
  "ACTMATH.z"                   ,
  "dist"
)

target_classes <- c("GRADEGPA") 

###########
## CLEAN ##
###########

cleanFilters <- !is.na(firstCourses$cleanGrade) & 
  !rareGrades
#!popCourses$cleanGrade %in% c("missing", "V", "I", "NC", "CR") 

# Outlier filters
zeroHSGPA_filter <- firstCourses$HSGPA == 0
highLoad_filter <- firstCourses$load > quantile(firstCourses$load, 0.99)
preMath_filter <- firstCourses$yr_diff < -0.5 # since all cohort dates start in September, I wanted to include the summer before the cohort date
outlierFilters <- !zeroHSGPA_filter & !highLoad_filter & !preMath_filter

# withdrawFilter
withdrawFilter <- firstCourses$cleanGrade != "W"

# Combined filters
cleanData <- firstCourses[cleanFilters & outlierFilters & course_levelFilter & recent_filter & ageFilter & withdrawFilter,]
cleanData <- cleanData[!is.na(cleanData$cleanGrade),]
cleanData$cleanGrade <- droplevels(cleanData$cleanGrade)

cleanData$course <- factor(cleanData$course)
cleanData$ETHNICITY <- factor(cleanData$ETHNICITY)

#############################################  
## DISTANCE FROM MEDIAN SUCCESSFUL STUDENT ##  
#############################################  

hiGrades <- aggregate(cbind(HSGPA,ACTMATH) ~ course + class_year, 
                      data = cleanData[cleanData$GRADEGPA > 2.4,], function(x){
                        c(median = median(x),
                          IQR = IQR(x),
                          stdev = sd(x),
                          Q = quantile(x, probs = c(0.1,0.25,0.5,0.75,0.9), na.rm = TRUE),
                          count = length(x)
                        ) 
                      }) |>
  (\(x){
    do.call(data.frame,x)
  })()  


# Merge to cleanData 

hiGrades$class_year_shift <- hiGrades$class_year + 1

cleanData <- merge(cleanData, hiGrades[,c("course", "class_year", "class_year_shift", "ACTMATH.median", "ACTMATH.stdev", "HSGPA.median", "HSGPA.stdev", "HSGPA.count")], by.x = c("course", "class_year"), by.y = c("course","class_year_shift"),  all.x=TRUE)

# Calculate distance as a z-score

cleanData$ACTMATH.z <- (cleanData$ACTMATH - cleanData$ACTMATH.median)/cleanData$ACTMATH.stdev
cleanData$HSGPA.z <- (cleanData$HSGPA - cleanData$HSGPA.median)/cleanData$HSGPA.stdev

  

cleanData$dist <- sqrt(
  ((cleanData$ACTMATH - cleanData$ACTMATH.median)/cleanData$ACTMATH.stdev)^2 +
    ((cleanData$HSGPA - cleanData$HSGPA.median)/cleanData$HSGPA.stdev)^2
)

```

```{r}

##############
## LOAD vH3 ##
##############

# load "hiGrades"

hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  list("GRADEGPA")
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  list("GRADEGPA")
)

```

```{r}

# Re-name 

vH3_data_list <- data_list  # test score
vH3_model_list <- model_list # test score

```

```{r}

#################
##      vH3    ## 
## PREDICTIONS ##
################# 

vH3_pred <- predict(vH3_model_list[["GRADEGPA"]], vH3_data_list[["GRADEGPA"]][["testing"]] ) 

thePredictions_vH3 <- list("GRADEGPA" = vH3_pred)

```

```{r}

#############
##   vH3   ## 
## RESULTS ##
############# 


vH3_results <- postResample(pred = vH3_pred,
                            obs = vH3_data_list[["GRADEGPA"]][["testing"]][["GRADEGPA"]] )

theResults_vH3 <- list("GRADEGPA" = vH3_results)

```

```{r}

# Align vH3 prediction

# First add the EMPLID back to the training and testing sets

vH3_data_list[["GRADEGPA"]][["training"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["trainID"]]

vH3_data_list[["GRADEGPA"]][["testing"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["testID"]]

#aligned_vH3 <- alignPrediction(prediction_list = thePredictions_vH3, reference_list = vH3_data_list)

aligned_vH3 <- vH3_data_list[["GRADEGPA"]][["testing"]]

aligned_vH3$pred <- vH3_pred

```

```{r}

#################
## ALL OPTIONS ## 
#################

testingGround <- aligned_vH3[,-which(colnames(aligned_vH3) %in% c("pred"))]  


courseGrid <- expand.grid(unique(testingGround$EMPLID), unique(testingGround$course))

testingGround <- merge(testingGround, courseGrid, by.x = "EMPLID", by.y="Var1", all = TRUE)
names(testingGround)[names(testingGround) == "course"] <- "course_orig"
names(testingGround)[names(testingGround) == "Var2"] <- "course"

```

```{r}

#########################
## CALCULATE DISTANCES ##
######################### 

# merge in the information from the year prior

tg2 <- merge(testingGround[,-which(colnames(testingGround) %in% c("HSGPA.z", "ACTMATH.z","dist") )],
                        hiGrades[,c("course", "class_year", "class_year_shift", "ACTMATH.median", "ACTMATH.stdev", "HSGPA.median", "HSGPA.stdev", "HSGPA.count")], by.x = c("course", "class_year"), by.y = c("course","class_year_shift"),  all.x=TRUE 
                        )

# Calculate the distance as a z-score 

tg2$ACTMATH.z <- (tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev
tg2$HSGPA.z <- (tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev

# Calculate distance 

tg2$dist <- sqrt(
  ((tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev)^2 +
    ((tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev)^2
)


```

```{r}

#################
## PREDICTIONS ##
#################

trainColumns <- vH3_model_list[["GRADEGPA"]]$trainingData |> colnames() |> (\(x){x[-which(x %in% ".outcome")]})()

completeCasesFilter <- tg2[,trainColumns] |>
    complete.cases()
 
#  complete.cases(tg2[,trainColumns])

  
tg2$pred.vH3 <- NA
tg2$pred.vH3[completeCasesFilter] <- predict(vH3_model_list[["GRADEGPA"]], tg2[completeCasesFilter,]               
                        ) 

#[,-which(colnames(tg2) %in% c("class_year.y", "ACTMATH.stdev","HSGPA.median","HSGPA.stdev","HSGPA.count"))]


testingGround$pred.vH3 <- predict(vH3_model_list[["GRADEGPA"]], testingGround) 


```

```{r}
#| label: fig-GPA
#| fig-cap: "GPA with distance"

# Grade decreases with distance

par(mar = c(5.1,6.1,4.1,2.1),
    bg = "lavenderblush"
    )

boxplot(dist ~ GRADEGPA, data = aligned_vH3, outline = FALSE,
        col = "rosybrown",
        xlab = "",
        ylab = "",
        las = 1)

mtext(side = 1,
      text = "Math grade (GPA)",
      font = 1,
      cex = 1.6,
      line = 3.5)

mtext(side = 2,
      text = "Distance",
      font = 1,
      cex = 1.6,
      line = 3.5)

mtext(side = 3,
      text = "GPA with distance",
      font  = 2,
      cex = 2.4,
      line = 2)

mtext(side = 3,
      text = "This plot shows that grade declines with distance",
      font  = 3,
      cex = 1.4,
      line = 0.7)

```

```{r}
#| label: fig-errors
#| fig-cap: "Prediction accuracy with distance"

# Errors increase with distance 

par(mar = c(5.1,6.1,4.1,2.1),
    bg = "lavenderblush"
    )

aligned_vH3$error <- aligned_vH3$pred - aligned_vH3$GRADEGPA

aligned_vH3$dist_band <- cut(aligned_vH3$dist, breaks = c(0,0.5, 1,1.5, 2, 2.5, 3,4,5,10,20,120))

boxplot(abs(error) ~ dist_band, data = aligned_vH3,
        xlab = "",
        ylab = "",
        col = "plum")

mtext(side = 1,
      text = "Distance from median successful student",
      font = 1,
      cex = 1.6,
      line = 3.5)

mtext(side = 2,
      text = "Absolute prediction error",
      font = 1,
      cex = 1.6,
      line = 3.5)


mtext(side = 3,
      text = "Prediction accuracy with distance",
      font  = 2,
      cex = 2.4,
      line = 2)

mtext(side = 3,
      text = "This plot shows that prediction error increases with distance",
      font  = 3,
      cex = 1.4,
      line = 0.8)

```

```{r}
#| label: fig-cuml-perc
#| fig-cap: "Distance as a cumulative percentage"

par(mar = c(5.1,6.1,4.1,2.1),
    bg = "lavenderblush"
    )

actualDist <- tg2$dist[tg2$course_orig == tg2$course] |>
  cut(breaks = c(0,0.5, 1,1.5, 2, 2.5, 3,4,5,10,20,120), right = TRUE, include.lowest = TRUE) |>
  table() 

actualDist |>
  proportions() |>
  cumsum() |>
  (\(x){
    plot(x,
         ylim = c(0, 1),
         las = 1,
         xlab = "",
         ylab = "",
         xaxt = "n",
         yaxt = "n")
  })()

rect(par("usr")[1] ,
     par("usr")[3] ,
     par("usr")[2] ,
     par("usr")[4] ,
     col = "gray92")

invisible(
lapply(seq(0,1, by = 0.05), function(x){
  abline(h = x, lwd = 2, lty=3, col = "gray99")
})
)


invisible(
lapply(1:length(actualDist), function(x){
  abline(v = x, lwd = 2, lty=5, col = "gray99")
})
)

axis(side = 1,
     at = 1:length(actualDist),
     labels = names(actualDist))

axis(side = 2,
     at = seq(from=0, to=1, by = 0.2 ),
     labels = paste(seq(from = 0, to = 100, by =20), "%", sep = ""),
     las =1
     )

actualDist |>
  proportions() |>
  cumsum() |>
  points(pch = 19,
         col = "purple",
         cex=1.15)

dist3 <- actualDist |>
  proportions() |>
  cumsum() |>
  (\(x){ 
    x[which(names(x) == "(2.5,3]")]
    })()
  

text(y = dist3,
     x = which(names(actualDist) == "(2.5,3]"),
     labels = paste(round(100*dist3,0), "%", sep = "" ),
     pos = 1
     )

mtext(side = 1,
      text = "Distance breaks",
      font = 1,
      cex = 1.6,
      line = 3.5)

mtext(side = 2,
      text = "Cumulative percentage",
      font = 1,
      cex = 1.6,
      line = 3.5)

mtext(side = 3,
      text = "Distance as a cumulative percentage",
      font  = 2,
      cex = 2.4,
      line = 2)

mtext(side = 3,
      text = "This plot shows that 91% of all students took a course with a distance less than three",
      font  = 3,
      cex = 1.4,
      line = 0.8)

# is this -ALL- students, or just the successful ones?

```

```{r}
#| eval: false

# I don't think plotting the cumulative percentage of the predictions is informative

 counterfactualDist <- tg2$dist |>
  cut(breaks = c(0,0.5, 1,1.5, 2, 2.5, 3,4,5,10,20,120), right = TRUE) |>
  table() 

 counterfactualDist |>
  proportions() |>
  cumsum() |>
  (\(x){
    points(x,
           pch = 13,
           col = "red")
  })()


```

# Walking through an example

An example is given for a first-time freshman who took Math 1090 in the Fall of 2024. This student had a high GPA coming from high school, but relatively low test scores. These qualifications are plotted and compared to the median qualifications of the successful student from the prior year (@fig-course-scatter and @fig-course-scatter-plotly). Both plots show the interquartile range plotted as an ellipse around Math 1090, the class he actually ended up taking, that roughly show the qualifications of about half of the successful students.

Our student is outside of that range, and it's not terribly clear which course is closest to him. Several courses seem reasonable.

A look at the counterfactuals shows a wide range of estimated grades as well (@fig-counterfactual). The counterfactuals are sorted from the closest to the farthest, with the three-unit cutoff shown as a horizontal orange line. His actual performance is shown as a circled red 'x'.

As it is, he performed better than expected with a grade of 3.7 against a prediction of 2.7. However, one wonders if he could have done even better in Math 1035, where he was predicted to achieve 3.2. In any case, it looks like he could have done as well in any of Math 1030, 1010, 1050, 1100, or 1035 as he did in 1090. 1040, where his predicted grade was lower, might be a questionable recommendation. Math 1070, 1060, 1080, and 1215 can hardly be recommended and Math 1210 can't be recommended at all as it is both far away from him and his predicted grade is so low it puts him at a higher risk of failing.

```{r}
#| label: fig-course-scatter
#| fig-cap: "Median qualifications of successful students per course (year 2023)"
#| fig-width: 6.5

courseScatter(data = hiGrades[hiGrades$class_year == 2023,],
            #  cutree_params = list(k=4), # sheesh, how do I match the course guide?
            showCourse = FALSE,
            ellipseFilter = which(hiGrades$course[hiGrades$class_year == 2023] == unique(tg2$course_orig[tg2$EMPLID ==  "01487355" ]) ),
            plot_params = list(ylim = c(15,36)),
            mtext_params = list(
text = c("HSGPA","ACT MATH", "Median qualifications of successful students", sep = "") 
            ),
            qrect_params = list(border = NA ),
            legend_params = list(plot = FALSE) # inset = c(-0.17,0
            )

points(x = unique(tg2$HSGPA[tg2$EMPLID == "01487355"]) ,
       y = unique(tg2$ACTMATH[tg2$EMPLID == "01487355"]),
       pch = 13,
       cex = 3,
       col = "firebrick")

text(
  x = unique(tg2$HSGPA[tg2$EMPLID == "01487355"]) ,
  y = unique(tg2$ACTMATH[tg2$EMPLID == "01487355"]),
  labels = "01487355",
  pos = 3,
  col = "firebrick",
  cex = 1.2
  
)

```


```{r}
#| label: fig-course-scatter-plotly
#| fig-cap: "Interactive median qualifications of successful students per course (year 2023)"

# plotly version

result <- courseScatterPlotly(
  data = hiGrades[hiGrades$class_year == 2023,],
                    ellipseFilter = which(hiGrades$course[hiGrades$class_year == 2023] == unique(tg2$course_orig[tg2$EMPLID ==  "01487355" ]) ),
                    plot_title = "Interactive median qualifications"
                    ) 

result$plot %>% 
  layout(yaxis = list(range = c(15,38))) %>%
  add_markers(
    x = unique(tg2$HSGPA[tg2$EMPLID == "01487355"]),
    y = unique(tg2$ACTMATH[tg2$EMPLID == "01487355"]),
    marker = list(
      symbol = "x-thin", 
      size = 20, 
      line = list(
        color = "firebrick", 
        width = 3
        )
                  ),
    inherit = FALSE,
    name = "01487355"
  ) 

```


```{r}
#| fig-height: 7
#| fig-width: 7
#| fig-cap: "Counterfactual grade predictions per course"
#| label: fig-counterfactual

displayCourseAlternatives(data = tg2, unid = "01487355", dist_cutoff = NA, closest_cutoff = NA, legend_params = list(inset = c(-0.45,0)), risk_legend_params = list(inset = c(-0.375,0)) ) 


abline(h = 8.5, col = "orange", lwd = 3, lty = 1)

axis(side = 4,
     at = 8,
     labels = "DISREGARD (TOO FAR)",
     tick = FALSE,
     line = -8.5,
     col.axis = "orange",
     font = 2,
     las = 1,
     outer = FALSE
)

points(y = 7.75,
       x = 3.75,
       pch = 25,
       bg = "orange",
       col = "orange",
       cex = 2
)

```


```{r}
#| label: tbl-student
#| tbl-cap: "Incoming student qualifications"

# Student qualifications  

unid <- "01487355"

emplidFilter <- ftfData$EMPLID == unid

ftfData |>
  (\(x){x[emplidFilter, c("HSGPA","ACTMATH")]})() |>
  unique() |>
  round(2)  |>
  kbl(caption = paste("Student ", unid),row.names = FALSE,
      col.names = c("HS GPA", "ACT Math"),  # nicer column names
      align = "c") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "left") 

```

```{r}
#| label: tbl-distance
#| tbl-cap: "Distance to median successful student and predicted grades"

# Table of predicted grades

tableFilter <- tg2 |>
  (\(x){x[x$EMPLID == unid, ]})() |>
  (\(x){rep(TRUE, nrow(x))})()
  
#  (\(x){x[order(x$dist),]})() |> # descending order
#  (\(x){x$dist <= params$dist_cutoff & !is.na(x$dist)})()

tg2 |>
  (\(x){x[x$EMPLID == "01487355", ]})() |> # filter to emplid
  (\(x){x[order(x$dist),]})() |> # descending order
  (\(x){x[tableFilter, ]})() |> # filter to distance cut-off 
  (\(x){row.names(x) <- x[,"course"]; return(x)})() |>
  (\(x){x[,c("dist", 
             "pred.vH3",
          #   "ACTMATH",
          #   "HSGPA",
             "ACTMATH.median", 
             "HSGPA.median")]})() |>
  round(2) |>
  kbl(caption = "Distance to median successful student and predicted grades",
      col.names = c("Distance", "Predicted Grade", "Median ACT Math", "Median HS GPA"),
      align = "c") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)


```

```{r}
#| eval: false

# Errors increase with distance 
# Sandbox

boxplot(pred ~ GRADEGPA, data = aligned_vH3)

boxplot(dist ~ GRADEGPA, data = aligned_vH3, outline = FALSE)

plot(x = aligned_vH3$dist, y = abs(aligned_vH3$error), xlim = c(0,5))
lines(loess.smooth(aligned_vH3$dist, abs(aligned_vH3$error)), col = "blue", lwd = 2)

boxplot(dist_band ~ GRADEGPA, data = aligned_vH3, outline = FALSE)

```
