---
title: "Counterfactual Exploration"
author: "Bill Prisbrey"
date: "2025-11-24"
output: html_document
---

## PURPOSE: 

The purpose of this report is to experiment with counter-factuals by examining the predictions for grades of all courses using a couple of models (vH2 (test scores) and vK0 (no test scores).)


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(ggplot2)
library(patchwork)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r}

##########
## LOAD ##
##########

# Strikes me that I want one list per target class, with multiple years per

# Build file paths
testYears <- 2007:2025
grid <- expand.grid("vK0", testYears,  list("GRADEGPA", "grade_quad"))

model_files <- here::here("Models", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " model.rds"))

# Read this into a list of lists by target class

model_list <- setNames(
  lapply(list("GRADEGPA", "grade_quad"), function(tClass){
    files <- model_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  list("GRADEGPA", "grade_quad")
)



# Build file paths
data_files <- here::here("Data", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(list("GRADEGPA", "grade_quad"), function(tClass){
    files <- data_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  list("GRADEGPA", "grade_quad")
)


```

```{r}

# Re-name 

vK0_data_list <- data_list    # no test
vK0_model_list <- model_list  # no test

``` 


```{r}

## PREDICTIONS ##

predictions_over_time <- lapply(list("GRADEGPA", "grade_quad"), function(tClass){
  perClass <- lapply(as.character(testYears), function(tYr){
    
#  courseColumn <- ifelse("course" %in% colnames(vK0_data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
    
    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["course"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  
            
      
    # predict 
    predict(vK0_model_list[[tClass]][[tYr]],
            newdata = vK0_data_list[[tClass]][[tYr]][["testing"]][missing,]
            )
  })
  names(perClass) <- testYears
  return(perClass)
})
names(predictions_over_time) <- list("GRADEGPA", "grade_quad")

# predictions_over_time[["GRADEGPA"]][["2024"]] |> hist() # looks o.k.



```


```{r}

theResults_vK0 <- lapply(list("GRADEGPA", "grade_quad"), function(tClass){
  
  perClass <- lapply(as.character(testYears), function(tYr){

    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["course"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  

       postResample(pred = predictions_over_time[[tClass]][[tYr]],
                  obs = vK0_data_list[[tClass]][[tYr]][["testing"]][[tClass]][missingCourseFilter]
               ) 
    
        
  })
  names(perClass) <- testYears
  return(perClass)
})
  
names(theResults_vK0) <- list("GRADEGPA", "grade_quad")

```




```{r}

##########
## LOAD ##
##########

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", "vH2", " ", list("GRADEGPA", "grade_quad"), " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  list("GRADEGPA", "grade_quad")
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", "vH2", " ", list("GRADEGPA", "grade_quad"), " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  list("GRADEGPA", "grade_quad")
)

```


```{r}

# Re-name 

vH2_data_list <- data_list  # test score
vH2_model_list <- model_list # test score

``` 


```{r}

## VH2 ##
## PREDICTIONS ##

thePredictions_vH2 <- lapply(list("GRADEGPA", "grade_quad"), function(x) {
  
  # identify course values in testing that aren't in training
  
  courseColumn <- ifelse("course" %in% colnames(vH2_data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(vH2_data_list[[x]][["testing"]][[courseColumn]],vH2_data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(vH2_data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !vH2_data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  # print(x)
  # print(table(missingFilter))
}
  
  predict(vH2_model_list[[x]],
          newdata = vH2_data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions_vH2) <- list("GRADEGPA", "grade_quad")

```

```{r}


theResults_vH2 <- lapply(list("GRADEGPA", "grade_quad"), function(x) {
  
  print(x)
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(vH2_data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(vH2_data_list[[x]][["testing"]][[courseColumn]],vH2_data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(vH2_data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !vH2_data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
#  print(x)
#  print(table(missingFilter))
}
  
  postResample(pred = thePredictions_vH2[[x]],
                  obs = vH2_data_list[[x]][["testing"]][[x]][missingFilter]
  
               )
})
names(theResults_vH2) <- list("GRADEGPA", "grade_quad")



```

```{r}

# Align vH2 prediction

aligned_vH2 <- alignPrediction(prediction_list = thePredictions_vH2, reference_list = vH2_data_list)


```


```{r}

# Align vK0

aligned_vK0 <- alignPrediction(prediction_list = predictions_over_time[["GRADEGPA"]], reference_list = vK0_data_list[["GRADEGPA"]])

``` 

# Combine predictions


```{r}

# Combine predictions

merge_columns <- colnames(aligned_vK0[["2024"]]) |>
  (\(x){x[-which(x %in% c("pred", "ACT_submit", "SAT_submit"))]})()

aligned_comb <- merge(aligned_vK0[["2024"]],
                        aligned_vH2[["GRADEGPA"]],
                        by = merge_columns,
                        all.x = TRUE
                        ) 
colnames(aligned_comb)[colnames(aligned_comb) %in% c("pred.x")] <- "pred.vK0"
colnames(aligned_comb)[colnames(aligned_comb) %in% c("pred.y")] <- "pred.vH2"

```



# Develop predictions for all courses

```{r}

############################################
## CREATE 'TESTING GROUND' COUNTERFACTUAL ##
############################################

testingGround <- aligned_comb[,-which(colnames(aligned_comb) %in% c("pred.vK0","pred.vH2"))]
testingGround$unid <- 10001:(10000+nrow(testingGround))

courseGrid <- expand.grid(testingGround$unid, unique(testingGround$course))

testingGround <- merge(testingGround, courseGrid, by.x = "unid", by.y="Var1", all = TRUE)
names(testingGround)[names(testingGround) == "course"] <- "course_orig"
names(testingGround)[names(testingGround) == "Var2"] <- "course"

## PREDICTIONS ##

testingGround$pred.vH2[!is.na(testingGround$ACTMATH)] <- predict(vH2_model_list[["GRADEGPA"]], testingGround[!is.na(testingGround$ACTMATH) ,-which(colnames(testingGround) %in% c("unid","course_orig"))])

testingGround$pred.vK0 <- predict(vK0_model_list[["GRADEGPA"]][["2024"]],                                  testingGround[,-which(colnames(testingGround) %in% c("unid","course_orig"))])


testingGround$course <- factor(testingGround$course)

```


Looking at the counterfactuals---
like, should I focus on the counterfactuals for the lowest grades?  See if there was a course that would have put them over the line?

Should I look at lowest actual grades? Largest differences between predictions?  

Lowest predictions?  Best grades despite lowest predictions?  

```{r}

# Lowest actual

bottomQuarter <-   testingGround$unid[testingGround$GRADEGPA < quantile(testingGround$GRADEGPA, 0.25)]

lowSample <- sample(bottomQuarter, 1)

displayCourseAlternatives(data = testingGround,
                          unid = lowSample)

# This is fun, but, like, also not that interesting.

lowPrediction <- testingGround$unid[testingGround$pred.vH2 <= 1.3 & !is.na(testingGround$pred.vH2)] |> unique()

displaySample <- sample(lowPrediction,1)
displayCourseAlternatives(data = testingGround,
                          unid = displaySample)

# so, like ---
# maybe it's the ranking not the absolute (?)
# maybe it's just -not- about the course selection

# and let's look at the linear model

# is the ranking the same for everyone?


aggregate(pred.vH2 ~ unid + course, 
          data = testingGround,
          rank
          ) |> 
  (\(x){ 
  aggregate(pred.vH2 ~ course, data = x, mean)
})()


testingGround$pred_rank <- ave(testingGround$pred.vH2,                          testingGround$unid, 
    FUN = rank)

courseRank <- aggregate(pred_rank ~ course, data = testingGround, mean)

courseRankStDev <- aggregate(pred_rank ~ course, data = testingGround, sd)

# Merge them
plot_data <- merge(courseRank, courseRankStDev, by = "course", suffixes = c("_mean", "_sd"))

# Create the plot
n_courses <- nrow(plot_data)

par(mar = c(3,6,2,1), bg = "bisque")
plot(x = plot_data$pred_rank_mean,
     y = 1:n_courses,
     xlim = range(c(plot_data$pred_rank_mean - plot_data$pred_rank_sd,
                    plot_data$pred_rank_mean + plot_data$pred_rank_sd)),
     ylim = c(1, n_courses),
     pch = 19,
     col = "dodgerblue2",
     xlab = "Average Rank",
     ylab = "",
     yaxt = "n")

# Add y-axis labels
axis(2, at = 1:n_courses, labels = plot_data$course, las = 1)

# Add error bars (mean ± 1 SD)
arrows(x0 = plot_data$pred_rank_mean - plot_data$pred_rank_sd,
       y0 = 1:n_courses,
       x1 = plot_data$pred_rank_mean + plot_data$pred_rank_sd,
       y1 = 1:n_courses,
       angle = 90,
       code = 3,  # arrows on both ends
       length = 0.05,
       col = "dodgerblue2")

segments(x0 = plot_data$pred_rank_mean - plot_data$pred_rank_sd,
         y0 = 1:n_courses,
         x1 = plot_data$pred_rank_mean + plot_data$pred_rank_sd,
         y1 = 1:n_courses,
         col = "gray60",
         lwd = 2)

points(x = plot_data$pred_rank_mean,
       y = 1:n_courses,
       pch = 19,
       col = "dodgerblue2",
       cex = 1.5)

``` 

Wow, so these really are mixed up according to the individual.   That makes me feel better, that the model is actually doing something.  

And, like --- how much of a difference is it?  How big of a difference would it have made? How much should I read into 0.05 GPA prediction difference?

Math 1210, Math 1050 -- why does it swing so much? 

For whom is 1210 high, and for whom is 1210 low?  What differentiates them?

Is that an invitation to do another ML model to differentiate them?  Would it give me new information?

Or should I just start using SHAP values? 







