---
title: "Counterfactual Exploration"
author: "Bill Prisbrey"
date: "2025-11-24"
output: html_document
---

## PURPOSE: 

The purpose of this report is to experiment with counter-factuals by examining the predictions for grades of all courses using several models (vH2 and vH3 (test scores) and vK0 (no test scores).)

It bugs me that the grade predictions by course are very uniform -- could "dist" separate "good" and "bad" predictions?


```{r include=FALSE}

############
## SET UP ##
############

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE,
                      fig.height=5, fig.width=7)

oldPar <- par(cex.main = 3,
    cex.lab = 3,
    cex.axis = 2,
    mar = c(5.1,4.1,4.1,2.1) # default is c(5.1,4.1,4.1,2.1)
    )

library(here)
library(skimr)
library(caret)
library(xgboost)
library(lubridate)
library(plotly)
library(ggplot2)
library(patchwork)
library(kableExtra)

source(here::here("Functions", "Predict Math Success Functions.R"))

source(here::here("Functions", "Draw Confusion Matrix vC6.R"))

```



```{r eval=FALSE}

##########
## LOAD ##
##########

# Strikes me that I want one list per target class, with multiple years per

# Build file paths
testYears <- 2007:2025
grid <- expand.grid("vK0", testYears,  list("GRADEGPA", "grade_quad"))

model_files <- here::here("Models", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " model.rds"))

# Read this into a list of lists by target class

model_list <- setNames(
  lapply(list("GRADEGPA", "grade_quad"), function(tClass){
    files <- model_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  list("GRADEGPA", "grade_quad")
)



# Build file paths
data_files <- here::here("Data", "Over Time", paste0("Decision Tree ", grid$Var1, " ", grid$Var2, " ", grid$Var3, " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(list("GRADEGPA", "grade_quad"), function(tClass){
    files <- data_files[grep(tClass, model_files)]
    
    # Extract year from each filename
    years <- gsub(".*Decision Tree.*?(\\d{4}).*", "\\1", files)
    
    setNames(
      lapply(files, readRDS),
      years  # Use extracted years
    )
  }),
  list("GRADEGPA", "grade_quad")
)


```

```{r eval=FALSE}

# Re-name 

vK0_data_list <- data_list    # no test
vK0_model_list <- model_list  # no test

``` 


```{r eval=FALSE}

## PREDICTIONS ##

predictions_over_time <- lapply(list("GRADEGPA", "grade_quad"), function(tClass){
  perClass <- lapply(as.character(testYears), function(tYr){
    
#  courseColumn <- ifelse("course" %in% colnames(vK0_data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
    
    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["course"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  
            
      
    # predict 
    predict(vK0_model_list[[tClass]][[tYr]],
            newdata = vK0_data_list[[tClass]][[tYr]][["testing"]][missing,]
            )
  })
  names(perClass) <- testYears
  return(perClass)
})
names(predictions_over_time) <- list("GRADEGPA", "grade_quad")

# predictions_over_time[["GRADEGPA"]][["2024"]] |> hist() # looks o.k.



```


```{r eval=FALSE}

theResults_vK0 <- lapply(list("GRADEGPA", "grade_quad"), function(tClass){
  
  perClass <- lapply(as.character(testYears), function(tYr){

    # Identify discrepancies between the testing and training sets and filter them out
    
      missingCourse <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["course"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["course"]])
    
      if (length(missingCourse) == 0) {
  # no missingCourse values — keep everything
  missingCourseFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingCourseFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["course"]] %in% missingCourse
  
#  print(x)
#  print(table(missingCourseFilter))
}

      missingEthnicity <- setdiff(vK0_data_list[[tClass]][[tYr]] [["testing"]][["ETHNICITY"]],vK0_data_list[[tClass]][[tYr]] [["training"]][["ETHNICITY"]])
    
      if (length(missingEthnicity) == 0) {
  # no missingCourse values — keep everything
  missingEthnicityFilter <- rep(TRUE, nrow(vK0_data_list[[tClass]][[tYr]][["testing"]]))
} else {
  # normal case — exclude the missingCourse courses
  missingEthnicityFilter <- !vK0_data_list[[tClass]][[tYr]][["testing"]][["ETHNICITY"]] %in% missingEthnicity
  
}
      
missing <- missingCourseFilter & missingEthnicityFilter  

       postResample(pred = predictions_over_time[[tClass]][[tYr]],
                  obs = vK0_data_list[[tClass]][[tYr]][["testing"]][[tClass]][missingCourseFilter]
               ) 
    
        
  })
  names(perClass) <- testYears
  return(perClass)
})
  
names(theResults_vK0) <- list("GRADEGPA", "grade_quad")

```




```{r eval=FALSE}

##########
## LOAD ##
##########

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", "vH2", " ", list("GRADEGPA", "grade_quad"), " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  list("GRADEGPA", "grade_quad")
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", "vH2", " ", list("GRADEGPA", "grade_quad"), " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  list("GRADEGPA", "grade_quad")
)

```


```{r eval=FALSE}

# Re-name 

vH2_data_list <- data_list  # test score
vH2_model_list <- model_list # test score

``` 


```{r eval=FALSE}

## VH2 ##
## PREDICTIONS ##

thePredictions_vH2 <- lapply(list("GRADEGPA", "grade_quad"), function(x) {
  
  # identify course values in testing that aren't in training
  
  courseColumn <- ifelse("course" %in% colnames(vH2_data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(vH2_data_list[[x]][["testing"]][[courseColumn]],vH2_data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(vH2_data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !vH2_data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
  # print(x)
  # print(table(missingFilter))
}
  
  predict(vH2_model_list[[x]],
          newdata = vH2_data_list[[x]][["testing"]][missingFilter,])
  
} )
names(thePredictions_vH2) <- list("GRADEGPA", "grade_quad")

```

```{r eval=FALSE}


theResults_vH2 <- lapply(list("GRADEGPA", "grade_quad"), function(x) {
  
  print(x)
  # identify course values in testing that aren't in training
  
    courseColumn <- ifelse("course" %in% colnames(vH2_data_list[[x]][["testing"]]), "course", "FIRST_MATH_COURSE_CATNBR")
  
  missing <- setdiff(vH2_data_list[[x]][["testing"]][[courseColumn]],vH2_data_list[[x]][["training"]][[courseColumn]])
  
  if (length(missing) == 0) {
  # no missing values — keep everything
  missingFilter <- rep(TRUE, nrow(vH2_data_list[[x]][["testing"]]))
} else {
  # normal case — exclude the missing courses
  missingFilter <- !vH2_data_list[[x]][["testing"]][[courseColumn]] %in% missing
  
#  print(x)
#  print(table(missingFilter))
}
  
  postResample(pred = thePredictions_vH2[[x]],
                  obs = vH2_data_list[[x]][["testing"]][[x]][missingFilter]
  
               )
})
names(theResults_vH2) <- list("GRADEGPA", "grade_quad")



```

```{r eval=FALSE}

# Align vH2 prediction

aligned_vH2 <- alignPrediction(prediction_list = thePredictions_vH2, reference_list = vH2_data_list)


```


```{r eval=FALSE}

# Align vK0

aligned_vK0 <- alignPrediction(prediction_list = predictions_over_time[["GRADEGPA"]], reference_list = vK0_data_list[["GRADEGPA"]])

``` 


```{r eval=FALSE}

# Combine predictions

merge_columns <- colnames(aligned_vK0[["2024"]]) |>
  (\(x){x[-which(x %in% c("pred", "ACT_submit", "SAT_submit"))]})()

aligned_comb <- merge(aligned_vK0[["2024"]],
                        aligned_vH2[["GRADEGPA"]],
                        by = merge_columns,
                        all.x = TRUE
                        ) 
colnames(aligned_comb)[colnames(aligned_comb) %in% c("pred.x")] <- "pred.vK0"
colnames(aligned_comb)[colnames(aligned_comb) %in% c("pred.y")] <- "pred.vH2"

```












# Start Here

```{r}

##############
## LOAD vH3 ##
##############

# load "hiGrades"

hiGrades <- readRDS(here::here("Data","hiGrades from vH3.rds"))

## MODELING RELATED ##

# target_classes <- c("grade_binary", "grade_trinary", "grade_quad", "GRADEGPA")


# Build file paths
model_files <- here::here("Models", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " model.rds"))

# Read each model into a named list
model_list <- setNames(
  lapply(model_files, readRDS),
  list("GRADEGPA")
)

# Build file paths
data_files <- here::here("Data", paste0("Decision Tree ", "vH3", " ", list("GRADEGPA"), " Data.rds"))

# Read each model into a named list
data_list <- setNames(
  lapply(data_files, readRDS),
  list("GRADEGPA")
)

```


```{r}

# Re-name 

vH3_data_list <- data_list  # test score
vH3_model_list <- model_list # test score

``` 

```{r}

#################
##      vH3    ## 
## PREDICTIONS ##
################# 

vH3_pred <- predict(vH3_model_list[["GRADEGPA"]], vH3_data_list[["GRADEGPA"]][["testing"]] ) 

thePredictions_vH3 <- list("GRADEGPA" = vH3_pred)

```

```{r}

#############
##   vH3   ## 
## RESULTS ##
############# 


vH3_results <- postResample(pred = vH3_pred,
                            obs = vH3_data_list[["GRADEGPA"]][["testing"]][["GRADEGPA"]] )

theResults_vH3 <- list("GRADEGPA" = vH3_results)

```



```{r}

# Align vH3 prediction

# First add the EMPLID back to the training and testing sets

vH3_data_list[["GRADEGPA"]][["training"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["trainID"]]

vH3_data_list[["GRADEGPA"]][["testing"]][["EMPLID"]] <- vH3_data_list[["GRADEGPA"]][["testID"]]

#aligned_vH3 <- alignPrediction(prediction_list = thePredictions_vH3, reference_list = vH3_data_list)

aligned_vH3 <- vH3_data_list[["GRADEGPA"]][["testing"]]

aligned_vH3$pred <- vH3_pred

```






# Develop predictions for all courses vH3

```{r}

#################
## ALL OPTIONS ## 
#################

testingGround <- aligned_vH3[,-which(colnames(aligned_vH3) %in% c("pred"))]  


courseGrid <- expand.grid(unique(testingGround$EMPLID), unique(testingGround$course))

testingGround <- merge(testingGround, courseGrid, by.x = "EMPLID", by.y="Var1", all = TRUE)
names(testingGround)[names(testingGround) == "course"] <- "course_orig"
names(testingGround)[names(testingGround) == "Var2"] <- "course"

```

```{r}

#########################
## CALCULATE DISTANCES ##
######################### 

# merge in the information from the year prior

tg2 <- merge(testingGround[,-which(colnames(testingGround) %in% c("HSGPA.z", "ACTMATH.z","dist") )],
                        hiGrades[,c("course", "class_year", "class_year_shift", "ACTMATH.median", "ACTMATH.stdev", "HSGPA.median", "HSGPA.stdev", "HSGPA.count")], by.x = c("course", "class_year"), by.y = c("course","class_year_shift"),  all.x=TRUE 
                        )

# Calculate the distance as a z-score 

tg2$ACTMATH.z <- (tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev
tg2$HSGPA.z <- (tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev

# Calculate distance 

tg2$dist <- sqrt(
  ((tg2$ACTMATH - tg2$ACTMATH.median)/tg2$ACTMATH.stdev)^2 +
    ((tg2$HSGPA - tg2$HSGPA.median)/tg2$HSGPA.stdev)^2
)


```


```{r}

#################
## PREDICTIONS ##
#################

trainColumns <- vH3_model_list[["GRADEGPA"]]$trainingData |> colnames() |> (\(x){x[-which(x %in% ".outcome")]})()

completeCasesFilter <- tg2[,trainColumns] |>
    complete.cases()
 
#  complete.cases(tg2[,trainColumns])

  
tg2$pred.vH3 <- NA
tg2$pred.vH3[completeCasesFilter] <- predict(vH3_model_list[["GRADEGPA"]], tg2[completeCasesFilter,]               
                        ) 

#[,-which(colnames(tg2) %in% c("class_year.y", "ACTMATH.stdev","HSGPA.median","HSGPA.stdev","HSGPA.count"))]


testingGround$pred.vH3 <- predict(vH3_model_list[["GRADEGPA"]], testingGround) 


```  

```{r eval=FALSE}

# COMPARISON SANDBOX
# This section is not repeatable because I fixed the "merge" and now I properly calculate the z-scores and distances per course.

# But now my displayCourseAlternatives won't work # FIXED

# Now I get a lot of variation in the predicted grade-- nice!

# And I'd like to compare with the courseScatter



```


# Alternatives vH3  

```{r} 

set.seed(123)
sampleEmplid <- sample(testingGround$EMPLID, 1) 
displayCourseAlternatives(testingGround, unid = sampleEmplid)

``` 

```{r}  

# Let's see a few more of these

sampleEmplid <- sample(testingGround$EMPLID, 1) 
displayCourseAlternatives(testingGround, unid = sampleEmplid)

# I should put text on here with their pred - actual
# and list, in a legend to the right, top courses, or something

# well, this one did less than the prediction

# Identify people predicted at less than 2.4

testingGround$cut_vH3 <- cut(testingGround$pred.vH3, breaks = c(0,2.4,4), include.lowest = TRUE, right = FALSE)

lowEmplids <- testingGround$EMPLID[testingGround$cut_vH3 == "[0,2.4)" & !is.na(testingGround$cut_vH3)]

set.seed(42)
sampleLowEmplids <- sample(lowEmplids, 10, replace = FALSE)

```


```{r}

#####################
## EXAMINE VISUALS ##
#####################

displayCourseAlternatives(testingGround, 
                          unid = sampleLowEmplids[1],
                          legend_params = list(plot=FALSE))  

displayCourseAlternatives(tg2, 
                          unid = sampleLowEmplids[1],
                          legend_params = list(plot=FALSE))  

emplidFilter <- tg2$EMPLID == sampleLowEmplids[1] 

nearest10courses <- tg2 |>
  (\(x){x[emplidFilter,]})() |>
  (\(x){x[order(x$dist, decreasing = FALSE),]})() |>
  (\(x){x$course[1:10]})()

distCutOff <- tg2$dist <= 3

cutoffCourses <- tg2$course[emplidFilter & distCutOff]

courseScatter(hiGrades[hiGrades$class_year == 2023 & hiGrades$course %in% cutoffCourses,],
              # ellipse_params = list(plot=FALSE),
              qrect_params = list(border = NA),
              showCourse = TRUE,
              plot_params = list(xlim = c(3.2,4.0), ylim = c(15,34)),
              legend_params = list(inset = c(-0.3,0), xpd = TRUE))


points(x = unique(tg2$HSGPA[emplidFilter]),
       y = unique(tg2$ACTMATH[emplidFilter]),
       col = "red",
       pch = 13, cex = 2
       )

plot(y = tg2$pred.vH3[emplidFilter & tg2$dist <= 3], 
     x = tg2$dist[emplidFilter  & tg2$dist <= 3],
     pch = 1:nrow(tg2[emplidFilter  & tg2$dist <= 3,]))




courseScatter(hiGrades[hiGrades$class_year == 2023,], 
              clusterColumn = "clust1",
              ellipse_params = list(plot = FALSE),
              qrect_params = list(border = NA),
              legend_params = list(plot = FALSE),
              plot_params = list(xlim = c(3.6,4.0), ylim = c(17,36)),
              mtext_params = list(side = c(1,2,3,3),
                    text = c("High School GPA","ACTMATH","Incoming qualifications of successful students by course", "Median student achieving at least B-"),
                    font = c(1,1,2,3),
                    cex = c(1.3, 1.3, 1.75, 1.25),
                    line = c(2.3, 2.3, 1.25, 0.30)
              )
              )

```  

```{r}

############################
## EXPLORING DIST CUT-OFF ##
############################

hist(tg2$dist)
hist(tg2$dist[tg2$dist <= 3], right = TRUE)

tg2$dist |>
  cut(breaks = c(seq(from=0,to = 9, by = 1), seq(from =10, to = 120, by = 10))) |>
  table() |>
  proportions() |>
  cumsum() |>
  plot()

# not a bad elbow at a distance of 5

# <=3 is 71% of the counterfactuals

# let's see if I only look at actual courses

actualFilter <- tg2$course == tg2$course_orig

tg2$dist[actualFilter] |>
  cut(breaks = c(seq(from=0,to = 9, by = 1), seq(from =10, to = 120, by = 10))) |>
  table() |>
  proportions() |>
  cumsum() |>
  plot()

# it's a lot tighter.
# <= 3 is 90% of the actual courses
# so <= 3 looks like a pretty reasonable cut-off

# In fact, you almost want to cut-off at <=2, with 74%

#    (0,1]     (1,2]     (2,3]     (3,4]     (4,5]     (5,6] 
# 0.3732465 0.7374749 0.9058116 0.9619238 0.9854709 0.9919840 
#    (6,7]     (7,8]     (8,9]    (9,10]   (10,20]   (20,30] 
# 0.9939880 0.9959920 0.9974950 0.9974950 0.9994990 0.9994990 
#  (30,40]   (40,50]   (50,60]   (60,70]   (70,80]   (80,90] 
# 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000 
# (90,100] (100,110] (110,120] 
# 1.0000000 1.0000000 1.0000000 

## Can I just assume closest is best?

cor(tg2$pred.vH3, tg2$dist, use = "complete.obs") # -0.17 # not nothing!

cor(tg2$pred.vH3[actualFilter], tg2$dist[actualFilter], use = "complete.obs")
# -0.33 # A lot stronger!

distFilter <- tg2$pred.vH3 <=3
cor(tg2$pred.vH3[distFilter], tg2$dist[distFilter], use = "complete.obs")
# -0.08 # huh, about disappears

cor(tg2$pred.vH3[actualFilter & distFilter], tg2$dist[actualFilter & distFilter], use = "complete.obs")
# -0.341 # as expected, the strongest one 

distFilter <- tg2$pred.vH3 <=2
cor(tg2$pred.vH3[actualFilter & distFilter], tg2$dist[actualFilter & distFilter], use = "complete.obs") # -0.345 

plot(y=tg2$pred.vH3[actualFilter],
     x = tg2$dist[actualFilter])

# I like this plot-- predicted grade drops with distance
boxplot(pred.vH3 ~ cut(dist, breaks = 0:5) , data = tg2[actualFilter,] )

# It's a good pattern that holds
# Plenty of overlap
# This graphic needs to be shown in my deep dive if it's not already there  

# Use these two graphics to explain the <=3 dist cut-off 
boxplot(GRADEGPA ~ cut(dist, breaks = 0:5) , data = tg2[actualFilter,] )


tg2$dist[actualFilter] |>
  cut(breaks = c(seq(from=0,to = 9, by = 1), seq(from =10, to = 120, by = 10))) |>
  table() |>
  proportions() |>
  cumsum() |>
  plot()

```



Using the z-score distances correctly:

I'm looking at the grade predictions for EMPLID = "01512029" and it predicts a high grade for MATH_1311 and 2270 --- in fact, it recommends the "highest" grades for extremely high distance courses, and courses that don't look like reasonable recommendations.

Why would it do that?

- "dist" is a bad predictor, and I should just remove it (and rely on z-scores alone, as they also have direction)

- I should cap "dist" to remove these extremely high values (But should I cap z-scores too, then?)

- I have a "matching" problem where very high ability and confidence students dare to take courses far higher than them, and the predictions reflect the counter-examples of students who could do this (when reasonable people self-select out, as they should)

- This shows the difference between "predicting" and "recommending" -- the model can accurately predict the high confidence/ability daredevils, who have some extra reason to believe they can (and they are right) but this doesn't make a good recommendation. 

- How could I manage this?  Should I truncate the predictions to a maximum distance?  Or limit course recommendations to the ten closest classes?

++ Later

This is how recommendation differs from prediction -- prediction is what I have observed.  In order for a valid recommendation, I need "positivity" or that each possible intervention (course choice) is possible.  

In other words, I can't accurately predict a result ( a course choice) that I never saw in my training.  This is because I used a decision tree, which split what I have.  A linear regression could extrapolate. 

What I can do is something like--

-- Exclude courses where less than 15-20 people matched my incoming qualifications (due to poor representation).
-- Use "dist" as a measure of fit, and exclude courses > some cut-off.
-- I am using Euclidean distance of z-scores, which is a special case of Mahalonobis distance for un-correlated variables.  Except-- my variables are correlated, which suggests I should calculate Mahalanobis distance.

-- But, in any case, roughly--- I'll try using dist <=3 as my cut-off, sorting highest to lowest GPA, and see where that gets me.  

-- It's tempting to plot dist x pred

-- Given the lack of precision in these predictions, I don't think the extra information of plotting dist x pred will be that interesting. 

```{r} 

###################################
## IMPROVED DISPLAY ALTERNATIVES ##
##       WITH DIST CUT-OFF       ##
###################################

# Developing in a concurrent sandbox

```






==

The following text refers to not using the z-score distances correctly.  I'll delete this before too long. 

What I mean to say is, this graphic is really sucky and frankly I don't believe it.  

I have it predicting the same grades for a wide variety of classes.

The "distance" graphic suggests that there is some difference in these courses.

Of course, I haven't calcualated the distance between passing and failing.  It might not be much.

But roughly I'm thinking things like -- 

 - focusing on the <2.7 group (high risk)
 - scrapping this graphic (as there is minimal separation)
 - Presenting the information in a new way  
   - Calling out the "baseline" grade and group of courses at this baseline 
   - Listing courses above and below
   - Don't you think? 

 - What would be really fascinating is finding people predicted to fail, and seeing if there was a course with a higher predicted grade.  
 
I'd like to compare this graphic with the distance graphic. 
   



# Develop predictions for all courses

```{r eval=FALSE}

############################################
## CREATE 'TESTING GROUND' COUNTERFACTUAL ##
############################################

testingGround <- aligned_comb[,-which(colnames(aligned_comb) %in% c("pred.vK0","pred.vH2"))]
testingGround$unid <- 10001:(10000+nrow(testingGround))

courseGrid <- expand.grid(testingGround$unid, unique(testingGround$course))

testingGround <- merge(testingGround, courseGrid, by.x = "unid", by.y="Var1", all = TRUE)
names(testingGround)[names(testingGround) == "course"] <- "course_orig"
names(testingGround)[names(testingGround) == "Var2"] <- "course"

## PREDICTIONS ##

testingGround$pred.vH2[!is.na(testingGround$ACTMATH)] <- predict(vH2_model_list[["GRADEGPA"]], testingGround[!is.na(testingGround$ACTMATH) ,-which(colnames(testingGround) %in% c("unid","course_orig"))])

testingGround$pred.vK0 <- predict(vK0_model_list[["GRADEGPA"]][["2024"]],                                  testingGround[,-which(colnames(testingGround) %in% c("unid","course_orig"))])


testingGround$course <- factor(testingGround$course)

```


Looking at the counterfactuals---
like, should I focus on the counterfactuals for the lowest grades?  See if there was a course that would have put them over the line?

Should I look at lowest actual grades? Largest differences between predictions?  

Lowest predictions?  Best grades despite lowest predictions?  

```{r eval=FALSE}

# Lowest actual

bottomQuarter <-   testingGround$unid[testingGround$GRADEGPA < quantile(testingGround$GRADEGPA, 0.25)]

lowSample <- sample(bottomQuarter, 1)

displayCourseAlternatives(data = testingGround,
                          unid = lowSample)

# This is fun, but, like, also not that interesting.

lowPrediction <- testingGround$unid[testingGround$pred.vH2 <= 1.3 & !is.na(testingGround$pred.vH2)] |> unique()

displaySample <- sample(lowPrediction,1)
displayCourseAlternatives(data = testingGround,
                          unid = displaySample)

# so, like ---
# maybe it's the ranking not the absolute (?)
# maybe it's just -not- about the course selection

# and let's look at the linear model

# is the ranking the same for everyone?


aggregate(pred.vH2 ~ unid + course, 
          data = testingGround,
          rank
          ) |> 
  (\(x){ 
  aggregate(pred.vH2 ~ course, data = x, mean)
})()


testingGround$pred_rank <- ave(testingGround$pred.vH2,                          testingGround$unid, 
    FUN = rank)

courseRank <- aggregate(pred_rank ~ course, data = testingGround, mean)

courseRankStDev <- aggregate(pred_rank ~ course, data = testingGround, sd)

# Merge them
plot_data <- merge(courseRank, courseRankStDev, by = "course", suffixes = c("_mean", "_sd"))

# Create the plot
n_courses <- nrow(plot_data)

par(mar = c(3,6,2,1), bg = "bisque")
plot(x = plot_data$pred_rank_mean,
     y = 1:n_courses,
     xlim = range(c(plot_data$pred_rank_mean - plot_data$pred_rank_sd,
                    plot_data$pred_rank_mean + plot_data$pred_rank_sd)),
     ylim = c(1, n_courses),
     pch = 19,
     col = "dodgerblue2",
     xlab = "Average Rank",
     ylab = "",
     yaxt = "n")

# Add y-axis labels
axis(2, at = 1:n_courses, labels = plot_data$course, las = 1)

# Add error bars (mean ± 1 SD)
arrows(x0 = plot_data$pred_rank_mean - plot_data$pred_rank_sd,
       y0 = 1:n_courses,
       x1 = plot_data$pred_rank_mean + plot_data$pred_rank_sd,
       y1 = 1:n_courses,
       angle = 90,
       code = 3,  # arrows on both ends
       length = 0.05,
       col = "dodgerblue2")

segments(x0 = plot_data$pred_rank_mean - plot_data$pred_rank_sd,
         y0 = 1:n_courses,
         x1 = plot_data$pred_rank_mean + plot_data$pred_rank_sd,
         y1 = 1:n_courses,
         col = "gray60",
         lwd = 2)

points(x = plot_data$pred_rank_mean,
       y = 1:n_courses,
       pch = 19,
       col = "dodgerblue2",
       cex = 1.5)

``` 

Wow, so these really are mixed up according to the individual.   That makes me feel better, that the model is actually doing something.  

And, like --- how much of a difference is it?  How big of a difference would it have made? How much should I read into 0.05 GPA prediction difference?

Math 1210, Math 1050 -- why does it swing so much? 

For whom is 1210 high, and for whom is 1210 low?  What differentiates them?

Is that an invitation to do another ML model to differentiate them?  Would it give me new information?

Or should I just start using SHAP values? 



```{r eval=FALSE}  

# I sh/could just delete this section.
# It partially preps "cleanData" then attempted a failed merge, which I managed by going back to the model train to save the EMPLID's.

# Recovering EMPLIDs

ftfData <- readRDS(here::here("Data", "Freshman_data.rds"))
mathCourses <- readRDS(here::here("Data", "FTF_math_data.rds"))

# convert to numeric
mathCourses$CATNBR <- as.numeric(as.character(mathCourses$CATNBR))

# identify unique classes
mathCourses$class <- paste(mathCourses$TERM, mathCourses$SUBJECT_CD, mathCourses$CATNBR, mathCourses$SECTION, sep= "_")

# identify unique courses
mathCourses$course <- paste(mathCourses$SUBJECT_CD, mathCourses$CATNBR, sep="_")

# re-arrange order
mathCourses <- mathCourses[,c("TERM", "course", "class", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME", "EOTDATE", colnames(mathCourses)[!colnames(mathCourses) %in% c("TERM", "course", "EOTDATE", "CATNBR", "EMPLID", "TITLE", "SECTION",  "UNITS", "GRADE", "GRADEGPA", "INSTEMPLID", "INSTNAME") ] )   ]

## IDENTIFY FIRST TERM OF MATH CLASSES ##

# Identify minimum term per student
minTerm <- aggregate(TERM ~ EMPLID, data = mathCourses, min, na.rm = TRUE)
minTerm$reference <- paste(minTerm$EMPLID, minTerm$TERM, sep = "_") # combine for reference
mathCourses$minTerm <- paste(mathCourses$EMPLID, mathCourses$TERM, sep = "_") %in% minTerm$reference # identify courses in minimum terms 

firstTermFilter <- mathCourses$minTerm == TRUE

## identify popular, high volume courses
# mostPopCourses <- c("MATH_1050", "MATH_1210", "MATH_1010", "MATH_1220", "MATH_1030",
#                    "MATH_1090", "MATH_1060", "MATH_2210", "MATH_1070", "MATH_2250") # identified via hierarchical clustering later in this report


# mostPopFilter <- mathCourses$course %in% mostPopCourses

mathLabFilter <- 
  mathCourses$GRADE %in% c(" ", NA) &
  mathCourses$UNITS == 0

# popCourses <- mathCourses[mostPopFilter &  !mathLabFilter,] 

firstCourses <- mathCourses[firstTermFilter & !mathLabFilter,]

# Merge in FTF data
student_demographics <- c("SEX",
                          "FIRST_GEN_STATUS_CD",
                          "ETHNICITY",
                          "RESSTAT",
                          "FA_PELL",
                          "AGE"
)
academic_prep <- c(
  "APCREDIT",
  "HSGPA",
  "HSPRIVATE",
  "HONORS",
  "ACTCOMP",
  "ACTENGL",
  "ACTMATH",
  "ACTSCI" #,
  #  SATMATH, # leaving out SAT because it's a smaller proportion
  #  SATVERBAL,
  #  SATWRTG        
)

firstCourses <- merge(firstCourses, ftfData[, c("EMPLID","COHORT_DT", student_demographics, academic_prep)], by = "EMPLID",  all.x = TRUE)

# Merge in credit load
# names(creditLoad)[names(creditLoad) == "UNITS"] <- "load"

# firstCourses <- merge(firstCourses, creditLoad, by = c("TERM", "EMPLID"), all.x = TRUE)



# Freshman year
firstCourses$cohort_year <- year(firstCourses$COHORT_DT)

# Class year
firstCourses$class_year <- year(firstCourses$EOTDATE)

# Year to class
firstCourses$yr_diff <- time_length(interval(firstCourses$COHORT_DT, firstCourses$EOTDATE), "years")

# Season

# Align season
season_map <- c("4" = "spring", "6" = "summer", "8" = "fall")

# Extract 4th digit and map
firstCourses$season <- season_map[substr(firstCourses$TERM, 4, 4)]

# Course levels column
firstCourses$course_level <- firstCourses$CATNBR %/% 1000

#############################  
### MERGE EMPLIDs BACK IN ### 
#############################  

merge_columns <- intersect(colnames(firstCourses), colnames(aligned_vH3[["GRADEGPA"]]))


aligned <- merge(aligned_vH3[["GRADEGPA"]], firstCourses[,c(merge_columns, "EMPLID")], by = merge_columns, all.x = TRUE)

``` 



